### 2018-01-23 soap.tex
initial version ported from vasilios

\documentclass{sig-alternate-05-2015}
	\pdfpagewidth=8.5truein
	\pdfpageheight=11truein

\usepackage{graphicx}
\usepackage{subfig}
\usepackage[bookmarks=false]{hyperref}
\usepackage{xspace}
\usepackage{listings}
\usepackage[usenames, dvipsnames]{color}

\usepackage{booktabs}

\graphicspath{{./img/}}

%\hyphenation{op-tical net-works semi-conduc-tor}

\input{useful_defines}

\begin{document}

\CopyrightYear{2018}
\setcopyright{acmcopyright}
\conferenceinfo{SAC 2018,}{April 09-13, 2018, Pau, France}
%\isbn{978-1-4503-5191-1/18/04}\acmPrice{\$15.00}
%\doi{DOI: http://dx.doi.org/xxx}

%
\title{A Low-Effort Analytics Platform for Visualizing Evolving Flask-Based Python Web Services}
% Alternative Titles: The Importance of Visualization in the Performance Monitoring of Python Web Services

\numberofauthors{4}
\author{
\alignauthor Patrick Vogel\\
\email{p.p.vogel@student.rug.nl}
%
\alignauthor Thijs Klooster\\
\email{t.klooster.1@student.rug.nl}
%
\and
\alignauthor Vasilios Andrikopoulos\\
\email{v.andrikopoulos@rug.nl}
%
\alignauthor Mircea Lungu\\
\email{m.f.lungu@rug.nl}
%
\sharedaffiliation
\affaddr{Johann Bernoulli Institute for Mathematics and Computer Science}\\
\affaddr{University of Groningen, the Netherlands}\\
}
\date{\today}

% make the title area
\maketitle

\begin{abstract}
\todo{Update abstract \& title}
  Tens of thousands of web applications are written in Flask, a Python-based web framework. Despite a rich ecosystem of extensions, there is none that supports the developer in gaining insight into the evolving performance of their service. In this paper, we introduce \tool, a library that addresses this problem. We present the ease with which the library can be integrated in an already existing web application, discuss some of the visualization perspectives that the library provides and point to some future challenges for similar libraries.

\end{abstract}

\begin{CCSXML}
	<ccs2012>
	<concept>
	<concept_id>10010520.10010553.10010554</concept_id>
	<concept_desc>Computer systems organization~Robotics</concept_desc>
	<concept_significance>100</concept_significance>
	</concept>
	<concept>
	<concept_id>10003033.10003083.10003095</concept_id>
	<concept_desc>Networks~Network reliability</concept_desc>
	<concept_significance>100</concept_significance>
	</concept>
	</ccs2012>  
\end{CCSXML}

\ccsdesc{Computer systems organization~Robotics}
\ccsdesc[100]{Networks~Network reliability}

\section{Introduction}

\todo{Intro to be rewritten; refer to VISSOFT paper as previous work, explain focus is on integration with version control/CI for evolution support, and use of unit tests as early performance indicator}

\todo{motivate with: monitoring in a world which is becoming more and more service-oriented is essential because it allows to perform three main types of actions: system adaptations to provide the request service at the desired level of quality, enabling flexibility in dealing with changing requirements and modes of operation, and enabling operational awareness through dashboards organizing the collected data~\cite{pernici2016monitoring}. In this work we focus on the last part and we discuss how a minimal effort probe-based solution for a particular type of web services can be used to facilitate the other two types of actions by involving the developer.}

\todo{explain: in~\cite{vogel2017low} we introduced the \tool an a high level with a focus on presenting its performance visualization aspects; in this work we only summarize these features by means of introducing some of the provided functionalities of the \tool. We focus on discussing how the tool is implemented and operating, provide a deeper look at its capabilities for integration with version control and continuous integration environments, introduce a mechanism for performance prediction, and provide an evaluation of the proposed approach based on a case study. }

%Every system is a distributed system nowadays \cite{cavage2013there}. Indeed a very large number of applications and web applications are nowadays implemented as two-tier architectures with a front-end implemented with web technologies and a service back-end.
%\ml{I'm not completely happy with this paragraph}
{\em There is no getting around it: you are building a distributed system} argues a recent article \cite{cavage2013there}. Indeed, even the simplest second-year student project is a web application implemented as two-tier architecture with a Javascript/HTML5 front-end a service backend, usually a REST API.

% \hfill mds
% Many contemporary programming languages are offering libraries, modules, or frameworks that facilitate the development of such architectures. 
Python is one of the most popular programming language choices for implementing the back-end of web applications. GitHub contains more than 500K open source Python projects and the Tiobe Index\footnote{TIOBE programming community index is a measure of popularity of programming languages, created and maintained by the TIOBE Company based in Eindhoven, the Netherlands} ranks Python as the 4th most popular programming language as of June 2016.

% possible flask summary
Within the Python community, Flask\footnote{\url{http://flask.pocoo.org/}} is a very popular web framework\footnote{More than 25K projects on GitHub (5\% of all Python projects) are implemented with Flask (cf. a GitHub search for ``language:Python Flask'')}. It provides simplicity and flexibility by implementing a bare-minimum web server, and thus advertises as a micro-framework. The Flask tutorial shows how setting up a simple Flask {\em ``Hello World''} web-service requires no more than 5 lines of Python code \cite{ flask:tutorial}.
% end of summary

Despite their popularity, to the best of our knowledge, there is no simple solution for monitoring the evolving performance of Flask web applications. Thus, every one of the developers of these projects faces one of the following options when confronted with the need of gathering insight into the runtime behavior of their implemented services:

  \begin{enumerate}

    \item Use a commercial monitoring tool which treats the subject API as a black-box (e.g. Pingdom, Runscope). 
    % , Graphite+Graphana+statd etc.

    \item Implement their own ad-hoc analytics solution, having to reinvent basic visualization and interaction strategies.

    \item Live without analytics insight into their services.

  \end{enumerate}

%\todo{For the first point in the list, we can also argue that analytics solutions like Google Analytics can be used, but they have no notion of versioning/integration with the development life cycle. Feel free to cite \cite{papazoglou2011managing} for service evolution purposes}

For projects on a budget (e.g. research, startups) the first and the second options are often not available due to time and financial constraints. Even when using 3rd-party analytics solutions, a critical insight into the evolution of the exposed services of the web application, is missing because such solutions have no notion of versioning and no integration with the development life cycle.~\cite{papazoglou2011managing}

To avoid projects ending up in the third situation, that of living without analytics, in this paper we present \tool~ --- a low-effort service monitoring library for Flask-based Python web services that is easy to integrate and enables the {\em agile assessment} of service evolution. \cite{Nier12b}

As a case study, on which we will illustrate our solution, we are going to use an open source API which, for several years, was in the third of the above-presented situations.

% In the next section, we will present a case study of an open source research API which was for a long time in the third situation presented above -- deployed without analytics insight.

\todo{Add signposting paragraph}

\section{Case Study}
\label{sec:case}

%\todo{Either leave as is or add more info on usage by college if possible}

  \zee\footnote{\url{https://github.com/zeeguu-ecosystem/}} is a platform and an ecosystem of applications for accelerating vocabulary acquisition in a foreign language \cite{Lungu16}. 
%
  The architecture of the ecosystem has at its core an API and a series of satellite applications that together offer to a learner three main inter-dependent features:

  \begin{enumerate}

    \item Reader applications that provide effortless translations for those texts which are too difficult for the readers.

    \item Interactive exercises personally generated based on the preferences and past likes of the learner.

    \item Article recommendations which are personalized for the interests of the learner and come with difficulty estimation that helps the learner find articles with the appropriate difficulty.

  \end{enumerate}

  The core API implemented with Flask and Python provides correspondingly three types of functionality: contextual translations, article recommendations, and personalized exercise suggestions. In total the API provides a bit less than 50 endpoints, out of which probably a dozen are very frequently used. 
  % The development of the core API itself is a research project.

  At the time of writing this article, the ecosystem consists of a reader web application, a web based exercises platform, and a smartwatch application, which are used at the moment of writing this article by more than \activeUserCount active users. The users come from a highschool, a language school, and some users are using it on their own, without any educational context. The highest load we observed until now on the API consisted of 12K requests in one day.

  We will use this \zee API as a case study for this paper. 
  All the figures in this paper are captured from the actual deployment of \tool in the context of the \zee API. The figures are interactive offering basic data exploration capabilities: filter, zoom, and details on demand\cite{Shne99a}. The \tool deployment for the case study can be accessed publicly\footnote{\url{https://zeeguu.unibe.ch/api/dashboard}; username: {\em guest}, password: {\em soap-sac}}.

 % \todo{create new account, update URL if necessary, update activeUserCount, add a couple of sentences of the current state of the case study with the college/language center}
% \ml{we should consider adding also one section in which the architecture/implementation and main features of the dashboard are presented before going on with discussing them in more depth in the following sections --- this should include a rundown on which views are provided from where (overview or per endpoint)}

\section{The Flask Dashboard}
\label{sec:tool}

%\todo{Refer to VISSOFT paper~\cite{vogel2017low} and rephrase where possible; add more info on implementation, e.g. annotation mechanism, database design, session handling etc.}

  As discussed in the introductory section, in previous work~\cite{vogel2017low} we introduced \tool, a drop-in Python library that allows developers to monitor their Flask-based Python web applications with minimal effort.
%
  The \tool is implemented for Python 3.6 and is available on the Python Package Index repository\footnote{\url{https://pypi.python.org/pypi/flask-monitoring-dashboard/1.8}} from where it can be installed by running \install from the command line. 
%  
  The source code of the \tool is published under a permissive MIT license and is available on GitHub\footnote{\url{https://github.com/flask-dashboard}}.

  \paragraph{Usage \& Configuration}

  The fact that the \tool itself is being developed in Python using Flask makes binding to the services of a to-be monitored application developed with the same technologies easy and intuitive. \ml{this is not what matters... rather, what matters is that with the right backend, we can do this minimal configuration for any technology... do you see what i mean? let me see if i can put it in words...}

  Our main goal with the \tool was to allow the integration of the dashboard with an API with minimal effort.
  To start using the \tool, one simply needs to add two lines of code to their Flask web service\footnote{\ins{ In this paper we present the integration with APIs written in Python and Flask hoping that this will not prevent the reader from seeing the more general idea; all the tools we show here for Flask can be applied to other API technologies (e.g. Django) by simply providing a few back-end adapters in the right places. }}:

  % caption=Configuring the \tool is straightforward,
  \begin{lstlisting}[style=custompython]
  import dashboard

  # flask_app is the Flask app object
  dashboard.bind(flask_app)
  \end{lstlisting}

  After (re)deployment of the application, the \tool becomes available at the \code{/dashboard} route of the Flask application. Further configuration is possible by adding additional statements before the binding definition, for example,

  \begin{lstlisting}[style=custompython]
  ...
  dashboard.config.link = 'mydashboard'
  dashboard.bind(flask_app)
  \end{lstlisting}

  allows for a custom route (\code{/mydashboard}) to the dashboard to be defined by the programmer. An external configuration file can be used instead, or in addition to this:

  \begin{lstlisting}[style=custompython]
  ...
  dashboard.config.from_file('config.cfg')
  ...
  \end{lstlisting}

  During binding, the \tool will search for all endpoints defined in the target application. These will be presented to the user in the tool web interface, where the user can select the ones that should be monitored, see \Fref{fig:sep}. The last access time of the endpoint by users, irrespective of whether it is selected by the user to be monitored or not, is also displayed as the means of identifying potentially obsolete endpoints.

    % \mltp{a small screenshot of how the dashboard allows one to select the interesting }
    \begin{figure}
      \centering
      \includegraphics[width=\linewidth]{selecting_endpoints.png}
      \caption{Endpoints that are available for monitoring in the \zee API}
      \label{fig:sep}
    \end{figure}

  In order to monitor an endpoint, the \tool creates a function wrapper for the API function that corresponds to the endpoint. This way, the wrapper will be executed whenever that API call is made. The wrapper contains the code that takes care of monitoring an endpoint. Data collected by the wrappers are persisted in a local database. The SQLAlchemy Object Relational Mapper\footnote{\url{https://www.sqlalchemy.org/}} on top of SQLite\footnote{\url{https://www.sqlite.org/}} is used for this purpose.

  \ml{we must insert at least one dashboard view here... otherwise people will lose their patience... also it will help with explaining the outliers... also, one should see that a large variety of views are already available at this point... I know this risks being  }

  \paragraph{Automated outlier detection and monitoring}

  \ins{When an API is called from within a highly interactive application (as it is the case with the Zeeguu ecosystem) 
  of particular interest to the API developers are performance {\em outliers}. Indeed, a translation request that takes three times more than expected can seriously decrease the perceived quality of the application. Thus,} identifying, collecting all appropriate data, and diagnosing the root causes of such outliers is especially critical in improving the quality of an application.

  % In the context of the RESTful services support by Flask, this means request serving times that deviate from the average to an unexpected degree.

  For this purpose the \tool tracks for every monitored endpoint a {\em running average} response time value\footnote{\ins{For performance reasons, we assume that the response times for the endpoints are normally distributed. Otherwise, more general density distribution information must be collected in real time.}}. When it detects that a given request is an outlier with respect to this past average running value, it triggers the {\em outlier data collection routine} which stores \ins{extra information} about the current execution environment. A configurable threshold with a default value of $2.5$ times the running average response time is used for this purpose.

  For every detected outlier request, the \tool collects information about the current Python stack trace, CPU load, memory consumption, request parameters, etc. in order to allow the maintainer to investigate the causes of these exceptionally slow response times. In this way it is possible to get \ins{detailed insight into the operation of the application in the extreme cases without unnecessarily burdening it with logging this information for every request}.

  \paragraph{User-level monitoring}

  \ins{Sometimes, the same service endpoint will have different execution times for different users, based on the amount of data that a user has in their account. Gmail engineers admitted in 2010 that the performance of their service was dependent on the amount of emails a user would have in their Inbox \footnote{https://techcrunch.com/2010/03/14/gmail-slow/, visited September 23 2017}. }

  \ins{If the API provider needs to understand the performance of their service for individual users, or for different user groups, the dashboard should have a way of distilling a user or a user group from the API request. For a large number of Flask applications that use the global \code{flask.request} object, the object} can be used to retrieve session information, \ins{as in the following snippet}: 
  % . This ability can be used in order to allow, for example, the identification of application users from the \tool, and the consequent aggregation of monitoring information on a per user basis as follows:

  \begin{lstlisting}[style=custompython]  
  # app specific way of extracting the user
  # from a flask request object    
  def get_user_id():
  sid = int(flask.request.args['session'])
  session = User.find_for_session(sid)
  return user_id

  # attaching the get_user_id function
  dashboard.config.get_group_by = 
      get_user_id

  \end{lstlisting}
  This feature, as discussed further in the following section, is important for the clustering of users depending on their application usage. It also provides a view on the distribution of the work load across the application endpoints on a per user basis that allows the identification of more or less popular endpoints for further performance improvements. \ml{this previous paragraph seems a bit reduntant. must revisit}

  \paragraph{Evolution monitoring for Git-versioned projects }

  Version control in \tool is supported in two ways.

\begin{enumerate}

  \item  \textit{Manual} version control requires the developer to tag each new version of the application with an appropriate version identifier~\cite{papazoglou2011managing} using the \code{APP\_VERSION} configuration parameter, for example by adding to the configuration file:

    \begin{lstlisting}[style=custompython]
    [dashboard]
    APP_VERSION=<versionID>
    \end{lstlisting}

  \item \textit{Automatic} version control, aimed for use in conjunction with a version control system and currently only supporting Git\footnote{\url{https://git-scm.com/}}, is enabled by linking to the Git folder in the application source code:

      \begin{lstlisting}[style=custompython]
      ....
      dashboard.config.git = '.git'
      ....
      \end{lstlisting}

    With this extra configuration parameter, the \tool automatically detects the current version of the project by reading the hash code of the deployed version as soon as the application is deployed and started. Information about the version history of the application is also persisted in the tool database and is used during the collection of measurements for the application.

\end{enumerate}

  \paragraph{Preemptive monitoring}

  \todo{the following might need a bit of rewrite}

  The concept of {\em preemptive monitoring} of the application performance by means of instrumenting integration unit testing as the synthetic load is similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}, i.e.~testing service-based applications by using dedicated test input in parallel to its normal use and operation. The difference is that we take advantage of the capability of the CI framework (i.e. Travis in this case) to create an emulated ``live'' environment for integration testing purposes, and use unit testing as the dedicated test input in order to measure performance.

  For example, \Fref{fig:builds} is a screenshot from the dashboard showing the measured response times for 5 consequent builds, with 180 iterations of the unit tests executed in total across all endpoints. The outliers on the right of the figure are due to initial requests \ins{which must wait for a boot-up phase of the API}.

    \begin{figure}[h!]
        \centering
        \includegraphics[width=\columnwidth]{travis_builds}
        \caption{Response times in 5 consequent Travis builds in the \zee case study}
        \label{fig:builds}
      \end{figure}

  While this integration testing environment is different from the actual production deployment one, and the load used is purely synthetic, as will show in Section~\ref{sec:evaluation}, it can actually be used successfully as an early performance indicator for the application developer. \ins{Although predicting the performance of the application would require advanced statistical work, it can be succesfully used as an advanced warning system for the overall performance trend for given endpoints}.

  \paragraph{Integration with Travis}
  \ml{added new section... also, this has to be moved after introducing the pre-emptive monitoring since it's about running tests. switched the order}

  \ins{The pre-emptive monitoring system} can be configured to work together with Continuous Integration (CI) frameworks like Travis\footnote{\url{https://travis-ci.org/}} that deal with automated integration testing.
  The developer needs to define the unit test folder of the project, the URL that Travis should use for reporting its results, and the number of iterations for each unit test should be executed in order to generate a load for the application.

  Each time a new build is created through Travis, the \tool automatically detects all available unit tests defined by the application developer and iterates through each one of them the number of times predefined by the developer while monitoring the response times for each request. The resulting measurements are persisted in a separate part of the tool database so as not to contaminate the ``live'' monitoring data. Beyond seeing the results of each Travis build in the \tool, this feature also allows for preemptive monitoring of the application, as we discuss in the following.

  \ins{Besides functioning as an early warning system for performance, tracking the evolving performance of API tests serves a second purpose. To function as an anchor when the developer analyzes performance degradations. For example, when a developer sees that a given endpoint has become less performant after in a newly deployed version, they are able to investigate whether the performance degradation is visible also with the synthetic load. This would correspond to a performance degradation which is due to ``algorithmic performance degradation''. If on the other hand, the performance of the tests does not change between versions, the developer might conclude that the performance degradation could be due to the workload on the machines, or maybe to the workload mix of the users}

  %By these means, an application developer can use unit tests as synthetic loads for driving the performance monitoring of the application before it reaches the production phase.  As we will discuss further in Section~\ref{sec:testing}, this feature is integral in enabling preemptive performance monitoring of the application.

\section{Performance Perspectives}
\label{sec:views}

  There are three main categories of visual perspectives that are available using \tool:
  \begin{enumerate}
    \item \textit{Service Utilization} presents information about the usage of all endpoints of interest to the developer,
    \item \textit{Endpoint Performance} presents response times of the various service endpoints, and
    \item \textit{User Experience} presents information about the user-perceived performance of monitored service endpoints.
  \end{enumerate}

  %In the remainder of the paper we present several of these perspectives.\footnote{We recommend obtaining a color version of this paper for better readability}

  % The second endpoint conists of two parts, one of them being a table that shows for every monitored endpoint the number of hits it has gotten, the time it was last accessed and its average execution time. The second part is a view with four graphs which show:
  % \begin{itemize} 
  %   \item A heatmap of the total number of requests to the monitored endpoints
  %   \item A stacked bar chart that shows the total number of requests to the monitored endpoints per endpoint per day
  %   \item A boxplot graph showing the average execution time per version of the web service
  %   \item A boxplot graph showing the average execution time for every monitored endpoint
  % \end{itemize}

\todo{Summarize the subsections extensively; evaluate which figures to keep and ideally take new screenshots for the ones we keep; consider adding two more views (like Fig. 3.2 of Patrick's thesis)}

\begin{figure*}[t]
	\centering
	\subfloat[number of requests per endpoint per day]{\includegraphics[width=.9\columnwidth]{number_of_requests_}\label{fig:aeu}}
	\subfloat[requests per hour heatmap]{\includegraphics[width=\columnwidth]{daily_patterns_}\label{fig:dp}}
	\qquad
	\subfloat[response time (in ms) per monitored endpoint view]{\includegraphics[width=.8\columnwidth]{endpoint_performance_}\label{fig:ep}}\qquad
	\subfloat[The Performance Evolution of the \epTranslations endpoint]{\includegraphics[width=.95\columnwidth]{translation_endpoint_evolution_}\label{fig:tee}}
	\caption{Some of the available views\label{fig:views}}
\end{figure*}

\paragraph{Service Utilization}
\label{sec:util}

The most fundamental insight that a service maintainer needs regards service utilization. %\vspace{0.5cm}

Figure \ref{fig:aeu} shows a first perspective on endpoint utilization that \tool provides: a stacked bar chart of the number of hits to various endpoints grouped by day\footnote{Endpoint colors are the same in different views}. Figure~\ref{fig:aeu} in particular shows that at its peak the API has about 12.000 hits per day. 
The way users interact with the platform can also be inferred since the endpoints are indicators of different activity types, e.g.:

\begin{enumerate}

	\item {\color{myred} \epTranslations} is an indicator of the amount of foreign language reading the users are doing, and

	\item {\color{mygreen} \epOutcome} is an indicator of the amount of foreign vocabulary practice the users are doing.

\end{enumerate}

%\begin{figure}[!ht]
%	\centering
%	\includegraphics[width=\linewidth]{number_of_requests_}
%	\caption{The number of requests per endpoint per day view shows the overall utilization of the monitored application}
%	\label{fig:aeu}
%\end{figure}

Besides showing the overall utilization, this endpoint provides the maintainer with information relevant for decisions regarding endpoint deprecation --- one of the most elementary ways of {\em understanding the needs of the downstream}\cite{Haen14a}. In our case study, the maintainer realized that one endpoint which they thought was not being used (i.e. \code{words\_to\_study}), contrary to their expectations, was actually being used\footnote{A complementary type of usage information can also be discovered in the view presented in Figure \ref{fig:sep} where seeing that an endpoint is never accessed can increase the confidence of the maintainer that a given endpoint is not used, although it can never be used a proof.}.

\niceseparator

%   \todo{Add the time series graph and discuss it before the heatmap? We can then sell the heatmap better} 
%   \ml{Not sure about which graph you refer to here V}

A second type of utilization question that the \tool can answer automatically regards {\em cyclic patterns of usage per hour of day} by means of a heatmap, as in \Fref{fig:dp}.

% \mltp{ can we add vertical lines that highlight the beginning of a new week (e.g. before Sunday): }
% Patrick: adding separators in the graph is unfortunately not supported by the library.

%\begin{figure}[!ht]
%	\centering
%	\includegraphics[width=0.9\linewidth]{daily_patterns_}
%	\caption{Usage patterns become easy to spot in the requests per hour heatmap}
%	\label{fig:dp}
%\end{figure}

Figure \ref{fig:dp} shows the API not being used during the early morning hours, with most of the activity focused around working hours and some light activity during the evening. This is consistent with the fact that the current users are all in the central European timezone. Also, the figure shows that the spike in utilization that was visible also in the previous graph happended in on afternoon/evening.

\paragraph{Endpoint Performance}
\label{sec:perf}

  The \tool also collects information regarding endpoint performance. The view in \Fref{fig:ep} summarizes the response times for various endpoints by using a box-and-whiskers plot.

  % \ml{Thijs and Patrick... the visualization people will complain when they see that we have different colors for the same endpoint in different graphs. Can we insure that there is consistency in colors? Simplest trick would be to obtain the color by hashing the name of the endpoint... in that case the same endpoint would have the same color in various graphs.}
  % This is supported in the latest version of the dashboard

%  \begin{figure}[!ht]
%    \centering
%    \includegraphics[width=0.95\linewidth]{endpoint_performance_}
%    \caption{The response time (in ms) per monitored endpoint view allows for identifying performance variability and balancing issues}
%    \label{fig:ep}
%  \end{figure}

  From this view it became clear to the maintainer that four of the endpoints had very large variation in performance.   The most critical for the application and consequently the one optimized first was the \epTranslations endpoint which was part of an interactive loop in the reader applications that relied on the Zeeguu API. Moreover, cf. \Fref{fig:aeu} this endpoint is one of the most used in the system.

  \niceseparator

  \todo{remove the part about how to configure git, the discussion has been subsumed by Section~\ref{sec:tool}.}

  However, with the current configuration of the tool, it would be impossible for the maintainer to see the improvements resulting from the optimization. One way to do this is to add an extra line of configuration to allow \tool to find the git\footnote{\url{https://git-scm.com/}} folder of the deployed service:

%  \begin{lstlisting}[caption=Configuring the \tool with the path to the .git folder enables the generation of evolutionary performance graphs, style=custompython]
%
%dashboard.config.git = 'path/to/.git'
%
%
%    \end{lstlisting}

    \begin{lstlisting}[style=custompython]

  dashboard.config.git = 'path/to/.git'

  \end{lstlisting}

  With this extra configuration, the \tool can now automatically detect the current version of the project, and group measurements by version\footnote{Alternatively, the maintainer can add version identifiers manually for the web application through a configuration file if the system does not use git.}. \Fref{fig:tee} is a zoomed-in version of such a view for \epTranslations with versions increasing from top to bottom

%    \begin{figure}[h!]
%      \centering
%      \includegraphics[width=0.95\linewidth]{translation_endpoint_evolution_}
%      \caption{Visualizing The Performance Evolution of the \epTranslations endpoint}
%      \label{fig:tee}
%    \end{figure}

  This view confirms that the performance of the translation endpoint improved in the recent versions: the median of the last three versions is constantly moving towards the left, and progresses from 1.4 seconds (in the top-most box plot in \Fref{fig:tee}) to 0.8 in the latest version (bottom-most box plot).

\paragraph{User Experience}
\label{sec:user}

For service endpoints which run computations in real time, the maintainer of a system might want to understand the endpoint performance on a per-user basis, especially for situations where the system response time is a function of some individual user load\footnote{E.g. in GMail some users have two emails while other have twenty thousand and this induces different response times for different users}.

\todo{remove the configuration part and rephrase the following, the discussion has been partially subsumed by Section~\ref{sec:tool}.}

To enable this, the \tool must be configured to associate an API call with a given user. The simplest way is to take advantage of the architecture of Flask applications in which a global \code{flask.request} object can be used to retrieve the session which can in turn lead to user identification:

%\begin{lstlisting}[float,caption=Simply define a custom app-specific function for user retrieval and pass it to the \tool to group information by user,style=custompython]
\begin{lstlisting}[style=custompython]  
# app specific way of extracting the user
# from a flask request object    
def get_user_id():
sid = int(flask.request.args['session'])
session = User.find_for_session(sid)
return user_id

# attaching the get_user_id function
dashboard.config.get_group_by = get_user_id

\end{lstlisting}

% \niceseparator

\todo{consider losing one or both of the figures --- or at least updating them}

In Zeeguu, \epFeedItems retrieves a list of recommended articles for a given user. Cf. \Fref{fig:ep} it is the endpoint with the slowest response time and highest variability. The reason for this is that a user can be subscribed to anything from one to three dozen article sources and for each of the sources the system must compute the personalized difficulty of each article at every request.

% In the Zeeguu case study, one of the slowest endpoints, and one with the highest variability as shown in \Fref{fig:ep} is \epFeedItems: it retrieves a list of recommended articles for a given user. However, since a user can be subscribed to anything from one to three dozen article sources, and since the computation of the difficulty is personalized and it is slow, the variability in time among users is likely to be very large.

\begin{figure}[!ht]
	\centering
	\includegraphics[width=0.85\linewidth]{time_per_user}
	\caption{The \epFeedItems shows a very high variability across users}
	\label{fig:tpu}
\end{figure}

A \perspective{Per-User Performance} perspective should show the different response times for different users. Figure \ref{fig:tpu} presents a subset of the corresponding view in the \tool. The figure shows that the response times for this endpoint can vary considerably for different users with some extreme cases where a user has to wait a full minute until their recommended articles are shown\footnote{After seeing this perspective, the maintainer refactored the architecture of the system to move part the difficulty computation out of the interactive loop}.

% \niceseparator

The limitation of the previous view is that it does not present the information also on a per version basis. To address this, a different visual perspective entitled \perspective{Multi-Version per-User Performance} can be defined. Figure \ref{fig:tuv} presents such a perspective by mapping the average execution time for a given user (lines) and given version (columns) on the area of the corresponding circle. The colors represent users. The figure shows average performance varying  across users and versions with no clear trend: this is probably because varying user workload (i.e. number of sources to which the user is registered) is the reason for the variation in response times.

\begin{figure}[!ht]
	\centering
	\includegraphics[width=0.9\linewidth]{time_per_user_per_version}
	\caption{This perspective shows that the evolution of response times for individual users (horizontal lines) across versions (the x-axis)}
	\label{fig:tuv}
\end{figure}

%\section{Preemptive Monitoring}
%\label{sec:testing}
%
%%\todo{discuss here idea of using unit testing as an early performance indicator for evolving services (`preemptive monitoring'?); add a side-by-side figure comparing unit testing vs live system performance per version, maybe calculate error in trend curves? find a better term than prediction for the section title (plus introduction)}
%
%\todo{Similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}, i.e.~testing service-based applications by using dedicated test input in parallel to its normal use and operation. The difference is that we take advantage of the capability of Travis to create an emulated live environment for integration testing purposes, and use unit testing as the dedicated test input in order to measure performance. What we are going provide evidence for in the following is that we can indeed use this preemptive monitoring concept as an early performance indicator for evolving services.}
%
%The developer needs to define the unit test folder of the project, the URL that Travis should use for reporting its results, and the number of iterations for each unit test should be executed in order to generate

\section{Evaluation}
\label{sec:evaluation}

%\todo{evaluation of the \tool across the following dimensions: ease of use, effectiveness, overhead to the system; for the last part we need to do the overhead measurement experiment, for the other two we base it on the case study and bring over whatever is necessary from the discussion section}

\subsection{Effectiveness \& Ease of Use}

\todo{@M: move here the war stories, and discuss how useful the whole thing was to the application maintainer}

\subsection{Efficiency of Preemptive Monitoring}

\todo{here goes the discussion about our forecasting capabilities}

\begin{figure}
	\centering
	\subfloat[The reported response time (in ms) per deployed version in the observation period]{\includegraphics[width=.5\columnwidth]{response_per_version_trunced_trend}}
	\quad
	\subfloat[The measured response time (in ms) using integration with Travis]{\includegraphics[width=.46\columnwidth]{travis_builds_no_outliers_trend}}
	\caption{Comparison of the response times per endpoint: actual production system versus preemptive monitoring data}
	\label{fig:preemptive}
\end{figure}

\begin{table}[h]
	\begin{tabular}{lll}
		\toprule
		Iteration & \bfseries Live (median) & \bfseries Travis (median)\\
		\midrule
		1 & 1349.41 & 764.87\\ 
		2 & 1466.13 & 992.87\\
		3 & 1256.65 & 760.87\\
		4 & 1266.42 & 813.89\\
		5 & 1080.68 & 303.4\\
		\bottomrule
	\end{tabular}
\end{table}

Pearson correlation $r(3)=.93, p=.02$

\subsection{Discussion}

\todo{use this subsection to discuss limitations (from the previous evaluation section) and add the need to be able to handle the horizontal scaling of the application; remove sectioning and summarize briefly limitations that have already been discussed in the VISSOFT paper/move material to other sections; consider renaming as limitations or something similar}

  \subsubsection{Automatically Monitoring System Evolution}

  The main goal of the \tool design was to allow analytics to be collected and insight to be gained by making the smallest possible changes to a running API. %To allow the collection of evolutionary information 
%
  This technique assumes that the web application code which is the target of the monitoring is deployed using \git in the following way:

  \begin{enumerate}
    \item The deployment engineer pulls the latest version of the code from the integration server; this will result in a new commit being pointed at by the HEAD pointer. %than previously
    \item The deployment engineer restarts the new version of the service. At this point, the \tool detects that a new HEAD is present in the local code base and consequently starts associating all the new data points with this new commit\footnote{The \tool detects the current version of the analyzed system the first time it is attached to the application object, and thus, assumes that the Flask application is restarted when a new version is deployed. This is in tune with the current version of Flask, but if the web server will support dynamic updates in the future, this might have to be taken into account}.
  \end{enumerate}

  The advantage of this approach is the need for minimal configuration effort, as discussed in the presentation of the tool. The disadvantage is that it will consider on equal ground the smallest of commits, even one that modifies a comment, and the shortest lived of commits, e.g.~a commit which was active only for a half an hour before a new version with a bug fix was deployed, with major and minor releases of the software. %as a distinct way of grouping the data points. 
  A mechanism to control which versions are important for monitoring purposes is therefore required to be added.
%
  A further possible extension point here is supporting other version control systems (e.g. Mercurial). However, this is a straightforward extension.

  \subsubsection{User-Awareness }

    For the situations in which the user information is not available, the \tool tracks by default information about different IPs and in some cases this might be a sufficiently good approximation of the user diversity and identity. 
    %

    The visualizations for the user experiene perspectives as presented in Section \ref{sec:user} have been tested with several hundred users (of which about \activeUserCount were active during the course of the study), but the scalability of the visualizations must be further investigated for web services with tens of thousands of users.

  \subsubsection{Other Possible Groupings}

    There are other groupings of service utilization and performance that could be important to the maintainer, that we did not explore in this paper. For example, if the service is using OAuth, then together with every request, in the header of the request there is information about the application which is sending a request. Grouping the information by application that sends the request could be important in such a context.

    In general, providing a mechanism that would allow very easy specification of groupings (either as code annotations, as normal code, or as configuration options) is an open problem that \tool and any other similar library will have to face.

\section{Related Work}
\label{sec:related}

\todo{Refer the reader to~\cite{ghezzi2007run} and ~\cite{metzger2010analytical} for a more extensive discussion}

There is a long tradition of using visualization for gaining insight into software performance. Tools like Jinsight \cite{Pauw02a} and Web Services Navigator \cite{Pauw05} pioneered such an approach for Java and for Web Services that communicate with SOAP messages. Both have an ``omniscient'' view of the services / objects and their interactions. As opposed to them, in our work we present an analytics platform which focuses on monitoring a single Python web service from its own point of view.

From the perspective of service monitoring, our work falls within the server-side run-time monitoring of services ~\cite{ghezzi2007run}. While we don't implement the more advanced features of related monitoring solutions like QoS policies driving the monitoring, it presents nevertheless an easy to use approach support improving the performance of web applications.

% \va{Mircea: Consider removing the rest for space...}
% An existing monitoring tool is Pingdom \footnote{https://www.pingdom.com/company/why-pingdom}, which monitors the uptime of an existing web-service. This tool works by pinging the websites (up to 60 times) every minute automatically. Thus this creates a lot of overhead and is bound to be noisy since it will also be influenced by the speed of the network connection\footnote{Another problem is that such a tool would }

% \todo{Runscope? Others?}

\section{Conclusion and Future Work}
\label{sec:conclusions}

\todo{update as appropriate}

In this paper we have shown that it is possible to create a monitoring solution which provides basic insight into web service utilization and performance  with very little effort from the developer. The user group that we are aiming for with this work is application developers using Flask and Python to build web applications with limited or no budget for implementing their own monitoring solutions. The emphasis is in allowing such users to gain insight into how the performance of the service evolves together with the application itself. We believe that the same architecture, and lessons can be applied to other frameworks and other languages.

In the future, we plan to perform case studies with other sytstems, with the goal of discover other needs and to wean out the less useful visualizations in the \tool. We plan to also extend the tool towards supporting multiple deployments of the same applications across multiple nodes (e.g. for the situations where the application is deployed together with a load balancer). Finally, we plan to integrate \tool with unit testing as a complementary source of information about performance evolution.

% references section

\bibliographystyle{abbrv}
\bibliography{soap}

% that's all folks
\end{document}

### 2018-01-23 paper.tex
renamed and moved to IEEE style

--- 2018-01-23 soap.tex
@@
-\documentclass{sig-alternate-05-2015}
+\documentclass[conference]{IEEEtran}
 	\pdfpagewidth=8.5truein
@@
 
-\CopyrightYear{2018}
-\setcopyright{acmcopyright}
-\conferenceinfo{SAC 2018,}{April 09-13, 2018, Pau, France}
-%\isbn{978-1-4503-5191-1/18/04}\acmPrice{\$15.00}
-%\doi{DOI: http://dx.doi.org/xxx}
-
 %
@@
 
-\numberofauthors{4}
 \author{
-\alignauthor Patrick Vogel\\
-\email{p.p.vogel@student.rug.nl}
-%
-\alignauthor Thijs Klooster\\
-\email{t.klooster.1@student.rug.nl}
-%
-\and
-\alignauthor Vasilios Andrikopoulos\\
-\email{v.andrikopoulos@rug.nl}
-%
-\alignauthor Mircea Lungu\\
-\email{m.f.lungu@rug.nl}
-%
-\sharedaffiliation
-\affaddr{Johann Bernoulli Institute for Mathematics and Computer Science}\\
-\affaddr{University of Groningen, the Netherlands}\\
+\IEEEauthorblockN{Patrick Vogel, Thijs Klooster, Vasilios Andrikopoulos, Mircea Lungu}\\
+Johann Bernoulli Institute for Mathematics and Computer Science\\
+University of Groningen, Netherland\\
+Email: \{t.klooster.1,p.p.vogel\}@student.rug.nl, \{v.andrikopoulos,m.f.lungu\}@rug.nl 
 }
-\date{\today}
 
@@
 \end{abstract}
-
-\begin{CCSXML}
-	<ccs2012>
-	<concept>
-	<concept_id>10010520.10010553.10010554</concept_id>
-	<concept_desc>Computer systems organization~Robotics</concept_desc>
-	<concept_significance>100</concept_significance>
-	</concept>
-	<concept>
-	<concept_id>10003033.10003083.10003095</concept_id>
-	<concept_desc>Networks~Network reliability</concept_desc>
-	<concept_significance>100</concept_significance>
-	</concept>
-	</ccs2012>  
-\end{CCSXML}
-
-\ccsdesc{Computer systems organization~Robotics}
-\ccsdesc[100]{Networks~Network reliability}
 

### 2018-02-22 paper.tex
updated biblio file name
Start of writing the introduction

@@
 %
-\title{A Low-Effort Analytics Platform for Visualizing Evolving Flask-Based Python Web Services}
-% Alternative Titles: The Importance of Visualization in the Performance Monitoring of Python Web Services
+\title{The Importance of Visualization in Performance Monitoring of Python Web Services}
+% Alternative Titles: A Low-Effort Analytics Platform for Visualizing Evolving Flask-Based Python Web Services
 
@@
 University of Groningen, Netherland\\
-Email: \{t.klooster.1,p.p.vogel\}@student.rug.nl, \{v.andrikopoulos,m.f.lungu\}@rug.nl 
+Email: \{t.klooster.1,p.p.vogel\}@student.rug.nl, 
+\{v.andrikopoulos,m.f.lungu\}@rug.nl 
 }
@@
 \begin{abstract}
-\todo{Update abstract \& title}
+% TODO: Update abstract
   Tens of thousands of web applications are written in Flask, a Python-based web framework.
@@
 We present the ease with which the library can be integrated in an already existing web application, discuss some of the visualization perspectives that the library provides and point to some future challenges for similar libraries.
-
 \end{abstract}
@@
 
-\todo{motivate with: monitoring in a world which is becoming more and more service-oriented is essential because it allows to perform three main types of actions: system adaptations to provide the request service at the desired level of quality, enabling flexibility in dealing with changing requirements and modes of operation, and enabling operational awareness through dashboards organizing the collected data~\cite{pernici2016monitoring}.
-In this work we focus on the last part and we discuss how a minimal effort probe-based solution for a particular type of web services can be used to facilitate the other two types of actions by involving the developer.}
+Networked computer systems play important roles in the operation of many businesses and organizations.
+The performance of a computer system providing services to a business or customers of a business is integral to the
+successful operation of the business. % TODO: Add citation
+Therefore, monitoring those services is essential because it allows to perform three main types of actions: system adaptations to provide the request service at the desired level of quality, enabling flexibility in dealing with changing requirements and modes of operation, and enabling operational awareness through dashboards organizing the collected data~\cite{pernici2016monitoring}.
+In this work we focus on the last part and we discuss how a minimal effort probe-based solution for a particular type of web services can be used to facilitate the other two types of actions by involving the developer.
 
-\todo{explain: in~\cite{vogel2017low} we introduced the \tool an a high level with a focus on presenting its performance visualization aspects; in this work we only summarize these features by means of introducing some of the provided functionalities of the \tool.
+In \cite{vogel2017low} we introduced the \tool on a high level with a focus on presenting its performance visualization aspects; in this work we only summarize these features by means of introducing some of the provided functionalities of the \tool.
 We focus on discussing how the tool is implemented and operating, provide a deeper look at its capabilities for integration with version control and continuous integration environments, introduce a mechanism for performance prediction, and provide an evaluation of the proposed approach based on a case study.
-}
 
@@
 \bibliographystyle{abbrv}
-\bibliography{soap}
+\bibliography{paper}
 

### 2018-02-23 paper.tex
Worked on the paper

@@
 \usepackage[usenames, dvipsnames]{color}
-
+\usepackage{cite}
+\usepackage{caption}
 \usepackage{booktabs}
@@
 \begin{document}
-
-%
 \title{The Importance of Visualization in Performance Monitoring of Python Web Services}
@@
 Johann Bernoulli Institute for Mathematics and Computer Science\\
-University of Groningen, Netherland\\
+University of Groningen, Netherlands\\
 Email: \{t.klooster.1,p.p.vogel\}@student.rug.nl, 
@@
 
-\todo{Intro to be rewritten; refer to VISSOFT paper as previous work, explain focus is on integration with version control/CI for evolution support, and use of unit tests as early performance indicator}
+% \todo{Intro to be rewritten; refer to VISSOFT paper as previous work, explain focus is on integration with version control/CI for evolution support, and use of unit tests as early performance indicator}
 
-Networked computer systems play important roles in the operation of many businesses and organizations.
-The performance of a computer system providing services to a business or customers of a business is integral to the
-successful operation of the business. % TODO: Add citation
+Networked computer systems play important roles in the operation of many businesses and organizations \cite{davenport1998working}. 
+The performance of a computer system providing services to a business or customers of a business is integral to the successful operation of the business.
+And, as every system is nowadays a distributed system \cite{cavage2013there}, it is often hard to achieve the desired quality of those systems. 
 Therefore, monitoring those services is essential because it allows to perform three main types of actions: system adaptations to provide the request service at the desired level of quality, enabling flexibility in dealing with changing requirements and modes of operation, and enabling operational awareness through dashboards organizing the collected data~\cite{pernici2016monitoring}.
+
 In this work we focus on the last part and we discuss how a minimal effort probe-based solution for a particular type of web services can be used to facilitate the other two types of actions by involving the developer.
@@
 
-%Every system is a distributed system nowadays \cite{cavage2013there}.
-Indeed a very large number of applications and web applications are nowadays implemented as two-tier architectures with a front-end implemented with web technologies and a service back-end.
-%\ml{I'm not completely happy with this paragraph}
-{\em There is no getting around it: you are building a distributed system} argues a recent article \cite{cavage2013there}.
+%{\em There is no getting around it: you are building a distributed system} argues a recent article \cite{cavage2013there}.
 Indeed, even the simplest second-year student project is a web application implemented as two-tier architecture with a Javascript/HTML5 front-end a service backend, usually a REST API.
@@
 Python is one of the most popular programming language choices for implementing the back-end of web applications.
-GitHub contains more than 500K open source Python projects and the Tiobe Index\footnote{TIOBE programming community index is a measure of popularity of programming languages, created and maintained by the TIOBE Company based in Eindhoven, the Netherlands} ranks Python as the 4th most popular programming language as of June 2016.
+GitHub contains more than 500K open source Python projects and the Tiobe Index\footnote{TIOBE programming community index is a measure of popularity of programming languages, created and maintained by the TIOBE Company based in Eindhoven, the Netherlands} ranks Python as the 4th most popular programming language as of February 2018.
 
-% possible flask summary
 Within the Python community, Flask\footnote{\url{http://flask.pocoo.org/}} is a very popular web framework\footnote{More than 25K projects on GitHub (5\% of all Python projects) are implemented with Flask (cf.
@@
 The Flask tutorial shows how setting up a simple Flask {\em ``Hello World''} web-service requires no more than 5 lines of Python code \cite{ flask:tutorial}.
-% end of summary
 
@@
 \label{sec:case}
-
-%\todo{Either leave as is or add more info on usage by college if possible}
-
   \zee\footnote{\url{https://github.com/zeeguu-ecosystem/}} is a platform and an ecosystem of applications for accelerating vocabulary acquisition in a foreign language \cite{Lungu16}. 
-%
 The architecture of the ecosystem has at its core an API and a series of satellite applications that together offer to a learner three main inter-dependent features:
@@
 The highest load we observed until now on the API consisted of 12K requests in one day.
+% Without the Dashboard, there is no possible way to measure that you've 12K requests per day, so maybe leave this out.
+% Furhtermore, it sounds like an advertisement for Zeeguu, which is not the purpose of this paper.
 
@@
 %
-  The \tool is implemented for Python 3.6 and is available on the Python Package Index repository\footnote{\url{https://pypi.python.org/pypi/flask-monitoring-dashboard/1.8}} from where it can be installed by running \install from the command line.
+  The \tool is implemented in Python and works in Python 2.7 up to Python 3.6 and is available on the Python Package Index repository\footnote{\url{https://pypi.python.org/pypi/flask-monitoring-dashboard/1.8}} from where it can be installed by running \install from the command line.
 %  
@@
 
-  During binding, the \tool will search for all endpoints defined in the target application.
+  During binding, the \tool will iterate through all endpoints defined in the target application.
 These will be presented to the user in the tool web interface, where the user can select the ones that should be monitored, see \Fref{fig:sep}.
@@
   In order to monitor an endpoint, the \tool creates a function wrapper for the API function that corresponds to the endpoint.
-This way, the wrapper will be executed whenever that API call is made.
+An alternative for monitoring endpoints would be to wrap the entire application into a middleware.
+However, this alternative results in an overhead for all endpoints, while using a function wrapper only provides overhead for the monitored endpoints.
+Thus, the wrapper will be executed whenever that API call is made.
 The wrapper contains the code that takes care of monitoring an endpoint.
@@
 
-  \ml{we must insert at least one dashboard view here...
-otherwise people will lose their patience...
-also it will help with explaining the outliers...
-also, one should see that a large variety of views are already available at this point...
-I know this risks being  }
+Together with the monitoring-API, a visualization-API is provided.
+The functionality of this API is to present the results that are collected in the former API to the user of the \tool.
+The visualization-API is further explained in Section \ref{sec:views}, but can already be seen in Figure \ref{fig:overview}.
+
+    \begin{figure}
+      \centering
+      \includegraphics[width=\linewidth]{overview.png}
+      \caption{Overview of the endpoints that are monitored in \zee API}
+      \label{fig:overview}
+    \end{figure}
 
@@
 
-  \paragraph{Evolution monitoring for Git-versioned projects }
+  \paragraph{Evolution monitoring for Git-versioned projects}
 
@@
 % references section
-
 \bibliographystyle{abbrv}

### 2018-03-03 paper.tex
Added my contributions to the paper (until the Evaluation-section)

@@
 Python is one of the most popular programming language choices for implementing the back-end of web applications.
-GitHub contains more than 500K open source Python projects and the Tiobe Index\footnote{TIOBE programming community index is a measure of popularity of programming languages, created and maintained by the TIOBE Company based in Eindhoven, the Netherlands} ranks Python as the 4th most popular programming language as of February 2018.
+GitHub contains more than 500K open source Python projects and the Tiobe Index\footnote{TIOBE programming community index is a measure of popularity of programming languages, created and maintained by the TIOBE Company based in Eindhoven, the Netherlands} ranks Python as the 4th most popular programming language as of February 2018 \cite{tiobe}.
 
-Within the Python community, Flask\footnote{\url{http://flask.pocoo.org/}} is a very popular web framework\footnote{More than 25K projects on GitHub (5\% of all Python projects) are implemented with Flask (cf.
+Within the Python community, Flask\footnote{\url{http://flask.pocoo.org/}} is a very popular web framework\footnote{More than 35K projects on GitHub are implemented with Flask (cf.
 a GitHub search for ``language:Python Flask'')}.
@@
 The highest load we observed until now on the API consisted of 12K requests in one day.
-% Without the Dashboard, there is no possible way to measure that you've 12K requests per day, so maybe leave this out.
-% Furhtermore, it sounds like an advertisement for Zeeguu, which is not the purpose of this paper.
 
@@
 
-  As discussed in the introductory section, in previous work~\cite{vogel2017low} we introduced \tool, a drop-in Python library that allows developers to monitor their Flask-based Python web applications with minimal effort.
+  As discussed in the introductory section, in previous work~\cite{vogel2017low} we introduced \tool, a drop-in Flask Extension in Python that allows developers to monitor their web applications with minimal effort.
 %
-  The \tool is implemented in Python and works in Python 2.7 up to Python 3.6 and is available on the Python Package Index repository\footnote{\url{https://pypi.python.org/pypi/flask-monitoring-dashboard/1.8}} from where it can be installed by running \install from the command line.
+  The \tool is implemented in Python and works in Python 2.7 up to Python 3.6 and is available on the Python Package Index repository\footnote{\url{https://pypi.org/project/Flask-MonitoringDashboard/}} from where it can be installed by running \install from the command line.
 %  
-  The source code of the \tool is published under a permissive MIT license and is available on GitHub\footnote{\url{https://github.com/flask-dashboard}}.
+  The source code of the \tool is published under a permissive MIT license and is available on GitHub\footnote{\url{https://github.com/flask-dashboard/flask_monitoringdashboard/}}.
 
@@
   The fact that the \tool itself is being developed in Python using Flask makes binding to the services of a to-be monitored application developed with the same technologies easy and intuitive.
-\ml{this is not what matters...
-rather, what matters is that with the right backend, we can do this minimal configuration for any technology...
-do you see what i mean?
-let me see if i can put it in words...}
+Thus, there are no external dependencies needed (apart from the \tool itself).
+This leads to minimal effort, as the configuration is the only thing that has be done.
 
-  Our main goal with the \tool was to allow the integration of the dashboard with an API with minimal effort.
-  To start using the \tool, one simply needs to add two lines of code to their Flask web service\footnote{\ins{ In this paper we present the integration with APIs written in Python and Flask hoping that this will not prevent the reader from seeing the more general idea; all the tools we show here for Flask can be applied to other API technologies (e.g.
-Django) by simply providing a few back-end adapters in the right places.
-}}:
+  To start using the \tool, one simply needs to add two lines of code to their Flask web service:
 
-  % caption=Configuring the \tool is straightforward,
-  \begin{lstlisting}[style=custompython]
-  import dashboard
+  \begin{lstlisting}[style=custompython, caption=Configuring the \tool is straightforward]
+  import flask_monitoringdashboard as dashboard
 
@@
   After (re)deployment of the application, the \tool becomes available at the \code{/dashboard} route of the Flask application.
-Further configuration is possible by adding additional statements before the binding definition, for example,
+Further configuration is possible by adding additional statements before the binding definition, for example:
 
@@
   ...
-  dashboard.config.from_file('config.cfg')
+  dashboard.config.init_form(file='config.cfg')
   ...
@@
 These will be presented to the user in the tool web interface, where the user can select the ones that should be monitored, see \Fref{fig:sep}.
-The last access time of the endpoint by users, irrespective of whether it is selected by the user to be monitored or not, is also displayed as the means of identifying potentially obsolete endpoints.
 
-    % \mltp{a small screenshot of how the dashboard allows one to select the interesting }
     \begin{figure}
@@
 
-  In order to monitor an endpoint, the \tool creates a function wrapper for the API function that corresponds to the endpoint.
-An alternative for monitoring endpoints would be to wrap the entire application into a middleware.
-However, this alternative results in an overhead for all endpoints, while using a function wrapper only provides overhead for the monitored endpoints.
-Thus, the wrapper will be executed whenever that API call is made.
+  In order to monitor an endpoint, the \tool attaches a function wrapper for the API function that corresponds to the endpoint.
+An alternative for monitoring endpoints would be to wrap the entire application into a middleware \footnote{This is presented in \cite{Zeng}.}.
+However, this alternative results in an overhead for the entire application, while using a function wrapper only provides overhead for the endpoints that are wrapped.
+Thus, the wrapper will be executed only whenever the corresponding API call is made.
 The wrapper contains the code that takes care of monitoring an endpoint.
@@
 
-Together with the monitoring-API, a visualization-API is provided.
+  In \Fref{fig:sep} the last access time of the endpoint by users, irrespective of whether it is selected by the user to be monitored or not, is also displayed.
+This has the means of identifying potentially obsolete endpoints.
+This functionality is implemented using a different wrapper.
+But, since this wrapper is attached to every endpoint, it introduces some overhead for the entire application.
+
+% PV: Maybe remove paragraph and figure, as it doesn't make really sense in this place.
+Apart form the monitoring-API, a visualization-API is provided.
 The functionality of this API is to present the results that are collected in the former API to the user of the \tool.
@@
 \paragraph{Automated outlier detection and monitoring}
+  When an API is called from within a highly interactive application (as it is the case with the Zeeguu ecosystem) of particular interest to the API developers are performance {\em outliers}.
+We consider a request an outlier, if processing that requests takes three times more than the {\em running average} for that endpoint\footnote{It is also possible to configure a different number for detecting outliers.}.
+Identifying and collecting all appropriate data, and diagnosing the root causes of such outliers is especially critical in improving the quality of an application.
 
-  \ins{When an API is called from within a highly interactive application (as it is the case with the Zeeguu ecosystem) 
-  of particular interest to the API developers are performance {\em outliers}.
-Indeed, a translation request that takes three times more than expected can seriously decrease the perceived quality of the application.
-Thus,} identifying, collecting all appropriate data, and diagnosing the root causes of such outliers is especially critical in improving the quality of an application.
-
-  % In the context of the RESTful services support by Flask, this means request serving times that deviate from the average to an unexpected degree.
-
-  For this purpose the \tool tracks for every monitored endpoint a {\em running average} response time value\footnote{\ins{For performance reasons, we assume that the response times for the endpoints are normally distributed.
-Otherwise, more general density distribution information must be collected in real time.}}.
-When it detects that a given request is an outlier with respect to this past average running value, it triggers the {\em outlier data collection routine} which stores \ins{extra information} about the current execution environment.
-A configurable threshold with a default value of $2.5$ times the running average response time is used for this purpose.
-
-  For every detected outlier request, the \tool collects information about the current Python stack trace, CPU load, memory consumption, request parameters, etc.
-in order to allow the maintainer to investigate the causes of these exceptionally slow response times.
-In this way it is possible to get \ins{detailed insight into the operation of the application in the extreme cases without unnecessarily burdening it with logging this information for every request}.
+  An assumption that have to be made is that the response time values are normally distributed, otherwise more general density distribution information must be collected in real time.
+When it detects that a given request is an outlier with respect to this past average running value, it triggers the {\em outlier data collection routine} which stores extra information, such as the current stack trace, all request variables, memory consumption, and CPU load at the moment of processing the request.
+This allow the maintainer to investigate the causes of these exceptionally slow response times.
+In this way it is possible to get detailed insight into the operation of the application in the extreme cases without unnecessarily burdening it with logging this information for every request.
 
   \paragraph{User-level monitoring}
+  \label{sec:user-monitoring}
+  Another possibility for the occurrence of outliers is due to differences in response time per user.
+For example, Gmail engineers admitted in 2010 that the performance of their service was dependent on the amount of emails a user would have in their Inbox \cite{gmail}.
+Therefore, it is important to keep track which request belongs to which user.
+This provides the functionality for the developer to understand the performance of their service for individual users, or for different user groups.
+However, since Flask itself support no functionality for user-management, it is not possible for the \tool to retrieve the corresponding user for a certain request.
+Neither it is an option for storing this in the configuration file, due to retrieving the user-information is only possible in the code itself.
 
-  \ins{Sometimes, the same service endpoint will have different execution times for different users, based on the amount of data that a user has in their account.
-Gmail engineers admitted in 2010 that the performance of their service was dependent on the amount of emails a user would have in their Inbox \footnote{https://techcrunch.com/2010/03/14/gmail-slow/, visited September 23 2017}.
-}
-
-  \ins{If the API provider needs to understand the performance of their service for individual users, or for different user groups, the dashboard should have a way of distilling a user or a user group from the API request.
-For a large number of Flask applications that use the global \code{flask.request} object, the object} can be used to retrieve session information, \ins{as in the following snippet}:
+This leads to a functionality in which the API developer can pass a function that retrieves the information per user.
+This implements a way of distilling a user or a user group from the API request.
+Another advantage of specifying the function by the API developer is that the results can be clustered on more than users only.
+In the example above, the developers might group on email-count, which allows them to find a relation between this and the execution time.
+For a large number of Flask applications that use the global \code{flask.request} object, the object can be used to retrieve session information, as in the following snippet:
   % .
@@
 
-  \begin{lstlisting}[style=custompython]  
-  # app specific way of extracting the user
-  # from a flask request object    
+  \begin{lstlisting}[style=custompython,caption=App specific way of extracting the user from a Flask request object]
   def get_user_id():
@@
   \end{lstlisting}
-  This feature, as discussed further in the following section, is important for the clustering of users depending on their application usage.
+  %This feature, as discussed further in the following section, is important for the clustering of users depending on their application usage.
 It also provides a view on the distribution of the work load across the application endpoints on a per user basis that allows the identification of more or less popular endpoints for further performance improvements.
@@
   \paragraph{Evolution monitoring for Git-versioned projects}
-
-  Version control in \tool is supported in two ways.
+  It is not only useful to find relations in processing endpoints across users, but it is also important to detect whether the overall performance of the web-service decreases during its evolution, which is likely to happen according to \cite{Kaminski:2006:DAW:1137677.1137694}.
+There are two ways in \tool to support version control:
 
 \begin{enumerate}
-
   \item  \textit{Manual} version control requires the developer to tag each new version of the application with an appropriate version identifier~\cite{papazoglou2011managing} using the \code{APP\_VERSION} configuration parameter, for example by adding to the configuration file:
 
-    \begin{lstlisting}[style=custompython]
+    \begin{lstlisting}[style=custompython,caption=Manually adding version control]
     [dashboard]
-    APP_VERSION=<versionID>
+    APP_VERSION=1.2.3
     \end{lstlisting}
@@
       ....
-      dashboard.config.git = '.git'
+      dashboard.config.git = '/path/to/.git'
       ....
@@
     With this extra configuration parameter, the \tool automatically detects the current version of the project by reading the hash code of the deployed version as soon as the application is deployed and started.
-Information about the version history of the application is also persisted in the tool database and is used during the collection of measurements for the application.
+Information about the version history of the application is also persisted in the tool database and is used for clustering the requests result per endpoint.
 
@@
 
-  \todo{the following might need a bit of rewrite}
-
-  The concept of {\em preemptive monitoring} of the application performance by means of instrumenting integration unit testing as the synthetic load is similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}, i.e.~testing service-based applications by using dedicated test input in parallel to its normal use and operation.
-The difference is that we take advantage of the capability of the CI framework (i.e.
-Travis in this case) to create an emulated ``live'' environment for integration testing purposes, and use unit testing as the dedicated test input in order to measure performance.
+  The concept of {\em preemptive monitoring} of the application performance by means of instrumenting integration unit testing as the synthetic load is similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}.
+Testing service-based application can be done in parallel to its normal use and operation.
+By using the capability of a Continuous Integration (CI) framework, a ``live'' environment for integration testing purposes can be simulated without affecting the performance of the deployed application.
+After the results on the testing framework have been collected, they are send to the deployed application, on which they are visualized in the \tool.
 
   For example, \Fref{fig:builds} is a screenshot from the dashboard showing the measured response times for 5 consequent builds, with 180 iterations of the unit tests executed in total across all endpoints.
-The outliers on the right of the figure are due to initial requests \ins{which must wait for a boot-up phase of the API}.
+The outliers on the right of the figure are due to initial requests, which must wait for a boot-up phase of the API.
 
@@
   While this integration testing environment is different from the actual production deployment one, and the load used is purely synthetic, as will show in Section~\ref{sec:evaluation}, it can actually be used successfully as an early performance indicator for the application developer.
-\ins{Although predicting the performance of the application would require advanced statistical work, it can be succesfully used as an advanced warning system for the overall performance trend for given endpoints}.
+Although predicting the performance of the application would require advanced statistical work, it can be successfully used as an advanced warning system for the overall performance trend for given endpoints.
 
-  \paragraph{Integration with Travis}
-  \ml{added new section... also, this has to be moved after introducing the pre-emptive monitoring since it's about running tests. switched the order}
+%  \paragraph{Integration with Travis}
+%  \ml{added new section... also, this has to be moved after introducing the pre-emptive monitoring since it's about running tests. switched the order}
 
-  \ins{The pre-emptive monitoring system} can be configured to work together with Continuous Integration (CI) frameworks like Travis\footnote{\url{https://travis-ci.org/}} that deal with automated integration testing.
+  The {\em pre-emptive monitoring system} can be configured to work together with CI frameworks like Travis\footnote{\url{https://travis-ci.org/}} that deal with automated integration testing.
 The developer needs to define the unit test folder of the project, the URL that Travis should use for reporting its results, and the number of iterations for each unit test should be executed in order to generate a load for the application.
@@
 
-  \ins{Besides functioning as an early warning system for performance, tracking the evolving performance of API tests serves a second purpose.
+  Besides functioning as an early warning system for performance, tracking the evolving performance of API tests serves a second purpose.
 To function as an anchor when the developer analyzes performance degradations.
@@
 This would correspond to a performance degradation which is due to ``algorithmic performance degradation''.
-If on the other hand, the performance of the tests does not change between versions, the developer might conclude that the performance degradation could be due to the workload on the machines, or maybe to the workload mix of the users}
+If on the other hand, the performance of the tests does not change between versions, the developer might conclude that the performance degradation could be due to the workload on the machines, or maybe to the workload mix of the users.
 
@@
 
-  %In the remainder of the paper we present several of these perspectives.\footnote{We recommend obtaining a color version of this paper for better readability}
+In the remainder of this section we present the views into more depth\footnote{We recommend obtaining a color version of this paper for better readability.}.
+% TODO: Elaborate on the introduction, as it is quite short
 
@@
 
-\todo{Summarize the subsections extensively; evaluate which figures to keep and ideally take new screenshots for the ones we keep; consider adding two more views (like Fig. 3.2 of Patrick's thesis)}
+% \todo{Summarize the subsections extensively; evaluate which figures to keep and ideally take new screenshots for the ones we keep; consider adding two more views (like Fig. 3.2 of Patrick's thesis)}
 
@@
 	\centering
-	\subfloat[number of requests per endpoint per day]{\includegraphics[width=.9\columnwidth]{number_of_requests_}\label{fig:aeu}}
-	\subfloat[requests per hour heatmap]{\includegraphics[width=\columnwidth]{daily_patterns_}\label{fig:dp}}
+	\subfloat[Number of requests per endpoint per day]{\includegraphics[width=.9\columnwidth]{number_of_requests_}\label{fig:aeu}}
+	\subfloat[Daily patterns for the \epOutcome endpoint]{\includegraphics[width=\columnwidth]{heatmap_requests}\label{fig:dp}}
 	\qquad
-	\subfloat[response time (in ms) per monitored endpoint view]{\includegraphics[width=.8\columnwidth]{endpoint_performance_}\label{fig:ep}}\qquad
-	\subfloat[The Performance Evolution of the \epTranslations endpoint]{\includegraphics[width=.95\columnwidth]{translation_endpoint_evolution_}\label{fig:tee}}
+\subfloat[The Performance Evolution of the \epTranslations endpoint]{\includegraphics[width=.95\columnwidth]{exec_time_reo}\label{fig:tee}}\qquad	
+\subfloat[response time (in ms) per monitored endpoint view]{\includegraphics[width=.8\columnwidth]{endpoint_performance_}\label{fig:ep}}
+%	\subfloat[The Performance Evolution of the \epTranslations endpoint]{\includegraphics[width=.95\columnwidth]{translation_endpoint_evolution_}\label{fig:tee}}
+
 	\caption{Some of the available views\label{fig:views}}
@@
 
-The most fundamental insight that a service maintainer needs regards service utilization. %\vspace{0.5cm}
-
-Figure \ref{fig:aeu} shows a first perspective on endpoint utilization that \tool provides: a stacked bar chart of the number of hits to various endpoints grouped by day\footnote{Endpoint colors are the same in different views}.
-Figure~\ref{fig:aeu} in particular shows that at its peak the API has about 12.000 hits per day.
-The way users interact with the platform can also be inferred since the endpoints are indicators of different activity types, e.g.:
+The most fundamental insight that a service maintainer needs regards service utilization. 
+\Fref{fig:aeu} shows a first perspective on endpoint utilization that \tool provides: a stacked bar chart of the number of hits to various endpoints grouped by day\footnote{Endpoint colors are the same in different views.}.
+This figure is interesting for two reasons.
+First, it provides a quick overview of the number of requests that the API processes.
+It shows that at its peak the API has about 12.000 requests per day.
+Second, it is easy to spot what the most popular endpoints are, which are in our case study the following three endpoints:
 
@@
 
-	\item {\color{myred} \epTranslations} is an indicator of the amount of foreign language reading the users are doing, and
+	\item \textbf{\color{myred} \epTranslations} is an indicator of the amount of foreign language reading the users are doing, and
 
-	\item {\color{mygreen} \epOutcome} is an indicator of the amount of foreign vocabulary practice the users are doing.
+	\item \textbf{\color{myblue} \epLearnedLanguage} is an indicator of ... % TODO for Mircea
+
+	\item \textbf{\color{mygreen} \epOutcome} is an indicator of the amount of foreign vocabulary practice the users are doing.
 
@@
 
-\niceseparator
-
 %   \todo{Add the time series graph and discuss it before the heatmap? We can then sell the heatmap better} 
@@
 
-A second type of utilization question that the \tool can answer automatically regards {\em cyclic patterns of usage per hour of day} by means of a heatmap, as in \Fref{fig:dp}.
+As \epOutcome is one of the endpoints that are most used in the API, it is interesting to have investigate the usage patterns into more depth.
+For this purpose, \Fref{fig:dp} has been created.
+This graph shows the usage of the endpoint in the number of requests per hour, which is presented in a color-scale from low usage (blue) to highly active (red).
+This figure shows the API is not being used during the early morning hours, with most of the activity focused around working hours and some light activity during the evening.
+This is consistent with the fact that the current users are all in the central European timezone.
+Also, the figure shows that the spike in utilization that was visible also in \Fref{fig:aeu} happened in the afternoon/evening.
+Another utilization that is visible in the \Fref{fig:dp} regards {\em cyclic patterns of usage per hour of day}, which is of interest to attack possible service overloads in the future.
+% PV I'm not sure whether this is a good argument, but the following source might be of interest: http://static.usenix.org/events/usits03/tech/full_papers/welsh/welsh_html/
 
-% \mltp{ can we add vertical lines that highlight the beginning of a new week (e.g. before Sunday): }
-% Patrick: adding separators in the graph is unfortunately not supported by the library.
+%A second type of utilization question that the \tool can answer automatically regards {\em cyclic patterns of usage per hour of day} by means of a heatmap, as in \Fref{fig:dp}.
 
@@
 
-Figure \ref{fig:dp} shows the API not being used during the early morning hours, with most of the activity focused around working hours and some light activity during the evening.
-This is consistent with the fact that the current users are all in the central European timezone.
-Also, the figure shows that the spike in utilization that was visible also in the previous graph happended in on afternoon/evening.
-
 \paragraph{Endpoint Performance}
 \label{sec:perf}
+The \tool also collects information regarding endpoint performance, and as discussed in the previous section, \epTranslations is one of the most used endpoints in the API.
+Therefore, this endpoint is used for further investigating the overall response time.
+In order to do so, one might ask himself whether the overall execution time remains constant during its evolution.
+In order to answer this question, the \tool creates a view which is presented in \Fref{fig:tee}.
+This Figure creates a horizontal boxplot for every version using the response times.
+It is easy to see that there is an increase in the performance times since version {\em 257e1a}.
+The maintainer has confirmed that this decrease in performance is due to a new algorithm that was deployed.
 
-  The \tool also collects information regarding endpoint performance. The view in \Fref{fig:ep} summarizes the response times for various endpoints by using a box-and-whiskers plot.
+Another useful view that the \tool generates is the overall response time clustered per endpoint.
+This is of interest to the maintainer of the API since it provides information about what endpoints might need optimization.
+The results of this view for the case study are presented in \Fref{fig:ep}.
+From this figure it is visible that the response time for \epLearnedLanguage and \epOutcome are in most cases almost instant, while the response time for \epTranslations is far from instant.
+Another observation that can be made is the fact that some endpoints show a huge variation in the response time, this is explained below:
 
-  % \ml{Thijs and Patrick...
-the visualization people will complain when they see that we have different colors for the same endpoint in different graphs.
-Can we insure that there is consistency in colors?
-Simplest trick would be to obtain the color by hashing the name of the endpoint...
-in that case the same endpoint would have the same color in various graphs.}
-  % This is supported in the latest version of the dashboard
-
-%  \begin{figure}[!ht]
-%    \centering
-%    \includegraphics[width=0.95\linewidth]{endpoint_performance_}
-%    \caption{The response time (in ms) per monitored endpoint view allows for identifying performance variability and balancing issues}
-%    \label{fig:ep}
-%  \end{figure}
-
-  From this view it became clear to the maintainer that four of the endpoints had very large variation in performance.
-The most critical for the application and consequently the one optimized first was the \epTranslations endpoint which was part of an interactive loop in the reader applications that relied on the Zeeguu API.
-Moreover, cf.
-\Fref{fig:aeu} this endpoint is one of the most used in the system.
-
-  \niceseparator
-
-  \todo{remove the part about how to configure git, the discussion has been subsumed by Section~\ref{sec:tool}.}
-
-  However, with the current configuration of the tool, it would be impossible for the maintainer to see the improvements resulting from the optimization.
-One way to do this is to add an extra line of configuration to allow \tool to find the git\footnote{\url{https://git-scm.com/}} folder of the deployed service:
-
-%  \begin{lstlisting}[caption=Configuring the \tool with the path to the .git folder enables the generation of evolutionary performance graphs, style=custompython]
-%
-%dashboard.config.git = 'path/to/.git'
-%
-%
-%    \end{lstlisting}
-
-    \begin{lstlisting}[style=custompython]
-
-  dashboard.config.git = 'path/to/.git'
-
-  \end{lstlisting}
-
-  With this extra configuration, the \tool can now automatically detect the current version of the project, and group measurements by version\footnote{Alternatively, the maintainer can add version identifiers manually for the web application through a configuration file if the system does not use git.}.
-\Fref{fig:tee} is a zoomed-in version of such a view for \epTranslations with versions increasing from top to bottom
-
-%    \begin{figure}[h!]
-%      \centering
-%      \includegraphics[width=0.95\linewidth]{translation_endpoint_evolution_}
-%      \caption{Visualizing The Performance Evolution of the \epTranslations endpoint}
-%      \label{fig:tee}
-%    \end{figure}
-
-  This view confirms that the performance of the translation endpoint improved in the recent versions: the median of the last three versions is constantly moving towards the left, and progresses from 1.4 seconds (in the top-most box plot in \Fref{fig:tee}) to 0.8 in the latest version (bottom-most box plot).
+\begin{figure*}[t]
+	\centering
+	\subfloat[User specific performance for the \epLearnedLanguage endpoint]{\includegraphics[width=\linewidth]{exec_time_user}\label{fig:etu}}\qquad
+	\subfloat[User specific performance per version for the \epLearnedLanguage endpoint]{\includegraphics[width=\linewidth]{exec_time_user_version}\label{fig:etuv}}
+	% TODO: Change to order of one of the two figures, such that the users are listed in the right order (they are now different listed).
+	\caption{Some of the available views\label{fig:user_experience}}
+\end{figure*}
 
@@
 \label{sec:user}
+According to \cite{gmail}, the performance might be user-specific.
+In such case, the maintainer of a system might want to understand the impact on the resulting response time.
+The \tool allows several ways of doing this, some of them are presented in \Fref{fig:user_experience}\footnote{Note that both figures show a subset of the corresponding view in the \tool.}.
+First of all, the view presented in \Fref{fig:etu} can be used to detect the impact per version across the entire evolution of the API.
+It shows a number of horizontal boxplots, one for each user (a user is represented by a user-id).
+From this view, it can be concluded that \epLearnedLanguage has a user-specific processing time.
 
-For service endpoints which run computations in real time, the maintainer of a system might want to understand the endpoint performance on a per-user basis, especially for situations where the system response time is a function of some individual user load\footnote{E.g.
-in GMail some users have two emails while other have twenty thousand and this induces different response times for different users}.
-
-\todo{remove the configuration part and rephrase the following, the discussion has been partially subsumed by Section~\ref{sec:tool}.}
-
-To enable this, the \tool must be configured to associate an API call with a given user.
-The simplest way is to take advantage of the architecture of Flask applications in which a global \code{flask.request} object can be used to retrieve the session which can in turn lead to user identification:
-
-%\begin{lstlisting}[float,caption=Simply define a custom app-specific function for user retrieval and pass it to the \tool to group information by user,style=custompython]
-\begin{lstlisting}[style=custompython]  
-# app specific way of extracting the user
-# from a flask request object    
-def get_user_id():
-sid = int(flask.request.args['session'])
-session = User.find_for_session(sid)
-return user_id
-
-# attaching the get_user_id function
-dashboard.config.get_group_by = get_user_id
-
-\end{lstlisting}
-
-% \niceseparator
-
-\todo{consider losing one or both of the figures --- or at least updating them}
-
-In Zeeguu, \epFeedItems retrieves a list of recommended articles for a given user.
-Cf.
-\Fref{fig:ep} it is the endpoint with the slowest response time and highest variability.
-The reason for this is that a user can be subscribed to anything from one to three dozen article sources and for each of the sources the system must compute the personalized difficulty of each article at every request.
-
-% In the Zeeguu case study, one of the slowest endpoints, and one with the highest variability as shown in \Fref{fig:ep} is \epFeedItems: it retrieves a list of recommended articles for a given user.
-However, since a user can be subscribed to anything from one to three dozen article sources, and since the computation of the difficulty is personalized and it is slow, the variability in time among users is likely to be very large.
-
-\begin{figure}[!ht]
-	\centering
-	\includegraphics[width=0.85\linewidth]{time_per_user}
-	\caption{The \epFeedItems shows a very high variability across users}
-	\label{fig:tpu}
-\end{figure}
-
-A \perspective{Per-User Performance} perspective should show the different response times for different users.
-Figure \ref{fig:tpu} presents a subset of the corresponding view in the \tool.
-The figure shows that the response times for this endpoint can vary considerably for different users with some extreme cases where a user has to wait a full minute until their recommended articles are shown\footnote{After seeing this perspective, the maintainer refactored the architecture of the system to move part the difficulty computation out of the interactive loop}.
-
-% \niceseparator
-
-The limitation of the previous view is that it does not present the information also on a per version basis.
-To address this, a different visual perspective entitled \perspective{Multi-Version per-User Performance} can be defined.
-Figure \ref{fig:tuv} presents such a perspective by mapping the average execution time for a given user (lines) and given version (columns) on the area of the corresponding circle.
-The colors represent users.
-The figure shows average performance varying  across users and versions with no clear trend: this is probably because varying user workload (i.e.
+The limitation of this view is that it is not possible to detect whether the performance increases per user throughout the evolution of the API.
+To address this, a different view can be defined, which is shown in \Fref{fig:etuv}.
+This view presents a {\em circle-plot}, which shows a circle for the average execution time for every user per version.
+A larger circle represents a larger execution time.
+While it is impossible to infer the actual value of this average execution time, this plot is useful for determining for which users the performance increases throughout its evolution.
+However, for the \epLearnedLanguage endpoint, there is no clear trend visible: this is probably because varying user workload (i.e.
 number of sources to which the user is registered) is the reason for the variation in response times.
 
-\begin{figure}[!ht]
-	\centering
-	\includegraphics[width=0.9\linewidth]{time_per_user_per_version}
-	\caption{This perspective shows that the evolution of response times for individual users (horizontal lines) across versions (the x-axis)}
-	\label{fig:tuv}
-\end{figure}
+% TODO: Add paragraph about Preemptive Monitoring
 

### 2018-03-20 paper.tex
.

--- 2018-02-22 paper.tex
@@
 %
-\title{The Importance of Visualization in Performance Monitoring of Python Web Services}
-% Alternative Titles: A Low-Effort Analytics Platform for Visualizing Evolving Flask-Based Python Web Services
+\title{A Low-Effort Analytics Platform for Visualizing Evolving Flask-Based Python Web Services}
+% Alternative Titles: The Importance of Visualization in the Performance Monitoring of Python Web Services
 
@@
 University of Groningen, Netherland\\
-Email: \{t.klooster.1,p.p.vogel\}@student.rug.nl, 
-\{v.andrikopoulos,m.f.lungu\}@rug.nl 
+Email: \{t.klooster.1,p.p.vogel\}@student.rug.nl, \{v.andrikopoulos,m.f.lungu\}@rug.nl 
 }
@@
 \begin{abstract}
-% TODO: Update abstract
+\todo{Update abstract \& title}
   Tens of thousands of web applications are written in Flask, a Python-based web framework.
@@
 We present the ease with which the library can be integrated in an already existing web application, discuss some of the visualization perspectives that the library provides and point to some future challenges for similar libraries.
+
 \end{abstract}
@@
 
-Networked computer systems play important roles in the operation of many businesses and organizations.
-The performance of a computer system providing services to a business or customers of a business is integral to the
-successful operation of the business. % TODO: Add citation
-Therefore, monitoring those services is essential because it allows to perform three main types of actions: system adaptations to provide the request service at the desired level of quality, enabling flexibility in dealing with changing requirements and modes of operation, and enabling operational awareness through dashboards organizing the collected data~\cite{pernici2016monitoring}.
-In this work we focus on the last part and we discuss how a minimal effort probe-based solution for a particular type of web services can be used to facilitate the other two types of actions by involving the developer.
+\todo{motivate with: monitoring in a world which is becoming more and more service-oriented is essential because it allows to perform three main types of actions: system adaptations to provide the request service at the desired level of quality, enabling flexibility in dealing with changing requirements and modes of operation, and enabling operational awareness through dashboards organizing the collected data~\cite{pernici2016monitoring}.
+In this work we focus on the last part and we discuss how a minimal effort probe-based solution for a particular type of web services can be used to facilitate the other two types of actions by involving the developer.}
 
-In \cite{vogel2017low} we introduced the \tool on a high level with a focus on presenting its performance visualization aspects; in this work we only summarize these features by means of introducing some of the provided functionalities of the \tool.
+\todo{explain: in~\cite{vogel2017low} we introduced the \tool an a high level with a focus on presenting its performance visualization aspects; in this work we only summarize these features by means of introducing some of the provided functionalities of the \tool.
 We focus on discussing how the tool is implemented and operating, provide a deeper look at its capabilities for integration with version control and continuous integration environments, introduce a mechanism for performance prediction, and provide an evaluation of the proposed approach based on a case study.
+}
 
@@
 
+  \subsubsection{NEW: API Renames}
+  \ml{just an idea}
+
+    The history would normally be lost in the case of an API rename. (also known as the {\em provenance issue} in source code evolution analysis)
+    In theory we could infer that an API was renamed if it is not to be found anymore, and another one with very similar timing characteristics would be found.
+Such a detection can never be completely sure, but it would help with meeting the needs of the API maintainer...
+
   \subsubsection{User-Awareness }

### 2018-03-20 user-or-app-level.tex
.

--- 2018-03-20 paper.tex
@@
-\documentclass[conference]{IEEEtran}
-	\pdfpagewidth=8.5truein
-	\pdfpageheight=11truein
-
-\usepackage{graphicx}
-\usepackage{subfig}
-\usepackage[bookmarks=false]{hyperref}
-\usepackage{xspace}
-\usepackage{listings}
-\usepackage[usenames, dvipsnames]{color}
-
-\usepackage{booktabs}
-
-\graphicspath{{./img/}}
-
-%\hyphenation{op-tical net-works semi-conduc-tor}
-
-\input{useful_defines}
-
-\begin{document}
-
-%
-\title{A Low-Effort Analytics Platform for Visualizing Evolving Flask-Based Python Web Services}
-% Alternative Titles: The Importance of Visualization in the Performance Monitoring of Python Web Services
-
-\author{
-\IEEEauthorblockN{Patrick Vogel, Thijs Klooster, Vasilios Andrikopoulos, Mircea Lungu}\\
-Johann Bernoulli Institute for Mathematics and Computer Science\\
-University of Groningen, Netherland\\
-Email: \{t.klooster.1,p.p.vogel\}@student.rug.nl, \{v.andrikopoulos,m.f.lungu\}@rug.nl 
-}
-
-% make the title area
-\maketitle
-
-\begin{abstract}
-\todo{Update abstract \& title}
-  Tens of thousands of web applications are written in Flask, a Python-based web framework.
-Despite a rich ecosystem of extensions, there is none that supports the developer in gaining insight into the evolving performance of their service.
-In this paper, we introduce \tool, a library that addresses this problem.
-We present the ease with which the library can be integrated in an already existing web application, discuss some of the visualization perspectives that the library provides and point to some future challenges for similar libraries.
-
-\end{abstract}
-
-\section{Introduction}
-
-\todo{Intro to be rewritten; refer to VISSOFT paper as previous work, explain focus is on integration with version control/CI for evolution support, and use of unit tests as early performance indicator}
-
-\todo{motivate with: monitoring in a world which is becoming more and more service-oriented is essential because it allows to perform three main types of actions: system adaptations to provide the request service at the desired level of quality, enabling flexibility in dealing with changing requirements and modes of operation, and enabling operational awareness through dashboards organizing the collected data~\cite{pernici2016monitoring}.
-In this work we focus on the last part and we discuss how a minimal effort probe-based solution for a particular type of web services can be used to facilitate the other two types of actions by involving the developer.}
-
-\todo{explain: in~\cite{vogel2017low} we introduced the \tool an a high level with a focus on presenting its performance visualization aspects; in this work we only summarize these features by means of introducing some of the provided functionalities of the \tool.
-We focus on discussing how the tool is implemented and operating, provide a deeper look at its capabilities for integration with version control and continuous integration environments, introduce a mechanism for performance prediction, and provide an evaluation of the proposed approach based on a case study.
-}
-
-%Every system is a distributed system nowadays \cite{cavage2013there}.
-Indeed a very large number of applications and web applications are nowadays implemented as two-tier architectures with a front-end implemented with web technologies and a service back-end.
-%\ml{I'm not completely happy with this paragraph}
-{\em There is no getting around it: you are building a distributed system} argues a recent article \cite{cavage2013there}.
-Indeed, even the simplest second-year student project is a web application implemented as two-tier architecture with a Javascript/HTML5 front-end a service backend, usually a REST API.
-
-% \hfill mds
-% Many contemporary programming languages are offering libraries, modules, or frameworks that facilitate the development of such architectures. 
-Python is one of the most popular programming language choices for implementing the back-end of web applications.
-GitHub contains more than 500K open source Python projects and the Tiobe Index\footnote{TIOBE programming community index is a measure of popularity of programming languages, created and maintained by the TIOBE Company based in Eindhoven, the Netherlands} ranks Python as the 4th most popular programming language as of June 2016.
-
-% possible flask summary
-Within the Python community, Flask\footnote{\url{http://flask.pocoo.org/}} is a very popular web framework\footnote{More than 25K projects on GitHub (5\% of all Python projects) are implemented with Flask (cf.
-a GitHub search for ``language:Python Flask'')}.
-It provides simplicity and flexibility by implementing a bare-minimum web server, and thus advertises as a micro-framework.
-The Flask tutorial shows how setting up a simple Flask {\em ``Hello World''} web-service requires no more than 5 lines of Python code \cite{ flask:tutorial}.
-% end of summary
-
-Despite their popularity, to the best of our knowledge, there is no simple solution for monitoring the evolving performance of Flask web applications.
-Thus, every one of the developers of these projects faces one of the following options when confronted with the need of gathering insight into the runtime behavior of their implemented services:
-
-  \begin{enumerate}
-
-    \item Use a commercial monitoring tool which treats the subject API as a black-box (e.g. Pingdom, Runscope). 
-    % , Graphite+Graphana+statd etc.
-
-    \item Implement their own ad-hoc analytics solution, having to reinvent basic visualization and interaction strategies.
-
-    \item Live without analytics insight into their services.
-
-  \end{enumerate}
-
-%\todo{For the first point in the list, we can also argue that analytics solutions like Google Analytics can be used, but they have no notion of versioning/integration with the development life cycle.
-Feel free to cite \cite{papazoglou2011managing} for service evolution purposes}
-
-For projects on a budget (e.g.
-research, startups) the first and the second options are often not available due to time and financial constraints.
-Even when using 3rd-party analytics solutions, a critical insight into the evolution of the exposed services of the web application, is missing because such solutions have no notion of versioning and no integration with the development life cycle.~\cite{papazoglou2011managing}
-
-To avoid projects ending up in the third situation, that of living without analytics, in this paper we present \tool~ --- a low-effort service monitoring library for Flask-based Python web services that is easy to integrate and enables the {\em agile assessment} of service evolution.
-\cite{Nier12b}
-
-As a case study, on which we will illustrate our solution, we are going to use an open source API which, for several years, was in the third of the above-presented situations.
-
-% In the next section, we will present a case study of an open source research API which was for a long time in the third situation presented above -- deployed without analytics insight.
-
-\todo{Add signposting paragraph}
-
-\section{Case Study}
-\label{sec:case}
-
-%\todo{Either leave as is or add more info on usage by college if possible}
-
-  \zee\footnote{\url{https://github.com/zeeguu-ecosystem/}} is a platform and an ecosystem of applications for accelerating vocabulary acquisition in a foreign language \cite{Lungu16}. 
-%
-  The architecture of the ecosystem has at its core an API and a series of satellite applications that together offer to a learner three main inter-dependent features:
-
-  \begin{enumerate}
-
-    \item Reader applications that provide effortless translations for those texts which are too difficult for the readers.
-
-    \item Interactive exercises personally generated based on the preferences and past likes of the learner.
-
-    \item Article recommendations which are personalized for the interests of the learner and come with difficulty estimation that helps the learner find articles with the appropriate difficulty.
-
-  \end{enumerate}
-
-  The core API implemented with Flask and Python provides correspondingly three types of functionality: contextual translations, article recommendations, and personalized exercise suggestions.
-In total the API provides a bit less than 50 endpoints, out of which probably a dozen are very frequently used.
-  % The development of the core API itself is a research project.
-
-  At the time of writing this article, the ecosystem consists of a reader web application, a web based exercises platform, and a smartwatch application, which are used at the moment of writing this article by more than \activeUserCount active users.
-The users come from a highschool, a language school, and some users are using it on their own, without any educational context.
-The highest load we observed until now on the API consisted of 12K requests in one day.
-
-  We will use this \zee API as a case study for this paper. 
-  All the figures in this paper are captured from the actual deployment of \tool in the context of the \zee API.
-The figures are interactive offering basic data exploration capabilities: filter, zoom, and details on demand\cite{Shne99a}.
-The \tool deployment for the case study can be accessed publicly\footnote{\url{https://zeeguu.unibe.ch/api/dashboard}; username: {\em guest}, password: {\em soap-sac}}.
-
- % \todo{create new account, update URL if necessary, update activeUserCount, add a couple of sentences of the current state of the case study with the college/language center}
-% \ml{we should consider adding also one section in which the architecture/implementation and main features of the dashboard are presented before going on with discussing them in more depth in the following sections --- this should include a rundown on which views are provided from where (overview or per endpoint)}
-
-\section{The Flask Dashboard}
-\label{sec:tool}
-
-%\todo{Refer to VISSOFT paper~\cite{vogel2017low} and rephrase where possible; add more info on implementation, e.g. annotation mechanism, database design, session handling etc.}
-
-  As discussed in the introductory section, in previous work~\cite{vogel2017low} we introduced \tool, a drop-in Python library that allows developers to monitor their Flask-based Python web applications with minimal effort.
-%
-  The \tool is implemented for Python 3.6 and is available on the Python Package Index repository\footnote{\url{https://pypi.python.org/pypi/flask-monitoring-dashboard/1.8}} from where it can be installed by running \install from the command line.
-%  
-  The source code of the \tool is published under a permissive MIT license and is available on GitHub\footnote{\url{https://github.com/flask-dashboard}}.
-
-  \paragraph{Usage \& Configuration}
-
-  The fact that the \tool itself is being developed in Python using Flask makes binding to the services of a to-be monitored application developed with the same technologies easy and intuitive.
-\ml{this is not what matters...
-rather, what matters is that with the right backend, we can do this minimal configuration for any technology...
-do you see what i mean?
-let me see if i can put it in words...}
-
-  Our main goal with the \tool was to allow the integration of the dashboard with an API with minimal effort.
-  To start using the \tool, one simply needs to add two lines of code to their Flask web service\footnote{\ins{ In this paper we present the integration with APIs written in Python and Flask hoping that this will not prevent the reader from seeing the more general idea; all the tools we show here for Flask can be applied to other API technologies (e.g.
-Django) by simply providing a few back-end adapters in the right places.
-}}:
-
-  % caption=Configuring the \tool is straightforward,
-  \begin{lstlisting}[style=custompython]
-  import dashboard
-
-  # flask_app is the Flask app object
-  dashboard.bind(flask_app)
-  \end{lstlisting}
-
-  After (re)deployment of the application, the \tool becomes available at the \code{/dashboard} route of the Flask application.
-Further configuration is possible by adding additional statements before the binding definition, for example,
-
-  \begin{lstlisting}[style=custompython]
-  ...
-  dashboard.config.link = 'mydashboard'
-  dashboard.bind(flask_app)
-  \end{lstlisting}
-
-  allows for a custom route (\code{/mydashboard}) to the dashboard to be defined by the programmer. An external configuration file can be used instead, or in addition to this:
-
-  \begin{lstlisting}[style=custompython]
-  ...
-  dashboard.config.from_file('config.cfg')
-  ...
-  \end{lstlisting}
-
-  During binding, the \tool will search for all endpoints defined in the target application.
-These will be presented to the user in the tool web interface, where the user can select the ones that should be monitored, see \Fref{fig:sep}.
-The last access time of the endpoint by users, irrespective of whether it is selected by the user to be monitored or not, is also displayed as the means of identifying potentially obsolete endpoints.
-
-    % \mltp{a small screenshot of how the dashboard allows one to select the interesting }
-    \begin{figure}
-      \centering
-      \includegraphics[width=\linewidth]{selecting_endpoints.png}
-      \caption{Endpoints that are available for monitoring in the \zee API}
-      \label{fig:sep}
-    \end{figure}
-
-  In order to monitor an endpoint, the \tool creates a function wrapper for the API function that corresponds to the endpoint.
-This way, the wrapper will be executed whenever that API call is made.
-The wrapper contains the code that takes care of monitoring an endpoint.
-Data collected by the wrappers are persisted in a local database.
-The SQLAlchemy Object Relational Mapper\footnote{\url{https://www.sqlalchemy.org/}} on top of SQLite\footnote{\url{https://www.sqlite.org/}} is used for this purpose.
-
-  \ml{we must insert at least one dashboard view here...
-otherwise people will lose their patience...
-also it will help with explaining the outliers...
-also, one should see that a large variety of views are already available at this point...
-I know this risks being  }
-
-  \paragraph{Automated outlier detection and monitoring}
-
-  \ins{When an API is called from within a highly interactive application (as it is the case with the Zeeguu ecosystem) 
-  of particular interest to the API developers are performance {\em outliers}.
-Indeed, a translation request that takes three times more than expected can seriously decrease the perceived quality of the application.
-Thus,} identifying, collecting all appropriate data, and diagnosing the root causes of such outliers is especially critical in improving the quality of an application.
-
-  % In the context of the RESTful services support by Flask, this means request serving times that deviate from the average to an unexpected degree.
-
-  For this purpose the \tool tracks for every monitored endpoint a {\em running average} response time value\footnote{\ins{For performance reasons, we assume that the response times for the endpoints are normally distributed.
-Otherwise, more general density distribution information must be collected in real time.}}.
-When it detects that a given request is an outlier with respect to this past average running value, it triggers the {\em outlier data collection routine} which stores \ins{extra information} about the current execution environment.
-A configurable threshold with a default value of $2.5$ times the running average response time is used for this purpose.
-
-  For every detected outlier request, the \tool collects information about the current Python stack trace, CPU load, memory consumption, request parameters, etc.
-in order to allow the maintainer to investigate the causes of these exceptionally slow response times.
-In this way it is possible to get \ins{detailed insight into the operation of the application in the extreme cases without unnecessarily burdening it with logging this information for every request}.
-
-  \paragraph{User-level monitoring}
+  \newpage
+  \section{User-level / App Level Monitoring}
 
@@
 must revisit}
-
-  \paragraph{Evolution monitoring for Git-versioned projects }
-
-  Version control in \tool is supported in two ways.
-
-\begin{enumerate}
-
-  \item  \textit{Manual} version control requires the developer to tag each new version of the application with an appropriate version identifier~\cite{papazoglou2011managing} using the \code{APP\_VERSION} configuration parameter, for example by adding to the configuration file:
-
-    \begin{lstlisting}[style=custompython]
-    [dashboard]
-    APP_VERSION=<versionID>
-    \end{lstlisting}
-
-  \item \textit{Automatic} version control, aimed for use in conjunction with a version control system and currently only supporting Git\footnote{\url{https://git-scm.com/}}, is enabled by linking to the Git folder in the application source code:
-
-      \begin{lstlisting}[style=custompython]
-      ....
-      dashboard.config.git = '.git'
-      ....
-      \end{lstlisting}
-
-    With this extra configuration parameter, the \tool automatically detects the current version of the project by reading the hash code of the deployed version as soon as the application is deployed and started.
-Information about the version history of the application is also persisted in the tool database and is used during the collection of measurements for the application.
-
-\end{enumerate}
-
-  \paragraph{Preemptive monitoring}
-
-  \todo{the following might need a bit of rewrite}
-
-  The concept of {\em preemptive monitoring} of the application performance by means of instrumenting integration unit testing as the synthetic load is similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}, i.e.~testing service-based applications by using dedicated test input in parallel to its normal use and operation.
-The difference is that we take advantage of the capability of the CI framework (i.e.
-Travis in this case) to create an emulated ``live'' environment for integration testing purposes, and use unit testing as the dedicated test input in order to measure performance.
-
-  For example, \Fref{fig:builds} is a screenshot from the dashboard showing the measured response times for 5 consequent builds, with 180 iterations of the unit tests executed in total across all endpoints.
-The outliers on the right of the figure are due to initial requests \ins{which must wait for a boot-up phase of the API}.
-
-    \begin{figure}[h!]
-        \centering
-        \includegraphics[width=\columnwidth]{travis_builds}
-        \caption{Response times in 5 consequent Travis builds in the \zee case study}
-        \label{fig:builds}
-      \end{figure}
-
-  While this integration testing environment is different from the actual production deployment one, and the load used is purely synthetic, as will show in Section~\ref{sec:evaluation}, it can actually be used successfully as an early performance indicator for the application developer.
-\ins{Although predicting the performance of the application would require advanced statistical work, it can be succesfully used as an advanced warning system for the overall performance trend for given endpoints}.
-
-  \paragraph{Integration with Travis}
-  \ml{added new section... also, this has to be moved after introducing the pre-emptive monitoring since it's about running tests. switched the order}
-
-  \ins{The pre-emptive monitoring system} can be configured to work together with Continuous Integration (CI) frameworks like Travis\footnote{\url{https://travis-ci.org/}} that deal with automated integration testing.
-  The developer needs to define the unit test folder of the project, the URL that Travis should use for reporting its results, and the number of iterations for each unit test should be executed in order to generate a load for the application.
-
-  Each time a new build is created through Travis, the \tool automatically detects all available unit tests defined by the application developer and iterates through each one of them the number of times predefined by the developer while monitoring the response times for each request.
-The resulting measurements are persisted in a separate part of the tool database so as not to contaminate the ``live'' monitoring data.
-Beyond seeing the results of each Travis build in the \tool, this feature also allows for preemptive monitoring of the application, as we discuss in the following.
-
-  \ins{Besides functioning as an early warning system for performance, tracking the evolving performance of API tests serves a second purpose.
-To function as an anchor when the developer analyzes performance degradations.
-For example, when a developer sees that a given endpoint has become less performant after in a newly deployed version, they are able to investigate whether the performance degradation is visible also with the synthetic load.
-This would correspond to a performance degradation which is due to ``algorithmic performance degradation''.
-If on the other hand, the performance of the tests does not change between versions, the developer might conclude that the performance degradation could be due to the workload on the machines, or maybe to the workload mix of the users}
-
-  %By these means, an application developer can use unit tests as synthetic loads for driving the performance monitoring of the application before it reaches the production phase.
-As we will discuss further in Section~\ref{sec:testing}, this feature is integral in enabling preemptive performance monitoring of the application.
-
-\section{Performance Perspectives}
-\label{sec:views}
-
-  There are three main categories of visual perspectives that are available using \tool:
-  \begin{enumerate}
-    \item \textit{Service Utilization} presents information about the usage of all endpoints of interest to the developer,
-    \item \textit{Endpoint Performance} presents response times of the various service endpoints, and
-    \item \textit{User Experience} presents information about the user-perceived performance of monitored service endpoints.
-  \end{enumerate}
-
-  %In the remainder of the paper we present several of these perspectives.\footnote{We recommend obtaining a color version of this paper for better readability}
-
-  % The second endpoint conists of two parts, one of them being a table that shows for every monitored endpoint the number of hits it has gotten, the time it was last accessed and its average execution time.
-The second part is a view with four graphs which show:
-  % \begin{itemize} 
-  %   \item A heatmap of the total number of requests to the monitored endpoints
-  %   \item A stacked bar chart that shows the total number of requests to the monitored endpoints per endpoint per day
-  %   \item A boxplot graph showing the average execution time per version of the web service
-  %   \item A boxplot graph showing the average execution time for every monitored endpoint
-  % \end{itemize}
-
-\todo{Summarize the subsections extensively; evaluate which figures to keep and ideally take new screenshots for the ones we keep; consider adding two more views (like Fig. 3.2 of Patrick's thesis)}
-
-\begin{figure*}[t]
-	\centering
-	\subfloat[number of requests per endpoint per day]{\includegraphics[width=.9\columnwidth]{number_of_requests_}\label{fig:aeu}}
-	\subfloat[requests per hour heatmap]{\includegraphics[width=\columnwidth]{daily_patterns_}\label{fig:dp}}
-	\qquad
-	\subfloat[response time (in ms) per monitored endpoint view]{\includegraphics[width=.8\columnwidth]{endpoint_performance_}\label{fig:ep}}\qquad
-	\subfloat[The Performance Evolution of the \epTranslations endpoint]{\includegraphics[width=.95\columnwidth]{translation_endpoint_evolution_}\label{fig:tee}}
-	\caption{Some of the available views\label{fig:views}}
-\end{figure*}
-
-\paragraph{Service Utilization}
-\label{sec:util}
-
-The most fundamental insight that a service maintainer needs regards service utilization. %\vspace{0.5cm}
-
-Figure \ref{fig:aeu} shows a first perspective on endpoint utilization that \tool provides: a stacked bar chart of the number of hits to various endpoints grouped by day\footnote{Endpoint colors are the same in different views}.
-Figure~\ref{fig:aeu} in particular shows that at its peak the API has about 12.000 hits per day.
-The way users interact with the platform can also be inferred since the endpoints are indicators of different activity types, e.g.:
-
-\begin{enumerate}
-
-	\item {\color{myred} \epTranslations} is an indicator of the amount of foreign language reading the users are doing, and
-
-	\item {\color{mygreen} \epOutcome} is an indicator of the amount of foreign vocabulary practice the users are doing.
-
-\end{enumerate}
-
-%\begin{figure}[!ht]
-%	\centering
-%	\includegraphics[width=\linewidth]{number_of_requests_}
-%	\caption{The number of requests per endpoint per day view shows the overall utilization of the monitored application}
-%	\label{fig:aeu}
-%\end{figure}
-
-Besides showing the overall utilization, this endpoint provides the maintainer with information relevant for decisions regarding endpoint deprecation --- one of the most elementary ways of {\em understanding the needs of the downstream}\cite{Haen14a}.
-In our case study, the maintainer realized that one endpoint which they thought was not being used (i.e.
-\code{words\_to\_study}), contrary to their expectations, was actually being used\footnote{A complementary type of usage information can also be discovered in the view presented in Figure \ref{fig:sep} where seeing that an endpoint is never accessed can increase the confidence of the maintainer that a given endpoint is not used, although it can never be used a proof.}.
-
-\niceseparator
-
-%   \todo{Add the time series graph and discuss it before the heatmap? We can then sell the heatmap better} 
-%   \ml{Not sure about which graph you refer to here V}
-
-A second type of utilization question that the \tool can answer automatically regards {\em cyclic patterns of usage per hour of day} by means of a heatmap, as in \Fref{fig:dp}.
-
-% \mltp{ can we add vertical lines that highlight the beginning of a new week (e.g. before Sunday): }
-% Patrick: adding separators in the graph is unfortunately not supported by the library.
-
-%\begin{figure}[!ht]
-%	\centering
-%	\includegraphics[width=0.9\linewidth]{daily_patterns_}
-%	\caption{Usage patterns become easy to spot in the requests per hour heatmap}
-%	\label{fig:dp}
-%\end{figure}
-
-Figure \ref{fig:dp} shows the API not being used during the early morning hours, with most of the activity focused around working hours and some light activity during the evening.
-This is consistent with the fact that the current users are all in the central European timezone.
-Also, the figure shows that the spike in utilization that was visible also in the previous graph happended in on afternoon/evening.
-
-\paragraph{Endpoint Performance}
-\label{sec:perf}
-
-  The \tool also collects information regarding endpoint performance. The view in \Fref{fig:ep} summarizes the response times for various endpoints by using a box-and-whiskers plot.
-
-  % \ml{Thijs and Patrick...
-the visualization people will complain when they see that we have different colors for the same endpoint in different graphs.
-Can we insure that there is consistency in colors?
-Simplest trick would be to obtain the color by hashing the name of the endpoint...
-in that case the same endpoint would have the same color in various graphs.}
-  % This is supported in the latest version of the dashboard
-
-%  \begin{figure}[!ht]
-%    \centering
-%    \includegraphics[width=0.95\linewidth]{endpoint_performance_}
-%    \caption{The response time (in ms) per monitored endpoint view allows for identifying performance variability and balancing issues}
-%    \label{fig:ep}
-%  \end{figure}
-
-  From this view it became clear to the maintainer that four of the endpoints had very large variation in performance.
-The most critical for the application and consequently the one optimized first was the \epTranslations endpoint which was part of an interactive loop in the reader applications that relied on the Zeeguu API.
-Moreover, cf.
-\Fref{fig:aeu} this endpoint is one of the most used in the system.
-
-  \niceseparator
-
-  \todo{remove the part about how to configure git, the discussion has been subsumed by Section~\ref{sec:tool}.}
-
-  However, with the current configuration of the tool, it would be impossible for the maintainer to see the improvements resulting from the optimization.
-One way to do this is to add an extra line of configuration to allow \tool to find the git\footnote{\url{https://git-scm.com/}} folder of the deployed service:
-
-%  \begin{lstlisting}[caption=Configuring the \tool with the path to the .git folder enables the generation of evolutionary performance graphs, style=custompython]
-%
-%dashboard.config.git = 'path/to/.git'
-%
-%
-%    \end{lstlisting}
-
-    \begin{lstlisting}[style=custompython]
-
-  dashboard.config.git = 'path/to/.git'
-
-  \end{lstlisting}
-
-  With this extra configuration, the \tool can now automatically detect the current version of the project, and group measurements by version\footnote{Alternatively, the maintainer can add version identifiers manually for the web application through a configuration file if the system does not use git.}.
-\Fref{fig:tee} is a zoomed-in version of such a view for \epTranslations with versions increasing from top to bottom
-
-%    \begin{figure}[h!]
-%      \centering
-%      \includegraphics[width=0.95\linewidth]{translation_endpoint_evolution_}
-%      \caption{Visualizing The Performance Evolution of the \epTranslations endpoint}
-%      \label{fig:tee}
-%    \end{figure}
-
-  This view confirms that the performance of the translation endpoint improved in the recent versions: the median of the last three versions is constantly moving towards the left, and progresses from 1.4 seconds (in the top-most box plot in \Fref{fig:tee}) to 0.8 in the latest version (bottom-most box plot).
-
-\paragraph{User Experience}
-\label{sec:user}
-
-For service endpoints which run computations in real time, the maintainer of a system might want to understand the endpoint performance on a per-user basis, especially for situations where the system response time is a function of some individual user load\footnote{E.g.
-in GMail some users have two emails while other have twenty thousand and this induces different response times for different users}.
-
-\todo{remove the configuration part and rephrase the following, the discussion has been partially subsumed by Section~\ref{sec:tool}.}
-
-To enable this, the \tool must be configured to associate an API call with a given user.
-The simplest way is to take advantage of the architecture of Flask applications in which a global \code{flask.request} object can be used to retrieve the session which can in turn lead to user identification:
-
-%\begin{lstlisting}[float,caption=Simply define a custom app-specific function for user retrieval and pass it to the \tool to group information by user,style=custompython]
-\begin{lstlisting}[style=custompython]  
-# app specific way of extracting the user
-# from a flask request object    
-def get_user_id():
-sid = int(flask.request.args['session'])
-session = User.find_for_session(sid)
-return user_id
-
-# attaching the get_user_id function
-dashboard.config.get_group_by = get_user_id
-
-\end{lstlisting}
-
-% \niceseparator
-
-\todo{consider losing one or both of the figures --- or at least updating them}
-
-In Zeeguu, \epFeedItems retrieves a list of recommended articles for a given user.
-Cf.
-\Fref{fig:ep} it is the endpoint with the slowest response time and highest variability.
-The reason for this is that a user can be subscribed to anything from one to three dozen article sources and for each of the sources the system must compute the personalized difficulty of each article at every request.
-
-% In the Zeeguu case study, one of the slowest endpoints, and one with the highest variability as shown in \Fref{fig:ep} is \epFeedItems: it retrieves a list of recommended articles for a given user.
-However, since a user can be subscribed to anything from one to three dozen article sources, and since the computation of the difficulty is personalized and it is slow, the variability in time among users is likely to be very large.
-
-\begin{figure}[!ht]
-	\centering
-	\includegraphics[width=0.85\linewidth]{time_per_user}
-	\caption{The \epFeedItems shows a very high variability across users}
-	\label{fig:tpu}
-\end{figure}
-
-A \perspective{Per-User Performance} perspective should show the different response times for different users.
-Figure \ref{fig:tpu} presents a subset of the corresponding view in the \tool.
-The figure shows that the response times for this endpoint can vary considerably for different users with some extreme cases where a user has to wait a full minute until their recommended articles are shown\footnote{After seeing this perspective, the maintainer refactored the architecture of the system to move part the difficulty computation out of the interactive loop}.
-
-% \niceseparator
-
-The limitation of the previous view is that it does not present the information also on a per version basis.
-To address this, a different visual perspective entitled \perspective{Multi-Version per-User Performance} can be defined.
-Figure \ref{fig:tuv} presents such a perspective by mapping the average execution time for a given user (lines) and given version (columns) on the area of the corresponding circle.
-The colors represent users.
-The figure shows average performance varying  across users and versions with no clear trend: this is probably because varying user workload (i.e.
-number of sources to which the user is registered) is the reason for the variation in response times.
-
-\begin{figure}[!ht]
-	\centering
-	\includegraphics[width=0.9\linewidth]{time_per_user_per_version}
-	\caption{This perspective shows that the evolution of response times for individual users (horizontal lines) across versions (the x-axis)}
-	\label{fig:tuv}
-\end{figure}
-
-%\section{Preemptive Monitoring}
-%\label{sec:testing}
-%
-%%\todo{discuss here idea of using unit testing as an early performance indicator for evolving services (`preemptive monitoring'?); add a side-by-side figure comparing unit testing vs live system performance per version, maybe calculate error in trend curves?
-find a better term than prediction for the section title (plus introduction)}
-%
-%\todo{Similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}, i.e.~testing service-based applications by using dedicated test input in parallel to its normal use and operation.
-The difference is that we take advantage of the capability of Travis to create an emulated live environment for integration testing purposes, and use unit testing as the dedicated test input in order to measure performance.
-What we are going provide evidence for in the following is that we can indeed use this preemptive monitoring concept as an early performance indicator for evolving services.}
-%
-%The developer needs to define the unit test folder of the project, the URL that Travis should use for reporting its results, and the number of iterations for each unit test should be executed in order to generate
-
-\section{Evaluation}
-\label{sec:evaluation}
-
-%\todo{evaluation of the \tool across the following dimensions: ease of use, effectiveness, overhead to the system; for the last part we need to do the overhead measurement experiment, for the other two we base it on the case study and bring over whatever is necessary from the discussion section}
-
-\subsection{Effectiveness \& Ease of Use}
-
-\todo{@M: move here the war stories, and discuss how useful the whole thing was to the application maintainer}
-
-\subsection{Efficiency of Preemptive Monitoring}
-
-\todo{here goes the discussion about our forecasting capabilities}
-
-\begin{figure}
-	\centering
-	\subfloat[The reported response time (in ms) per deployed version in the observation period]{\includegraphics[width=.5\columnwidth]{response_per_version_trunced_trend}}
-	\quad
-	\subfloat[The measured response time (in ms) using integration with Travis]{\includegraphics[width=.46\columnwidth]{travis_builds_no_outliers_trend}}
-	\caption{Comparison of the response times per endpoint: actual production system versus preemptive monitoring data}
-	\label{fig:preemptive}
-\end{figure}
-
-\begin{table}[h]
-	\begin{tabular}{lll}
-		\toprule
-		Iteration & \bfseries Live (median) & \bfseries Travis (median)\\
-		\midrule
-		1 & 1349.41 & 764.87\\ 
-		2 & 1466.13 & 992.87\\
-		3 & 1256.65 & 760.87\\
-		4 & 1266.42 & 813.89\\
-		5 & 1080.68 & 303.4\\
-		\bottomrule
-	\end{tabular}
-\end{table}
-
-Pearson correlation $r(3)=.93, p=.02$
-
-\subsection{Discussion}
-
-\todo{use this subsection to discuss limitations (from the previous evaluation section) and add the need to be able to handle the horizontal scaling of the application; remove sectioning and summarize briefly limitations that have already been discussed in the VISSOFT paper/move material to other sections; consider renaming as limitations or something similar}
-
-  \subsubsection{Automatically Monitoring System Evolution}
-
-  The main goal of the \tool design was to allow analytics to be collected and insight to be gained by making the smallest possible changes to a running API.
-%To allow the collection of evolutionary information
-%
-  This technique assumes that the web application code which is the target of the monitoring is deployed using \git in the following way:
-
-  \begin{enumerate}
-    \item The deployment engineer pulls the latest version of the code from the integration server; this will result in a new commit being pointed at by the HEAD pointer. %than previously
-    \item The deployment engineer restarts the new version of the service.
-At this point, the \tool detects that a new HEAD is present in the local code base and consequently starts associating all the new data points with this new commit\footnote{The \tool detects the current version of the analyzed system the first time it is attached to the application object, and thus, assumes that the Flask application is restarted when a new version is deployed.
-This is in tune with the current version of Flask, but if the web server will support dynamic updates in the future, this might have to be taken into account}.
-  \end{enumerate}
-
-  The advantage of this approach is the need for minimal configuration effort, as discussed in the presentation of the tool.
-The disadvantage is that it will consider on equal ground the smallest of commits, even one that modifies a comment, and the shortest lived of commits, e.g.~a commit which was active only for a half an hour before a new version with a bug fix was deployed, with major and minor releases of the software.
-%as a distinct way of grouping the data points.
-  A mechanism to control which versions are important for monitoring purposes is therefore required to be added.
-%
-  A further possible extension point here is supporting other version control systems (e.g. Mercurial). However, this is a straightforward extension.
-
-  \subsubsection{NEW: API Renames}
-  \ml{just an idea}
-
-    The history would normally be lost in the case of an API rename. (also known as the {\em provenance issue} in source code evolution analysis)
-    In theory we could infer that an API was renamed if it is not to be found anymore, and another one with very similar timing characteristics would be found.
-Such a detection can never be completely sure, but it would help with meeting the needs of the API maintainer...
-
-  \subsubsection{User-Awareness }
-
-    For the situations in which the user information is not available, the \tool tracks by default information about different IPs and in some cases this might be a sufficiently good approximation of the user diversity and identity.
-    %
-
-    The visualizations for the user experiene perspectives as presented in Section \ref{sec:user} have been tested with several hundred users (of which about \activeUserCount were active during the course of the study), but the scalability of the visualizations must be further investigated for web services with tens of thousands of users.
-
-  \subsubsection{Other Possible Groupings}
-
-    There are other groupings of service utilization and performance that could be important to the maintainer, that we did not explore in this paper.
-For example, if the service is using OAuth, then together with every request, in the header of the request there is information about the application which is sending a request.
-Grouping the information by application that sends the request could be important in such a context.
-
-    In general, providing a mechanism that would allow very easy specification of groupings (either as code annotations, as normal code, or as configuration options) is an open problem that \tool and any other similar library will have to face.
-
-\section{Related Work}
-\label{sec:related}
-
-\todo{Refer the reader to~\cite{ghezzi2007run} and ~\cite{metzger2010analytical} for a more extensive discussion}
-
-There is a long tradition of using visualization for gaining insight into software performance.
-Tools like Jinsight \cite{Pauw02a} and Web Services Navigator \cite{Pauw05} pioneered such an approach for Java and for Web Services that communicate with SOAP messages.
-Both have an ``omniscient'' view of the services / objects and their interactions.
-As opposed to them, in our work we present an analytics platform which focuses on monitoring a single Python web service from its own point of view.
-
-From the perspective of service monitoring, our work falls within the server-side run-time monitoring of services ~\cite{ghezzi2007run}.
-While we don't implement the more advanced features of related monitoring solutions like QoS policies driving the monitoring, it presents nevertheless an easy to use approach support improving the performance of web applications.
-
-% \va{Mircea: Consider removing the rest for space...}
-% An existing monitoring tool is Pingdom \footnote{https://www.pingdom.com/company/why-pingdom}, which monitors the uptime of an existing web-service.
-This tool works by pinging the websites (up to 60 times) every minute automatically.
-Thus this creates a lot of overhead and is bound to be noisy since it will also be influenced by the speed of the network connection\footnote{Another problem is that such a tool would }
-
-% \todo{Runscope? Others?}
-
-\section{Conclusion and Future Work}
-\label{sec:conclusions}
-
-\todo{update as appropriate}
-
-In this paper we have shown that it is possible to create a monitoring solution which provides basic insight into web service utilization and performance  with very little effort from the developer.
-The user group that we are aiming for with this work is application developers using Flask and Python to build web applications with limited or no budget for implementing their own monitoring solutions.
-The emphasis is in allowing such users to gain insight into how the performance of the service evolves together with the application itself.
-We believe that the same architecture, and lessons can be applied to other frameworks and other languages.
-
-In the future, we plan to perform case studies with other sytstems, with the goal of discover other needs and to wean out the less useful visualizations in the \tool.
-We plan to also extend the tool towards supporting multiple deployments of the same applications across multiple nodes (e.g.
-for the situations where the application is deployed together with a load balancer).
-Finally, we plan to integrate \tool with unit testing as a complementary source of information about performance evolution.
-
-% references section
-
-\bibliographystyle{abbrv}
-\bibliography{paper}
-
-% that's all folks
-\end{document}

### 2018-03-21 restructured.tex
restructuring a bit...

--- 2018-03-20 paper.tex
@@
 %
-\title{A Low-Effort Analytics Platform for Visualizing Evolving Flask-Based Python Web Services}
-% Alternative Titles: The Importance of Visualization in the Performance Monitoring of Python Web Services
+\title{Agile Monitoring of Evolving Web Services}
 
@@
 
-\section{Case Study}
+\section{Case Study: The API (Zeeguu)}
 \label{sec:case}
@@
 
-\section{The Flask Dashboard}
+\newpage
+
+\section{Case Study: The Technology (Flask)}
 \label{sec:tool}
 
-%\todo{Refer to VISSOFT paper~\cite{vogel2017low} and rephrase where possible; add more info on implementation, e.g. annotation mechanism, database design, session handling etc.}
+ Flask is a microframework for Python. It IS USED BY THOUSANDS OF PROJECTS. ELABORATE MORE ON THIS... 
+ \vspace{2cm}
+
+ Flask depends on two external libraries: the Jinja2 template engine and the Werkzeug WSGI toolkit.
+
+ “Micro” does not mean that your whole web application has to fit into a single Python file (although it certainly can), nor does it mean that Flask is lacking in functionality.
+The “micro” in microframework means Flask aims to keep the core simple but extensible.
+Flask won’t make many decisions for you, such as what database to use.
+Those decisions that it does make, such as what templating engine to use, are easy to change.
+Everything else is up to you, so that Flask can be everything you need and nothing you don’t.
+
+ Flask has many configuration values, with sensible defaults, and a few conventions when getting started.
+By convention, templates and static files are stored in subdirectories within the application’s Python source tree, with the names templates and static respectively.
+While this can be changed, you usually don’t have to, especially when getting started.
+
+ A Python web application based on WSGI has to have one central callable object that implements the actual application.
+In Flask this is an instance of the Flask class.
+Each Flask application has to create an instance of this class itself and pass it the name of the module, but why can’t Flask do that itself?
+
+\begin{lstlisting}[style=custompython]
+
+from flask import Flask
+app = Flask(__name__)
+
+@app.route('/')
+def index():
+    return 'Hello World!'
+
+\end{lstlisting}
+
+\subsection*{NOTHING PREVENTS GENERALIZATION}
+In this article we take as an example the Flask case study.
+But there is nothing that prevents the generalization of 
+our approach.
+
+MAKE SURE THAT THIS IS CLEAR.
+
+    \begin{figure*}[ht!]
+      \centering
+      \includegraphics[width=0.8\linewidth]{interceptor}
+      \caption{The first thing that needs to be done, is to decorate the application object with an INTERCEPTOR}
+      \label{fig:sep}
+    \end{figure*}
+
+\newpage
+
+\section{Monitoring Service Utilization}
+
+REMEMBER THE GOAL: Our main goal with the \tool issues to allow the integration of the dashboard with an API with minimal effort.
+
+\ins{The first important question that such a 
+service monitoring system should provide is 
+information about what endpoints are being
+used and by whom}.
+
+TODO: Find references, arguments, related work
+that provide more details about why understanding
+who uses endpoints, and which are used are important.
+
+This is clearly the case in static analysis (see paper
+by haenni and lungu...) but should even more be the
+case in APIs.
+
+  \newpage
+TODO: TALK ABOUT THE DATABASE...
 
@@
 
-  \paragraph{Usage \& Configuration}
+  \newpage
+  \section{One Line Of Code Configuration With Technology Specific Interceptors}
 
   The fact that the \tool itself is being developed in Python using Flask makes binding to the services of a to-be monitored application developed with the same technologies easy and intuitive.
-\ml{this is not what matters...
-rather, what matters is that with the right backend, we can do this minimal configuration for any technology...
-do you see what i mean?
-let me see if i can put it in words...}
 
-  Our main goal with the \tool was to allow the integration of the dashboard with an API with minimal effort.
-  To start using the \tool, one simply needs to add two lines of code to their Flask web service\footnote{\ins{ In this paper we present the integration with APIs written in Python and Flask hoping that this will not prevent the reader from seeing the more general idea; all the tools we show here for Flask can be applied to other API technologies (e.g.
+  \ins{In general, what matters is that with the right backend, we can do this minimal configuration for any technology... }
+
+  To start using the \tool, one simply needs to one lines of code\footnote{we don't count the import statement that enables that line...} to their Flask web service\footnote{\ins{ In this paper we present the integration with APIs written in Python and Flask hoping that this will not prevent the reader from seeing the more general idea; all the tools we show here for Flask can be applied to other API technologies (e.g.
 Django) by simply providing a few back-end adapters in the right places.
@@
 
-  After (re)deployment of the application, the \tool becomes available at the \code{/dashboard} route of the Flask application.
-Further configuration is possible by adding additional statements before the binding definition, for example,
-
-  \begin{lstlisting}[style=custompython]
-  ...
-  dashboard.config.link = 'mydashboard'
-  dashboard.bind(flask_app)
-  \end{lstlisting}
-
-  allows for a custom route (\code{/mydashboard}) to the dashboard to be defined by the programmer. An external configuration file can be used instead, or in addition to this:
-
-  \begin{lstlisting}[style=custompython]
-  ...
-  dashboard.config.from_file('config.cfg')
-  ...
-  \end{lstlisting}
+  \ins{The dashboard can take advantage of the fact that it is monitoring an API...
+so it can simply just add one extra endpoint there.} After (re)deployment of the application, the interactive visualizations of the \tool become available at the \code{/dashboard} route of the Flask application.
 
   During binding, the \tool will search for all endpoints defined in the target application.
+\ins{PERFORMANCE DISCUSSION: this introspection, which looks up endpoints, happens only once at the startup of the service, so it is going to affect the API startup performance, but not the actual endpoint performance.
+In our case, the overhead here was quite small: TO MEASURE}.
+
+  In order to monitor an endpoint, the \tool creates a function wrapper for the API function that corresponds to the endpoint.
+This way, the wrapper will be executed whenever that API call is made before the actual function is called.
+The wrapper contains the code that takes care of monitoring an endpoint.
+Data collected by the wrappers are persisted in a local database.
+The SQLAlchemy Object Relational Mapper\footnote{\url{https://www.sqlalchemy.org/}} on top of SQLite\footnote{\url{https://www.sqlite.org/}} is used for this purpose.
+  \ins{DISCUSSION about how do to this in other languages? Java: AspectJ probably, etc.}
+
   These will be presented to the user in the tool web interface, where the user can select the ones that should be monitored, see \Fref{fig:sep}.
-The last access time of the endpoint by users, irrespective of whether it is selected by the user to be monitored or not, is also displayed as the means of identifying potentially obsolete endpoints.
 
-    % \mltp{a small screenshot of how the dashboard allows one to select the interesting }
-    \begin{figure}
+    \begin{figure}[h!]
       \centering
       \includegraphics[width=\linewidth]{selecting_endpoints.png}
-      \caption{Endpoints that are available for monitoring in the \zee API}
+      \caption{Once connected to an API the Dashboard presents the endpoints that are available for monitoring}
       \label{fig:sep}
@@
 
-  In order to monitor an endpoint, the \tool creates a function wrapper for the API function that corresponds to the endpoint.
-This way, the wrapper will be executed whenever that API call is made.
-The wrapper contains the code that takes care of monitoring an endpoint.
-Data collected by the wrappers are persisted in a local database.
-The SQLAlchemy Object Relational Mapper\footnote{\url{https://www.sqlalchemy.org/}} on top of SQLite\footnote{\url{https://www.sqlite.org/}} is used for this purpose.
+  This is the alternative to using annotations like other white-box tools (e.g. the other flask performance plugin).
+
+  \ins{PROMISE:} The last access time of the endpoint by users, irrespective of whether it is selected by the user to be monitored or not, is also displayed as the means of identifying potentially obsolete endpoints.
 
@@
 
-  \paragraph{Automated outlier detection and monitoring}
+\begin{figure*}[h!]
+  \centering
+  \subfloat[overview of the tracked endpoints]{\includegraphics[width=0.8\columnwidth]{overview}\label{fig:overview}}
+  \subfloat[number of requests per endpoint per day]{\includegraphics[width=.7\columnwidth]{number_of_requests_}\label{fig:aeu}}
+  \qquad
+  \subfloat[requests per hour heatmap]{\includegraphics[width=0.8\columnwidth]{daily_patterns_}\label{fig:dp}}
+  \subfloat[response time (in ms) per monitored endpoint view]{\includegraphics[width=.7\columnwidth]{endpoint_performance_}\label{fig:ep}}\qquad
+  \caption{Some of the available views\label{fig:views}}
+\end{figure*}
 
-  \ins{When an API is called from within a highly interactive application (as it is the case with the Zeeguu ecosystem) 
-  of particular interest to the API developers are performance {\em outliers}.
-Indeed, a translation request that takes three times more than expected can seriously decrease the perceived quality of the application.
-Thus,} identifying, collecting all appropriate data, and diagnosing the root causes of such outliers is especially critical in improving the quality of an application.
+\newpage
+\section{How Far Can We Go With One Line of Configuration?}
 
-  % In the context of the RESTful services support by Flask, this means request serving times that deviate from the average to an unexpected degree.
-
-  For this purpose the \tool tracks for every monitored endpoint a {\em running average} response time value\footnote{\ins{For performance reasons, we assume that the response times for the endpoints are normally distributed.
-Otherwise, more general density distribution information must be collected in real time.}}.
-When it detects that a given request is an outlier with respect to this past average running value, it triggers the {\em outlier data collection routine} which stores \ins{extra information} about the current execution environment.
-A configurable threshold with a default value of $2.5$ times the running average response time is used for this purpose.
-
-  For every detected outlier request, the \tool collects information about the current Python stack trace, CPU load, memory consumption, request parameters, etc.
-in order to allow the maintainer to investigate the causes of these exceptionally slow response times.
-In this way it is possible to get \ins{detailed insight into the operation of the application in the extreme cases without unnecessarily burdening it with logging this information for every request}.
-
-  \paragraph{User-level monitoring}
-
-  \ins{Sometimes, the same service endpoint will have different execution times for different users, based on the amount of data that a user has in their account.
-Gmail engineers admitted in 2010 that the performance of their service was dependent on the amount of emails a user would have in their Inbox \footnote{https://techcrunch.com/2010/03/14/gmail-slow/, visited September 23 2017}.
-}
-
-  \ins{If the API provider needs to understand the performance of their service for individual users, or for different user groups, the dashboard should have a way of distilling a user or a user group from the API request.
-For a large number of Flask applications that use the global \code{flask.request} object, the object} can be used to retrieve session information, \ins{as in the following snippet}:
-  % .
-This ability can be used in order to allow, for example, the identification of application users from the \tool, and the consequent aggregation of monitoring information on a per user basis as follows:
-
-  \begin{lstlisting}[style=custompython]  
-  # app specific way of extracting the user
-  # from a flask request object    
-  def get_user_id():
-  sid = int(flask.request.args['session'])
-  session = User.find_for_session(sid)
-  return user_id
-
-  # attaching the get_user_id function
-  dashboard.config.get_group_by = 
-      get_user_id
-
-  \end{lstlisting}
-  This feature, as discussed further in the following section, is important for the clustering of users depending on their application usage.
-It also provides a view on the distribution of the work load across the application endpoints on a per user basis that allows the identification of more or less popular endpoints for further performance improvements.
-\ml{this previous paragraph seems a bit reduntant.
-must revisit}
-
-  \paragraph{Evolution monitoring for Git-versioned projects }
-
-  Version control in \tool is supported in two ways.
-
-\begin{enumerate}
-
-  \item  \textit{Manual} version control requires the developer to tag each new version of the application with an appropriate version identifier~\cite{papazoglou2011managing} using the \code{APP\_VERSION} configuration parameter, for example by adding to the configuration file:
-
-    \begin{lstlisting}[style=custompython]
-    [dashboard]
-    APP_VERSION=<versionID>
-    \end{lstlisting}
-
-  \item \textit{Automatic} version control, aimed for use in conjunction with a version control system and currently only supporting Git\footnote{\url{https://git-scm.com/}}, is enabled by linking to the Git folder in the application source code:
-
-      \begin{lstlisting}[style=custompython]
-      ....
-      dashboard.config.git = '.git'
-      ....
-      \end{lstlisting}
-
-    With this extra configuration parameter, the \tool automatically detects the current version of the project by reading the hash code of the deployed version as soon as the application is deployed and started.
-Information about the version history of the application is also persisted in the tool database and is used during the collection of measurements for the application.
-
-\end{enumerate}
-
-  \paragraph{Preemptive monitoring}
-
-  \todo{the following might need a bit of rewrite}
-
-  The concept of {\em preemptive monitoring} of the application performance by means of instrumenting integration unit testing as the synthetic load is similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}, i.e.~testing service-based applications by using dedicated test input in parallel to its normal use and operation.
-The difference is that we take advantage of the capability of the CI framework (i.e.
-Travis in this case) to create an emulated ``live'' environment for integration testing purposes, and use unit testing as the dedicated test input in order to measure performance.
-
-  For example, \Fref{fig:builds} is a screenshot from the dashboard showing the measured response times for 5 consequent builds, with 180 iterations of the unit tests executed in total across all endpoints.
-The outliers on the right of the figure are due to initial requests \ins{which must wait for a boot-up phase of the API}.
-
-    \begin{figure}[h!]
-        \centering
-        \includegraphics[width=\columnwidth]{travis_builds}
-        \caption{Response times in 5 consequent Travis builds in the \zee case study}
-        \label{fig:builds}
-      \end{figure}
-
-  While this integration testing environment is different from the actual production deployment one, and the load used is purely synthetic, as will show in Section~\ref{sec:evaluation}, it can actually be used successfully as an early performance indicator for the application developer.
-\ins{Although predicting the performance of the application would require advanced statistical work, it can be succesfully used as an advanced warning system for the overall performance trend for given endpoints}.
-
-  \paragraph{Integration with Travis}
-  \ml{added new section... also, this has to be moved after introducing the pre-emptive monitoring since it's about running tests. switched the order}
-
-  \ins{The pre-emptive monitoring system} can be configured to work together with Continuous Integration (CI) frameworks like Travis\footnote{\url{https://travis-ci.org/}} that deal with automated integration testing.
-  The developer needs to define the unit test folder of the project, the URL that Travis should use for reporting its results, and the number of iterations for each unit test should be executed in order to generate a load for the application.
-
-  Each time a new build is created through Travis, the \tool automatically detects all available unit tests defined by the application developer and iterates through each one of them the number of times predefined by the developer while monitoring the response times for each request.
-The resulting measurements are persisted in a separate part of the tool database so as not to contaminate the ``live'' monitoring data.
-Beyond seeing the results of each Travis build in the \tool, this feature also allows for preemptive monitoring of the application, as we discuss in the following.
-
-  \ins{Besides functioning as an early warning system for performance, tracking the evolving performance of API tests serves a second purpose.
-To function as an anchor when the developer analyzes performance degradations.
-For example, when a developer sees that a given endpoint has become less performant after in a newly deployed version, they are able to investigate whether the performance degradation is visible also with the synthetic load.
-This would correspond to a performance degradation which is due to ``algorithmic performance degradation''.
-If on the other hand, the performance of the tests does not change between versions, the developer might conclude that the performance degradation could be due to the workload on the machines, or maybe to the workload mix of the users}
-
-  %By these means, an application developer can use unit tests as synthetic loads for driving the performance monitoring of the application before it reaches the production phase.
-As we will discuss further in Section~\ref{sec:testing}, this feature is integral in enabling preemptive performance monitoring of the application.
-
-\section{Performance Perspectives}
-\label{sec:views}
+A discussion must be added here about the types of useful information and problems that can be solved with these views.
 
@@
 
-  %In the remainder of the paper we present several of these perspectives.\footnote{We recommend obtaining a color version of this paper for better readability}
-
-  % The second endpoint conists of two parts, one of them being a table that shows for every monitored endpoint the number of hits it has gotten, the time it was last accessed and its average execution time.
-The second part is a view with four graphs which show:
-  % \begin{itemize} 
-  %   \item A heatmap of the total number of requests to the monitored endpoints
-  %   \item A stacked bar chart that shows the total number of requests to the monitored endpoints per endpoint per day
-  %   \item A boxplot graph showing the average execution time per version of the web service
-  %   \item A boxplot graph showing the average execution time for every monitored endpoint
-  % \end{itemize}
-
-\todo{Summarize the subsections extensively; evaluate which figures to keep and ideally take new screenshots for the ones we keep; consider adding two more views (like Fig. 3.2 of Patrick's thesis)}
-
-\begin{figure*}[t]
-	\centering
-	\subfloat[number of requests per endpoint per day]{\includegraphics[width=.9\columnwidth]{number_of_requests_}\label{fig:aeu}}
-	\subfloat[requests per hour heatmap]{\includegraphics[width=\columnwidth]{daily_patterns_}\label{fig:dp}}
-	\qquad
-	\subfloat[response time (in ms) per monitored endpoint view]{\includegraphics[width=.8\columnwidth]{endpoint_performance_}\label{fig:ep}}\qquad
-	\subfloat[The Performance Evolution of the \epTranslations endpoint]{\includegraphics[width=.95\columnwidth]{translation_endpoint_evolution_}\label{fig:tee}}
-	\caption{Some of the available views\label{fig:views}}
-\end{figure*}
+REQUIREMENT: Adding a time interval is important. 
+Otherwise the performance numbers in chart 1 are not good enough... also an outlier very far in the past, will make all the heatmap data points to look very pale in comparison forever after.
 
@@
 
-  \niceseparator
+      \begin{figure*}[h!]
+    \centering
+    \includegraphics[width=\linewidth]{outliers}
+    \caption{Automatically collected outlier information}
+    \label{fig:figure1}
+  \end{figure*}
 
-  \todo{remove the part about how to configure git, the discussion has been subsumed by Section~\ref{sec:tool}.}
+  \newpage
+
+  \section{Automated outlier detection and monitoring}
+
+  When an API is called from within a highly interactive application (as it is the case with the case study in this paper) 
+  of particular interest to the API developers are performance {\em outliers}.
+
+  Indeed, a translation request that takes three times more than expected can seriously decrease the perceived quality of the application.
+Thus, identifying, collecting all appropriate data, and diagnosing the root causes of such outliers is especially critical in improving the quality of an application.
+
+  % In the context of the RESTful services support by Flask, this means request serving times that deviate from the average to an unexpected degree.
+
+  For this purpose the \tool tracks for every monitored endpoint a {\em running average} response time value\footnote{\ins{For performance reasons, we assume that the response times for the endpoints are normally distributed.
+Otherwise, more general density distribution information must be collected in real time.}}.
+When it detects that a given request is an outlier with respect to this past average running value, it triggers the {\em outlier data collection routine} which stores \ins{extra information} about the current execution environment.
+A configurable threshold with a default value of $2.5$ times the running average response time is used for this purpose.
+
+  For every detected outlier request, the \tool collects information about the current Python stack trace, CPU load, memory consumption, request parameters, etc.
+in order to allow the maintainer to investigate the causes of these exceptionally slow response times.
+In this way it is possible to get \ins{detailed insight into the operation of the application in the extreme cases without unnecessarily burdening it with logging this information for every request}.
+
+The bottom panel shows the stack trace. 
+In this particular case, it is revealing for the developer to learn that at the time of the stack trace snapshot, the code was in the microsoft\_translator.py: indeed, the system uses as back-end multiple translators, and it has been observed that many of the outliers happen to be waiting in the microsoft translator.
+
+This information has to be corroborated with the observations that neither the memory nor the processor are overloaded at the moment.
+Thus this functionality in microsoft\_translator is really slow in itself, and this is not a result of the machine being overloaded for example.
+
+\subsection {A Special Type of Outlier: Exceptions}
+
+   TODO: probably something for the journal extension ...
+  statistics about which endpoints fail most often might be useful too.
+  same information as the outlier maybe?
+
+\begin{figure*}[h!]
+  \centering
+  \includegraphics[width=0.6\linewidth]{translation_endpoint_evolution_}
+  \caption{The Performance Evolution of the \epTranslations endpoint}
+  \label{fig:tee}
+\end{figure*}
+
+  \newpage
+  \section{Evolution monitoring }
+
+  When thinking about performance, 
+  one can not avoid thinking about the performance evolution.
+
+  Version control in \tool can be supported in two ways,
+  the developer explicitly states the current version, 
+  or the current version is automatically detected based
+  on some version control system.
 
   However, with the current configuration of the tool, it would be impossible for the maintainer to see the improvements resulting from the optimization.
-One way to do this is to add an extra line of configuration to allow \tool to find the git\footnote{\url{https://git-scm.com/}} folder of the deployed service:
 
-%  \begin{lstlisting}[caption=Configuring the \tool with the path to the .git folder enables the generation of evolutionary performance graphs, style=custompython]
-%
-%dashboard.config.git = 'path/to/.git'
-%
-%
-%    \end{lstlisting}
+  \ins{The latest STack Overflow Developer Survey: 88.4\% of **professional** developers use git.}  
+  If assume that the code that they run is deployed using .git, then with an extra line of configuration they can allow \tool to find the git\footnote{\url{https://git-scm.com/}} folder of the deployed service and automatically detect the version of the project that is running:
 
@@
 
-  With this extra configuration, the \tool can now automatically detect the current version of the project, and group measurements by version\footnote{Alternatively, the maintainer can add version identifiers manually for the web application through a configuration file if the system does not use git.}.
+  If the \tool can automatically detect the current version of the project by reading the .git configuration as soon as the API is started and can then group measurements by version\footnote{Alternatively, the maintainer can add version identifiers manually for the web application through a configuration file if the system does not use git.}.
 \Fref{fig:tee} is a zoomed-in version of such a view for \epTranslations with versions increasing from top to bottom
-
-%    \begin{figure}[h!]
-%      \centering
-%      \includegraphics[width=0.95\linewidth]{translation_endpoint_evolution_}
-%      \caption{Visualizing The Performance Evolution of the \epTranslations endpoint}
-%      \label{fig:tee}
-%    \end{figure}
 
@@
 
-\paragraph{User Experience}
+  CHALLENGE: Supporting other types of version control... not real challenge, as we said 88percent of devs use git. 
+  CHALLENGE 2: detecting minor versions which don't need to be tracked, since they didn't touch the performance of the system. E.g. a modification of the README, etc.
+
+\todo{TODO: endpoint evolution }
+
+\begin{figure*}[!ht]
+  \centering
+  \includegraphics[width=0.4\linewidth]{time_per_user}
+  \caption{The \epFeedItems shows a very high variability across users}
+  \label{fig:tpu}
+\end{figure*}
+
+\newpage
+\section{Grouping Requests}
 \label{sec:user}
@@
 
-% In the Zeeguu case study, one of the slowest endpoints, and one with the highest variability as shown in \Fref{fig:ep} is \epFeedItems: it retrieves a list of recommended articles for a given user.
-However, since a user can be subscribed to anything from one to three dozen article sources, and since the computation of the difficulty is personalized and it is slow, the variability in time among users is likely to be very large.
-
-\begin{figure}[!ht]
-	\centering
-	\includegraphics[width=0.85\linewidth]{time_per_user}
-	\caption{The \epFeedItems shows a very high variability across users}
-	\label{fig:tpu}
-\end{figure}
-
 A \perspective{Per-User Performance} perspective should show the different response times for different users.
@@
 
-% \niceseparator
-
 The limitation of the previous view is that it does not present the information also on a per version basis.
@@
 Figure \ref{fig:tuv} presents such a perspective by mapping the average execution time for a given user (lines) and given version (columns) on the area of the corresponding circle.
+
 The colors represent users.
@@
 number of sources to which the user is registered) is the reason for the variation in response times.
+\ins{one can see that for user 1 performance degrades over the }
 
@@
   \centering
-	\includegraphics[width=0.9\linewidth]{time_per_user_per_version}
+  \includegraphics[width=0.7\linewidth]{time_per_user_per_version}
   \caption{This perspective shows that the evolution of response times for individual users (horizontal lines) across versions (the x-axis)}
@@
 
-%\section{Preemptive Monitoring}
-%\label{sec:testing}
-%
-%%\todo{discuss here idea of using unit testing as an early performance indicator for evolving services (`preemptive monitoring'?); add a side-by-side figure comparing unit testing vs live system performance per version, maybe calculate error in trend curves?
-find a better term than prediction for the section title (plus introduction)}
-%
-%\todo{Similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}, i.e.~testing service-based applications by using dedicated test input in parallel to its normal use and operation.
-The difference is that we take advantage of the capability of Travis to create an emulated live environment for integration testing purposes, and use unit testing as the dedicated test input in order to measure performance.
-What we are going provide evidence for in the following is that we can indeed use this preemptive monitoring concept as an early performance indicator for evolving services.}
-%
-%The developer needs to define the unit test folder of the project, the URL that Travis should use for reporting its results, and the number of iterations for each unit test should be executed in order to generate
+  \newpage
+  \section{Preemptive monitoring}
 
-\section{Evaluation}
-\label{sec:evaluation}
+  \ins{INTRO - LINKING WITH PREVIOUS PARAGRAPH: the previous evolutionary view crucial for 
+  understanding the impact of software evolution on performance.
+  However, just by observing this one can't be sure whether 
+  performance changes because of the code or because of the payload.}
 
-%\todo{evaluation of the \tool across the following dimensions: ease of use, effectiveness, overhead to the system; for the last part we need to do the overhead measurement experiment, for the other two we base it on the case study and bring over whatever is necessary from the discussion section}
+  \todo{the following might need a bit of rewrite}
 
-\subsection{Effectiveness \& Ease of Use}
+  The concept of {\em preemptive monitoring} of the application performance by means of instrumenting integration unit testing as the synthetic load is similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}, i.e.~testing service-based applications by using dedicated test input in parallel to its normal use and operation.
+The difference is that we take advantage of the capability of the CI framework (i.e.
+Travis in this case) to create an emulated ``live'' environment for integration testing purposes, and use unit testing as the dedicated test input in order to measure performance.
 
-\todo{@M: move here the war stories, and discuss how useful the whole thing was to the application maintainer}
+  For example, \Fref{fig:builds} is a screenshot from the dashboard showing the measured response times for 5 consequent builds, with 180 iterations of the unit tests executed in total across all endpoints.
+The outliers on the right of the figure are due to initial requests \ins{which must wait for a boot-up phase of the API}.
 
-\subsection{Efficiency of Preemptive Monitoring}
+    \begin{figure}[h!]
+        \centering
+        \includegraphics[width=\columnwidth]{travis_builds}
+        \caption{Response times in 5 consequent Travis builds in the \zee case study}
+        \label{fig:builds}
+      \end{figure}
+
+  While this integration testing environment is different from the actual production deployment one, and the load used is purely synthetic, as will show in Section~\ref{sec:evaluation}, it can actually be used successfully as an early performance indicator for the application developer.
+\ins{Although predicting the performance of the application would require advanced statistical work, it can be succesfully used as an advanced warning system for the overall performance trend for given endpoints}.
+
+  \paragraph{Integration with Travis}
+  \ml{added new section... also, this has to be moved after introducing the pre-emptive monitoring since it's about running tests. switched the order}
+
+  \ins{The pre-emptive monitoring system} can be configured to work together with Continuous Integration (CI) frameworks like Travis\footnote{\url{https://travis-ci.org/}} that deal with automated integration testing.
+  The developer needs to define the unit test folder of the project, the URL that Travis should use for reporting its results, and the number of iterations for each unit test should be executed in order to generate a load for the application.
+
+  Each time a new build is created through Travis, the \tool automatically detects all available unit tests defined by the application developer and iterates through each one of them the number of times predefined by the developer while monitoring the response times for each request.
+The resulting measurements are persisted in a separate part of the tool database so as not to contaminate the ``live'' monitoring data.
+Beyond seeing the results of each Travis build in the \tool, this feature also allows for preemptive monitoring of the application, as we discuss in the following.
+
+  \ins{Besides functioning as an early warning system for performance, tracking the evolving performance of API tests serves a second purpose.
+To function as an anchor when the developer analyzes performance degradations.
+For example, when a developer sees that a given endpoint has become less performant after in a newly deployed version, they are able to investigate whether the performance degradation is visible also with the synthetic load.
+This would correspond to a performance degradation which is due to ``algorithmic performance degradation''.
+If on the other hand, the performance of the tests does not change between versions, the developer might conclude that the performance degradation could be due to the workload on the machines, or maybe to the workload mix of the users}
+
+  %By these means, an application developer can use unit tests as synthetic loads for driving the performance monitoring of the application before it reaches the production phase.
+As we will discuss further in Section~\ref{sec:testing}, this feature is integral in enabling preemptive performance monitoring of the application.
 
@@
 
-\begin{figure}
+\begin{figure}[h!]
   \centering
@@
 
-\subsection{Discussion}
+\section{Performance Perspectives}
+\label{sec:views}
+
+  %In the remainder of the paper we present several of these perspectives.\footnote{We recommend obtaining a color version of this paper for better readability}
+
+  % The second endpoint conists of two parts, one of them being a table that shows for every monitored endpoint the number of hits it has gotten, the time it was last accessed and its average execution time.
+The second part is a view with four graphs which show:
+  % \begin{itemize} 
+  %   \item A heatmap of the total number of requests to the monitored endpoints
+  %   \item A stacked bar chart that shows the total number of requests to the monitored endpoints per endpoint per day
+  %   \item A boxplot graph showing the average execution time per version of the web service
+  %   \item A boxplot graph showing the average execution time for every monitored endpoint
+  % \end{itemize}
+
+\todo{Summarize the subsections extensively; evaluate which figures to keep and ideally take new screenshots for the ones we keep; consider adding two more views (like Fig. 3.2 of Patrick's thesis)}
+
+% \begin{figure*}[t]
+% 	\centering
+% 	\subfloat[number of requests per endpoint per day]{\includegraphics[width=.9\columnwidth]{number_of_requests_}\label{fig:aeu}}
+% 	\subfloat[requests per hour heatmap]{\includegraphics[width=\columnwidth]{daily_patterns_}\label{fig:dp}}
+% 	\qquad
+% 	\subfloat[response time (in ms) per monitored endpoint view]{\includegraphics[width=.8\columnwidth]{endpoint_performance_}\label{fig:ep}}\qquad
+% 	\subfloat[The Performance Evolution of the \epTranslations endpoint]{\includegraphics[width=.95\columnwidth]{translation_endpoint_evolution_}\label{fig:tee}}
+% 	\caption{Some of the available views\label{fig:views}}
+% \end{figure*}
+
+% In the Zeeguu case study, one of the slowest endpoints, and one with the highest variability as shown in \Fref{fig:ep} is \epFeedItems: it retrieves a list of recommended articles for a given user.
+However, since a user can be subscribed to anything from one to three dozen article sources, and since the computation of the difficulty is personalized and it is slow, the variability in time among users is likely to be very large.
+
+% \niceseparator
+
+%\section{Preemptive Monitoring}
+%\label{sec:testing}
+%
+%%\todo{discuss here idea of using unit testing as an early performance indicator for evolving services (`preemptive monitoring'?); add a side-by-side figure comparing unit testing vs live system performance per version, maybe calculate error in trend curves?
+find a better term than prediction for the section title (plus introduction)}
+%
+%\todo{Similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}, i.e.~testing service-based applications by using dedicated test input in parallel to its normal use and operation.
+The difference is that we take advantage of the capability of Travis to create an emulated live environment for integration testing purposes, and use unit testing as the dedicated test input in order to measure performance.
+What we are going provide evidence for in the following is that we can indeed use this preemptive monitoring concept as an early performance indicator for evolving services.}
+%
+%The developer needs to define the unit test folder of the project, the URL that Travis should use for reporting its results, and the number of iterations for each unit test should be executed in order to generate
+
+\section{Evaluation}
+\label{sec:evaluation}
+
+%\todo{evaluation of the \tool across the following dimensions: ease of use, effectiveness, overhead to the system; for the last part we need to do the overhead measurement experiment, for the other two we base it on the case study and bring over whatever is necessary from the discussion section}
+
+\section{Discussion}
 
 \todo{use this subsection to discuss limitations (from the previous evaluation section) and add the need to be able to handle the horizontal scaling of the application; remove sectioning and summarize briefly limitations that have already been discussed in the VISSOFT paper/move material to other sections; consider renaming as limitations or something similar}
+
+  \subsection{Distributed System Monitoring}
+  \todo{To think about}
+  Supporting situations where the API is deployed across multiple containers for example.
 
@@
 
+\newpage
+\appendix
+
+  \subsection{Changing the Default /dashboard Endpoint }
+
+  THIS STUFF SHOULD BE MOVED TO THE APPENDIX.
+And in a footnote we just mention that ...
+\ins{REQUIREMENT}: The system must provide an easy way to add more specific configurations if needed.
+A way of overwriting the common sense defaults.
+
+  Further configuration is not needed possible by adding additional statements before the binding definition, for example,
+
+  \begin{lstlisting}[style=custompython]
+  ...
+  dashboard.config.link = 'mydashboard'
+  dashboard.bind(flask_app)
+  \end{lstlisting}
+
+  allows for a custom route (\code{/mydashboard}) to the dashboard to be defined by the programmer. An external configuration file can be used instead, or in addition to this:
+
+  \begin{lstlisting}[style=custompython]
+  ...
+  dashboard.config.from_file('config.cfg')
+  ...
+  \end{lstlisting}
+
+  \subsection{For those who don't use GIT}
+
+    \textit{Manual} version control requires the developer to tag each new version of the application with an appropriate version identifier~\cite{papazoglou2011managing} using the \code{APP\_VERSION} configuration parameter, for example by adding to the configuration file:
+
+      \begin{lstlisting}[style=custompython]
+      [dashboard]
+      APP_VERSION=<versionID>
+      \end{lstlisting}
+
 % references section

### 2018-03-21 restructured.tex
small restructuring

@@
   During binding, the \tool will search for all endpoints defined in the target application.
-\ins{PERFORMANCE DISCUSSION: this introspection, which looks up endpoints, happens only once at the startup of the service, so it is going to affect the API startup performance, but not the actual endpoint performance.
-In our case, the overhead here was quite small: TO MEASURE}.
 
@@
 
+In the Zeeguu case study, one of the slowest endpoints, and one with the highest variability as shown in \Fref{fig:ep} is \epFeedItems: it retrieves a list of recommended articles for a given user.
+However, since a user can be subscribed to anything from one to three dozen article sources, and since the computation of the difficulty is personalized and it is slow, the variability in time among users is likely to be very large.
+
 \begin{figure*}[h!]
   \centering
-  \subfloat[overview of the tracked endpoints]{\includegraphics[width=0.8\columnwidth]{overview}\label{fig:overview}}
-  \subfloat[number of requests per endpoint per day]{\includegraphics[width=.7\columnwidth]{number_of_requests_}\label{fig:aeu}}
-  \qquad
-  \subfloat[requests per hour heatmap]{\includegraphics[width=0.8\columnwidth]{daily_patterns_}\label{fig:dp}}
-  \subfloat[response time (in ms) per monitored endpoint view]{\includegraphics[width=.7\columnwidth]{endpoint_performance_}\label{fig:ep}}\qquad
+  \subfloat[number of requests per endpoint per day]{\includegraphics[width=.9\columnwidth]{number_of_requests_}\label{fig:aeu}}
+  \subfloat[requests per hour heatmap]{\includegraphics[width=\columnwidth]{daily_patterns_}\label{fig:dp}}
   \caption{Some of the available views\label{fig:views}}
@@
 \newpage
-\section{How Far Can We Go With One Line of Configuration?}
+\section{Service Utilization}
+\label{sec:util}
 
-A discussion must be added here about the types of useful information and problems that can be solved with these views.
+\ins{A discussion must be added here about the types of useful information and problems that can be solved with these views.}
 
-  There are three main categories of visual perspectives that are available using \tool:
-  \begin{enumerate}
-    \item \textit{Service Utilization} presents information about the usage of all endpoints of interest to the developer,
-    \item \textit{Endpoint Performance} presents response times of the various service endpoints, and
-    \item \textit{User Experience} presents information about the user-perceived performance of monitored service endpoints.
-  \end{enumerate}
-
-REQUIREMENT: Adding a time interval is important. 
-Otherwise the performance numbers in chart 1 are not good enough... also an outlier very far in the past, will make all the heatmap data points to look very pale in comparison forever after.
-
-\paragraph{Service Utilization}
-\label{sec:util}
+\textit{Service Utilization} presents information about the usage of all endpoints of interest to the developer,
 
@@
 
+Simplest utilization perspective:
+
+\begin{figure}[h!]
+  \centering
+  \includegraphics[width=0.5\linewidth]{endpoint-call-count}
+  \caption{Caption here}
+  \label{fig:figure1}
+\end{figure}
+
+% REQUIREMENT: Adding a time interval is important. 
+% Otherwise the performance numbers in chart 1 are not good enough... also an outlier very far in the past, will make all the heatmap data points to look very pale in comparison forever after.
+
 Figure \ref{fig:aeu} shows a first perspective on endpoint utilization that \tool provides: a stacked bar chart of the number of hits to various endpoints grouped by day\footnote{Endpoint colors are the same in different views}.
 Figure~\ref{fig:aeu} in particular shows that at its peak the API has about 12.000 hits per day.
+
 The way users interact with the platform can also be inferred since the endpoints are indicators of different activity types, e.g.:
@@
 
-\paragraph{Endpoint Performance}
+\newpage
+\section{Endpoint Performance}
 \label{sec:perf}
@@
 
-  % \ml{Thijs and Patrick...
-the visualization people will complain when they see that we have different colors for the same endpoint in different graphs.
-Can we insure that there is consistency in colors?
-Simplest trick would be to obtain the color by hashing the name of the endpoint...
-in that case the same endpoint would have the same color in various graphs.}
-  % This is supported in the latest version of the dashboard
-
-%  \begin{figure}[!ht]
-%    \centering
-%    \includegraphics[width=0.95\linewidth]{endpoint_performance_}
-%    \caption{The response time (in ms) per monitored endpoint view allows for identifying performance variability and balancing issues}
-%    \label{fig:ep}
-%  \end{figure}
+ \begin{figure}[!ht]
+   \centering
+   \includegraphics[width=0.95\linewidth]{endpoint_performance_}
+   \caption{The response time (in ms) per monitored endpoint view allows for identifying performance variability and balancing issues}
+   \label{fig:ep}
+ \end{figure}
 
@@
 \Fref{fig:aeu} this endpoint is one of the most used in the system.
+
+\begin{figure}[h!]
+  \centering
+  \subfloat[overview of the tracked endpoints]{\includegraphics[width=0.8\columnwidth]{overview}\label{fig:overview}}
+  \caption{Some of the available views\label{fig:views}}
+\end{figure}
+
+  \newpage
+  \section{Evolution monitoring }
+
+  \begin{figure}[h!]
+  \centering
+  \includegraphics[width=0.8\linewidth]{translation_endpoint_evolution_}
+  \caption{The Performance Evolution of the \epTranslations endpoint}
+  \label{fig:tee}
+\end{figure}
+
+  When thinking about performance, 
+  one can not avoid thinking about the performance evolution.
+
+  Version control in \tool can be supported in two ways,
+  the developer explicitly states the current version, 
+  or the current version is automatically detected based
+  on some version control system.
+
+  However, with the current configuration of the tool, it would be impossible for the maintainer to see the improvements resulting from the optimization.
+
+  \ins{The latest STack Overflow Developer Survey: 88.4\% of **professional** developers use git.}  
+  If assume that the code that they run is deployed using .git, then with an extra line of configuration they can allow \tool to find the git\footnote{\url{https://git-scm.com/}} folder of the deployed service and automatically detect the version of the project that is running:
+
+    \begin{lstlisting}[style=custompython]
+
+      dashboard.config.git = 'path/to/.git'
+
+    \end{lstlisting}
+
+  If the \tool can automatically detect the current version of the project by reading the .git configuration as soon as the API is started and can then group measurements by version\footnote{Alternatively, the maintainer can add version identifiers manually for the web application through a configuration file if the system does not use git.}.
+\Fref{fig:tee} is a zoomed-in version of such a view for \epTranslations with versions increasing from top to bottom
+
+  This view confirms that the performance of the translation endpoint improved in the recent versions: the median of the last three versions is constantly moving towards the left, and progresses from 1.4 seconds (in the top-most box plot in \Fref{fig:tee}) to 0.8 in the latest version (bottom-most box plot).
+
+  CHALLENGE: Supporting other types of version control... not real challenge, as we said 88percent of devs use git. 
+  CHALLENGE 2: detecting minor versions which don't need to be tracked, since they didn't touch the performance of the system. E.g. a modification of the README, etc.
+
+\todo{TODO: endpoint evolution }
 
@@
   \end{figure*}
-
   \newpage
-
   \section{Automated outlier detection and monitoring}
@@
   same information as the outlier maybe?
-
-\begin{figure*}[h!]
-  \centering
-  \includegraphics[width=0.6\linewidth]{translation_endpoint_evolution_}
-  \caption{The Performance Evolution of the \epTranslations endpoint}
-  \label{fig:tee}
-\end{figure*}
-
-  \newpage
-  \section{Evolution monitoring }
-
-  When thinking about performance, 
-  one can not avoid thinking about the performance evolution.
-
-  Version control in \tool can be supported in two ways,
-  the developer explicitly states the current version, 
-  or the current version is automatically detected based
-  on some version control system.
-
-  However, with the current configuration of the tool, it would be impossible for the maintainer to see the improvements resulting from the optimization.
-
-  \ins{The latest STack Overflow Developer Survey: 88.4\% of **professional** developers use git.}  
-  If assume that the code that they run is deployed using .git, then with an extra line of configuration they can allow \tool to find the git\footnote{\url{https://git-scm.com/}} folder of the deployed service and automatically detect the version of the project that is running:
-
-    \begin{lstlisting}[style=custompython]
-
-      dashboard.config.git = 'path/to/.git'
-
-    \end{lstlisting}
-
-  If the \tool can automatically detect the current version of the project by reading the .git configuration as soon as the API is started and can then group measurements by version\footnote{Alternatively, the maintainer can add version identifiers manually for the web application through a configuration file if the system does not use git.}.
-\Fref{fig:tee} is a zoomed-in version of such a view for \epTranslations with versions increasing from top to bottom
-
-  This view confirms that the performance of the translation endpoint improved in the recent versions: the median of the last three versions is constantly moving towards the left, and progresses from 1.4 seconds (in the top-most box plot in \Fref{fig:tee}) to 0.8 in the latest version (bottom-most box plot).
-
-  CHALLENGE: Supporting other types of version control... not real challenge, as we said 88percent of devs use git. 
-  CHALLENGE 2: detecting minor versions which don't need to be tracked, since they didn't touch the performance of the system. E.g. a modification of the README, etc.
-
-\todo{TODO: endpoint evolution }
 
@@
 
+\ins{PERFORMANCE DISCUSSION: this introspection, which looks up endpoints, happens only once at the startup of the service, so it is going to affect the API startup performance, but not the actual endpoint performance.
+In our case, the overhead here was quite small: TO MEASURE}.
+
   %In the remainder of the paper we present several of these perspectives.\footnote{We recommend obtaining a color version of this paper for better readability}
@@
 
-\todo{Summarize the subsections extensively; evaluate which figures to keep and ideally take new screenshots for the ones we keep; consider adding two more views (like Fig. 3.2 of Patrick's thesis)}
-
-% \begin{figure*}[t]
-% 	\centering
-% 	\subfloat[number of requests per endpoint per day]{\includegraphics[width=.9\columnwidth]{number_of_requests_}\label{fig:aeu}}
-% 	\subfloat[requests per hour heatmap]{\includegraphics[width=\columnwidth]{daily_patterns_}\label{fig:dp}}
-% 	\qquad
-% 	\subfloat[response time (in ms) per monitored endpoint view]{\includegraphics[width=.8\columnwidth]{endpoint_performance_}\label{fig:ep}}\qquad
-% 	\subfloat[The Performance Evolution of the \epTranslations endpoint]{\includegraphics[width=.95\columnwidth]{translation_endpoint_evolution_}\label{fig:tee}}
-% 	\caption{Some of the available views\label{fig:views}}
-% \end{figure*}
+\todo{evaluate which figures to keep and ideally take new screenshots for the ones we keep; consider adding two more views (like Fig. 3.2 of Patrick's thesis)}
 

### 2018-03-28 restructured.tex
.

@@
 
-    \begin{figure*}[ht!]
-      \centering
-      \includegraphics[width=0.8\linewidth]{interceptor}
-      \caption{The first thing that needs to be done, is to decorate the application object with an INTERCEPTOR}
-      \label{fig:sep}
-    \end{figure*}
-
-\newpage
-
-\section{Monitoring Service Utilization}
-
-REMEMBER THE GOAL: Our main goal with the \tool issues to allow the integration of the dashboard with an API with minimal effort.
-
-\ins{The first important question that such a 
-service monitoring system should provide is 
-information about what endpoints are being
-used and by whom}.
-
-TODO: Find references, arguments, related work
-that provide more details about why understanding
-who uses endpoints, and which are used are important.
-
-This is clearly the case in static analysis (see paper
-by haenni and lungu...) but should even more be the
-case in APIs.
-
-  \newpage
-TODO: TALK ABOUT THE DATABASE...
-
-  As discussed in the introductory section, in previous work~\cite{vogel2017low} we introduced \tool, a drop-in Python library that allows developers to monitor their Flask-based Python web applications with minimal effort.
-%
-  The \tool is implemented for Python 3.6 and is available on the Python Package Index repository\footnote{\url{https://pypi.python.org/pypi/flask-monitoring-dashboard/1.8}} from where it can be installed by running \install from the command line.
-%  
-  The source code of the \tool is published under a permissive MIT license and is available on GitHub\footnote{\url{https://github.com/flask-dashboard}}.
-
   \newpage
@@
 
+\begin{figure*}[h!]
+  \centering
+  \subfloat[overview of the tracked endpoints]{\includegraphics[width=1.5\columnwidth]{overview}\label{fig:overview}}
+  \caption{Some of the available views\label{fig:views}}
+\end{figure*}
+
 \newpage
@@
 
-\begin{figure}[h!]
+  \begin{figure*}[h!]
     \centering
-  \subfloat[overview of the tracked endpoints]{\includegraphics[width=0.8\columnwidth]{overview}\label{fig:overview}}
-  \caption{Some of the available views\label{fig:views}}
-\end{figure}
+    \includegraphics[width=\linewidth]{outliers}
+    \caption{Automatically collected outlier information}
+    \label{fig:figure1}
+  \end{figure*}
+  \newpage
+  \subsection{Performance Insight: Automated outlier detection and monitoring}
+
+  When an API is called from within a highly interactive application (as it is the case with the case study in this paper) 
+  of particular interest to the API developers are performance {\em outliers}.
+
+  Indeed, a translation request that takes three times more than expected can seriously decrease the perceived quality of the application.
+Thus, identifying, collecting all appropriate data, and diagnosing the root causes of such outliers is especially critical in improving the quality of an application.
+
+  % In the context of the RESTful services support by Flask, this means request serving times that deviate from the average to an unexpected degree.
+
+  For this purpose the \tool tracks for every monitored endpoint a {\em running average} response time value\footnote{\ins{For performance reasons, we assume that the response times for the endpoints are normally distributed.
+Otherwise, more general density distribution information must be collected in real time.}}.
+When it detects that a given request is an outlier with respect to this past average running value, it triggers the {\em outlier data collection routine} which stores \ins{extra information} about the current execution environment.
+A configurable threshold with a default value of $2.5$ times the running average response time is used for this purpose.
+
+  For every detected outlier request, the \tool collects information about the current Python stack trace, CPU load, memory consumption, request parameters, etc.
+in order to allow the maintainer to investigate the causes of these exceptionally slow response times.
+In this way it is possible to get \ins{detailed insight into the operation of the application in the extreme cases without unnecessarily burdening it with logging this information for every request}.
+
+The bottom panel shows the stack trace. 
+In this particular case, it is revealing for the developer to learn that at the time of the stack trace snapshot, the code was in the microsoft\_translator.py: indeed, the system uses as back-end multiple translators, and it has been observed that many of the outliers happen to be waiting in the microsoft translator.
+
+This information has to be corroborated with the observations that neither the memory nor the processor are overloaded at the moment.
+Thus this functionality in microsoft\_translator is really slow in itself, and this is not a result of the machine being overloaded for example.
+
+\subsection {A Special Type of Outlier: Exceptions}
+
+   TODO: probably something for the journal extension ...
+  statistics about which endpoints fail most often might be useful too.
+  same information as the outlier maybe?
+
+\begin{figure*}[h!]
+  \centering
+  \includegraphics[width=0.8\linewidth]{utilization-evolution}
+  \caption{The Performance Evolution of the \epTranslations endpoint}
+  \label{fig:tee}
+\end{figure*}
 
@@
 \todo{TODO: endpoint evolution }
-
-  \begin{figure*}[h!]
-    \centering
-    \includegraphics[width=\linewidth]{outliers}
-    \caption{Automatically collected outlier information}
-    \label{fig:figure1}
-  \end{figure*}
-  \newpage
-  \section{Automated outlier detection and monitoring}
-
-  When an API is called from within a highly interactive application (as it is the case with the case study in this paper) 
-  of particular interest to the API developers are performance {\em outliers}.
-
-  Indeed, a translation request that takes three times more than expected can seriously decrease the perceived quality of the application.
-Thus, identifying, collecting all appropriate data, and diagnosing the root causes of such outliers is especially critical in improving the quality of an application.
-
-  % In the context of the RESTful services support by Flask, this means request serving times that deviate from the average to an unexpected degree.
-
-  For this purpose the \tool tracks for every monitored endpoint a {\em running average} response time value\footnote{\ins{For performance reasons, we assume that the response times for the endpoints are normally distributed.
-Otherwise, more general density distribution information must be collected in real time.}}.
-When it detects that a given request is an outlier with respect to this past average running value, it triggers the {\em outlier data collection routine} which stores \ins{extra information} about the current execution environment.
-A configurable threshold with a default value of $2.5$ times the running average response time is used for this purpose.
-
-  For every detected outlier request, the \tool collects information about the current Python stack trace, CPU load, memory consumption, request parameters, etc.
-in order to allow the maintainer to investigate the causes of these exceptionally slow response times.
-In this way it is possible to get \ins{detailed insight into the operation of the application in the extreme cases without unnecessarily burdening it with logging this information for every request}.
-
-The bottom panel shows the stack trace. 
-In this particular case, it is revealing for the developer to learn that at the time of the stack trace snapshot, the code was in the microsoft\_translator.py: indeed, the system uses as back-end multiple translators, and it has been observed that many of the outliers happen to be waiting in the microsoft translator.
-
-This information has to be corroborated with the observations that neither the memory nor the processor are overloaded at the moment.
-Thus this functionality in microsoft\_translator is really slow in itself, and this is not a result of the machine being overloaded for example.
-
-\subsection {A Special Type of Outlier: Exceptions}
-
-   TODO: probably something for the journal extension ...
-  statistics about which endpoints fail most often might be useful too.
-  same information as the outlier maybe?
 
@@
 Pearson correlation $r(3)=.93, p=.02$
+
+    \begin{figure*}[ht!]
+      \centering
+      \includegraphics[width=0.8\linewidth]{interceptor}
+      \caption{The first thing that needs to be done, is to decorate the application object with an INTERCEPTOR}
+      \label{fig:sep}
+    \end{figure*}
+
+\newpage
+
+\section{Monitoring Service Utilization}
+
+REMEMBER THE GOAL: Our main goal with the \tool issues to allow the integration of the dashboard with an API with minimal effort.
+
+\ins{The first important question that such a 
+service monitoring system should provide is 
+information about what endpoints are being
+used and by whom}.
+
+TODO: Find references, arguments, related work
+that provide more details about why understanding
+who uses endpoints, and which are used are important.
+
+This is clearly the case in static analysis (see paper
+by haenni and lungu...) but should even more be the
+case in APIs.
+
+  \newpage
+TODO: TALK ABOUT THE METAMODEL...
+
+    \begin{figure}[ht!]
+      \centering
+        \includegraphics[width=0.8\linewidth]{db_schema}
+        \caption{The model that supports the views presented in this paper}
+        \label{fig:sep}
+    \end{figure}
+
+  As discussed in the introductory section, in previous work~\cite{vogel2017low} we introduced \tool, a drop-in Python library that allows developers to monitor their Flask-based Python web applications with minimal effort.
+%
+  The \tool is implemented for Python 3.6 and is available on the Python Package Index repository\footnote{\url{https://pypi.python.org/pypi/flask-monitoring-dashboard/1.8}} from where it can be installed by running \install from the command line.
+%  
+  The source code of the \tool is published under a permissive MIT license and is available on GitHub\footnote{\url{https://github.com/flask-dashboard}}.
 

### 2018-04-03 architecture.tex
split the huge file in smaller ones.

--- 2018-03-28 restructured.tex
@@
-\documentclass[conference]{IEEEtran}
-	\pdfpagewidth=8.5truein
-	\pdfpageheight=11truein
+%!TEX root=restructured.tex
 
-\usepackage{graphicx}
-\usepackage{subfig}
-\usepackage[bookmarks=false]{hyperref}
-\usepackage{xspace}
-\usepackage{listings}
-\usepackage[usenames, dvipsnames]{color}
-
-\usepackage{booktabs}
-
-\graphicspath{{./img/}}
-
-%\hyphenation{op-tical net-works semi-conduc-tor}
-
-\input{useful_defines}
-
-\begin{document}
-
-%
-\title{Agile Monitoring of Evolving Web Services}
-
-\author{
-\IEEEauthorblockN{Patrick Vogel, Thijs Klooster, Vasilios Andrikopoulos, Mircea Lungu}\\
-Johann Bernoulli Institute for Mathematics and Computer Science\\
-University of Groningen, Netherland\\
-Email: \{t.klooster.1,p.p.vogel\}@student.rug.nl, \{v.andrikopoulos,m.f.lungu\}@rug.nl 
-}
-
-% make the title area
-\maketitle
-
-\begin{abstract}
-\todo{Update abstract \& title}
-  Tens of thousands of web applications are written in Flask, a Python-based web framework.
-Despite a rich ecosystem of extensions, there is none that supports the developer in gaining insight into the evolving performance of their service.
-In this paper, we introduce \tool, a library that addresses this problem.
-We present the ease with which the library can be integrated in an already existing web application, discuss some of the visualization perspectives that the library provides and point to some future challenges for similar libraries.
-
-\end{abstract}
-
-\section{Introduction}
-
-\todo{Intro to be rewritten; refer to VISSOFT paper as previous work, explain focus is on integration with version control/CI for evolution support, and use of unit tests as early performance indicator}
-
-\todo{motivate with: monitoring in a world which is becoming more and more service-oriented is essential because it allows to perform three main types of actions: system adaptations to provide the request service at the desired level of quality, enabling flexibility in dealing with changing requirements and modes of operation, and enabling operational awareness through dashboards organizing the collected data~\cite{pernici2016monitoring}.
-In this work we focus on the last part and we discuss how a minimal effort probe-based solution for a particular type of web services can be used to facilitate the other two types of actions by involving the developer.}
-
-\todo{explain: in~\cite{vogel2017low} we introduced the \tool an a high level with a focus on presenting its performance visualization aspects; in this work we only summarize these features by means of introducing some of the provided functionalities of the \tool.
-We focus on discussing how the tool is implemented and operating, provide a deeper look at its capabilities for integration with version control and continuous integration environments, introduce a mechanism for performance prediction, and provide an evaluation of the proposed approach based on a case study.
-}
-
-%Every system is a distributed system nowadays \cite{cavage2013there}.
-Indeed a very large number of applications and web applications are nowadays implemented as two-tier architectures with a front-end implemented with web technologies and a service back-end.
-%\ml{I'm not completely happy with this paragraph}
-{\em There is no getting around it: you are building a distributed system} argues a recent article \cite{cavage2013there}.
-Indeed, even the simplest second-year student project is a web application implemented as two-tier architecture with a Javascript/HTML5 front-end a service backend, usually a REST API.
-
-% \hfill mds
-% Many contemporary programming languages are offering libraries, modules, or frameworks that facilitate the development of such architectures. 
-Python is one of the most popular programming language choices for implementing the back-end of web applications.
-GitHub contains more than 500K open source Python projects and the Tiobe Index\footnote{TIOBE programming community index is a measure of popularity of programming languages, created and maintained by the TIOBE Company based in Eindhoven, the Netherlands} ranks Python as the 4th most popular programming language as of June 2016.
-
-% possible flask summary
-Within the Python community, Flask\footnote{\url{http://flask.pocoo.org/}} is a very popular web framework\footnote{More than 25K projects on GitHub (5\% of all Python projects) are implemented with Flask (cf.
-a GitHub search for ``language:Python Flask'')}.
-It provides simplicity and flexibility by implementing a bare-minimum web server, and thus advertises as a micro-framework.
-The Flask tutorial shows how setting up a simple Flask {\em ``Hello World''} web-service requires no more than 5 lines of Python code \cite{ flask:tutorial}.
-% end of summary
-
-Despite their popularity, to the best of our knowledge, there is no simple solution for monitoring the evolving performance of Flask web applications.
-Thus, every one of the developers of these projects faces one of the following options when confronted with the need of gathering insight into the runtime behavior of their implemented services:
-
-  \begin{enumerate}
-
-    \item Use a commercial monitoring tool which treats the subject API as a black-box (e.g. Pingdom, Runscope). 
-    % , Graphite+Graphana+statd etc.
-
-    \item Implement their own ad-hoc analytics solution, having to reinvent basic visualization and interaction strategies.
-
-    \item Live without analytics insight into their services.
-
-  \end{enumerate}
-
-%\todo{For the first point in the list, we can also argue that analytics solutions like Google Analytics can be used, but they have no notion of versioning/integration with the development life cycle.
-Feel free to cite \cite{papazoglou2011managing} for service evolution purposes}
-
-For projects on a budget (e.g.
-research, startups) the first and the second options are often not available due to time and financial constraints.
-Even when using 3rd-party analytics solutions, a critical insight into the evolution of the exposed services of the web application, is missing because such solutions have no notion of versioning and no integration with the development life cycle.~\cite{papazoglou2011managing}
-
-To avoid projects ending up in the third situation, that of living without analytics, in this paper we present \tool~ --- a low-effort service monitoring library for Flask-based Python web services that is easy to integrate and enables the {\em agile assessment} of service evolution.
-\cite{Nier12b}
-
-As a case study, on which we will illustrate our solution, we are going to use an open source API which, for several years, was in the third of the above-presented situations.
-
-% In the next section, we will present a case study of an open source research API which was for a long time in the third situation presented above -- deployed without analytics insight.
-
-\todo{Add signposting paragraph}
-
-\section{Case Study: The API (Zeeguu)}
-\label{sec:case}
-
-%\todo{Either leave as is or add more info on usage by college if possible}
-
-  \zee\footnote{\url{https://github.com/zeeguu-ecosystem/}} is a platform and an ecosystem of applications for accelerating vocabulary acquisition in a foreign language \cite{Lungu16}. 
-%
-  The architecture of the ecosystem has at its core an API and a series of satellite applications that together offer to a learner three main inter-dependent features:
-
-  \begin{enumerate}
-
-    \item Reader applications that provide effortless translations for those texts which are too difficult for the readers.
-
-    \item Interactive exercises personally generated based on the preferences and past likes of the learner.
-
-    \item Article recommendations which are personalized for the interests of the learner and come with difficulty estimation that helps the learner find articles with the appropriate difficulty.
-
-  \end{enumerate}
-
-  The core API implemented with Flask and Python provides correspondingly three types of functionality: contextual translations, article recommendations, and personalized exercise suggestions.
-In total the API provides a bit less than 50 endpoints, out of which probably a dozen are very frequently used.
-  % The development of the core API itself is a research project.
-
-  At the time of writing this article, the ecosystem consists of a reader web application, a web based exercises platform, and a smartwatch application, which are used at the moment of writing this article by more than \activeUserCount active users.
-The users come from a highschool, a language school, and some users are using it on their own, without any educational context.
-The highest load we observed until now on the API consisted of 12K requests in one day.
-
-  We will use this \zee API as a case study for this paper. 
-  All the figures in this paper are captured from the actual deployment of \tool in the context of the \zee API.
-The figures are interactive offering basic data exploration capabilities: filter, zoom, and details on demand\cite{Shne99a}.
-The \tool deployment for the case study can be accessed publicly\footnote{\url{https://zeeguu.unibe.ch/api/dashboard}; username: {\em guest}, password: {\em soap-sac}}.
-
- % \todo{create new account, update URL if necessary, update activeUserCount, add a couple of sentences of the current state of the case study with the college/language center}
-% \ml{we should consider adding also one section in which the architecture/implementation and main features of the dashboard are presented before going on with discussing them in more depth in the following sections --- this should include a rundown on which views are provided from where (overview or per endpoint)}
-
-\newpage
-
-\section{Case Study: The Technology (Flask)}
-\label{sec:tool}
-
- Flask is a microframework for Python. It IS USED BY THOUSANDS OF PROJECTS. ELABORATE MORE ON THIS... 
- \vspace{2cm}
-
- Flask depends on two external libraries: the Jinja2 template engine and the Werkzeug WSGI toolkit.
-
- “Micro” does not mean that your whole web application has to fit into a single Python file (although it certainly can), nor does it mean that Flask is lacking in functionality.
-The “micro” in microframework means Flask aims to keep the core simple but extensible.
-Flask won’t make many decisions for you, such as what database to use.
-Those decisions that it does make, such as what templating engine to use, are easy to change.
-Everything else is up to you, so that Flask can be everything you need and nothing you don’t.
-
- Flask has many configuration values, with sensible defaults, and a few conventions when getting started.
-By convention, templates and static files are stored in subdirectories within the application’s Python source tree, with the names templates and static respectively.
-While this can be changed, you usually don’t have to, especially when getting started.
-
- A Python web application based on WSGI has to have one central callable object that implements the actual application.
-In Flask this is an instance of the Flask class.
-Each Flask application has to create an instance of this class itself and pass it the name of the module, but why can’t Flask do that itself?
-
-\begin{lstlisting}[style=custompython]
-
-from flask import Flask
-app = Flask(__name__)
-
-@app.route('/')
-def index():
-    return 'Hello World!'
-
-\end{lstlisting}
-
-\subsection*{NOTHING PREVENTS GENERALIZATION}
-In this article we take as an example the Flask case study.
-But there is nothing that prevents the generalization of 
-our approach.
-
-MAKE SURE THAT THIS IS CLEAR.
-
-  \newpage
-  \section{One Line Of Code Configuration With Technology Specific Interceptors}
-
-  The fact that the \tool itself is being developed in Python using Flask makes binding to the services of a to-be monitored application developed with the same technologies easy and intuitive.
-
-  \ins{In general, what matters is that with the right backend, we can do this minimal configuration for any technology... }
-
-  To start using the \tool, one simply needs to one lines of code\footnote{we don't count the import statement that enables that line...} to their Flask web service\footnote{\ins{ In this paper we present the integration with APIs written in Python and Flask hoping that this will not prevent the reader from seeing the more general idea; all the tools we show here for Flask can be applied to other API technologies (e.g.
-Django) by simply providing a few back-end adapters in the right places.
-}}:
-
-  % caption=Configuring the \tool is straightforward,
-  \begin{lstlisting}[style=custompython]
-  import dashboard
-
-  # flask_app is the Flask app object
-  dashboard.bind(flask_app)
-  \end{lstlisting}
-
-  \ins{The dashboard can take advantage of the fact that it is monitoring an API...
-so it can simply just add one extra endpoint there.} After (re)deployment of the application, the interactive visualizations of the \tool become available at the \code{/dashboard} route of the Flask application.
-
-  During binding, the \tool will search for all endpoints defined in the target application.
-
-  In order to monitor an endpoint, the \tool creates a function wrapper for the API function that corresponds to the endpoint.
-This way, the wrapper will be executed whenever that API call is made before the actual function is called.
-The wrapper contains the code that takes care of monitoring an endpoint.
-Data collected by the wrappers are persisted in a local database.
-The SQLAlchemy Object Relational Mapper\footnote{\url{https://www.sqlalchemy.org/}} on top of SQLite\footnote{\url{https://www.sqlite.org/}} is used for this purpose.
-  \ins{DISCUSSION about how do to this in other languages? Java: AspectJ probably, etc.}
-
-  These will be presented to the user in the tool web interface, where the user can select the ones that should be monitored, see \Fref{fig:sep}.
-
-    \begin{figure}[h!]
-      \centering
-      \includegraphics[width=\linewidth]{selecting_endpoints.png}
-      \caption{Once connected to an API the Dashboard presents the endpoints that are available for monitoring}
-      \label{fig:sep}
-    \end{figure}
-
-  This is the alternative to using annotations like other white-box tools (e.g. the other flask performance plugin).
-
-  \ins{PROMISE:} The last access time of the endpoint by users, irrespective of whether it is selected by the user to be monitored or not, is also displayed as the means of identifying potentially obsolete endpoints.
-
-  \ml{we must insert at least one dashboard view here...
-otherwise people will lose their patience...
-also it will help with explaining the outliers...
-also, one should see that a large variety of views are already available at this point...
-I know this risks being  }
-
-In the Zeeguu case study, one of the slowest endpoints, and one with the highest variability as shown in \Fref{fig:ep} is \epFeedItems: it retrieves a list of recommended articles for a given user.
-However, since a user can be subscribed to anything from one to three dozen article sources, and since the computation of the difficulty is personalized and it is slow, the variability in time among users is likely to be very large.
-
-\begin{figure*}[h!]
-  \centering
-  \subfloat[number of requests per endpoint per day]{\includegraphics[width=.9\columnwidth]{number_of_requests_}\label{fig:aeu}}
-  \subfloat[requests per hour heatmap]{\includegraphics[width=\columnwidth]{daily_patterns_}\label{fig:dp}}
-  \caption{Some of the available views\label{fig:views}}
-\end{figure*}
-
-\newpage
-\section{Service Utilization}
-\label{sec:util}
-
-\ins{A discussion must be added here about the types of useful information and problems that can be solved with these views.}
-
-\textit{Service Utilization} presents information about the usage of all endpoints of interest to the developer,
-
-The most fundamental insight that a service maintainer needs regards service utilization. %\vspace{0.5cm}
-
-Simplest utilization perspective:
-
-\begin{figure}[h!]
-  \centering
-  \includegraphics[width=0.5\linewidth]{endpoint-call-count}
-  \caption{Caption here}
-  \label{fig:figure1}
-\end{figure}
-
-% REQUIREMENT: Adding a time interval is important. 
-% Otherwise the performance numbers in chart 1 are not good enough... also an outlier very far in the past, will make all the heatmap data points to look very pale in comparison forever after.
-
-Figure \ref{fig:aeu} shows a first perspective on endpoint utilization that \tool provides: a stacked bar chart of the number of hits to various endpoints grouped by day\footnote{Endpoint colors are the same in different views}.
-Figure~\ref{fig:aeu} in particular shows that at its peak the API has about 12.000 hits per day.
-
-The way users interact with the platform can also be inferred since the endpoints are indicators of different activity types, e.g.:
-
-\begin{enumerate}
-
-  \item {\color{myred} \epTranslations} is an indicator of the amount of foreign language reading the users are doing, and
-
-  \item {\color{mygreen} \epOutcome} is an indicator of the amount of foreign vocabulary practice the users are doing.
-
-\end{enumerate}
-
-%\begin{figure}[!ht]
-% \centering
-% \includegraphics[width=\linewidth]{number_of_requests_}
-% \caption{The number of requests per endpoint per day view shows the overall utilization of the monitored application}
-% \label{fig:aeu}
-%\end{figure}
-
-Besides showing the overall utilization, this endpoint provides the maintainer with information relevant for decisions regarding endpoint deprecation --- one of the most elementary ways of {\em understanding the needs of the downstream}\cite{Haen14a}.
-In our case study, the maintainer realized that one endpoint which they thought was not being used (i.e.
-\code{words\_to\_study}), contrary to their expectations, was actually being used\footnote{A complementary type of usage information can also be discovered in the view presented in Figure \ref{fig:sep} where seeing that an endpoint is never accessed can increase the confidence of the maintainer that a given endpoint is not used, although it can never be used a proof.}.
-
-\niceseparator
-
-%   \todo{Add the time series graph and discuss it before the heatmap? We can then sell the heatmap better} 
-%   \ml{Not sure about which graph you refer to here V}
-
-A second type of utilization question that the \tool can answer automatically regards {\em cyclic patterns of usage per hour of day} by means of a heatmap, as in \Fref{fig:dp}.
-
-% \mltp{ can we add vertical lines that highlight the beginning of a new week (e.g. before Sunday): }
-% Patrick: adding separators in the graph is unfortunately not supported by the library.
-
-%\begin{figure}[!ht]
-% \centering
-% \includegraphics[width=0.9\linewidth]{daily_patterns_}
-% \caption{Usage patterns become easy to spot in the requests per hour heatmap}
-% \label{fig:dp}
-%\end{figure}
-
-Figure \ref{fig:dp} shows the API not being used during the early morning hours, with most of the activity focused around working hours and some light activity during the evening.
-This is consistent with the fact that the current users are all in the central European timezone.
-Also, the figure shows that the spike in utilization that was visible also in the previous graph happended in on afternoon/evening.
-
-\begin{figure*}[h!]
-  \centering
-  \subfloat[overview of the tracked endpoints]{\includegraphics[width=1.5\columnwidth]{overview}\label{fig:overview}}
-  \caption{Some of the available views\label{fig:views}}
-\end{figure*}
-
-\newpage
-\section{Endpoint Performance}
-\label{sec:perf}
-
-  The \tool also collects information regarding endpoint performance. The view in \Fref{fig:ep} summarizes the response times for various endpoints by using a box-and-whiskers plot.
-
- \begin{figure}[!ht]
-   \centering
-   \includegraphics[width=0.95\linewidth]{endpoint_performance_}
-   \caption{The response time (in ms) per monitored endpoint view allows for identifying performance variability and balancing issues}
-   \label{fig:ep}
- \end{figure}
-
-  From this view it became clear to the maintainer that four of the endpoints had very large variation in performance.
-The most critical for the application and consequently the one optimized first was the \epTranslations endpoint which was part of an interactive loop in the reader applications that relied on the Zeeguu API.
-Moreover, cf.
-\Fref{fig:aeu} this endpoint is one of the most used in the system.
-
-  \begin{figure*}[h!]
-    \centering
-    \includegraphics[width=\linewidth]{outliers}
-    \caption{Automatically collected outlier information}
-    \label{fig:figure1}
-  \end{figure*}
-  \newpage
-  \subsection{Performance Insight: Automated outlier detection and monitoring}
-
-  When an API is called from within a highly interactive application (as it is the case with the case study in this paper) 
-  of particular interest to the API developers are performance {\em outliers}.
-
-  Indeed, a translation request that takes three times more than expected can seriously decrease the perceived quality of the application.
-Thus, identifying, collecting all appropriate data, and diagnosing the root causes of such outliers is especially critical in improving the quality of an application.
-
-  % In the context of the RESTful services support by Flask, this means request serving times that deviate from the average to an unexpected degree.
-
-  For this purpose the \tool tracks for every monitored endpoint a {\em running average} response time value\footnote{\ins{For performance reasons, we assume that the response times for the endpoints are normally distributed.
-Otherwise, more general density distribution information must be collected in real time.}}.
-When it detects that a given request is an outlier with respect to this past average running value, it triggers the {\em outlier data collection routine} which stores \ins{extra information} about the current execution environment.
-A configurable threshold with a default value of $2.5$ times the running average response time is used for this purpose.
-
-  For every detected outlier request, the \tool collects information about the current Python stack trace, CPU load, memory consumption, request parameters, etc.
-in order to allow the maintainer to investigate the causes of these exceptionally slow response times.
-In this way it is possible to get \ins{detailed insight into the operation of the application in the extreme cases without unnecessarily burdening it with logging this information for every request}.
-
-The bottom panel shows the stack trace. 
-In this particular case, it is revealing for the developer to learn that at the time of the stack trace snapshot, the code was in the microsoft\_translator.py: indeed, the system uses as back-end multiple translators, and it has been observed that many of the outliers happen to be waiting in the microsoft translator.
-
-This information has to be corroborated with the observations that neither the memory nor the processor are overloaded at the moment.
-Thus this functionality in microsoft\_translator is really slow in itself, and this is not a result of the machine being overloaded for example.
-
-\subsection {A Special Type of Outlier: Exceptions}
-
-   TODO: probably something for the journal extension ...
-  statistics about which endpoints fail most often might be useful too.
-  same information as the outlier maybe?
-
-\begin{figure*}[h!]
-  \centering
-  \includegraphics[width=0.8\linewidth]{utilization-evolution}
-  \caption{The Performance Evolution of the \epTranslations endpoint}
-  \label{fig:tee}
-\end{figure*}
-
-  \newpage
-  \section{Evolution monitoring }
-
-  \begin{figure}[h!]
-  \centering
-  \includegraphics[width=0.8\linewidth]{translation_endpoint_evolution_}
-  \caption{The Performance Evolution of the \epTranslations endpoint}
-  \label{fig:tee}
-\end{figure}
-
-  When thinking about performance, 
-  one can not avoid thinking about the performance evolution.
-
-  Version control in \tool can be supported in two ways,
-  the developer explicitly states the current version, 
-  or the current version is automatically detected based
-  on some version control system.
-
-  However, with the current configuration of the tool, it would be impossible for the maintainer to see the improvements resulting from the optimization.
-
-  \ins{The latest STack Overflow Developer Survey: 88.4\% of **professional** developers use git.}  
-  If assume that the code that they run is deployed using .git, then with an extra line of configuration they can allow \tool to find the git\footnote{\url{https://git-scm.com/}} folder of the deployed service and automatically detect the version of the project that is running:
-
-    \begin{lstlisting}[style=custompython]
-
-      dashboard.config.git = 'path/to/.git'
-
-    \end{lstlisting}
-
-  If the \tool can automatically detect the current version of the project by reading the .git configuration as soon as the API is started and can then group measurements by version\footnote{Alternatively, the maintainer can add version identifiers manually for the web application through a configuration file if the system does not use git.}.
-\Fref{fig:tee} is a zoomed-in version of such a view for \epTranslations with versions increasing from top to bottom
-
-  This view confirms that the performance of the translation endpoint improved in the recent versions: the median of the last three versions is constantly moving towards the left, and progresses from 1.4 seconds (in the top-most box plot in \Fref{fig:tee}) to 0.8 in the latest version (bottom-most box plot).
-
-  CHALLENGE: Supporting other types of version control... not real challenge, as we said 88percent of devs use git. 
-  CHALLENGE 2: detecting minor versions which don't need to be tracked, since they didn't touch the performance of the system. E.g. a modification of the README, etc.
-
-\todo{TODO: endpoint evolution }
-
-\begin{figure*}[!ht]
-  \centering
-  \includegraphics[width=0.4\linewidth]{time_per_user}
-  \caption{The \epFeedItems shows a very high variability across users}
-  \label{fig:tpu}
-\end{figure*}
-
-\newpage
-\section{Grouping Requests}
-\label{sec:user}
-
-For service endpoints which run computations in real time, the maintainer of a system might want to understand the endpoint performance on a per-user basis, especially for situations where the system response time is a function of some individual user load\footnote{E.g.
-in GMail some users have two emails while other have twenty thousand and this induces different response times for different users}.
-
-\todo{remove the configuration part and rephrase the following, the discussion has been partially subsumed by Section~\ref{sec:tool}.}
-
-To enable this, the \tool must be configured to associate an API call with a given user.
-The simplest way is to take advantage of the architecture of Flask applications in which a global \code{flask.request} object can be used to retrieve the session which can in turn lead to user identification:
-
-%\begin{lstlisting}[float,caption=Simply define a custom app-specific function for user retrieval and pass it to the \tool to group information by user,style=custompython]
-\begin{lstlisting}[style=custompython]  
-# app specific way of extracting the user
-# from a flask request object    
-def get_user_id():
-sid = int(flask.request.args['session'])
-session = User.find_for_session(sid)
-return user_id
-
-# attaching the get_user_id function
-dashboard.config.get_group_by = get_user_id
-
-\end{lstlisting}
-
-% \niceseparator
-
-\todo{consider losing one or both of the figures --- or at least updating them}
-
-In Zeeguu, \epFeedItems retrieves a list of recommended articles for a given user.
-Cf.
-\Fref{fig:ep} it is the endpoint with the slowest response time and highest variability.
-The reason for this is that a user can be subscribed to anything from one to three dozen article sources and for each of the sources the system must compute the personalized difficulty of each article at every request.
-
-A \perspective{Per-User Performance} perspective should show the different response times for different users.
-Figure \ref{fig:tpu} presents a subset of the corresponding view in the \tool.
-The figure shows that the response times for this endpoint can vary considerably for different users with some extreme cases where a user has to wait a full minute until their recommended articles are shown\footnote{After seeing this perspective, the maintainer refactored the architecture of the system to move part the difficulty computation out of the interactive loop}.
-
-The limitation of the previous view is that it does not present the information also on a per version basis.
-To address this, a different visual perspective entitled \perspective{Multi-Version per-User Performance} can be defined.
-Figure \ref{fig:tuv} presents such a perspective by mapping the average execution time for a given user (lines) and given version (columns) on the area of the corresponding circle.
-
-The colors represent users.
-The figure shows average performance varying  across users and versions with no clear trend: this is probably because varying user workload (i.e.
-number of sources to which the user is registered) is the reason for the variation in response times.
-\ins{one can see that for user 1 performance degrades over the }
-
-\begin{figure}[!ht]
-  \centering
-  \includegraphics[width=0.7\linewidth]{time_per_user_per_version}
-  \caption{This perspective shows that the evolution of response times for individual users (horizontal lines) across versions (the x-axis)}
-  \label{fig:tuv}
-\end{figure}
-
-  \newpage
-  \section{Preemptive monitoring}
-
-  \ins{INTRO - LINKING WITH PREVIOUS PARAGRAPH: the previous evolutionary view crucial for 
-  understanding the impact of software evolution on performance.
-  However, just by observing this one can't be sure whether 
-  performance changes because of the code or because of the payload.}
-
-  \todo{the following might need a bit of rewrite}
-
-  The concept of {\em preemptive monitoring} of the application performance by means of instrumenting integration unit testing as the synthetic load is similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}, i.e.~testing service-based applications by using dedicated test input in parallel to its normal use and operation.
-The difference is that we take advantage of the capability of the CI framework (i.e.
-Travis in this case) to create an emulated ``live'' environment for integration testing purposes, and use unit testing as the dedicated test input in order to measure performance.
-
-  For example, \Fref{fig:builds} is a screenshot from the dashboard showing the measured response times for 5 consequent builds, with 180 iterations of the unit tests executed in total across all endpoints.
-The outliers on the right of the figure are due to initial requests \ins{which must wait for a boot-up phase of the API}.
-
-    \begin{figure}[h!]
-        \centering
-        \includegraphics[width=\columnwidth]{travis_builds}
-        \caption{Response times in 5 consequent Travis builds in the \zee case study}
-        \label{fig:builds}
-      \end{figure}
-
-  While this integration testing environment is different from the actual production deployment one, and the load used is purely synthetic, as will show in Section~\ref{sec:evaluation}, it can actually be used successfully as an early performance indicator for the application developer.
-\ins{Although predicting the performance of the application would require advanced statistical work, it can be succesfully used as an advanced warning system for the overall performance trend for given endpoints}.
-
-  \paragraph{Integration with Travis}
-  \ml{added new section... also, this has to be moved after introducing the pre-emptive monitoring since it's about running tests. switched the order}
-
-  \ins{The pre-emptive monitoring system} can be configured to work together with Continuous Integration (CI) frameworks like Travis\footnote{\url{https://travis-ci.org/}} that deal with automated integration testing.
-  The developer needs to define the unit test folder of the project, the URL that Travis should use for reporting its results, and the number of iterations for each unit test should be executed in order to generate a load for the application.
-
-  Each time a new build is created through Travis, the \tool automatically detects all available unit tests defined by the application developer and iterates through each one of them the number of times predefined by the developer while monitoring the response times for each request.
-The resulting measurements are persisted in a separate part of the tool database so as not to contaminate the ``live'' monitoring data.
-Beyond seeing the results of each Travis build in the \tool, this feature also allows for preemptive monitoring of the application, as we discuss in the following.
-
-  \ins{Besides functioning as an early warning system for performance, tracking the evolving performance of API tests serves a second purpose.
-To function as an anchor when the developer analyzes performance degradations.
-For example, when a developer sees that a given endpoint has become less performant after in a newly deployed version, they are able to investigate whether the performance degradation is visible also with the synthetic load.
-This would correspond to a performance degradation which is due to ``algorithmic performance degradation''.
-If on the other hand, the performance of the tests does not change between versions, the developer might conclude that the performance degradation could be due to the workload on the machines, or maybe to the workload mix of the users}
-
-  %By these means, an application developer can use unit tests as synthetic loads for driving the performance monitoring of the application before it reaches the production phase.
-As we will discuss further in Section~\ref{sec:testing}, this feature is integral in enabling preemptive performance monitoring of the application.
-
-\todo{here goes the discussion about our forecasting capabilities}
-
-\begin{figure}[h!]
-  \centering
-  \subfloat[The reported response time (in ms) per deployed version in the observation period]{\includegraphics[width=.5\columnwidth]{response_per_version_trunced_trend}}
-  \quad
-  \subfloat[The measured response time (in ms) using integration with Travis]{\includegraphics[width=.46\columnwidth]{travis_builds_no_outliers_trend}}
-  \caption{Comparison of the response times per endpoint: actual production system versus preemptive monitoring data}
-  \label{fig:preemptive}
-\end{figure}
-
-\begin{table}[h]
-  \begin{tabular}{lll}
-    \toprule
-    Iteration & \bfseries Live (median) & \bfseries Travis (median)\\
-    \midrule
-    1 & 1349.41 & 764.87\\ 
-    2 & 1466.13 & 992.87\\
-    3 & 1256.65 & 760.87\\
-    4 & 1266.42 & 813.89\\
-    5 & 1080.68 & 303.4\\
-    \bottomrule
-  \end{tabular}
-\end{table}
-
-Pearson correlation $r(3)=.93, p=.02$
-
-    \begin{figure*}[ht!]
-      \centering
-      \includegraphics[width=0.8\linewidth]{interceptor}
-      \caption{The first thing that needs to be done, is to decorate the application object with an INTERCEPTOR}
-      \label{fig:sep}
-    \end{figure*}
-
-\newpage
-
-\section{Monitoring Service Utilization}
+\section{Architecture}
 
@@
   \newpage
+
+  % This used to be in the first section... 
+  % Maybe it fits better localized here: 
+  Data collected by the wrappers are persisted in a local database.
+The SQLAlchemy Object Relational Mapper\footnote{\url{https://www.sqlalchemy.org/}} on top of SQLite\footnote{\url{https://www.sqlite.org/}} is used for this purpose.
+
 TODO: TALK ABOUT THE METAMODEL...
@@
   The source code of the \tool is published under a permissive MIT license and is available on GitHub\footnote{\url{https://github.com/flask-dashboard}}.
-
-\section{Performance Perspectives}
-\label{sec:views}
-
-\ins{PERFORMANCE DISCUSSION: this introspection, which looks up endpoints, happens only once at the startup of the service, so it is going to affect the API startup performance, but not the actual endpoint performance.
-In our case, the overhead here was quite small: TO MEASURE}.
-
-  %In the remainder of the paper we present several of these perspectives.\footnote{We recommend obtaining a color version of this paper for better readability}
-
-  % The second endpoint conists of two parts, one of them being a table that shows for every monitored endpoint the number of hits it has gotten, the time it was last accessed and its average execution time.
-The second part is a view with four graphs which show:
-  % \begin{itemize} 
-  %   \item A heatmap of the total number of requests to the monitored endpoints
-  %   \item A stacked bar chart that shows the total number of requests to the monitored endpoints per endpoint per day
-  %   \item A boxplot graph showing the average execution time per version of the web service
-  %   \item A boxplot graph showing the average execution time for every monitored endpoint
-  % \end{itemize}
-
-\todo{evaluate which figures to keep and ideally take new screenshots for the ones we keep; consider adding two more views (like Fig. 3.2 of Patrick's thesis)}
-
-% In the Zeeguu case study, one of the slowest endpoints, and one with the highest variability as shown in \Fref{fig:ep} is \epFeedItems: it retrieves a list of recommended articles for a given user.
-However, since a user can be subscribed to anything from one to three dozen article sources, and since the computation of the difficulty is personalized and it is slow, the variability in time among users is likely to be very large.
-
-% \niceseparator
-
-%\section{Preemptive Monitoring}
-%\label{sec:testing}
-%
-%%\todo{discuss here idea of using unit testing as an early performance indicator for evolving services (`preemptive monitoring'?); add a side-by-side figure comparing unit testing vs live system performance per version, maybe calculate error in trend curves?
-find a better term than prediction for the section title (plus introduction)}
-%
-%\todo{Similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}, i.e.~testing service-based applications by using dedicated test input in parallel to its normal use and operation.
-The difference is that we take advantage of the capability of Travis to create an emulated live environment for integration testing purposes, and use unit testing as the dedicated test input in order to measure performance.
-What we are going provide evidence for in the following is that we can indeed use this preemptive monitoring concept as an early performance indicator for evolving services.}
-%
-%The developer needs to define the unit test folder of the project, the URL that Travis should use for reporting its results, and the number of iterations for each unit test should be executed in order to generate
-
-\section{Evaluation}
-\label{sec:evaluation}
-
-%\todo{evaluation of the \tool across the following dimensions: ease of use, effectiveness, overhead to the system; for the last part we need to do the overhead measurement experiment, for the other two we base it on the case study and bring over whatever is necessary from the discussion section}
-
-\section{Discussion}
-
-\todo{use this subsection to discuss limitations (from the previous evaluation section) and add the need to be able to handle the horizontal scaling of the application; remove sectioning and summarize briefly limitations that have already been discussed in the VISSOFT paper/move material to other sections; consider renaming as limitations or something similar}
-
-  \subsection{Distributed System Monitoring}
-  \todo{To think about}
-  Supporting situations where the API is deployed across multiple containers for example.
-
-  \subsubsection{Automatically Monitoring System Evolution}
-
-  The main goal of the \tool design was to allow analytics to be collected and insight to be gained by making the smallest possible changes to a running API.
-%To allow the collection of evolutionary information
-%
-  This technique assumes that the web application code which is the target of the monitoring is deployed using \git in the following way:
-
-  \begin{enumerate}
-    \item The deployment engineer pulls the latest version of the code from the integration server; this will result in a new commit being pointed at by the HEAD pointer. %than previously
-    \item The deployment engineer restarts the new version of the service.
-At this point, the \tool detects that a new HEAD is present in the local code base and consequently starts associating all the new data points with this new commit\footnote{The \tool detects the current version of the analyzed system the first time it is attached to the application object, and thus, assumes that the Flask application is restarted when a new version is deployed.
-This is in tune with the current version of Flask, but if the web server will support dynamic updates in the future, this might have to be taken into account}.
-  \end{enumerate}
-
-  The advantage of this approach is the need for minimal configuration effort, as discussed in the presentation of the tool.
-The disadvantage is that it will consider on equal ground the smallest of commits, even one that modifies a comment, and the shortest lived of commits, e.g.~a commit which was active only for a half an hour before a new version with a bug fix was deployed, with major and minor releases of the software.
-%as a distinct way of grouping the data points.
-  A mechanism to control which versions are important for monitoring purposes is therefore required to be added.
-%
-  A further possible extension point here is supporting other version control systems (e.g. Mercurial). However, this is a straightforward extension.
-
-  \subsubsection{NEW: API Renames}
-  \ml{just an idea}
-
-    The history would normally be lost in the case of an API rename. (also known as the {\em provenance issue} in source code evolution analysis)
-    In theory we could infer that an API was renamed if it is not to be found anymore, and another one with very similar timing characteristics would be found.
-Such a detection can never be completely sure, but it would help with meeting the needs of the API maintainer...
-
-  \subsubsection{User-Awareness }
-
-    For the situations in which the user information is not available, the \tool tracks by default information about different IPs and in some cases this might be a sufficiently good approximation of the user diversity and identity.
-    %
-
-    The visualizations for the user experiene perspectives as presented in Section \ref{sec:user} have been tested with several hundred users (of which about \activeUserCount were active during the course of the study), but the scalability of the visualizations must be further investigated for web services with tens of thousands of users.
-
-  \subsubsection{Other Possible Groupings}
-
-    There are other groupings of service utilization and performance that could be important to the maintainer, that we did not explore in this paper.
-For example, if the service is using OAuth, then together with every request, in the header of the request there is information about the application which is sending a request.
-Grouping the information by application that sends the request could be important in such a context.
-
-    In general, providing a mechanism that would allow very easy specification of groupings (either as code annotations, as normal code, or as configuration options) is an open problem that \tool and any other similar library will have to face.
-
-\section{Related Work}
-\label{sec:related}
-
-\todo{Refer the reader to~\cite{ghezzi2007run} and ~\cite{metzger2010analytical} for a more extensive discussion}
-
-There is a long tradition of using visualization for gaining insight into software performance.
-Tools like Jinsight \cite{Pauw02a} and Web Services Navigator \cite{Pauw05} pioneered such an approach for Java and for Web Services that communicate with SOAP messages.
-Both have an ``omniscient'' view of the services / objects and their interactions.
-As opposed to them, in our work we present an analytics platform which focuses on monitoring a single Python web service from its own point of view.
-
-From the perspective of service monitoring, our work falls within the server-side run-time monitoring of services ~\cite{ghezzi2007run}.
-While we don't implement the more advanced features of related monitoring solutions like QoS policies driving the monitoring, it presents nevertheless an easy to use approach support improving the performance of web applications.
-
-% \va{Mircea: Consider removing the rest for space...}
-% An existing monitoring tool is Pingdom \footnote{https://www.pingdom.com/company/why-pingdom}, which monitors the uptime of an existing web-service.
-This tool works by pinging the websites (up to 60 times) every minute automatically.
-Thus this creates a lot of overhead and is bound to be noisy since it will also be influenced by the speed of the network connection\footnote{Another problem is that such a tool would }
-
-% \todo{Runscope? Others?}
-
-\section{Conclusion and Future Work}
-\label{sec:conclusions}
-
-\todo{update as appropriate}
-
-In this paper we have shown that it is possible to create a monitoring solution which provides basic insight into web service utilization and performance  with very little effort from the developer.
-The user group that we are aiming for with this work is application developers using Flask and Python to build web applications with limited or no budget for implementing their own monitoring solutions.
-The emphasis is in allowing such users to gain insight into how the performance of the service evolves together with the application itself.
-We believe that the same architecture, and lessons can be applied to other frameworks and other languages.
-
-In the future, we plan to perform case studies with other sytstems, with the goal of discover other needs and to wean out the less useful visualizations in the \tool.
-We plan to also extend the tool towards supporting multiple deployments of the same applications across multiple nodes (e.g.
-for the situations where the application is deployed together with a load balancer).
-Finally, we plan to integrate \tool with unit testing as a complementary source of information about performance evolution.
-
-\newpage
-\appendix
-
-  \subsection{Changing the Default /dashboard Endpoint }
-
-  THIS STUFF SHOULD BE MOVED TO THE APPENDIX.
-And in a footnote we just mention that ...
-\ins{REQUIREMENT}: The system must provide an easy way to add more specific configurations if needed.
-A way of overwriting the common sense defaults.
-
-  Further configuration is not needed possible by adding additional statements before the binding definition, for example,
-
-  \begin{lstlisting}[style=custompython]
-  ...
-  dashboard.config.link = 'mydashboard'
-  dashboard.bind(flask_app)
-  \end{lstlisting}
-
-  allows for a custom route (\code{/mydashboard}) to the dashboard to be defined by the programmer. An external configuration file can be used instead, or in addition to this:
-
-  \begin{lstlisting}[style=custompython]
-  ...
-  dashboard.config.from_file('config.cfg')
-  ...
-  \end{lstlisting}
-
-  \subsection{For those who don't use GIT}
-
-    \textit{Manual} version control requires the developer to tag each new version of the application with an appropriate version identifier~\cite{papazoglou2011managing} using the \code{APP\_VERSION} configuration parameter, for example by adding to the configuration file:
-
-      \begin{lstlisting}[style=custompython]
-      [dashboard]
-      APP_VERSION=<versionID>
-      \end{lstlisting}
-
-% references section
-
-\bibliographystyle{abbrv}
-\bibliography{paper}
-
-% that's all folks
-\end{document}

### 2018-04-03 discussion.tex
split the huge file in smaller ones.

--- 2018-03-21 restructured.tex
@@
-\documentclass[conference]{IEEEtran}
-	\pdfpagewidth=8.5truein
-	\pdfpageheight=11truein
+%!TEX root=restructured.tex
 
-\usepackage{graphicx}
-\usepackage{subfig}
-\usepackage[bookmarks=false]{hyperref}
-\usepackage{xspace}
-\usepackage{listings}
-\usepackage[usenames, dvipsnames]{color}
+\section{Discussion}
 
-\usepackage{booktabs}
+-   \ins{We can do this minimal configuration for any technology.
+We showed here how to do it for Python and Flask because they are some of the most popular web development technologies at the moment.}
 
-\graphicspath{{./img/}}
+-   The fact that the \tool itself is being developed in Python using Flask makes binding to the services of a to-be monitored application developed with the same technologies easy and intuitive.
+However, by writing different backends that do the measurement and monitoring, the frontend can be reused.
 
-%\hyphenation{op-tical net-works semi-conduc-tor}
-
-\input{useful_defines}
-
-\begin{document}
-
-%
-\title{Agile Monitoring of Evolving Web Services}
-
-\author{
-\IEEEauthorblockN{Patrick Vogel, Thijs Klooster, Vasilios Andrikopoulos, Mircea Lungu}\\
-Johann Bernoulli Institute for Mathematics and Computer Science\\
-University of Groningen, Netherland\\
-Email: \{t.klooster.1,p.p.vogel\}@student.rug.nl, \{v.andrikopoulos,m.f.lungu\}@rug.nl 
-}
-
-% make the title area
-\maketitle
-
-\begin{abstract}
-\todo{Update abstract \& title}
-  Tens of thousands of web applications are written in Flask, a Python-based web framework.
-Despite a rich ecosystem of extensions, there is none that supports the developer in gaining insight into the evolving performance of their service.
-In this paper, we introduce \tool, a library that addresses this problem.
-We present the ease with which the library can be integrated in an already existing web application, discuss some of the visualization perspectives that the library provides and point to some future challenges for similar libraries.
-
-\end{abstract}
-
-\section{Introduction}
-
-\todo{Intro to be rewritten; refer to VISSOFT paper as previous work, explain focus is on integration with version control/CI for evolution support, and use of unit tests as early performance indicator}
-
-\todo{motivate with: monitoring in a world which is becoming more and more service-oriented is essential because it allows to perform three main types of actions: system adaptations to provide the request service at the desired level of quality, enabling flexibility in dealing with changing requirements and modes of operation, and enabling operational awareness through dashboards organizing the collected data~\cite{pernici2016monitoring}.
-In this work we focus on the last part and we discuss how a minimal effort probe-based solution for a particular type of web services can be used to facilitate the other two types of actions by involving the developer.}
-
-\todo{explain: in~\cite{vogel2017low} we introduced the \tool an a high level with a focus on presenting its performance visualization aspects; in this work we only summarize these features by means of introducing some of the provided functionalities of the \tool.
-We focus on discussing how the tool is implemented and operating, provide a deeper look at its capabilities for integration with version control and continuous integration environments, introduce a mechanism for performance prediction, and provide an evaluation of the proposed approach based on a case study.
-}
-
-%Every system is a distributed system nowadays \cite{cavage2013there}.
-Indeed a very large number of applications and web applications are nowadays implemented as two-tier architectures with a front-end implemented with web technologies and a service back-end.
-%\ml{I'm not completely happy with this paragraph}
-{\em There is no getting around it: you are building a distributed system} argues a recent article \cite{cavage2013there}.
-Indeed, even the simplest second-year student project is a web application implemented as two-tier architecture with a Javascript/HTML5 front-end a service backend, usually a REST API.
-
-% \hfill mds
-% Many contemporary programming languages are offering libraries, modules, or frameworks that facilitate the development of such architectures. 
-Python is one of the most popular programming language choices for implementing the back-end of web applications.
-GitHub contains more than 500K open source Python projects and the Tiobe Index\footnote{TIOBE programming community index is a measure of popularity of programming languages, created and maintained by the TIOBE Company based in Eindhoven, the Netherlands} ranks Python as the 4th most popular programming language as of June 2016.
-
-% possible flask summary
-Within the Python community, Flask\footnote{\url{http://flask.pocoo.org/}} is a very popular web framework\footnote{More than 25K projects on GitHub (5\% of all Python projects) are implemented with Flask (cf.
-a GitHub search for ``language:Python Flask'')}.
-It provides simplicity and flexibility by implementing a bare-minimum web server, and thus advertises as a micro-framework.
-The Flask tutorial shows how setting up a simple Flask {\em ``Hello World''} web-service requires no more than 5 lines of Python code \cite{ flask:tutorial}.
-% end of summary
-
-Despite their popularity, to the best of our knowledge, there is no simple solution for monitoring the evolving performance of Flask web applications.
-Thus, every one of the developers of these projects faces one of the following options when confronted with the need of gathering insight into the runtime behavior of their implemented services:
-
-  \begin{enumerate}
-
-    \item Use a commercial monitoring tool which treats the subject API as a black-box (e.g. Pingdom, Runscope). 
-    % , Graphite+Graphana+statd etc.
-
-    \item Implement their own ad-hoc analytics solution, having to reinvent basic visualization and interaction strategies.
-
-    \item Live without analytics insight into their services.
-
-  \end{enumerate}
-
-%\todo{For the first point in the list, we can also argue that analytics solutions like Google Analytics can be used, but they have no notion of versioning/integration with the development life cycle.
-Feel free to cite \cite{papazoglou2011managing} for service evolution purposes}
-
-For projects on a budget (e.g.
-research, startups) the first and the second options are often not available due to time and financial constraints.
-Even when using 3rd-party analytics solutions, a critical insight into the evolution of the exposed services of the web application, is missing because such solutions have no notion of versioning and no integration with the development life cycle.~\cite{papazoglou2011managing}
-
-To avoid projects ending up in the third situation, that of living without analytics, in this paper we present \tool~ --- a low-effort service monitoring library for Flask-based Python web services that is easy to integrate and enables the {\em agile assessment} of service evolution.
-\cite{Nier12b}
-
-As a case study, on which we will illustrate our solution, we are going to use an open source API which, for several years, was in the third of the above-presented situations.
-
-% In the next section, we will present a case study of an open source research API which was for a long time in the third situation presented above -- deployed without analytics insight.
-
-\todo{Add signposting paragraph}
-
-\section{Case Study: The API (Zeeguu)}
-\label{sec:case}
-
-%\todo{Either leave as is or add more info on usage by college if possible}
-
-  \zee\footnote{\url{https://github.com/zeeguu-ecosystem/}} is a platform and an ecosystem of applications for accelerating vocabulary acquisition in a foreign language \cite{Lungu16}. 
-%
-  The architecture of the ecosystem has at its core an API and a series of satellite applications that together offer to a learner three main inter-dependent features:
-
-  \begin{enumerate}
-
-    \item Reader applications that provide effortless translations for those texts which are too difficult for the readers.
-
-    \item Interactive exercises personally generated based on the preferences and past likes of the learner.
-
-    \item Article recommendations which are personalized for the interests of the learner and come with difficulty estimation that helps the learner find articles with the appropriate difficulty.
-
-  \end{enumerate}
-
-  The core API implemented with Flask and Python provides correspondingly three types of functionality: contextual translations, article recommendations, and personalized exercise suggestions.
-In total the API provides a bit less than 50 endpoints, out of which probably a dozen are very frequently used.
-  % The development of the core API itself is a research project.
-
-  At the time of writing this article, the ecosystem consists of a reader web application, a web based exercises platform, and a smartwatch application, which are used at the moment of writing this article by more than \activeUserCount active users.
-The users come from a highschool, a language school, and some users are using it on their own, without any educational context.
-The highest load we observed until now on the API consisted of 12K requests in one day.
-
-  We will use this \zee API as a case study for this paper. 
-  All the figures in this paper are captured from the actual deployment of \tool in the context of the \zee API.
-The figures are interactive offering basic data exploration capabilities: filter, zoom, and details on demand\cite{Shne99a}.
-The \tool deployment for the case study can be accessed publicly\footnote{\url{https://zeeguu.unibe.ch/api/dashboard}; username: {\em guest}, password: {\em soap-sac}}.
-
- % \todo{create new account, update URL if necessary, update activeUserCount, add a couple of sentences of the current state of the case study with the college/language center}
-% \ml{we should consider adding also one section in which the architecture/implementation and main features of the dashboard are presented before going on with discussing them in more depth in the following sections --- this should include a rundown on which views are provided from where (overview or per endpoint)}
-
-\newpage
-
-\section{Case Study: The Technology (Flask)}
-\label{sec:tool}
-
- Flask is a microframework for Python. It IS USED BY THOUSANDS OF PROJECTS. ELABORATE MORE ON THIS... 
- \vspace{2cm}
-
- Flask depends on two external libraries: the Jinja2 template engine and the Werkzeug WSGI toolkit.
-
- “Micro” does not mean that your whole web application has to fit into a single Python file (although it certainly can), nor does it mean that Flask is lacking in functionality.
-The “micro” in microframework means Flask aims to keep the core simple but extensible.
-Flask won’t make many decisions for you, such as what database to use.
-Those decisions that it does make, such as what templating engine to use, are easy to change.
-Everything else is up to you, so that Flask can be everything you need and nothing you don’t.
-
- Flask has many configuration values, with sensible defaults, and a few conventions when getting started.
-By convention, templates and static files are stored in subdirectories within the application’s Python source tree, with the names templates and static respectively.
-While this can be changed, you usually don’t have to, especially when getting started.
-
- A Python web application based on WSGI has to have one central callable object that implements the actual application.
-In Flask this is an instance of the Flask class.
-Each Flask application has to create an instance of this class itself and pass it the name of the module, but why can’t Flask do that itself?
-
-\begin{lstlisting}[style=custompython]
-
-from flask import Flask
-app = Flask(__name__)
-
-@app.route('/')
-def index():
-    return 'Hello World!'
-
-\end{lstlisting}
-
-\subsection*{NOTHING PREVENTS GENERALIZATION}
-In this article we take as an example the Flask case study.
-But there is nothing that prevents the generalization of 
-our approach.
-
-MAKE SURE THAT THIS IS CLEAR.
-
-    \begin{figure*}[ht!]
-      \centering
-      \includegraphics[width=0.8\linewidth]{interceptor}
-      \caption{The first thing that needs to be done, is to decorate the application object with an INTERCEPTOR}
-      \label{fig:sep}
-    \end{figure*}
-
-\newpage
-
-\section{Monitoring Service Utilization}
-
-REMEMBER THE GOAL: Our main goal with the \tool issues to allow the integration of the dashboard with an API with minimal effort.
-
-\ins{The first important question that such a 
-service monitoring system should provide is 
-information about what endpoints are being
-used and by whom}.
-
-TODO: Find references, arguments, related work
-that provide more details about why understanding
-who uses endpoints, and which are used are important.
-
-This is clearly the case in static analysis (see paper
-by haenni and lungu...) but should even more be the
-case in APIs.
-
-  \newpage
-TODO: TALK ABOUT THE DATABASE...
-
-  As discussed in the introductory section, in previous work~\cite{vogel2017low} we introduced \tool, a drop-in Python library that allows developers to monitor their Flask-based Python web applications with minimal effort.
-%
-  The \tool is implemented for Python 3.6 and is available on the Python Package Index repository\footnote{\url{https://pypi.python.org/pypi/flask-monitoring-dashboard/1.8}} from where it can be installed by running \install from the command line.
-%  
-  The source code of the \tool is published under a permissive MIT license and is available on GitHub\footnote{\url{https://github.com/flask-dashboard}}.
-
-  \newpage
-  \section{One Line Of Code Configuration With Technology Specific Interceptors}
-
-  The fact that the \tool itself is being developed in Python using Flask makes binding to the services of a to-be monitored application developed with the same technologies easy and intuitive.
-
-  \ins{In general, what matters is that with the right backend, we can do this minimal configuration for any technology... }
-
-  To start using the \tool, one simply needs to one lines of code\footnote{we don't count the import statement that enables that line...} to their Flask web service\footnote{\ins{ In this paper we present the integration with APIs written in Python and Flask hoping that this will not prevent the reader from seeing the more general idea; all the tools we show here for Flask can be applied to other API technologies (e.g.
-Django) by simply providing a few back-end adapters in the right places.
-}}:
-
-  % caption=Configuring the \tool is straightforward,
-  \begin{lstlisting}[style=custompython]
-  import dashboard
-
-  # flask_app is the Flask app object
-  dashboard.bind(flask_app)
-  \end{lstlisting}
-
-  \ins{The dashboard can take advantage of the fact that it is monitoring an API...
-so it can simply just add one extra endpoint there.} After (re)deployment of the application, the interactive visualizations of the \tool become available at the \code{/dashboard} route of the Flask application.
-
-  During binding, the \tool will search for all endpoints defined in the target application.
-
-  In order to monitor an endpoint, the \tool creates a function wrapper for the API function that corresponds to the endpoint.
-This way, the wrapper will be executed whenever that API call is made before the actual function is called.
-The wrapper contains the code that takes care of monitoring an endpoint.
-Data collected by the wrappers are persisted in a local database.
-The SQLAlchemy Object Relational Mapper\footnote{\url{https://www.sqlalchemy.org/}} on top of SQLite\footnote{\url{https://www.sqlite.org/}} is used for this purpose.
-  \ins{DISCUSSION about how do to this in other languages? Java: AspectJ probably, etc.}
-
-  These will be presented to the user in the tool web interface, where the user can select the ones that should be monitored, see \Fref{fig:sep}.
-
-    \begin{figure}[h!]
-      \centering
-      \includegraphics[width=\linewidth]{selecting_endpoints.png}
-      \caption{Once connected to an API the Dashboard presents the endpoints that are available for monitoring}
-      \label{fig:sep}
-    \end{figure}
-
-  This is the alternative to using annotations like other white-box tools (e.g. the other flask performance plugin).
-
-  \ins{PROMISE:} The last access time of the endpoint by users, irrespective of whether it is selected by the user to be monitored or not, is also displayed as the means of identifying potentially obsolete endpoints.
-
-  \ml{we must insert at least one dashboard view here...
-otherwise people will lose their patience...
-also it will help with explaining the outliers...
-also, one should see that a large variety of views are already available at this point...
-I know this risks being  }
-
-In the Zeeguu case study, one of the slowest endpoints, and one with the highest variability as shown in \Fref{fig:ep} is \epFeedItems: it retrieves a list of recommended articles for a given user.
-However, since a user can be subscribed to anything from one to three dozen article sources, and since the computation of the difficulty is personalized and it is slow, the variability in time among users is likely to be very large.
-
-\begin{figure*}[h!]
-  \centering
-  \subfloat[number of requests per endpoint per day]{\includegraphics[width=.9\columnwidth]{number_of_requests_}\label{fig:aeu}}
-  \subfloat[requests per hour heatmap]{\includegraphics[width=\columnwidth]{daily_patterns_}\label{fig:dp}}
-  \caption{Some of the available views\label{fig:views}}
-\end{figure*}
-
-\newpage
-\section{Service Utilization}
-\label{sec:util}
-
-\ins{A discussion must be added here about the types of useful information and problems that can be solved with these views.}
-
-\textit{Service Utilization} presents information about the usage of all endpoints of interest to the developer,
-
-The most fundamental insight that a service maintainer needs regards service utilization. %\vspace{0.5cm}
-
-Simplest utilization perspective:
-
-\begin{figure}[h!]
-  \centering
-  \includegraphics[width=0.5\linewidth]{endpoint-call-count}
-  \caption{Caption here}
-  \label{fig:figure1}
-\end{figure}
-
-% REQUIREMENT: Adding a time interval is important. 
-% Otherwise the performance numbers in chart 1 are not good enough... also an outlier very far in the past, will make all the heatmap data points to look very pale in comparison forever after.
-
-Figure \ref{fig:aeu} shows a first perspective on endpoint utilization that \tool provides: a stacked bar chart of the number of hits to various endpoints grouped by day\footnote{Endpoint colors are the same in different views}.
-Figure~\ref{fig:aeu} in particular shows that at its peak the API has about 12.000 hits per day.
-
-The way users interact with the platform can also be inferred since the endpoints are indicators of different activity types, e.g.:
-
-\begin{enumerate}
-
-  \item {\color{myred} \epTranslations} is an indicator of the amount of foreign language reading the users are doing, and
-
-  \item {\color{mygreen} \epOutcome} is an indicator of the amount of foreign vocabulary practice the users are doing.
-
-\end{enumerate}
-
-%\begin{figure}[!ht]
-% \centering
-% \includegraphics[width=\linewidth]{number_of_requests_}
-% \caption{The number of requests per endpoint per day view shows the overall utilization of the monitored application}
-% \label{fig:aeu}
-%\end{figure}
-
-Besides showing the overall utilization, this endpoint provides the maintainer with information relevant for decisions regarding endpoint deprecation --- one of the most elementary ways of {\em understanding the needs of the downstream}\cite{Haen14a}.
-In our case study, the maintainer realized that one endpoint which they thought was not being used (i.e.
-\code{words\_to\_study}), contrary to their expectations, was actually being used\footnote{A complementary type of usage information can also be discovered in the view presented in Figure \ref{fig:sep} where seeing that an endpoint is never accessed can increase the confidence of the maintainer that a given endpoint is not used, although it can never be used a proof.}.
-
-\niceseparator
-
-%   \todo{Add the time series graph and discuss it before the heatmap? We can then sell the heatmap better} 
-%   \ml{Not sure about which graph you refer to here V}
-
-A second type of utilization question that the \tool can answer automatically regards {\em cyclic patterns of usage per hour of day} by means of a heatmap, as in \Fref{fig:dp}.
-
-% \mltp{ can we add vertical lines that highlight the beginning of a new week (e.g. before Sunday): }
-% Patrick: adding separators in the graph is unfortunately not supported by the library.
-
-%\begin{figure}[!ht]
-% \centering
-% \includegraphics[width=0.9\linewidth]{daily_patterns_}
-% \caption{Usage patterns become easy to spot in the requests per hour heatmap}
-% \label{fig:dp}
-%\end{figure}
-
-Figure \ref{fig:dp} shows the API not being used during the early morning hours, with most of the activity focused around working hours and some light activity during the evening.
-This is consistent with the fact that the current users are all in the central European timezone.
-Also, the figure shows that the spike in utilization that was visible also in the previous graph happended in on afternoon/evening.
-
-\newpage
-\section{Endpoint Performance}
-\label{sec:perf}
-
-  The \tool also collects information regarding endpoint performance. The view in \Fref{fig:ep} summarizes the response times for various endpoints by using a box-and-whiskers plot.
-
- \begin{figure}[!ht]
-   \centering
-   \includegraphics[width=0.95\linewidth]{endpoint_performance_}
-   \caption{The response time (in ms) per monitored endpoint view allows for identifying performance variability and balancing issues}
-   \label{fig:ep}
- \end{figure}
-
-  From this view it became clear to the maintainer that four of the endpoints had very large variation in performance.
-The most critical for the application and consequently the one optimized first was the \epTranslations endpoint which was part of an interactive loop in the reader applications that relied on the Zeeguu API.
-Moreover, cf.
-\Fref{fig:aeu} this endpoint is one of the most used in the system.
-
-\begin{figure}[h!]
-  \centering
-  \subfloat[overview of the tracked endpoints]{\includegraphics[width=0.8\columnwidth]{overview}\label{fig:overview}}
-  \caption{Some of the available views\label{fig:views}}
-\end{figure}
-
-  \newpage
-  \section{Evolution monitoring }
-
-  \begin{figure}[h!]
-  \centering
-  \includegraphics[width=0.8\linewidth]{translation_endpoint_evolution_}
-  \caption{The Performance Evolution of the \epTranslations endpoint}
-  \label{fig:tee}
-\end{figure}
-
-  When thinking about performance, 
-  one can not avoid thinking about the performance evolution.
-
-  Version control in \tool can be supported in two ways,
-  the developer explicitly states the current version, 
-  or the current version is automatically detected based
-  on some version control system.
-
-  However, with the current configuration of the tool, it would be impossible for the maintainer to see the improvements resulting from the optimization.
-
-  \ins{The latest STack Overflow Developer Survey: 88.4\% of **professional** developers use git.}  
-  If assume that the code that they run is deployed using .git, then with an extra line of configuration they can allow \tool to find the git\footnote{\url{https://git-scm.com/}} folder of the deployed service and automatically detect the version of the project that is running:
-
-    \begin{lstlisting}[style=custompython]
-
-      dashboard.config.git = 'path/to/.git'
-
-    \end{lstlisting}
-
-  If the \tool can automatically detect the current version of the project by reading the .git configuration as soon as the API is started and can then group measurements by version\footnote{Alternatively, the maintainer can add version identifiers manually for the web application through a configuration file if the system does not use git.}.
-\Fref{fig:tee} is a zoomed-in version of such a view for \epTranslations with versions increasing from top to bottom
-
-  This view confirms that the performance of the translation endpoint improved in the recent versions: the median of the last three versions is constantly moving towards the left, and progresses from 1.4 seconds (in the top-most box plot in \Fref{fig:tee}) to 0.8 in the latest version (bottom-most box plot).
-
-  CHALLENGE: Supporting other types of version control... not real challenge, as we said 88percent of devs use git. 
-  CHALLENGE 2: detecting minor versions which don't need to be tracked, since they didn't touch the performance of the system. E.g. a modification of the README, etc.
-
-\todo{TODO: endpoint evolution }
-
-  \begin{figure*}[h!]
-    \centering
-    \includegraphics[width=\linewidth]{outliers}
-    \caption{Automatically collected outlier information}
-    \label{fig:figure1}
-  \end{figure*}
-  \newpage
-  \section{Automated outlier detection and monitoring}
-
-  When an API is called from within a highly interactive application (as it is the case with the case study in this paper) 
-  of particular interest to the API developers are performance {\em outliers}.
-
-  Indeed, a translation request that takes three times more than expected can seriously decrease the perceived quality of the application.
-Thus, identifying, collecting all appropriate data, and diagnosing the root causes of such outliers is especially critical in improving the quality of an application.
-
-  % In the context of the RESTful services support by Flask, this means request serving times that deviate from the average to an unexpected degree.
-
-  For this purpose the \tool tracks for every monitored endpoint a {\em running average} response time value\footnote{\ins{For performance reasons, we assume that the response times for the endpoints are normally distributed.
-Otherwise, more general density distribution information must be collected in real time.}}.
-When it detects that a given request is an outlier with respect to this past average running value, it triggers the {\em outlier data collection routine} which stores \ins{extra information} about the current execution environment.
-A configurable threshold with a default value of $2.5$ times the running average response time is used for this purpose.
-
-  For every detected outlier request, the \tool collects information about the current Python stack trace, CPU load, memory consumption, request parameters, etc.
-in order to allow the maintainer to investigate the causes of these exceptionally slow response times.
-In this way it is possible to get \ins{detailed insight into the operation of the application in the extreme cases without unnecessarily burdening it with logging this information for every request}.
-
-The bottom panel shows the stack trace. 
-In this particular case, it is revealing for the developer to learn that at the time of the stack trace snapshot, the code was in the microsoft\_translator.py: indeed, the system uses as back-end multiple translators, and it has been observed that many of the outliers happen to be waiting in the microsoft translator.
-
-This information has to be corroborated with the observations that neither the memory nor the processor are overloaded at the moment.
-Thus this functionality in microsoft\_translator is really slow in itself, and this is not a result of the machine being overloaded for example.
-
-\subsection {A Special Type of Outlier: Exceptions}
-
-   TODO: probably something for the journal extension ...
+- A Special Type of Outlier: Exceptions. TODO: probably something for the journal extension ...
   statistics about which endpoints fail most often might be useful too.
   same information as the outlier maybe?
-
-\begin{figure*}[!ht]
-  \centering
-  \includegraphics[width=0.4\linewidth]{time_per_user}
-  \caption{The \epFeedItems shows a very high variability across users}
-  \label{fig:tpu}
-\end{figure*}
-
-\newpage
-\section{Grouping Requests}
-\label{sec:user}
-
-For service endpoints which run computations in real time, the maintainer of a system might want to understand the endpoint performance on a per-user basis, especially for situations where the system response time is a function of some individual user load\footnote{E.g.
-in GMail some users have two emails while other have twenty thousand and this induces different response times for different users}.
-
-\todo{remove the configuration part and rephrase the following, the discussion has been partially subsumed by Section~\ref{sec:tool}.}
-
-To enable this, the \tool must be configured to associate an API call with a given user.
-The simplest way is to take advantage of the architecture of Flask applications in which a global \code{flask.request} object can be used to retrieve the session which can in turn lead to user identification:
-
-%\begin{lstlisting}[float,caption=Simply define a custom app-specific function for user retrieval and pass it to the \tool to group information by user,style=custompython]
-\begin{lstlisting}[style=custompython]  
-# app specific way of extracting the user
-# from a flask request object    
-def get_user_id():
-sid = int(flask.request.args['session'])
-session = User.find_for_session(sid)
-return user_id
-
-# attaching the get_user_id function
-dashboard.config.get_group_by = get_user_id
-
-\end{lstlisting}
-
-% \niceseparator
-
-\todo{consider losing one or both of the figures --- or at least updating them}
-
-In Zeeguu, \epFeedItems retrieves a list of recommended articles for a given user.
-Cf.
-\Fref{fig:ep} it is the endpoint with the slowest response time and highest variability.
-The reason for this is that a user can be subscribed to anything from one to three dozen article sources and for each of the sources the system must compute the personalized difficulty of each article at every request.
-
-A \perspective{Per-User Performance} perspective should show the different response times for different users.
-Figure \ref{fig:tpu} presents a subset of the corresponding view in the \tool.
-The figure shows that the response times for this endpoint can vary considerably for different users with some extreme cases where a user has to wait a full minute until their recommended articles are shown\footnote{After seeing this perspective, the maintainer refactored the architecture of the system to move part the difficulty computation out of the interactive loop}.
-
-The limitation of the previous view is that it does not present the information also on a per version basis.
-To address this, a different visual perspective entitled \perspective{Multi-Version per-User Performance} can be defined.
-Figure \ref{fig:tuv} presents such a perspective by mapping the average execution time for a given user (lines) and given version (columns) on the area of the corresponding circle.
-
-The colors represent users.
-The figure shows average performance varying  across users and versions with no clear trend: this is probably because varying user workload (i.e.
-number of sources to which the user is registered) is the reason for the variation in response times.
-\ins{one can see that for user 1 performance degrades over the }
-
-\begin{figure}[!ht]
-  \centering
-  \includegraphics[width=0.7\linewidth]{time_per_user_per_version}
-  \caption{This perspective shows that the evolution of response times for individual users (horizontal lines) across versions (the x-axis)}
-  \label{fig:tuv}
-\end{figure}
-
-  \newpage
-  \section{Preemptive monitoring}
-
-  \ins{INTRO - LINKING WITH PREVIOUS PARAGRAPH: the previous evolutionary view crucial for 
-  understanding the impact of software evolution on performance.
-  However, just by observing this one can't be sure whether 
-  performance changes because of the code or because of the payload.}
-
-  \todo{the following might need a bit of rewrite}
-
-  The concept of {\em preemptive monitoring} of the application performance by means of instrumenting integration unit testing as the synthetic load is similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}, i.e.~testing service-based applications by using dedicated test input in parallel to its normal use and operation.
-The difference is that we take advantage of the capability of the CI framework (i.e.
-Travis in this case) to create an emulated ``live'' environment for integration testing purposes, and use unit testing as the dedicated test input in order to measure performance.
-
-  For example, \Fref{fig:builds} is a screenshot from the dashboard showing the measured response times for 5 consequent builds, with 180 iterations of the unit tests executed in total across all endpoints.
-The outliers on the right of the figure are due to initial requests \ins{which must wait for a boot-up phase of the API}.
-
-    \begin{figure}[h!]
-        \centering
-        \includegraphics[width=\columnwidth]{travis_builds}
-        \caption{Response times in 5 consequent Travis builds in the \zee case study}
-        \label{fig:builds}
-      \end{figure}
-
-  While this integration testing environment is different from the actual production deployment one, and the load used is purely synthetic, as will show in Section~\ref{sec:evaluation}, it can actually be used successfully as an early performance indicator for the application developer.
-\ins{Although predicting the performance of the application would require advanced statistical work, it can be succesfully used as an advanced warning system for the overall performance trend for given endpoints}.
-
-  \paragraph{Integration with Travis}
-  \ml{added new section... also, this has to be moved after introducing the pre-emptive monitoring since it's about running tests. switched the order}
-
-  \ins{The pre-emptive monitoring system} can be configured to work together with Continuous Integration (CI) frameworks like Travis\footnote{\url{https://travis-ci.org/}} that deal with automated integration testing.
-  The developer needs to define the unit test folder of the project, the URL that Travis should use for reporting its results, and the number of iterations for each unit test should be executed in order to generate a load for the application.
-
-  Each time a new build is created through Travis, the \tool automatically detects all available unit tests defined by the application developer and iterates through each one of them the number of times predefined by the developer while monitoring the response times for each request.
-The resulting measurements are persisted in a separate part of the tool database so as not to contaminate the ``live'' monitoring data.
-Beyond seeing the results of each Travis build in the \tool, this feature also allows for preemptive monitoring of the application, as we discuss in the following.
-
-  \ins{Besides functioning as an early warning system for performance, tracking the evolving performance of API tests serves a second purpose.
-To function as an anchor when the developer analyzes performance degradations.
-For example, when a developer sees that a given endpoint has become less performant after in a newly deployed version, they are able to investigate whether the performance degradation is visible also with the synthetic load.
-This would correspond to a performance degradation which is due to ``algorithmic performance degradation''.
-If on the other hand, the performance of the tests does not change between versions, the developer might conclude that the performance degradation could be due to the workload on the machines, or maybe to the workload mix of the users}
-
-  %By these means, an application developer can use unit tests as synthetic loads for driving the performance monitoring of the application before it reaches the production phase.
-As we will discuss further in Section~\ref{sec:testing}, this feature is integral in enabling preemptive performance monitoring of the application.
-
-\todo{here goes the discussion about our forecasting capabilities}
-
-\begin{figure}[h!]
-  \centering
-  \subfloat[The reported response time (in ms) per deployed version in the observation period]{\includegraphics[width=.5\columnwidth]{response_per_version_trunced_trend}}
-  \quad
-  \subfloat[The measured response time (in ms) using integration with Travis]{\includegraphics[width=.46\columnwidth]{travis_builds_no_outliers_trend}}
-  \caption{Comparison of the response times per endpoint: actual production system versus preemptive monitoring data}
-  \label{fig:preemptive}
-\end{figure}
-
-\begin{table}[h]
-  \begin{tabular}{lll}
-    \toprule
-    Iteration & \bfseries Live (median) & \bfseries Travis (median)\\
-    \midrule
-    1 & 1349.41 & 764.87\\ 
-    2 & 1466.13 & 992.87\\
-    3 & 1256.65 & 760.87\\
-    4 & 1266.42 & 813.89\\
-    5 & 1080.68 & 303.4\\
-    \bottomrule
-  \end{tabular}
-\end{table}
-
-Pearson correlation $r(3)=.93, p=.02$
-
-\section{Performance Perspectives}
-\label{sec:views}
-
-\ins{PERFORMANCE DISCUSSION: this introspection, which looks up endpoints, happens only once at the startup of the service, so it is going to affect the API startup performance, but not the actual endpoint performance.
-In our case, the overhead here was quite small: TO MEASURE}.
-
-  %In the remainder of the paper we present several of these perspectives.\footnote{We recommend obtaining a color version of this paper for better readability}
-
-  % The second endpoint conists of two parts, one of them being a table that shows for every monitored endpoint the number of hits it has gotten, the time it was last accessed and its average execution time.
-The second part is a view with four graphs which show:
-  % \begin{itemize} 
-  %   \item A heatmap of the total number of requests to the monitored endpoints
-  %   \item A stacked bar chart that shows the total number of requests to the monitored endpoints per endpoint per day
-  %   \item A boxplot graph showing the average execution time per version of the web service
-  %   \item A boxplot graph showing the average execution time for every monitored endpoint
-  % \end{itemize}
-
-\todo{evaluate which figures to keep and ideally take new screenshots for the ones we keep; consider adding two more views (like Fig. 3.2 of Patrick's thesis)}
-
-% In the Zeeguu case study, one of the slowest endpoints, and one with the highest variability as shown in \Fref{fig:ep} is \epFeedItems: it retrieves a list of recommended articles for a given user.
-However, since a user can be subscribed to anything from one to three dozen article sources, and since the computation of the difficulty is personalized and it is slow, the variability in time among users is likely to be very large.
-
-% \niceseparator
-
-%\section{Preemptive Monitoring}
-%\label{sec:testing}
-%
-%%\todo{discuss here idea of using unit testing as an early performance indicator for evolving services (`preemptive monitoring'?); add a side-by-side figure comparing unit testing vs live system performance per version, maybe calculate error in trend curves?
-find a better term than prediction for the section title (plus introduction)}
-%
-%\todo{Similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}, i.e.~testing service-based applications by using dedicated test input in parallel to its normal use and operation.
-The difference is that we take advantage of the capability of Travis to create an emulated live environment for integration testing purposes, and use unit testing as the dedicated test input in order to measure performance.
-What we are going provide evidence for in the following is that we can indeed use this preemptive monitoring concept as an early performance indicator for evolving services.}
-%
-%The developer needs to define the unit test folder of the project, the URL that Travis should use for reporting its results, and the number of iterations for each unit test should be executed in order to generate
-
-\section{Evaluation}
-\label{sec:evaluation}
-
-%\todo{evaluation of the \tool across the following dimensions: ease of use, effectiveness, overhead to the system; for the last part we need to do the overhead measurement experiment, for the other two we base it on the case study and bring over whatever is necessary from the discussion section}
-
-\section{Discussion}
 
@@
     In general, providing a mechanism that would allow very easy specification of groupings (either as code annotations, as normal code, or as configuration options) is an open problem that \tool and any other similar library will have to face.
-
-\section{Related Work}
-\label{sec:related}
-
-\todo{Refer the reader to~\cite{ghezzi2007run} and ~\cite{metzger2010analytical} for a more extensive discussion}
-
-There is a long tradition of using visualization for gaining insight into software performance.
-Tools like Jinsight \cite{Pauw02a} and Web Services Navigator \cite{Pauw05} pioneered such an approach for Java and for Web Services that communicate with SOAP messages.
-Both have an ``omniscient'' view of the services / objects and their interactions.
-As opposed to them, in our work we present an analytics platform which focuses on monitoring a single Python web service from its own point of view.
-
-From the perspective of service monitoring, our work falls within the server-side run-time monitoring of services ~\cite{ghezzi2007run}.
-While we don't implement the more advanced features of related monitoring solutions like QoS policies driving the monitoring, it presents nevertheless an easy to use approach support improving the performance of web applications.
-
-% \va{Mircea: Consider removing the rest for space...}
-% An existing monitoring tool is Pingdom \footnote{https://www.pingdom.com/company/why-pingdom}, which monitors the uptime of an existing web-service.
-This tool works by pinging the websites (up to 60 times) every minute automatically.
-Thus this creates a lot of overhead and is bound to be noisy since it will also be influenced by the speed of the network connection\footnote{Another problem is that such a tool would }
-
-% \todo{Runscope? Others?}
-
-\section{Conclusion and Future Work}
-\label{sec:conclusions}
-
-\todo{update as appropriate}
-
-In this paper we have shown that it is possible to create a monitoring solution which provides basic insight into web service utilization and performance  with very little effort from the developer.
-The user group that we are aiming for with this work is application developers using Flask and Python to build web applications with limited or no budget for implementing their own monitoring solutions.
-The emphasis is in allowing such users to gain insight into how the performance of the service evolves together with the application itself.
-We believe that the same architecture, and lessons can be applied to other frameworks and other languages.
-
-In the future, we plan to perform case studies with other sytstems, with the goal of discover other needs and to wean out the less useful visualizations in the \tool.
-We plan to also extend the tool towards supporting multiple deployments of the same applications across multiple nodes (e.g.
-for the situations where the application is deployed together with a load balancer).
-Finally, we plan to integrate \tool with unit testing as a complementary source of information about performance evolution.
-
-\newpage
-\appendix
-
-  \subsection{Changing the Default /dashboard Endpoint }
-
-  THIS STUFF SHOULD BE MOVED TO THE APPENDIX.
-And in a footnote we just mention that ...
-\ins{REQUIREMENT}: The system must provide an easy way to add more specific configurations if needed.
-A way of overwriting the common sense defaults.
-
-  Further configuration is not needed possible by adding additional statements before the binding definition, for example,
-
-  \begin{lstlisting}[style=custompython]
-  ...
-  dashboard.config.link = 'mydashboard'
-  dashboard.bind(flask_app)
-  \end{lstlisting}
-
-  allows for a custom route (\code{/mydashboard}) to the dashboard to be defined by the programmer. An external configuration file can be used instead, or in addition to this:
-
-  \begin{lstlisting}[style=custompython]
-  ...
-  dashboard.config.from_file('config.cfg')
-  ...
-  \end{lstlisting}
-
-  \subsection{For those who don't use GIT}
-
-    \textit{Manual} version control requires the developer to tag each new version of the application with an appropriate version identifier~\cite{papazoglou2011managing} using the \code{APP\_VERSION} configuration parameter, for example by adding to the configuration file:
-
-      \begin{lstlisting}[style=custompython]
-      [dashboard]
-      APP_VERSION=<versionID>
-      \end{lstlisting}
-
-% references section
-
-\bibliographystyle{abbrv}
-\bibliography{paper}
-
-% that's all folks
-\end{document}

### 2018-04-03 evolution.tex
split the huge file in smaller ones.

--- 2018-03-28 restructured.tex
@@
-\documentclass[conference]{IEEEtran}
-	\pdfpagewidth=8.5truein
-	\pdfpageheight=11truein
-
-\usepackage{graphicx}
-\usepackage{subfig}
-\usepackage[bookmarks=false]{hyperref}
-\usepackage{xspace}
-\usepackage{listings}
-\usepackage[usenames, dvipsnames]{color}
-
-\usepackage{booktabs}
-
-\graphicspath{{./img/}}
-
-%\hyphenation{op-tical net-works semi-conduc-tor}
-
-\input{useful_defines}
-
-\begin{document}
-
-%
-\title{Agile Monitoring of Evolving Web Services}
-
-\author{
-\IEEEauthorblockN{Patrick Vogel, Thijs Klooster, Vasilios Andrikopoulos, Mircea Lungu}\\
-Johann Bernoulli Institute for Mathematics and Computer Science\\
-University of Groningen, Netherland\\
-Email: \{t.klooster.1,p.p.vogel\}@student.rug.nl, \{v.andrikopoulos,m.f.lungu\}@rug.nl 
-}
-
-% make the title area
-\maketitle
-
-\begin{abstract}
-\todo{Update abstract \& title}
-  Tens of thousands of web applications are written in Flask, a Python-based web framework.
-Despite a rich ecosystem of extensions, there is none that supports the developer in gaining insight into the evolving performance of their service.
-In this paper, we introduce \tool, a library that addresses this problem.
-We present the ease with which the library can be integrated in an already existing web application, discuss some of the visualization perspectives that the library provides and point to some future challenges for similar libraries.
-
-\end{abstract}
-
-\section{Introduction}
-
-\todo{Intro to be rewritten; refer to VISSOFT paper as previous work, explain focus is on integration with version control/CI for evolution support, and use of unit tests as early performance indicator}
-
-\todo{motivate with: monitoring in a world which is becoming more and more service-oriented is essential because it allows to perform three main types of actions: system adaptations to provide the request service at the desired level of quality, enabling flexibility in dealing with changing requirements and modes of operation, and enabling operational awareness through dashboards organizing the collected data~\cite{pernici2016monitoring}.
-In this work we focus on the last part and we discuss how a minimal effort probe-based solution for a particular type of web services can be used to facilitate the other two types of actions by involving the developer.}
-
-\todo{explain: in~\cite{vogel2017low} we introduced the \tool an a high level with a focus on presenting its performance visualization aspects; in this work we only summarize these features by means of introducing some of the provided functionalities of the \tool.
-We focus on discussing how the tool is implemented and operating, provide a deeper look at its capabilities for integration with version control and continuous integration environments, introduce a mechanism for performance prediction, and provide an evaluation of the proposed approach based on a case study.
-}
-
-%Every system is a distributed system nowadays \cite{cavage2013there}.
-Indeed a very large number of applications and web applications are nowadays implemented as two-tier architectures with a front-end implemented with web technologies and a service back-end.
-%\ml{I'm not completely happy with this paragraph}
-{\em There is no getting around it: you are building a distributed system} argues a recent article \cite{cavage2013there}.
-Indeed, even the simplest second-year student project is a web application implemented as two-tier architecture with a Javascript/HTML5 front-end a service backend, usually a REST API.
-
-% \hfill mds
-% Many contemporary programming languages are offering libraries, modules, or frameworks that facilitate the development of such architectures. 
-Python is one of the most popular programming language choices for implementing the back-end of web applications.
-GitHub contains more than 500K open source Python projects and the Tiobe Index\footnote{TIOBE programming community index is a measure of popularity of programming languages, created and maintained by the TIOBE Company based in Eindhoven, the Netherlands} ranks Python as the 4th most popular programming language as of June 2016.
-
-% possible flask summary
-Within the Python community, Flask\footnote{\url{http://flask.pocoo.org/}} is a very popular web framework\footnote{More than 25K projects on GitHub (5\% of all Python projects) are implemented with Flask (cf.
-a GitHub search for ``language:Python Flask'')}.
-It provides simplicity and flexibility by implementing a bare-minimum web server, and thus advertises as a micro-framework.
-The Flask tutorial shows how setting up a simple Flask {\em ``Hello World''} web-service requires no more than 5 lines of Python code \cite{ flask:tutorial}.
-% end of summary
-
-Despite their popularity, to the best of our knowledge, there is no simple solution for monitoring the evolving performance of Flask web applications.
-Thus, every one of the developers of these projects faces one of the following options when confronted with the need of gathering insight into the runtime behavior of their implemented services:
-
-  \begin{enumerate}
-
-    \item Use a commercial monitoring tool which treats the subject API as a black-box (e.g. Pingdom, Runscope). 
-    % , Graphite+Graphana+statd etc.
-
-    \item Implement their own ad-hoc analytics solution, having to reinvent basic visualization and interaction strategies.
-
-    \item Live without analytics insight into their services.
-
-  \end{enumerate}
-
-%\todo{For the first point in the list, we can also argue that analytics solutions like Google Analytics can be used, but they have no notion of versioning/integration with the development life cycle.
-Feel free to cite \cite{papazoglou2011managing} for service evolution purposes}
-
-For projects on a budget (e.g.
-research, startups) the first and the second options are often not available due to time and financial constraints.
-Even when using 3rd-party analytics solutions, a critical insight into the evolution of the exposed services of the web application, is missing because such solutions have no notion of versioning and no integration with the development life cycle.~\cite{papazoglou2011managing}
-
-To avoid projects ending up in the third situation, that of living without analytics, in this paper we present \tool~ --- a low-effort service monitoring library for Flask-based Python web services that is easy to integrate and enables the {\em agile assessment} of service evolution.
-\cite{Nier12b}
-
-As a case study, on which we will illustrate our solution, we are going to use an open source API which, for several years, was in the third of the above-presented situations.
-
-% In the next section, we will present a case study of an open source research API which was for a long time in the third situation presented above -- deployed without analytics insight.
-
-\todo{Add signposting paragraph}
-
-\section{Case Study: The API (Zeeguu)}
-\label{sec:case}
-
-%\todo{Either leave as is or add more info on usage by college if possible}
-
-  \zee\footnote{\url{https://github.com/zeeguu-ecosystem/}} is a platform and an ecosystem of applications for accelerating vocabulary acquisition in a foreign language \cite{Lungu16}. 
-%
-  The architecture of the ecosystem has at its core an API and a series of satellite applications that together offer to a learner three main inter-dependent features:
-
-  \begin{enumerate}
-
-    \item Reader applications that provide effortless translations for those texts which are too difficult for the readers.
-
-    \item Interactive exercises personally generated based on the preferences and past likes of the learner.
-
-    \item Article recommendations which are personalized for the interests of the learner and come with difficulty estimation that helps the learner find articles with the appropriate difficulty.
-
-  \end{enumerate}
-
-  The core API implemented with Flask and Python provides correspondingly three types of functionality: contextual translations, article recommendations, and personalized exercise suggestions.
-In total the API provides a bit less than 50 endpoints, out of which probably a dozen are very frequently used.
-  % The development of the core API itself is a research project.
-
-  At the time of writing this article, the ecosystem consists of a reader web application, a web based exercises platform, and a smartwatch application, which are used at the moment of writing this article by more than \activeUserCount active users.
-The users come from a highschool, a language school, and some users are using it on their own, without any educational context.
-The highest load we observed until now on the API consisted of 12K requests in one day.
-
-  We will use this \zee API as a case study for this paper. 
-  All the figures in this paper are captured from the actual deployment of \tool in the context of the \zee API.
-The figures are interactive offering basic data exploration capabilities: filter, zoom, and details on demand\cite{Shne99a}.
-The \tool deployment for the case study can be accessed publicly\footnote{\url{https://zeeguu.unibe.ch/api/dashboard}; username: {\em guest}, password: {\em soap-sac}}.
-
- % \todo{create new account, update URL if necessary, update activeUserCount, add a couple of sentences of the current state of the case study with the college/language center}
-% \ml{we should consider adding also one section in which the architecture/implementation and main features of the dashboard are presented before going on with discussing them in more depth in the following sections --- this should include a rundown on which views are provided from where (overview or per endpoint)}
-
-\newpage
-
-\section{Case Study: The Technology (Flask)}
-\label{sec:tool}
-
- Flask is a microframework for Python. It IS USED BY THOUSANDS OF PROJECTS. ELABORATE MORE ON THIS... 
- \vspace{2cm}
-
- Flask depends on two external libraries: the Jinja2 template engine and the Werkzeug WSGI toolkit.
-
- “Micro” does not mean that your whole web application has to fit into a single Python file (although it certainly can), nor does it mean that Flask is lacking in functionality.
-The “micro” in microframework means Flask aims to keep the core simple but extensible.
-Flask won’t make many decisions for you, such as what database to use.
-Those decisions that it does make, such as what templating engine to use, are easy to change.
-Everything else is up to you, so that Flask can be everything you need and nothing you don’t.
-
- Flask has many configuration values, with sensible defaults, and a few conventions when getting started.
-By convention, templates and static files are stored in subdirectories within the application’s Python source tree, with the names templates and static respectively.
-While this can be changed, you usually don’t have to, especially when getting started.
-
- A Python web application based on WSGI has to have one central callable object that implements the actual application.
-In Flask this is an instance of the Flask class.
-Each Flask application has to create an instance of this class itself and pass it the name of the module, but why can’t Flask do that itself?
-
-\begin{lstlisting}[style=custompython]
-
-from flask import Flask
-app = Flask(__name__)
-
-@app.route('/')
-def index():
-    return 'Hello World!'
-
-\end{lstlisting}
-
-\subsection*{NOTHING PREVENTS GENERALIZATION}
-In this article we take as an example the Flask case study.
-But there is nothing that prevents the generalization of 
-our approach.
-
-MAKE SURE THAT THIS IS CLEAR.
-
-  \newpage
-  \section{One Line Of Code Configuration With Technology Specific Interceptors}
-
-  The fact that the \tool itself is being developed in Python using Flask makes binding to the services of a to-be monitored application developed with the same technologies easy and intuitive.
-
-  \ins{In general, what matters is that with the right backend, we can do this minimal configuration for any technology... }
-
-  To start using the \tool, one simply needs to one lines of code\footnote{we don't count the import statement that enables that line...} to their Flask web service\footnote{\ins{ In this paper we present the integration with APIs written in Python and Flask hoping that this will not prevent the reader from seeing the more general idea; all the tools we show here for Flask can be applied to other API technologies (e.g.
-Django) by simply providing a few back-end adapters in the right places.
-}}:
-
-  % caption=Configuring the \tool is straightforward,
-  \begin{lstlisting}[style=custompython]
-  import dashboard
-
-  # flask_app is the Flask app object
-  dashboard.bind(flask_app)
-  \end{lstlisting}
-
-  \ins{The dashboard can take advantage of the fact that it is monitoring an API...
-so it can simply just add one extra endpoint there.} After (re)deployment of the application, the interactive visualizations of the \tool become available at the \code{/dashboard} route of the Flask application.
-
-  During binding, the \tool will search for all endpoints defined in the target application.
-
-  In order to monitor an endpoint, the \tool creates a function wrapper for the API function that corresponds to the endpoint.
-This way, the wrapper will be executed whenever that API call is made before the actual function is called.
-The wrapper contains the code that takes care of monitoring an endpoint.
-Data collected by the wrappers are persisted in a local database.
-The SQLAlchemy Object Relational Mapper\footnote{\url{https://www.sqlalchemy.org/}} on top of SQLite\footnote{\url{https://www.sqlite.org/}} is used for this purpose.
-  \ins{DISCUSSION about how do to this in other languages? Java: AspectJ probably, etc.}
-
-  These will be presented to the user in the tool web interface, where the user can select the ones that should be monitored, see \Fref{fig:sep}.
-
-    \begin{figure}[h!]
-      \centering
-      \includegraphics[width=\linewidth]{selecting_endpoints.png}
-      \caption{Once connected to an API the Dashboard presents the endpoints that are available for monitoring}
-      \label{fig:sep}
-    \end{figure}
-
-  This is the alternative to using annotations like other white-box tools (e.g. the other flask performance plugin).
-
-  \ins{PROMISE:} The last access time of the endpoint by users, irrespective of whether it is selected by the user to be monitored or not, is also displayed as the means of identifying potentially obsolete endpoints.
-
-  \ml{we must insert at least one dashboard view here...
-otherwise people will lose their patience...
-also it will help with explaining the outliers...
-also, one should see that a large variety of views are already available at this point...
-I know this risks being  }
-
-In the Zeeguu case study, one of the slowest endpoints, and one with the highest variability as shown in \Fref{fig:ep} is \epFeedItems: it retrieves a list of recommended articles for a given user.
-However, since a user can be subscribed to anything from one to three dozen article sources, and since the computation of the difficulty is personalized and it is slow, the variability in time among users is likely to be very large.
-
-\begin{figure*}[h!]
-  \centering
-  \subfloat[number of requests per endpoint per day]{\includegraphics[width=.9\columnwidth]{number_of_requests_}\label{fig:aeu}}
-  \subfloat[requests per hour heatmap]{\includegraphics[width=\columnwidth]{daily_patterns_}\label{fig:dp}}
-  \caption{Some of the available views\label{fig:views}}
-\end{figure*}
-
-\newpage
-\section{Service Utilization}
-\label{sec:util}
-
-\ins{A discussion must be added here about the types of useful information and problems that can be solved with these views.}
-
-\textit{Service Utilization} presents information about the usage of all endpoints of interest to the developer,
-
-The most fundamental insight that a service maintainer needs regards service utilization. %\vspace{0.5cm}
-
-Simplest utilization perspective:
-
-\begin{figure}[h!]
-  \centering
-  \includegraphics[width=0.5\linewidth]{endpoint-call-count}
-  \caption{Caption here}
-  \label{fig:figure1}
-\end{figure}
-
-% REQUIREMENT: Adding a time interval is important. 
-% Otherwise the performance numbers in chart 1 are not good enough... also an outlier very far in the past, will make all the heatmap data points to look very pale in comparison forever after.
-
-Figure \ref{fig:aeu} shows a first perspective on endpoint utilization that \tool provides: a stacked bar chart of the number of hits to various endpoints grouped by day\footnote{Endpoint colors are the same in different views}.
-Figure~\ref{fig:aeu} in particular shows that at its peak the API has about 12.000 hits per day.
-
-The way users interact with the platform can also be inferred since the endpoints are indicators of different activity types, e.g.:
-
-\begin{enumerate}
-
-  \item {\color{myred} \epTranslations} is an indicator of the amount of foreign language reading the users are doing, and
-
-  \item {\color{mygreen} \epOutcome} is an indicator of the amount of foreign vocabulary practice the users are doing.
-
-\end{enumerate}
-
-%\begin{figure}[!ht]
-% \centering
-% \includegraphics[width=\linewidth]{number_of_requests_}
-% \caption{The number of requests per endpoint per day view shows the overall utilization of the monitored application}
-% \label{fig:aeu}
-%\end{figure}
-
-Besides showing the overall utilization, this endpoint provides the maintainer with information relevant for decisions regarding endpoint deprecation --- one of the most elementary ways of {\em understanding the needs of the downstream}\cite{Haen14a}.
-In our case study, the maintainer realized that one endpoint which they thought was not being used (i.e.
-\code{words\_to\_study}), contrary to their expectations, was actually being used\footnote{A complementary type of usage information can also be discovered in the view presented in Figure \ref{fig:sep} where seeing that an endpoint is never accessed can increase the confidence of the maintainer that a given endpoint is not used, although it can never be used a proof.}.
-
-\niceseparator
-
-%   \todo{Add the time series graph and discuss it before the heatmap? We can then sell the heatmap better} 
-%   \ml{Not sure about which graph you refer to here V}
-
-A second type of utilization question that the \tool can answer automatically regards {\em cyclic patterns of usage per hour of day} by means of a heatmap, as in \Fref{fig:dp}.
-
-% \mltp{ can we add vertical lines that highlight the beginning of a new week (e.g. before Sunday): }
-% Patrick: adding separators in the graph is unfortunately not supported by the library.
-
-%\begin{figure}[!ht]
-% \centering
-% \includegraphics[width=0.9\linewidth]{daily_patterns_}
-% \caption{Usage patterns become easy to spot in the requests per hour heatmap}
-% \label{fig:dp}
-%\end{figure}
-
-Figure \ref{fig:dp} shows the API not being used during the early morning hours, with most of the activity focused around working hours and some light activity during the evening.
-This is consistent with the fact that the current users are all in the central European timezone.
-Also, the figure shows that the spike in utilization that was visible also in the previous graph happended in on afternoon/evening.
+%!TEX root=restructured.tex
 
@@
   \newpage
-\section{Endpoint Performance}
-\label{sec:perf}
-
-  The \tool also collects information regarding endpoint performance. The view in \Fref{fig:ep} summarizes the response times for various endpoints by using a box-and-whiskers plot.
-
- \begin{figure}[!ht]
-   \centering
-   \includegraphics[width=0.95\linewidth]{endpoint_performance_}
-   \caption{The response time (in ms) per monitored endpoint view allows for identifying performance variability and balancing issues}
-   \label{fig:ep}
- \end{figure}
-
-  From this view it became clear to the maintainer that four of the endpoints had very large variation in performance.
-The most critical for the application and consequently the one optimized first was the \epTranslations endpoint which was part of an interactive loop in the reader applications that relied on the Zeeguu API.
-Moreover, cf.
-\Fref{fig:aeu} this endpoint is one of the most used in the system.
-
-  \begin{figure*}[h!]
-    \centering
-    \includegraphics[width=\linewidth]{outliers}
-    \caption{Automatically collected outlier information}
-    \label{fig:figure1}
-  \end{figure*}
-  \newpage
-  \subsection{Performance Insight: Automated outlier detection and monitoring}
-
-  When an API is called from within a highly interactive application (as it is the case with the case study in this paper) 
-  of particular interest to the API developers are performance {\em outliers}.
-
-  Indeed, a translation request that takes three times more than expected can seriously decrease the perceived quality of the application.
-Thus, identifying, collecting all appropriate data, and diagnosing the root causes of such outliers is especially critical in improving the quality of an application.
-
-  % In the context of the RESTful services support by Flask, this means request serving times that deviate from the average to an unexpected degree.
-
-  For this purpose the \tool tracks for every monitored endpoint a {\em running average} response time value\footnote{\ins{For performance reasons, we assume that the response times for the endpoints are normally distributed.
-Otherwise, more general density distribution information must be collected in real time.}}.
-When it detects that a given request is an outlier with respect to this past average running value, it triggers the {\em outlier data collection routine} which stores \ins{extra information} about the current execution environment.
-A configurable threshold with a default value of $2.5$ times the running average response time is used for this purpose.
-
-  For every detected outlier request, the \tool collects information about the current Python stack trace, CPU load, memory consumption, request parameters, etc.
-in order to allow the maintainer to investigate the causes of these exceptionally slow response times.
-In this way it is possible to get \ins{detailed insight into the operation of the application in the extreme cases without unnecessarily burdening it with logging this information for every request}.
-
-The bottom panel shows the stack trace. 
-In this particular case, it is revealing for the developer to learn that at the time of the stack trace snapshot, the code was in the microsoft\_translator.py: indeed, the system uses as back-end multiple translators, and it has been observed that many of the outliers happen to be waiting in the microsoft translator.
-
-This information has to be corroborated with the observations that neither the memory nor the processor are overloaded at the moment.
-Thus this functionality in microsoft\_translator is really slow in itself, and this is not a result of the machine being overloaded for example.
-
-\subsection {A Special Type of Outlier: Exceptions}
-
-   TODO: probably something for the journal extension ...
-  statistics about which endpoints fail most often might be useful too.
-  same information as the outlier maybe?
-
-\begin{figure*}[h!]
-  \centering
-  \includegraphics[width=0.8\linewidth]{utilization-evolution}
-  \caption{The Performance Evolution of the \epTranslations endpoint}
-  \label{fig:tee}
-\end{figure*}
-
-  \newpage
   \section{Evolution monitoring }
-
-  \begin{figure}[h!]
-  \centering
-  \includegraphics[width=0.8\linewidth]{translation_endpoint_evolution_}
-  \caption{The Performance Evolution of the \epTranslations endpoint}
-  \label{fig:tee}
-\end{figure}
 
@@
 
+  \begin{figure}[h!]
+  \centering
+  \includegraphics[width=0.9\linewidth]{translation_endpoint_evolution_}
+  \caption{The Performance Evolution of the \epTranslations endpoint}
+  \label{fig:tee}
+\end{figure}
+
   This view confirms that the performance of the translation endpoint improved in the recent versions: the median of the last three versions is constantly moving towards the left, and progresses from 1.4 seconds (in the top-most box plot in \Fref{fig:tee}) to 0.8 in the latest version (bottom-most box plot).
@@
 \end{figure*}
-
-\newpage
-\section{Grouping Requests}
-\label{sec:user}
-
-For service endpoints which run computations in real time, the maintainer of a system might want to understand the endpoint performance on a per-user basis, especially for situations where the system response time is a function of some individual user load\footnote{E.g.
-in GMail some users have two emails while other have twenty thousand and this induces different response times for different users}.
-
-\todo{remove the configuration part and rephrase the following, the discussion has been partially subsumed by Section~\ref{sec:tool}.}
-
-To enable this, the \tool must be configured to associate an API call with a given user.
-The simplest way is to take advantage of the architecture of Flask applications in which a global \code{flask.request} object can be used to retrieve the session which can in turn lead to user identification:
-
-%\begin{lstlisting}[float,caption=Simply define a custom app-specific function for user retrieval and pass it to the \tool to group information by user,style=custompython]
-\begin{lstlisting}[style=custompython]  
-# app specific way of extracting the user
-# from a flask request object    
-def get_user_id():
-sid = int(flask.request.args['session'])
-session = User.find_for_session(sid)
-return user_id
-
-# attaching the get_user_id function
-dashboard.config.get_group_by = get_user_id
-
-\end{lstlisting}
-
-% \niceseparator
-
-\todo{consider losing one or both of the figures --- or at least updating them}
-
-In Zeeguu, \epFeedItems retrieves a list of recommended articles for a given user.
-Cf.
-\Fref{fig:ep} it is the endpoint with the slowest response time and highest variability.
-The reason for this is that a user can be subscribed to anything from one to three dozen article sources and for each of the sources the system must compute the personalized difficulty of each article at every request.
-
-A \perspective{Per-User Performance} perspective should show the different response times for different users.
-Figure \ref{fig:tpu} presents a subset of the corresponding view in the \tool.
-The figure shows that the response times for this endpoint can vary considerably for different users with some extreme cases where a user has to wait a full minute until their recommended articles are shown\footnote{After seeing this perspective, the maintainer refactored the architecture of the system to move part the difficulty computation out of the interactive loop}.
-
-The limitation of the previous view is that it does not present the information also on a per version basis.
-To address this, a different visual perspective entitled \perspective{Multi-Version per-User Performance} can be defined.
-Figure \ref{fig:tuv} presents such a perspective by mapping the average execution time for a given user (lines) and given version (columns) on the area of the corresponding circle.
-
-The colors represent users.
-The figure shows average performance varying  across users and versions with no clear trend: this is probably because varying user workload (i.e.
-number of sources to which the user is registered) is the reason for the variation in response times.
-\ins{one can see that for user 1 performance degrades over the }
-
-\begin{figure}[!ht]
-  \centering
-  \includegraphics[width=0.7\linewidth]{time_per_user_per_version}
-  \caption{This perspective shows that the evolution of response times for individual users (horizontal lines) across versions (the x-axis)}
-  \label{fig:tuv}
-\end{figure}
-
-  \newpage
-  \section{Preemptive monitoring}
-
-  \ins{INTRO - LINKING WITH PREVIOUS PARAGRAPH: the previous evolutionary view crucial for 
-  understanding the impact of software evolution on performance.
-  However, just by observing this one can't be sure whether 
-  performance changes because of the code or because of the payload.}
-
-  \todo{the following might need a bit of rewrite}
-
-  The concept of {\em preemptive monitoring} of the application performance by means of instrumenting integration unit testing as the synthetic load is similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}, i.e.~testing service-based applications by using dedicated test input in parallel to its normal use and operation.
-The difference is that we take advantage of the capability of the CI framework (i.e.
-Travis in this case) to create an emulated ``live'' environment for integration testing purposes, and use unit testing as the dedicated test input in order to measure performance.
-
-  For example, \Fref{fig:builds} is a screenshot from the dashboard showing the measured response times for 5 consequent builds, with 180 iterations of the unit tests executed in total across all endpoints.
-The outliers on the right of the figure are due to initial requests \ins{which must wait for a boot-up phase of the API}.
-
-    \begin{figure}[h!]
-        \centering
-        \includegraphics[width=\columnwidth]{travis_builds}
-        \caption{Response times in 5 consequent Travis builds in the \zee case study}
-        \label{fig:builds}
-      \end{figure}
-
-  While this integration testing environment is different from the actual production deployment one, and the load used is purely synthetic, as will show in Section~\ref{sec:evaluation}, it can actually be used successfully as an early performance indicator for the application developer.
-\ins{Although predicting the performance of the application would require advanced statistical work, it can be succesfully used as an advanced warning system for the overall performance trend for given endpoints}.
-
-  \paragraph{Integration with Travis}
-  \ml{added new section... also, this has to be moved after introducing the pre-emptive monitoring since it's about running tests. switched the order}
-
-  \ins{The pre-emptive monitoring system} can be configured to work together with Continuous Integration (CI) frameworks like Travis\footnote{\url{https://travis-ci.org/}} that deal with automated integration testing.
-  The developer needs to define the unit test folder of the project, the URL that Travis should use for reporting its results, and the number of iterations for each unit test should be executed in order to generate a load for the application.
-
-  Each time a new build is created through Travis, the \tool automatically detects all available unit tests defined by the application developer and iterates through each one of them the number of times predefined by the developer while monitoring the response times for each request.
-The resulting measurements are persisted in a separate part of the tool database so as not to contaminate the ``live'' monitoring data.
-Beyond seeing the results of each Travis build in the \tool, this feature also allows for preemptive monitoring of the application, as we discuss in the following.
-
-  \ins{Besides functioning as an early warning system for performance, tracking the evolving performance of API tests serves a second purpose.
-To function as an anchor when the developer analyzes performance degradations.
-For example, when a developer sees that a given endpoint has become less performant after in a newly deployed version, they are able to investigate whether the performance degradation is visible also with the synthetic load.
-This would correspond to a performance degradation which is due to ``algorithmic performance degradation''.
-If on the other hand, the performance of the tests does not change between versions, the developer might conclude that the performance degradation could be due to the workload on the machines, or maybe to the workload mix of the users}
-
-  %By these means, an application developer can use unit tests as synthetic loads for driving the performance monitoring of the application before it reaches the production phase.
-As we will discuss further in Section~\ref{sec:testing}, this feature is integral in enabling preemptive performance monitoring of the application.
-
-\todo{here goes the discussion about our forecasting capabilities}
-
-\begin{figure}[h!]
-  \centering
-  \subfloat[The reported response time (in ms) per deployed version in the observation period]{\includegraphics[width=.5\columnwidth]{response_per_version_trunced_trend}}
-  \quad
-  \subfloat[The measured response time (in ms) using integration with Travis]{\includegraphics[width=.46\columnwidth]{travis_builds_no_outliers_trend}}
-  \caption{Comparison of the response times per endpoint: actual production system versus preemptive monitoring data}
-  \label{fig:preemptive}
-\end{figure}
-
-\begin{table}[h]
-  \begin{tabular}{lll}
-    \toprule
-    Iteration & \bfseries Live (median) & \bfseries Travis (median)\\
-    \midrule
-    1 & 1349.41 & 764.87\\ 
-    2 & 1466.13 & 992.87\\
-    3 & 1256.65 & 760.87\\
-    4 & 1266.42 & 813.89\\
-    5 & 1080.68 & 303.4\\
-    \bottomrule
-  \end{tabular}
-\end{table}
-
-Pearson correlation $r(3)=.93, p=.02$
-
-    \begin{figure*}[ht!]
-      \centering
-      \includegraphics[width=0.8\linewidth]{interceptor}
-      \caption{The first thing that needs to be done, is to decorate the application object with an INTERCEPTOR}
-      \label{fig:sep}
-    \end{figure*}
-
-\newpage
-
-\section{Monitoring Service Utilization}
-
-REMEMBER THE GOAL: Our main goal with the \tool issues to allow the integration of the dashboard with an API with minimal effort.
-
-\ins{The first important question that such a 
-service monitoring system should provide is 
-information about what endpoints are being
-used and by whom}.
-
-TODO: Find references, arguments, related work
-that provide more details about why understanding
-who uses endpoints, and which are used are important.
-
-This is clearly the case in static analysis (see paper
-by haenni and lungu...) but should even more be the
-case in APIs.
-
-  \newpage
-TODO: TALK ABOUT THE METAMODEL...
-
-    \begin{figure}[ht!]
-      \centering
-        \includegraphics[width=0.8\linewidth]{db_schema}
-        \caption{The model that supports the views presented in this paper}
-        \label{fig:sep}
-    \end{figure}
-
-  As discussed in the introductory section, in previous work~\cite{vogel2017low} we introduced \tool, a drop-in Python library that allows developers to monitor their Flask-based Python web applications with minimal effort.
-%
-  The \tool is implemented for Python 3.6 and is available on the Python Package Index repository\footnote{\url{https://pypi.python.org/pypi/flask-monitoring-dashboard/1.8}} from where it can be installed by running \install from the command line.
-%  
-  The source code of the \tool is published under a permissive MIT license and is available on GitHub\footnote{\url{https://github.com/flask-dashboard}}.
-
-\section{Performance Perspectives}
-\label{sec:views}
-
-\ins{PERFORMANCE DISCUSSION: this introspection, which looks up endpoints, happens only once at the startup of the service, so it is going to affect the API startup performance, but not the actual endpoint performance.
-In our case, the overhead here was quite small: TO MEASURE}.
-
-  %In the remainder of the paper we present several of these perspectives.\footnote{We recommend obtaining a color version of this paper for better readability}
-
-  % The second endpoint conists of two parts, one of them being a table that shows for every monitored endpoint the number of hits it has gotten, the time it was last accessed and its average execution time.
-The second part is a view with four graphs which show:
-  % \begin{itemize} 
-  %   \item A heatmap of the total number of requests to the monitored endpoints
-  %   \item A stacked bar chart that shows the total number of requests to the monitored endpoints per endpoint per day
-  %   \item A boxplot graph showing the average execution time per version of the web service
-  %   \item A boxplot graph showing the average execution time for every monitored endpoint
-  % \end{itemize}
-
-\todo{evaluate which figures to keep and ideally take new screenshots for the ones we keep; consider adding two more views (like Fig. 3.2 of Patrick's thesis)}
-
-% In the Zeeguu case study, one of the slowest endpoints, and one with the highest variability as shown in \Fref{fig:ep} is \epFeedItems: it retrieves a list of recommended articles for a given user.
-However, since a user can be subscribed to anything from one to three dozen article sources, and since the computation of the difficulty is personalized and it is slow, the variability in time among users is likely to be very large.
-
-% \niceseparator
-
-%\section{Preemptive Monitoring}
-%\label{sec:testing}
-%
-%%\todo{discuss here idea of using unit testing as an early performance indicator for evolving services (`preemptive monitoring'?); add a side-by-side figure comparing unit testing vs live system performance per version, maybe calculate error in trend curves?
-find a better term than prediction for the section title (plus introduction)}
-%
-%\todo{Similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}, i.e.~testing service-based applications by using dedicated test input in parallel to its normal use and operation.
-The difference is that we take advantage of the capability of Travis to create an emulated live environment for integration testing purposes, and use unit testing as the dedicated test input in order to measure performance.
-What we are going provide evidence for in the following is that we can indeed use this preemptive monitoring concept as an early performance indicator for evolving services.}
-%
-%The developer needs to define the unit test folder of the project, the URL that Travis should use for reporting its results, and the number of iterations for each unit test should be executed in order to generate
-
-\section{Evaluation}
-\label{sec:evaluation}
-
-%\todo{evaluation of the \tool across the following dimensions: ease of use, effectiveness, overhead to the system; for the last part we need to do the overhead measurement experiment, for the other two we base it on the case study and bring over whatever is necessary from the discussion section}
-
-\section{Discussion}
-
-\todo{use this subsection to discuss limitations (from the previous evaluation section) and add the need to be able to handle the horizontal scaling of the application; remove sectioning and summarize briefly limitations that have already been discussed in the VISSOFT paper/move material to other sections; consider renaming as limitations or something similar}
-
-  \subsection{Distributed System Monitoring}
-  \todo{To think about}
-  Supporting situations where the API is deployed across multiple containers for example.
-
-  \subsubsection{Automatically Monitoring System Evolution}
-
-  The main goal of the \tool design was to allow analytics to be collected and insight to be gained by making the smallest possible changes to a running API.
-%To allow the collection of evolutionary information
-%
-  This technique assumes that the web application code which is the target of the monitoring is deployed using \git in the following way:
-
-  \begin{enumerate}
-    \item The deployment engineer pulls the latest version of the code from the integration server; this will result in a new commit being pointed at by the HEAD pointer. %than previously
-    \item The deployment engineer restarts the new version of the service.
-At this point, the \tool detects that a new HEAD is present in the local code base and consequently starts associating all the new data points with this new commit\footnote{The \tool detects the current version of the analyzed system the first time it is attached to the application object, and thus, assumes that the Flask application is restarted when a new version is deployed.
-This is in tune with the current version of Flask, but if the web server will support dynamic updates in the future, this might have to be taken into account}.
-  \end{enumerate}
-
-  The advantage of this approach is the need for minimal configuration effort, as discussed in the presentation of the tool.
-The disadvantage is that it will consider on equal ground the smallest of commits, even one that modifies a comment, and the shortest lived of commits, e.g.~a commit which was active only for a half an hour before a new version with a bug fix was deployed, with major and minor releases of the software.
-%as a distinct way of grouping the data points.
-  A mechanism to control which versions are important for monitoring purposes is therefore required to be added.
-%
-  A further possible extension point here is supporting other version control systems (e.g. Mercurial). However, this is a straightforward extension.
-
-  \subsubsection{NEW: API Renames}
-  \ml{just an idea}
-
-    The history would normally be lost in the case of an API rename. (also known as the {\em provenance issue} in source code evolution analysis)
-    In theory we could infer that an API was renamed if it is not to be found anymore, and another one with very similar timing characteristics would be found.
-Such a detection can never be completely sure, but it would help with meeting the needs of the API maintainer...
-
-  \subsubsection{User-Awareness }
-
-    For the situations in which the user information is not available, the \tool tracks by default information about different IPs and in some cases this might be a sufficiently good approximation of the user diversity and identity.
-    %
-
-    The visualizations for the user experiene perspectives as presented in Section \ref{sec:user} have been tested with several hundred users (of which about \activeUserCount were active during the course of the study), but the scalability of the visualizations must be further investigated for web services with tens of thousands of users.
-
-  \subsubsection{Other Possible Groupings}
-
-    There are other groupings of service utilization and performance that could be important to the maintainer, that we did not explore in this paper.
-For example, if the service is using OAuth, then together with every request, in the header of the request there is information about the application which is sending a request.
-Grouping the information by application that sends the request could be important in such a context.
-
-    In general, providing a mechanism that would allow very easy specification of groupings (either as code annotations, as normal code, or as configuration options) is an open problem that \tool and any other similar library will have to face.
-
-\section{Related Work}
-\label{sec:related}
-
-\todo{Refer the reader to~\cite{ghezzi2007run} and ~\cite{metzger2010analytical} for a more extensive discussion}
-
-There is a long tradition of using visualization for gaining insight into software performance.
-Tools like Jinsight \cite{Pauw02a} and Web Services Navigator \cite{Pauw05} pioneered such an approach for Java and for Web Services that communicate with SOAP messages.
-Both have an ``omniscient'' view of the services / objects and their interactions.
-As opposed to them, in our work we present an analytics platform which focuses on monitoring a single Python web service from its own point of view.
-
-From the perspective of service monitoring, our work falls within the server-side run-time monitoring of services ~\cite{ghezzi2007run}.
-While we don't implement the more advanced features of related monitoring solutions like QoS policies driving the monitoring, it presents nevertheless an easy to use approach support improving the performance of web applications.
-
-% \va{Mircea: Consider removing the rest for space...}
-% An existing monitoring tool is Pingdom \footnote{https://www.pingdom.com/company/why-pingdom}, which monitors the uptime of an existing web-service.
-This tool works by pinging the websites (up to 60 times) every minute automatically.
-Thus this creates a lot of overhead and is bound to be noisy since it will also be influenced by the speed of the network connection\footnote{Another problem is that such a tool would }
-
-% \todo{Runscope? Others?}
-
-\section{Conclusion and Future Work}
-\label{sec:conclusions}
-
-\todo{update as appropriate}
-
-In this paper we have shown that it is possible to create a monitoring solution which provides basic insight into web service utilization and performance  with very little effort from the developer.
-The user group that we are aiming for with this work is application developers using Flask and Python to build web applications with limited or no budget for implementing their own monitoring solutions.
-The emphasis is in allowing such users to gain insight into how the performance of the service evolves together with the application itself.
-We believe that the same architecture, and lessons can be applied to other frameworks and other languages.
-
-In the future, we plan to perform case studies with other sytstems, with the goal of discover other needs and to wean out the less useful visualizations in the \tool.
-We plan to also extend the tool towards supporting multiple deployments of the same applications across multiple nodes (e.g.
-for the situations where the application is deployed together with a load balancer).
-Finally, we plan to integrate \tool with unit testing as a complementary source of information about performance evolution.
-
-\newpage
-\appendix
-
-  \subsection{Changing the Default /dashboard Endpoint }
-
-  THIS STUFF SHOULD BE MOVED TO THE APPENDIX.
-And in a footnote we just mention that ...
-\ins{REQUIREMENT}: The system must provide an easy way to add more specific configurations if needed.
-A way of overwriting the common sense defaults.
-
-  Further configuration is not needed possible by adding additional statements before the binding definition, for example,
-
-  \begin{lstlisting}[style=custompython]
-  ...
-  dashboard.config.link = 'mydashboard'
-  dashboard.bind(flask_app)
-  \end{lstlisting}
-
-  allows for a custom route (\code{/mydashboard}) to the dashboard to be defined by the programmer. An external configuration file can be used instead, or in addition to this:
-
-  \begin{lstlisting}[style=custompython]
-  ...
-  dashboard.config.from_file('config.cfg')
-  ...
-  \end{lstlisting}
-
-  \subsection{For those who don't use GIT}
-
-    \textit{Manual} version control requires the developer to tag each new version of the application with an appropriate version identifier~\cite{papazoglou2011managing} using the \code{APP\_VERSION} configuration parameter, for example by adding to the configuration file:
-
-      \begin{lstlisting}[style=custompython]
-      [dashboard]
-      APP_VERSION=<versionID>
-      \end{lstlisting}
-
-% references section
-
-\bibliographystyle{abbrv}
-\bibliography{paper}
-
-% that's all folks
-\end{document}

### 2018-04-03 grouping.tex
split the huge file in smaller ones.

--- 2018-03-21 restructured.tex
@@
-\documentclass[conference]{IEEEtran}
-	\pdfpagewidth=8.5truein
-	\pdfpageheight=11truein
-
-\usepackage{graphicx}
-\usepackage{subfig}
-\usepackage[bookmarks=false]{hyperref}
-\usepackage{xspace}
-\usepackage{listings}
-\usepackage[usenames, dvipsnames]{color}
-
-\usepackage{booktabs}
-
-\graphicspath{{./img/}}
-
-%\hyphenation{op-tical net-works semi-conduc-tor}
-
-\input{useful_defines}
-
-\begin{document}
-
-%
-\title{Agile Monitoring of Evolving Web Services}
-
-\author{
-\IEEEauthorblockN{Patrick Vogel, Thijs Klooster, Vasilios Andrikopoulos, Mircea Lungu}\\
-Johann Bernoulli Institute for Mathematics and Computer Science\\
-University of Groningen, Netherland\\
-Email: \{t.klooster.1,p.p.vogel\}@student.rug.nl, \{v.andrikopoulos,m.f.lungu\}@rug.nl 
-}
-
-% make the title area
-\maketitle
-
-\begin{abstract}
-\todo{Update abstract \& title}
-  Tens of thousands of web applications are written in Flask, a Python-based web framework.
-Despite a rich ecosystem of extensions, there is none that supports the developer in gaining insight into the evolving performance of their service.
-In this paper, we introduce \tool, a library that addresses this problem.
-We present the ease with which the library can be integrated in an already existing web application, discuss some of the visualization perspectives that the library provides and point to some future challenges for similar libraries.
-
-\end{abstract}
-
-\section{Introduction}
-
-\todo{Intro to be rewritten; refer to VISSOFT paper as previous work, explain focus is on integration with version control/CI for evolution support, and use of unit tests as early performance indicator}
-
-\todo{motivate with: monitoring in a world which is becoming more and more service-oriented is essential because it allows to perform three main types of actions: system adaptations to provide the request service at the desired level of quality, enabling flexibility in dealing with changing requirements and modes of operation, and enabling operational awareness through dashboards organizing the collected data~\cite{pernici2016monitoring}.
-In this work we focus on the last part and we discuss how a minimal effort probe-based solution for a particular type of web services can be used to facilitate the other two types of actions by involving the developer.}
-
-\todo{explain: in~\cite{vogel2017low} we introduced the \tool an a high level with a focus on presenting its performance visualization aspects; in this work we only summarize these features by means of introducing some of the provided functionalities of the \tool.
-We focus on discussing how the tool is implemented and operating, provide a deeper look at its capabilities for integration with version control and continuous integration environments, introduce a mechanism for performance prediction, and provide an evaluation of the proposed approach based on a case study.
-}
-
-%Every system is a distributed system nowadays \cite{cavage2013there}.
-Indeed a very large number of applications and web applications are nowadays implemented as two-tier architectures with a front-end implemented with web technologies and a service back-end.
-%\ml{I'm not completely happy with this paragraph}
-{\em There is no getting around it: you are building a distributed system} argues a recent article \cite{cavage2013there}.
-Indeed, even the simplest second-year student project is a web application implemented as two-tier architecture with a Javascript/HTML5 front-end a service backend, usually a REST API.
-
-% \hfill mds
-% Many contemporary programming languages are offering libraries, modules, or frameworks that facilitate the development of such architectures. 
-Python is one of the most popular programming language choices for implementing the back-end of web applications.
-GitHub contains more than 500K open source Python projects and the Tiobe Index\footnote{TIOBE programming community index is a measure of popularity of programming languages, created and maintained by the TIOBE Company based in Eindhoven, the Netherlands} ranks Python as the 4th most popular programming language as of June 2016.
-
-% possible flask summary
-Within the Python community, Flask\footnote{\url{http://flask.pocoo.org/}} is a very popular web framework\footnote{More than 25K projects on GitHub (5\% of all Python projects) are implemented with Flask (cf.
-a GitHub search for ``language:Python Flask'')}.
-It provides simplicity and flexibility by implementing a bare-minimum web server, and thus advertises as a micro-framework.
-The Flask tutorial shows how setting up a simple Flask {\em ``Hello World''} web-service requires no more than 5 lines of Python code \cite{ flask:tutorial}.
-% end of summary
-
-Despite their popularity, to the best of our knowledge, there is no simple solution for monitoring the evolving performance of Flask web applications.
-Thus, every one of the developers of these projects faces one of the following options when confronted with the need of gathering insight into the runtime behavior of their implemented services:
-
-  \begin{enumerate}
-
-    \item Use a commercial monitoring tool which treats the subject API as a black-box (e.g. Pingdom, Runscope). 
-    % , Graphite+Graphana+statd etc.
-
-    \item Implement their own ad-hoc analytics solution, having to reinvent basic visualization and interaction strategies.
-
-    \item Live without analytics insight into their services.
-
-  \end{enumerate}
-
-%\todo{For the first point in the list, we can also argue that analytics solutions like Google Analytics can be used, but they have no notion of versioning/integration with the development life cycle.
-Feel free to cite \cite{papazoglou2011managing} for service evolution purposes}
-
-For projects on a budget (e.g.
-research, startups) the first and the second options are often not available due to time and financial constraints.
-Even when using 3rd-party analytics solutions, a critical insight into the evolution of the exposed services of the web application, is missing because such solutions have no notion of versioning and no integration with the development life cycle.~\cite{papazoglou2011managing}
-
-To avoid projects ending up in the third situation, that of living without analytics, in this paper we present \tool~ --- a low-effort service monitoring library for Flask-based Python web services that is easy to integrate and enables the {\em agile assessment} of service evolution.
-\cite{Nier12b}
-
-As a case study, on which we will illustrate our solution, we are going to use an open source API which, for several years, was in the third of the above-presented situations.
-
-% In the next section, we will present a case study of an open source research API which was for a long time in the third situation presented above -- deployed without analytics insight.
-
-\todo{Add signposting paragraph}
-
-\section{Case Study: The API (Zeeguu)}
-\label{sec:case}
-
-%\todo{Either leave as is or add more info on usage by college if possible}
-
-  \zee\footnote{\url{https://github.com/zeeguu-ecosystem/}} is a platform and an ecosystem of applications for accelerating vocabulary acquisition in a foreign language \cite{Lungu16}. 
-%
-  The architecture of the ecosystem has at its core an API and a series of satellite applications that together offer to a learner three main inter-dependent features:
-
-  \begin{enumerate}
-
-    \item Reader applications that provide effortless translations for those texts which are too difficult for the readers.
-
-    \item Interactive exercises personally generated based on the preferences and past likes of the learner.
-
-    \item Article recommendations which are personalized for the interests of the learner and come with difficulty estimation that helps the learner find articles with the appropriate difficulty.
-
-  \end{enumerate}
-
-  The core API implemented with Flask and Python provides correspondingly three types of functionality: contextual translations, article recommendations, and personalized exercise suggestions.
-In total the API provides a bit less than 50 endpoints, out of which probably a dozen are very frequently used.
-  % The development of the core API itself is a research project.
-
-  At the time of writing this article, the ecosystem consists of a reader web application, a web based exercises platform, and a smartwatch application, which are used at the moment of writing this article by more than \activeUserCount active users.
-The users come from a highschool, a language school, and some users are using it on their own, without any educational context.
-The highest load we observed until now on the API consisted of 12K requests in one day.
-
-  We will use this \zee API as a case study for this paper. 
-  All the figures in this paper are captured from the actual deployment of \tool in the context of the \zee API.
-The figures are interactive offering basic data exploration capabilities: filter, zoom, and details on demand\cite{Shne99a}.
-The \tool deployment for the case study can be accessed publicly\footnote{\url{https://zeeguu.unibe.ch/api/dashboard}; username: {\em guest}, password: {\em soap-sac}}.
-
- % \todo{create new account, update URL if necessary, update activeUserCount, add a couple of sentences of the current state of the case study with the college/language center}
-% \ml{we should consider adding also one section in which the architecture/implementation and main features of the dashboard are presented before going on with discussing them in more depth in the following sections --- this should include a rundown on which views are provided from where (overview or per endpoint)}
-
-\newpage
-
-\section{Case Study: The Technology (Flask)}
-\label{sec:tool}
-
- Flask is a microframework for Python. It IS USED BY THOUSANDS OF PROJECTS. ELABORATE MORE ON THIS... 
- \vspace{2cm}
-
- Flask depends on two external libraries: the Jinja2 template engine and the Werkzeug WSGI toolkit.
-
- “Micro” does not mean that your whole web application has to fit into a single Python file (although it certainly can), nor does it mean that Flask is lacking in functionality.
-The “micro” in microframework means Flask aims to keep the core simple but extensible.
-Flask won’t make many decisions for you, such as what database to use.
-Those decisions that it does make, such as what templating engine to use, are easy to change.
-Everything else is up to you, so that Flask can be everything you need and nothing you don’t.
-
- Flask has many configuration values, with sensible defaults, and a few conventions when getting started.
-By convention, templates and static files are stored in subdirectories within the application’s Python source tree, with the names templates and static respectively.
-While this can be changed, you usually don’t have to, especially when getting started.
-
- A Python web application based on WSGI has to have one central callable object that implements the actual application.
-In Flask this is an instance of the Flask class.
-Each Flask application has to create an instance of this class itself and pass it the name of the module, but why can’t Flask do that itself?
-
-\begin{lstlisting}[style=custompython]
-
-from flask import Flask
-app = Flask(__name__)
-
-@app.route('/')
-def index():
-    return 'Hello World!'
-
-\end{lstlisting}
-
-\subsection*{NOTHING PREVENTS GENERALIZATION}
-In this article we take as an example the Flask case study.
-But there is nothing that prevents the generalization of 
-our approach.
-
-MAKE SURE THAT THIS IS CLEAR.
-
-    \begin{figure*}[ht!]
-      \centering
-      \includegraphics[width=0.8\linewidth]{interceptor}
-      \caption{The first thing that needs to be done, is to decorate the application object with an INTERCEPTOR}
-      \label{fig:sep}
-    \end{figure*}
-
-\newpage
-
-\section{Monitoring Service Utilization}
-
-REMEMBER THE GOAL: Our main goal with the \tool issues to allow the integration of the dashboard with an API with minimal effort.
-
-\ins{The first important question that such a 
-service monitoring system should provide is 
-information about what endpoints are being
-used and by whom}.
-
-TODO: Find references, arguments, related work
-that provide more details about why understanding
-who uses endpoints, and which are used are important.
-
-This is clearly the case in static analysis (see paper
-by haenni and lungu...) but should even more be the
-case in APIs.
-
-  \newpage
-TODO: TALK ABOUT THE DATABASE...
-
-  As discussed in the introductory section, in previous work~\cite{vogel2017low} we introduced \tool, a drop-in Python library that allows developers to monitor their Flask-based Python web applications with minimal effort.
-%
-  The \tool is implemented for Python 3.6 and is available on the Python Package Index repository\footnote{\url{https://pypi.python.org/pypi/flask-monitoring-dashboard/1.8}} from where it can be installed by running \install from the command line.
-%  
-  The source code of the \tool is published under a permissive MIT license and is available on GitHub\footnote{\url{https://github.com/flask-dashboard}}.
-
-  \newpage
-  \section{One Line Of Code Configuration With Technology Specific Interceptors}
-
-  The fact that the \tool itself is being developed in Python using Flask makes binding to the services of a to-be monitored application developed with the same technologies easy and intuitive.
-
-  \ins{In general, what matters is that with the right backend, we can do this minimal configuration for any technology... }
-
-  To start using the \tool, one simply needs to one lines of code\footnote{we don't count the import statement that enables that line...} to their Flask web service\footnote{\ins{ In this paper we present the integration with APIs written in Python and Flask hoping that this will not prevent the reader from seeing the more general idea; all the tools we show here for Flask can be applied to other API technologies (e.g.
-Django) by simply providing a few back-end adapters in the right places.
-}}:
-
-  % caption=Configuring the \tool is straightforward,
-  \begin{lstlisting}[style=custompython]
-  import dashboard
-
-  # flask_app is the Flask app object
-  dashboard.bind(flask_app)
-  \end{lstlisting}
-
-  \ins{The dashboard can take advantage of the fact that it is monitoring an API...
-so it can simply just add one extra endpoint there.} After (re)deployment of the application, the interactive visualizations of the \tool become available at the \code{/dashboard} route of the Flask application.
-
-  During binding, the \tool will search for all endpoints defined in the target application.
-
-  In order to monitor an endpoint, the \tool creates a function wrapper for the API function that corresponds to the endpoint.
-This way, the wrapper will be executed whenever that API call is made before the actual function is called.
-The wrapper contains the code that takes care of monitoring an endpoint.
-Data collected by the wrappers are persisted in a local database.
-The SQLAlchemy Object Relational Mapper\footnote{\url{https://www.sqlalchemy.org/}} on top of SQLite\footnote{\url{https://www.sqlite.org/}} is used for this purpose.
-  \ins{DISCUSSION about how do to this in other languages? Java: AspectJ probably, etc.}
-
-  These will be presented to the user in the tool web interface, where the user can select the ones that should be monitored, see \Fref{fig:sep}.
-
-    \begin{figure}[h!]
-      \centering
-      \includegraphics[width=\linewidth]{selecting_endpoints.png}
-      \caption{Once connected to an API the Dashboard presents the endpoints that are available for monitoring}
-      \label{fig:sep}
-    \end{figure}
-
-  This is the alternative to using annotations like other white-box tools (e.g. the other flask performance plugin).
-
-  \ins{PROMISE:} The last access time of the endpoint by users, irrespective of whether it is selected by the user to be monitored or not, is also displayed as the means of identifying potentially obsolete endpoints.
-
-  \ml{we must insert at least one dashboard view here...
-otherwise people will lose their patience...
-also it will help with explaining the outliers...
-also, one should see that a large variety of views are already available at this point...
-I know this risks being  }
-
-In the Zeeguu case study, one of the slowest endpoints, and one with the highest variability as shown in \Fref{fig:ep} is \epFeedItems: it retrieves a list of recommended articles for a given user.
-However, since a user can be subscribed to anything from one to three dozen article sources, and since the computation of the difficulty is personalized and it is slow, the variability in time among users is likely to be very large.
-
-\begin{figure*}[h!]
-  \centering
-  \subfloat[number of requests per endpoint per day]{\includegraphics[width=.9\columnwidth]{number_of_requests_}\label{fig:aeu}}
-  \subfloat[requests per hour heatmap]{\includegraphics[width=\columnwidth]{daily_patterns_}\label{fig:dp}}
-  \caption{Some of the available views\label{fig:views}}
-\end{figure*}
-
-\newpage
-\section{Service Utilization}
-\label{sec:util}
-
-\ins{A discussion must be added here about the types of useful information and problems that can be solved with these views.}
-
-\textit{Service Utilization} presents information about the usage of all endpoints of interest to the developer,
-
-The most fundamental insight that a service maintainer needs regards service utilization. %\vspace{0.5cm}
-
-Simplest utilization perspective:
-
-\begin{figure}[h!]
-  \centering
-  \includegraphics[width=0.5\linewidth]{endpoint-call-count}
-  \caption{Caption here}
-  \label{fig:figure1}
-\end{figure}
-
-% REQUIREMENT: Adding a time interval is important. 
-% Otherwise the performance numbers in chart 1 are not good enough... also an outlier very far in the past, will make all the heatmap data points to look very pale in comparison forever after.
-
-Figure \ref{fig:aeu} shows a first perspective on endpoint utilization that \tool provides: a stacked bar chart of the number of hits to various endpoints grouped by day\footnote{Endpoint colors are the same in different views}.
-Figure~\ref{fig:aeu} in particular shows that at its peak the API has about 12.000 hits per day.
-
-The way users interact with the platform can also be inferred since the endpoints are indicators of different activity types, e.g.:
-
-\begin{enumerate}
-
-  \item {\color{myred} \epTranslations} is an indicator of the amount of foreign language reading the users are doing, and
-
-  \item {\color{mygreen} \epOutcome} is an indicator of the amount of foreign vocabulary practice the users are doing.
-
-\end{enumerate}
-
-%\begin{figure}[!ht]
-% \centering
-% \includegraphics[width=\linewidth]{number_of_requests_}
-% \caption{The number of requests per endpoint per day view shows the overall utilization of the monitored application}
-% \label{fig:aeu}
-%\end{figure}
-
-Besides showing the overall utilization, this endpoint provides the maintainer with information relevant for decisions regarding endpoint deprecation --- one of the most elementary ways of {\em understanding the needs of the downstream}\cite{Haen14a}.
-In our case study, the maintainer realized that one endpoint which they thought was not being used (i.e.
-\code{words\_to\_study}), contrary to their expectations, was actually being used\footnote{A complementary type of usage information can also be discovered in the view presented in Figure \ref{fig:sep} where seeing that an endpoint is never accessed can increase the confidence of the maintainer that a given endpoint is not used, although it can never be used a proof.}.
-
-\niceseparator
-
-%   \todo{Add the time series graph and discuss it before the heatmap? We can then sell the heatmap better} 
-%   \ml{Not sure about which graph you refer to here V}
-
-A second type of utilization question that the \tool can answer automatically regards {\em cyclic patterns of usage per hour of day} by means of a heatmap, as in \Fref{fig:dp}.
-
-% \mltp{ can we add vertical lines that highlight the beginning of a new week (e.g. before Sunday): }
-% Patrick: adding separators in the graph is unfortunately not supported by the library.
-
-%\begin{figure}[!ht]
-% \centering
-% \includegraphics[width=0.9\linewidth]{daily_patterns_}
-% \caption{Usage patterns become easy to spot in the requests per hour heatmap}
-% \label{fig:dp}
-%\end{figure}
-
-Figure \ref{fig:dp} shows the API not being used during the early morning hours, with most of the activity focused around working hours and some light activity during the evening.
-This is consistent with the fact that the current users are all in the central European timezone.
-Also, the figure shows that the spike in utilization that was visible also in the previous graph happended in on afternoon/evening.
-
-\newpage
-\section{Endpoint Performance}
-\label{sec:perf}
-
-  The \tool also collects information regarding endpoint performance. The view in \Fref{fig:ep} summarizes the response times for various endpoints by using a box-and-whiskers plot.
-
- \begin{figure}[!ht]
-   \centering
-   \includegraphics[width=0.95\linewidth]{endpoint_performance_}
-   \caption{The response time (in ms) per monitored endpoint view allows for identifying performance variability and balancing issues}
-   \label{fig:ep}
- \end{figure}
-
-  From this view it became clear to the maintainer that four of the endpoints had very large variation in performance.
-The most critical for the application and consequently the one optimized first was the \epTranslations endpoint which was part of an interactive loop in the reader applications that relied on the Zeeguu API.
-Moreover, cf.
-\Fref{fig:aeu} this endpoint is one of the most used in the system.
-
-\begin{figure}[h!]
-  \centering
-  \subfloat[overview of the tracked endpoints]{\includegraphics[width=0.8\columnwidth]{overview}\label{fig:overview}}
-  \caption{Some of the available views\label{fig:views}}
-\end{figure}
-
-  \newpage
-  \section{Evolution monitoring }
-
-  \begin{figure}[h!]
-  \centering
-  \includegraphics[width=0.8\linewidth]{translation_endpoint_evolution_}
-  \caption{The Performance Evolution of the \epTranslations endpoint}
-  \label{fig:tee}
-\end{figure}
-
-  When thinking about performance, 
-  one can not avoid thinking about the performance evolution.
-
-  Version control in \tool can be supported in two ways,
-  the developer explicitly states the current version, 
-  or the current version is automatically detected based
-  on some version control system.
-
-  However, with the current configuration of the tool, it would be impossible for the maintainer to see the improvements resulting from the optimization.
-
-  \ins{The latest STack Overflow Developer Survey: 88.4\% of **professional** developers use git.}  
-  If assume that the code that they run is deployed using .git, then with an extra line of configuration they can allow \tool to find the git\footnote{\url{https://git-scm.com/}} folder of the deployed service and automatically detect the version of the project that is running:
-
-    \begin{lstlisting}[style=custompython]
-
-      dashboard.config.git = 'path/to/.git'
-
-    \end{lstlisting}
-
-  If the \tool can automatically detect the current version of the project by reading the .git configuration as soon as the API is started and can then group measurements by version\footnote{Alternatively, the maintainer can add version identifiers manually for the web application through a configuration file if the system does not use git.}.
-\Fref{fig:tee} is a zoomed-in version of such a view for \epTranslations with versions increasing from top to bottom
-
-  This view confirms that the performance of the translation endpoint improved in the recent versions: the median of the last three versions is constantly moving towards the left, and progresses from 1.4 seconds (in the top-most box plot in \Fref{fig:tee}) to 0.8 in the latest version (bottom-most box plot).
-
-  CHALLENGE: Supporting other types of version control... not real challenge, as we said 88percent of devs use git. 
-  CHALLENGE 2: detecting minor versions which don't need to be tracked, since they didn't touch the performance of the system. E.g. a modification of the README, etc.
-
-\todo{TODO: endpoint evolution }
-
-  \begin{figure*}[h!]
-    \centering
-    \includegraphics[width=\linewidth]{outliers}
-    \caption{Automatically collected outlier information}
-    \label{fig:figure1}
-  \end{figure*}
-  \newpage
-  \section{Automated outlier detection and monitoring}
-
-  When an API is called from within a highly interactive application (as it is the case with the case study in this paper) 
-  of particular interest to the API developers are performance {\em outliers}.
-
-  Indeed, a translation request that takes three times more than expected can seriously decrease the perceived quality of the application.
-Thus, identifying, collecting all appropriate data, and diagnosing the root causes of such outliers is especially critical in improving the quality of an application.
-
-  % In the context of the RESTful services support by Flask, this means request serving times that deviate from the average to an unexpected degree.
-
-  For this purpose the \tool tracks for every monitored endpoint a {\em running average} response time value\footnote{\ins{For performance reasons, we assume that the response times for the endpoints are normally distributed.
-Otherwise, more general density distribution information must be collected in real time.}}.
-When it detects that a given request is an outlier with respect to this past average running value, it triggers the {\em outlier data collection routine} which stores \ins{extra information} about the current execution environment.
-A configurable threshold with a default value of $2.5$ times the running average response time is used for this purpose.
-
-  For every detected outlier request, the \tool collects information about the current Python stack trace, CPU load, memory consumption, request parameters, etc.
-in order to allow the maintainer to investigate the causes of these exceptionally slow response times.
-In this way it is possible to get \ins{detailed insight into the operation of the application in the extreme cases without unnecessarily burdening it with logging this information for every request}.
-
-The bottom panel shows the stack trace. 
-In this particular case, it is revealing for the developer to learn that at the time of the stack trace snapshot, the code was in the microsoft\_translator.py: indeed, the system uses as back-end multiple translators, and it has been observed that many of the outliers happen to be waiting in the microsoft translator.
-
-This information has to be corroborated with the observations that neither the memory nor the processor are overloaded at the moment.
-Thus this functionality in microsoft\_translator is really slow in itself, and this is not a result of the machine being overloaded for example.
-
-\subsection {A Special Type of Outlier: Exceptions}
-
-   TODO: probably something for the journal extension ...
-  statistics about which endpoints fail most often might be useful too.
-  same information as the outlier maybe?
-
-\begin{figure*}[!ht]
-  \centering
-  \includegraphics[width=0.4\linewidth]{time_per_user}
-  \caption{The \epFeedItems shows a very high variability across users}
-  \label{fig:tpu}
-\end{figure*}
+%!TEX root=restructured.tex
 
@@
 \end{figure}
-
-  \newpage
-  \section{Preemptive monitoring}
-
-  \ins{INTRO - LINKING WITH PREVIOUS PARAGRAPH: the previous evolutionary view crucial for 
-  understanding the impact of software evolution on performance.
-  However, just by observing this one can't be sure whether 
-  performance changes because of the code or because of the payload.}
-
-  \todo{the following might need a bit of rewrite}
-
-  The concept of {\em preemptive monitoring} of the application performance by means of instrumenting integration unit testing as the synthetic load is similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}, i.e.~testing service-based applications by using dedicated test input in parallel to its normal use and operation.
-The difference is that we take advantage of the capability of the CI framework (i.e.
-Travis in this case) to create an emulated ``live'' environment for integration testing purposes, and use unit testing as the dedicated test input in order to measure performance.
-
-  For example, \Fref{fig:builds} is a screenshot from the dashboard showing the measured response times for 5 consequent builds, with 180 iterations of the unit tests executed in total across all endpoints.
-The outliers on the right of the figure are due to initial requests \ins{which must wait for a boot-up phase of the API}.
-
-    \begin{figure}[h!]
-        \centering
-        \includegraphics[width=\columnwidth]{travis_builds}
-        \caption{Response times in 5 consequent Travis builds in the \zee case study}
-        \label{fig:builds}
-      \end{figure}
-
-  While this integration testing environment is different from the actual production deployment one, and the load used is purely synthetic, as will show in Section~\ref{sec:evaluation}, it can actually be used successfully as an early performance indicator for the application developer.
-\ins{Although predicting the performance of the application would require advanced statistical work, it can be succesfully used as an advanced warning system for the overall performance trend for given endpoints}.
-
-  \paragraph{Integration with Travis}
-  \ml{added new section... also, this has to be moved after introducing the pre-emptive monitoring since it's about running tests. switched the order}
-
-  \ins{The pre-emptive monitoring system} can be configured to work together with Continuous Integration (CI) frameworks like Travis\footnote{\url{https://travis-ci.org/}} that deal with automated integration testing.
-  The developer needs to define the unit test folder of the project, the URL that Travis should use for reporting its results, and the number of iterations for each unit test should be executed in order to generate a load for the application.
-
-  Each time a new build is created through Travis, the \tool automatically detects all available unit tests defined by the application developer and iterates through each one of them the number of times predefined by the developer while monitoring the response times for each request.
-The resulting measurements are persisted in a separate part of the tool database so as not to contaminate the ``live'' monitoring data.
-Beyond seeing the results of each Travis build in the \tool, this feature also allows for preemptive monitoring of the application, as we discuss in the following.
-
-  \ins{Besides functioning as an early warning system for performance, tracking the evolving performance of API tests serves a second purpose.
-To function as an anchor when the developer analyzes performance degradations.
-For example, when a developer sees that a given endpoint has become less performant after in a newly deployed version, they are able to investigate whether the performance degradation is visible also with the synthetic load.
-This would correspond to a performance degradation which is due to ``algorithmic performance degradation''.
-If on the other hand, the performance of the tests does not change between versions, the developer might conclude that the performance degradation could be due to the workload on the machines, or maybe to the workload mix of the users}
-
-  %By these means, an application developer can use unit tests as synthetic loads for driving the performance monitoring of the application before it reaches the production phase.
-As we will discuss further in Section~\ref{sec:testing}, this feature is integral in enabling preemptive performance monitoring of the application.
-
-\todo{here goes the discussion about our forecasting capabilities}
-
-\begin{figure}[h!]
-  \centering
-  \subfloat[The reported response time (in ms) per deployed version in the observation period]{\includegraphics[width=.5\columnwidth]{response_per_version_trunced_trend}}
-  \quad
-  \subfloat[The measured response time (in ms) using integration with Travis]{\includegraphics[width=.46\columnwidth]{travis_builds_no_outliers_trend}}
-  \caption{Comparison of the response times per endpoint: actual production system versus preemptive monitoring data}
-  \label{fig:preemptive}
-\end{figure}
-
-\begin{table}[h]
-  \begin{tabular}{lll}
-    \toprule
-    Iteration & \bfseries Live (median) & \bfseries Travis (median)\\
-    \midrule
-    1 & 1349.41 & 764.87\\ 
-    2 & 1466.13 & 992.87\\
-    3 & 1256.65 & 760.87\\
-    4 & 1266.42 & 813.89\\
-    5 & 1080.68 & 303.4\\
-    \bottomrule
-  \end{tabular}
-\end{table}
-
-Pearson correlation $r(3)=.93, p=.02$
-
-\section{Performance Perspectives}
-\label{sec:views}
-
-\ins{PERFORMANCE DISCUSSION: this introspection, which looks up endpoints, happens only once at the startup of the service, so it is going to affect the API startup performance, but not the actual endpoint performance.
-In our case, the overhead here was quite small: TO MEASURE}.
-
-  %In the remainder of the paper we present several of these perspectives.\footnote{We recommend obtaining a color version of this paper for better readability}
-
-  % The second endpoint conists of two parts, one of them being a table that shows for every monitored endpoint the number of hits it has gotten, the time it was last accessed and its average execution time.
-The second part is a view with four graphs which show:
-  % \begin{itemize} 
-  %   \item A heatmap of the total number of requests to the monitored endpoints
-  %   \item A stacked bar chart that shows the total number of requests to the monitored endpoints per endpoint per day
-  %   \item A boxplot graph showing the average execution time per version of the web service
-  %   \item A boxplot graph showing the average execution time for every monitored endpoint
-  % \end{itemize}
-
-\todo{evaluate which figures to keep and ideally take new screenshots for the ones we keep; consider adding two more views (like Fig. 3.2 of Patrick's thesis)}
-
-% In the Zeeguu case study, one of the slowest endpoints, and one with the highest variability as shown in \Fref{fig:ep} is \epFeedItems: it retrieves a list of recommended articles for a given user.
-However, since a user can be subscribed to anything from one to three dozen article sources, and since the computation of the difficulty is personalized and it is slow, the variability in time among users is likely to be very large.
-
-% \niceseparator
-
-%\section{Preemptive Monitoring}
-%\label{sec:testing}
-%
-%%\todo{discuss here idea of using unit testing as an early performance indicator for evolving services (`preemptive monitoring'?); add a side-by-side figure comparing unit testing vs live system performance per version, maybe calculate error in trend curves?
-find a better term than prediction for the section title (plus introduction)}
-%
-%\todo{Similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}, i.e.~testing service-based applications by using dedicated test input in parallel to its normal use and operation.
-The difference is that we take advantage of the capability of Travis to create an emulated live environment for integration testing purposes, and use unit testing as the dedicated test input in order to measure performance.
-What we are going provide evidence for in the following is that we can indeed use this preemptive monitoring concept as an early performance indicator for evolving services.}
-%
-%The developer needs to define the unit test folder of the project, the URL that Travis should use for reporting its results, and the number of iterations for each unit test should be executed in order to generate
-
-\section{Evaluation}
-\label{sec:evaluation}
-
-%\todo{evaluation of the \tool across the following dimensions: ease of use, effectiveness, overhead to the system; for the last part we need to do the overhead measurement experiment, for the other two we base it on the case study and bring over whatever is necessary from the discussion section}
-
-\section{Discussion}
-
-\todo{use this subsection to discuss limitations (from the previous evaluation section) and add the need to be able to handle the horizontal scaling of the application; remove sectioning and summarize briefly limitations that have already been discussed in the VISSOFT paper/move material to other sections; consider renaming as limitations or something similar}
-
-  \subsection{Distributed System Monitoring}
-  \todo{To think about}
-  Supporting situations where the API is deployed across multiple containers for example.
-
-  \subsubsection{Automatically Monitoring System Evolution}
-
-  The main goal of the \tool design was to allow analytics to be collected and insight to be gained by making the smallest possible changes to a running API.
-%To allow the collection of evolutionary information
-%
-  This technique assumes that the web application code which is the target of the monitoring is deployed using \git in the following way:
-
-  \begin{enumerate}
-    \item The deployment engineer pulls the latest version of the code from the integration server; this will result in a new commit being pointed at by the HEAD pointer. %than previously
-    \item The deployment engineer restarts the new version of the service.
-At this point, the \tool detects that a new HEAD is present in the local code base and consequently starts associating all the new data points with this new commit\footnote{The \tool detects the current version of the analyzed system the first time it is attached to the application object, and thus, assumes that the Flask application is restarted when a new version is deployed.
-This is in tune with the current version of Flask, but if the web server will support dynamic updates in the future, this might have to be taken into account}.
-  \end{enumerate}
-
-  The advantage of this approach is the need for minimal configuration effort, as discussed in the presentation of the tool.
-The disadvantage is that it will consider on equal ground the smallest of commits, even one that modifies a comment, and the shortest lived of commits, e.g.~a commit which was active only for a half an hour before a new version with a bug fix was deployed, with major and minor releases of the software.
-%as a distinct way of grouping the data points.
-  A mechanism to control which versions are important for monitoring purposes is therefore required to be added.
-%
-  A further possible extension point here is supporting other version control systems (e.g. Mercurial). However, this is a straightforward extension.
-
-  \subsubsection{NEW: API Renames}
-  \ml{just an idea}
-
-    The history would normally be lost in the case of an API rename. (also known as the {\em provenance issue} in source code evolution analysis)
-    In theory we could infer that an API was renamed if it is not to be found anymore, and another one with very similar timing characteristics would be found.
-Such a detection can never be completely sure, but it would help with meeting the needs of the API maintainer...
-
-  \subsubsection{User-Awareness }
-
-    For the situations in which the user information is not available, the \tool tracks by default information about different IPs and in some cases this might be a sufficiently good approximation of the user diversity and identity.
-    %
-
-    The visualizations for the user experiene perspectives as presented in Section \ref{sec:user} have been tested with several hundred users (of which about \activeUserCount were active during the course of the study), but the scalability of the visualizations must be further investigated for web services with tens of thousands of users.
-
-  \subsubsection{Other Possible Groupings}
-
-    There are other groupings of service utilization and performance that could be important to the maintainer, that we did not explore in this paper.
-For example, if the service is using OAuth, then together with every request, in the header of the request there is information about the application which is sending a request.
-Grouping the information by application that sends the request could be important in such a context.
-
-    In general, providing a mechanism that would allow very easy specification of groupings (either as code annotations, as normal code, or as configuration options) is an open problem that \tool and any other similar library will have to face.
-
-\section{Related Work}
-\label{sec:related}
-
-\todo{Refer the reader to~\cite{ghezzi2007run} and ~\cite{metzger2010analytical} for a more extensive discussion}
-
-There is a long tradition of using visualization for gaining insight into software performance.
-Tools like Jinsight \cite{Pauw02a} and Web Services Navigator \cite{Pauw05} pioneered such an approach for Java and for Web Services that communicate with SOAP messages.
-Both have an ``omniscient'' view of the services / objects and their interactions.
-As opposed to them, in our work we present an analytics platform which focuses on monitoring a single Python web service from its own point of view.
-
-From the perspective of service monitoring, our work falls within the server-side run-time monitoring of services ~\cite{ghezzi2007run}.
-While we don't implement the more advanced features of related monitoring solutions like QoS policies driving the monitoring, it presents nevertheless an easy to use approach support improving the performance of web applications.
-
-% \va{Mircea: Consider removing the rest for space...}
-% An existing monitoring tool is Pingdom \footnote{https://www.pingdom.com/company/why-pingdom}, which monitors the uptime of an existing web-service.
-This tool works by pinging the websites (up to 60 times) every minute automatically.
-Thus this creates a lot of overhead and is bound to be noisy since it will also be influenced by the speed of the network connection\footnote{Another problem is that such a tool would }
-
-% \todo{Runscope? Others?}
-
-\section{Conclusion and Future Work}
-\label{sec:conclusions}
-
-\todo{update as appropriate}
-
-In this paper we have shown that it is possible to create a monitoring solution which provides basic insight into web service utilization and performance  with very little effort from the developer.
-The user group that we are aiming for with this work is application developers using Flask and Python to build web applications with limited or no budget for implementing their own monitoring solutions.
-The emphasis is in allowing such users to gain insight into how the performance of the service evolves together with the application itself.
-We believe that the same architecture, and lessons can be applied to other frameworks and other languages.
-
-In the future, we plan to perform case studies with other sytstems, with the goal of discover other needs and to wean out the less useful visualizations in the \tool.
-We plan to also extend the tool towards supporting multiple deployments of the same applications across multiple nodes (e.g.
-for the situations where the application is deployed together with a load balancer).
-Finally, we plan to integrate \tool with unit testing as a complementary source of information about performance evolution.
-
-\newpage
-\appendix
-
-  \subsection{Changing the Default /dashboard Endpoint }
-
-  THIS STUFF SHOULD BE MOVED TO THE APPENDIX.
-And in a footnote we just mention that ...
-\ins{REQUIREMENT}: The system must provide an easy way to add more specific configurations if needed.
-A way of overwriting the common sense defaults.
-
-  Further configuration is not needed possible by adding additional statements before the binding definition, for example,
-
-  \begin{lstlisting}[style=custompython]
-  ...
-  dashboard.config.link = 'mydashboard'
-  dashboard.bind(flask_app)
-  \end{lstlisting}
-
-  allows for a custom route (\code{/mydashboard}) to the dashboard to be defined by the programmer. An external configuration file can be used instead, or in addition to this:
-
-  \begin{lstlisting}[style=custompython]
-  ...
-  dashboard.config.from_file('config.cfg')
-  ...
-  \end{lstlisting}
-
-  \subsection{For those who don't use GIT}
-
-    \textit{Manual} version control requires the developer to tag each new version of the application with an appropriate version identifier~\cite{papazoglou2011managing} using the \code{APP\_VERSION} configuration parameter, for example by adding to the configuration file:
-
-      \begin{lstlisting}[style=custompython]
-      [dashboard]
-      APP_VERSION=<versionID>
-      \end{lstlisting}
-
-% references section
-
-\bibliographystyle{abbrv}
-\bibliography{paper}
-
-% that's all folks
-\end{document}

### 2018-04-03 introductory.tex
split the huge file in smaller ones.

--- 2018-03-21 restructured.tex
@@
-\documentclass[conference]{IEEEtran}
-	\pdfpagewidth=8.5truein
-	\pdfpageheight=11truein
-
-\usepackage{graphicx}
-\usepackage{subfig}
-\usepackage[bookmarks=false]{hyperref}
-\usepackage{xspace}
-\usepackage{listings}
-\usepackage[usenames, dvipsnames]{color}
-
-\usepackage{booktabs}
-
-\graphicspath{{./img/}}
-
-%\hyphenation{op-tical net-works semi-conduc-tor}
-
-\input{useful_defines}
-
-\begin{document}
-
-%
-\title{Agile Monitoring of Evolving Web Services}
-
-\author{
-\IEEEauthorblockN{Patrick Vogel, Thijs Klooster, Vasilios Andrikopoulos, Mircea Lungu}\\
-Johann Bernoulli Institute for Mathematics and Computer Science\\
-University of Groningen, Netherland\\
-Email: \{t.klooster.1,p.p.vogel\}@student.rug.nl, \{v.andrikopoulos,m.f.lungu\}@rug.nl 
-}
-
-% make the title area
-\maketitle
+%!TEX root=restructured.tex
 
@@
 MAKE SURE THAT THIS IS CLEAR.
-
-    \begin{figure*}[ht!]
-      \centering
-      \includegraphics[width=0.8\linewidth]{interceptor}
-      \caption{The first thing that needs to be done, is to decorate the application object with an INTERCEPTOR}
-      \label{fig:sep}
-    \end{figure*}
-
-\newpage
-
-\section{Monitoring Service Utilization}
-
-REMEMBER THE GOAL: Our main goal with the \tool issues to allow the integration of the dashboard with an API with minimal effort.
-
-\ins{The first important question that such a 
-service monitoring system should provide is 
-information about what endpoints are being
-used and by whom}.
-
-TODO: Find references, arguments, related work
-that provide more details about why understanding
-who uses endpoints, and which are used are important.
-
-This is clearly the case in static analysis (see paper
-by haenni and lungu...) but should even more be the
-case in APIs.
-
-  \newpage
-TODO: TALK ABOUT THE DATABASE...
-
-  As discussed in the introductory section, in previous work~\cite{vogel2017low} we introduced \tool, a drop-in Python library that allows developers to monitor their Flask-based Python web applications with minimal effort.
-%
-  The \tool is implemented for Python 3.6 and is available on the Python Package Index repository\footnote{\url{https://pypi.python.org/pypi/flask-monitoring-dashboard/1.8}} from where it can be installed by running \install from the command line.
-%  
-  The source code of the \tool is published under a permissive MIT license and is available on GitHub\footnote{\url{https://github.com/flask-dashboard}}.
-
-  \newpage
-  \section{One Line Of Code Configuration With Technology Specific Interceptors}
-
-  The fact that the \tool itself is being developed in Python using Flask makes binding to the services of a to-be monitored application developed with the same technologies easy and intuitive.
-
-  \ins{In general, what matters is that with the right backend, we can do this minimal configuration for any technology... }
-
-  To start using the \tool, one simply needs to one lines of code\footnote{we don't count the import statement that enables that line...} to their Flask web service\footnote{\ins{ In this paper we present the integration with APIs written in Python and Flask hoping that this will not prevent the reader from seeing the more general idea; all the tools we show here for Flask can be applied to other API technologies (e.g.
-Django) by simply providing a few back-end adapters in the right places.
-}}:
-
-  % caption=Configuring the \tool is straightforward,
-  \begin{lstlisting}[style=custompython]
-  import dashboard
-
-  # flask_app is the Flask app object
-  dashboard.bind(flask_app)
-  \end{lstlisting}
-
-  \ins{The dashboard can take advantage of the fact that it is monitoring an API...
-so it can simply just add one extra endpoint there.} After (re)deployment of the application, the interactive visualizations of the \tool become available at the \code{/dashboard} route of the Flask application.
-
-  During binding, the \tool will search for all endpoints defined in the target application.
-
-  In order to monitor an endpoint, the \tool creates a function wrapper for the API function that corresponds to the endpoint.
-This way, the wrapper will be executed whenever that API call is made before the actual function is called.
-The wrapper contains the code that takes care of monitoring an endpoint.
-Data collected by the wrappers are persisted in a local database.
-The SQLAlchemy Object Relational Mapper\footnote{\url{https://www.sqlalchemy.org/}} on top of SQLite\footnote{\url{https://www.sqlite.org/}} is used for this purpose.
-  \ins{DISCUSSION about how do to this in other languages? Java: AspectJ probably, etc.}
-
-  These will be presented to the user in the tool web interface, where the user can select the ones that should be monitored, see \Fref{fig:sep}.
-
-    \begin{figure}[h!]
-      \centering
-      \includegraphics[width=\linewidth]{selecting_endpoints.png}
-      \caption{Once connected to an API the Dashboard presents the endpoints that are available for monitoring}
-      \label{fig:sep}
-    \end{figure}
-
-  This is the alternative to using annotations like other white-box tools (e.g. the other flask performance plugin).
-
-  \ins{PROMISE:} The last access time of the endpoint by users, irrespective of whether it is selected by the user to be monitored or not, is also displayed as the means of identifying potentially obsolete endpoints.
-
-  \ml{we must insert at least one dashboard view here...
-otherwise people will lose their patience...
-also it will help with explaining the outliers...
-also, one should see that a large variety of views are already available at this point...
-I know this risks being  }
-
-In the Zeeguu case study, one of the slowest endpoints, and one with the highest variability as shown in \Fref{fig:ep} is \epFeedItems: it retrieves a list of recommended articles for a given user.
-However, since a user can be subscribed to anything from one to three dozen article sources, and since the computation of the difficulty is personalized and it is slow, the variability in time among users is likely to be very large.
-
-\begin{figure*}[h!]
-  \centering
-  \subfloat[number of requests per endpoint per day]{\includegraphics[width=.9\columnwidth]{number_of_requests_}\label{fig:aeu}}
-  \subfloat[requests per hour heatmap]{\includegraphics[width=\columnwidth]{daily_patterns_}\label{fig:dp}}
-  \caption{Some of the available views\label{fig:views}}
-\end{figure*}
-
-\newpage
-\section{Service Utilization}
-\label{sec:util}
-
-\ins{A discussion must be added here about the types of useful information and problems that can be solved with these views.}
-
-\textit{Service Utilization} presents information about the usage of all endpoints of interest to the developer,
-
-The most fundamental insight that a service maintainer needs regards service utilization. %\vspace{0.5cm}
-
-Simplest utilization perspective:
-
-\begin{figure}[h!]
-  \centering
-  \includegraphics[width=0.5\linewidth]{endpoint-call-count}
-  \caption{Caption here}
-  \label{fig:figure1}
-\end{figure}
-
-% REQUIREMENT: Adding a time interval is important. 
-% Otherwise the performance numbers in chart 1 are not good enough... also an outlier very far in the past, will make all the heatmap data points to look very pale in comparison forever after.
-
-Figure \ref{fig:aeu} shows a first perspective on endpoint utilization that \tool provides: a stacked bar chart of the number of hits to various endpoints grouped by day\footnote{Endpoint colors are the same in different views}.
-Figure~\ref{fig:aeu} in particular shows that at its peak the API has about 12.000 hits per day.
-
-The way users interact with the platform can also be inferred since the endpoints are indicators of different activity types, e.g.:
-
-\begin{enumerate}
-
-  \item {\color{myred} \epTranslations} is an indicator of the amount of foreign language reading the users are doing, and
-
-  \item {\color{mygreen} \epOutcome} is an indicator of the amount of foreign vocabulary practice the users are doing.
-
-\end{enumerate}
-
-%\begin{figure}[!ht]
-% \centering
-% \includegraphics[width=\linewidth]{number_of_requests_}
-% \caption{The number of requests per endpoint per day view shows the overall utilization of the monitored application}
-% \label{fig:aeu}
-%\end{figure}
-
-Besides showing the overall utilization, this endpoint provides the maintainer with information relevant for decisions regarding endpoint deprecation --- one of the most elementary ways of {\em understanding the needs of the downstream}\cite{Haen14a}.
-In our case study, the maintainer realized that one endpoint which they thought was not being used (i.e.
-\code{words\_to\_study}), contrary to their expectations, was actually being used\footnote{A complementary type of usage information can also be discovered in the view presented in Figure \ref{fig:sep} where seeing that an endpoint is never accessed can increase the confidence of the maintainer that a given endpoint is not used, although it can never be used a proof.}.
-
-\niceseparator
-
-%   \todo{Add the time series graph and discuss it before the heatmap? We can then sell the heatmap better} 
-%   \ml{Not sure about which graph you refer to here V}
-
-A second type of utilization question that the \tool can answer automatically regards {\em cyclic patterns of usage per hour of day} by means of a heatmap, as in \Fref{fig:dp}.
-
-% \mltp{ can we add vertical lines that highlight the beginning of a new week (e.g. before Sunday): }
-% Patrick: adding separators in the graph is unfortunately not supported by the library.
-
-%\begin{figure}[!ht]
-% \centering
-% \includegraphics[width=0.9\linewidth]{daily_patterns_}
-% \caption{Usage patterns become easy to spot in the requests per hour heatmap}
-% \label{fig:dp}
-%\end{figure}
-
-Figure \ref{fig:dp} shows the API not being used during the early morning hours, with most of the activity focused around working hours and some light activity during the evening.
-This is consistent with the fact that the current users are all in the central European timezone.
-Also, the figure shows that the spike in utilization that was visible also in the previous graph happended in on afternoon/evening.
-
-\newpage
-\section{Endpoint Performance}
-\label{sec:perf}
-
-  The \tool also collects information regarding endpoint performance. The view in \Fref{fig:ep} summarizes the response times for various endpoints by using a box-and-whiskers plot.
-
- \begin{figure}[!ht]
-   \centering
-   \includegraphics[width=0.95\linewidth]{endpoint_performance_}
-   \caption{The response time (in ms) per monitored endpoint view allows for identifying performance variability and balancing issues}
-   \label{fig:ep}
- \end{figure}
-
-  From this view it became clear to the maintainer that four of the endpoints had very large variation in performance.
-The most critical for the application and consequently the one optimized first was the \epTranslations endpoint which was part of an interactive loop in the reader applications that relied on the Zeeguu API.
-Moreover, cf.
-\Fref{fig:aeu} this endpoint is one of the most used in the system.
-
-\begin{figure}[h!]
-  \centering
-  \subfloat[overview of the tracked endpoints]{\includegraphics[width=0.8\columnwidth]{overview}\label{fig:overview}}
-  \caption{Some of the available views\label{fig:views}}
-\end{figure}
-
-  \newpage
-  \section{Evolution monitoring }
-
-  \begin{figure}[h!]
-  \centering
-  \includegraphics[width=0.8\linewidth]{translation_endpoint_evolution_}
-  \caption{The Performance Evolution of the \epTranslations endpoint}
-  \label{fig:tee}
-\end{figure}
-
-  When thinking about performance, 
-  one can not avoid thinking about the performance evolution.
-
-  Version control in \tool can be supported in two ways,
-  the developer explicitly states the current version, 
-  or the current version is automatically detected based
-  on some version control system.
-
-  However, with the current configuration of the tool, it would be impossible for the maintainer to see the improvements resulting from the optimization.
-
-  \ins{The latest STack Overflow Developer Survey: 88.4\% of **professional** developers use git.}  
-  If assume that the code that they run is deployed using .git, then with an extra line of configuration they can allow \tool to find the git\footnote{\url{https://git-scm.com/}} folder of the deployed service and automatically detect the version of the project that is running:
-
-    \begin{lstlisting}[style=custompython]
-
-      dashboard.config.git = 'path/to/.git'
-
-    \end{lstlisting}
-
-  If the \tool can automatically detect the current version of the project by reading the .git configuration as soon as the API is started and can then group measurements by version\footnote{Alternatively, the maintainer can add version identifiers manually for the web application through a configuration file if the system does not use git.}.
-\Fref{fig:tee} is a zoomed-in version of such a view for \epTranslations with versions increasing from top to bottom
-
-  This view confirms that the performance of the translation endpoint improved in the recent versions: the median of the last three versions is constantly moving towards the left, and progresses from 1.4 seconds (in the top-most box plot in \Fref{fig:tee}) to 0.8 in the latest version (bottom-most box plot).
-
-  CHALLENGE: Supporting other types of version control... not real challenge, as we said 88percent of devs use git. 
-  CHALLENGE 2: detecting minor versions which don't need to be tracked, since they didn't touch the performance of the system. E.g. a modification of the README, etc.
-
-\todo{TODO: endpoint evolution }
-
-  \begin{figure*}[h!]
-    \centering
-    \includegraphics[width=\linewidth]{outliers}
-    \caption{Automatically collected outlier information}
-    \label{fig:figure1}
-  \end{figure*}
-  \newpage
-  \section{Automated outlier detection and monitoring}
-
-  When an API is called from within a highly interactive application (as it is the case with the case study in this paper) 
-  of particular interest to the API developers are performance {\em outliers}.
-
-  Indeed, a translation request that takes three times more than expected can seriously decrease the perceived quality of the application.
-Thus, identifying, collecting all appropriate data, and diagnosing the root causes of such outliers is especially critical in improving the quality of an application.
-
-  % In the context of the RESTful services support by Flask, this means request serving times that deviate from the average to an unexpected degree.
-
-  For this purpose the \tool tracks for every monitored endpoint a {\em running average} response time value\footnote{\ins{For performance reasons, we assume that the response times for the endpoints are normally distributed.
-Otherwise, more general density distribution information must be collected in real time.}}.
-When it detects that a given request is an outlier with respect to this past average running value, it triggers the {\em outlier data collection routine} which stores \ins{extra information} about the current execution environment.
-A configurable threshold with a default value of $2.5$ times the running average response time is used for this purpose.
-
-  For every detected outlier request, the \tool collects information about the current Python stack trace, CPU load, memory consumption, request parameters, etc.
-in order to allow the maintainer to investigate the causes of these exceptionally slow response times.
-In this way it is possible to get \ins{detailed insight into the operation of the application in the extreme cases without unnecessarily burdening it with logging this information for every request}.
-
-The bottom panel shows the stack trace. 
-In this particular case, it is revealing for the developer to learn that at the time of the stack trace snapshot, the code was in the microsoft\_translator.py: indeed, the system uses as back-end multiple translators, and it has been observed that many of the outliers happen to be waiting in the microsoft translator.
-
-This information has to be corroborated with the observations that neither the memory nor the processor are overloaded at the moment.
-Thus this functionality in microsoft\_translator is really slow in itself, and this is not a result of the machine being overloaded for example.
-
-\subsection {A Special Type of Outlier: Exceptions}
-
-   TODO: probably something for the journal extension ...
-  statistics about which endpoints fail most often might be useful too.
-  same information as the outlier maybe?
-
-\begin{figure*}[!ht]
-  \centering
-  \includegraphics[width=0.4\linewidth]{time_per_user}
-  \caption{The \epFeedItems shows a very high variability across users}
-  \label{fig:tpu}
-\end{figure*}
-
-\newpage
-\section{Grouping Requests}
-\label{sec:user}
-
-For service endpoints which run computations in real time, the maintainer of a system might want to understand the endpoint performance on a per-user basis, especially for situations where the system response time is a function of some individual user load\footnote{E.g.
-in GMail some users have two emails while other have twenty thousand and this induces different response times for different users}.
-
-\todo{remove the configuration part and rephrase the following, the discussion has been partially subsumed by Section~\ref{sec:tool}.}
-
-To enable this, the \tool must be configured to associate an API call with a given user.
-The simplest way is to take advantage of the architecture of Flask applications in which a global \code{flask.request} object can be used to retrieve the session which can in turn lead to user identification:
-
-%\begin{lstlisting}[float,caption=Simply define a custom app-specific function for user retrieval and pass it to the \tool to group information by user,style=custompython]
-\begin{lstlisting}[style=custompython]  
-# app specific way of extracting the user
-# from a flask request object    
-def get_user_id():
-sid = int(flask.request.args['session'])
-session = User.find_for_session(sid)
-return user_id
-
-# attaching the get_user_id function
-dashboard.config.get_group_by = get_user_id
-
-\end{lstlisting}
-
-% \niceseparator
-
-\todo{consider losing one or both of the figures --- or at least updating them}
-
-In Zeeguu, \epFeedItems retrieves a list of recommended articles for a given user.
-Cf.
-\Fref{fig:ep} it is the endpoint with the slowest response time and highest variability.
-The reason for this is that a user can be subscribed to anything from one to three dozen article sources and for each of the sources the system must compute the personalized difficulty of each article at every request.
-
-A \perspective{Per-User Performance} perspective should show the different response times for different users.
-Figure \ref{fig:tpu} presents a subset of the corresponding view in the \tool.
-The figure shows that the response times for this endpoint can vary considerably for different users with some extreme cases where a user has to wait a full minute until their recommended articles are shown\footnote{After seeing this perspective, the maintainer refactored the architecture of the system to move part the difficulty computation out of the interactive loop}.
-
-The limitation of the previous view is that it does not present the information also on a per version basis.
-To address this, a different visual perspective entitled \perspective{Multi-Version per-User Performance} can be defined.
-Figure \ref{fig:tuv} presents such a perspective by mapping the average execution time for a given user (lines) and given version (columns) on the area of the corresponding circle.
-
-The colors represent users.
-The figure shows average performance varying  across users and versions with no clear trend: this is probably because varying user workload (i.e.
-number of sources to which the user is registered) is the reason for the variation in response times.
-\ins{one can see that for user 1 performance degrades over the }
-
-\begin{figure}[!ht]
-  \centering
-  \includegraphics[width=0.7\linewidth]{time_per_user_per_version}
-  \caption{This perspective shows that the evolution of response times for individual users (horizontal lines) across versions (the x-axis)}
-  \label{fig:tuv}
-\end{figure}
-
-  \newpage
-  \section{Preemptive monitoring}
-
-  \ins{INTRO - LINKING WITH PREVIOUS PARAGRAPH: the previous evolutionary view crucial for 
-  understanding the impact of software evolution on performance.
-  However, just by observing this one can't be sure whether 
-  performance changes because of the code or because of the payload.}
-
-  \todo{the following might need a bit of rewrite}
-
-  The concept of {\em preemptive monitoring} of the application performance by means of instrumenting integration unit testing as the synthetic load is similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}, i.e.~testing service-based applications by using dedicated test input in parallel to its normal use and operation.
-The difference is that we take advantage of the capability of the CI framework (i.e.
-Travis in this case) to create an emulated ``live'' environment for integration testing purposes, and use unit testing as the dedicated test input in order to measure performance.
-
-  For example, \Fref{fig:builds} is a screenshot from the dashboard showing the measured response times for 5 consequent builds, with 180 iterations of the unit tests executed in total across all endpoints.
-The outliers on the right of the figure are due to initial requests \ins{which must wait for a boot-up phase of the API}.
-
-    \begin{figure}[h!]
-        \centering
-        \includegraphics[width=\columnwidth]{travis_builds}
-        \caption{Response times in 5 consequent Travis builds in the \zee case study}
-        \label{fig:builds}
-      \end{figure}
-
-  While this integration testing environment is different from the actual production deployment one, and the load used is purely synthetic, as will show in Section~\ref{sec:evaluation}, it can actually be used successfully as an early performance indicator for the application developer.
-\ins{Although predicting the performance of the application would require advanced statistical work, it can be succesfully used as an advanced warning system for the overall performance trend for given endpoints}.
-
-  \paragraph{Integration with Travis}
-  \ml{added new section... also, this has to be moved after introducing the pre-emptive monitoring since it's about running tests. switched the order}
-
-  \ins{The pre-emptive monitoring system} can be configured to work together with Continuous Integration (CI) frameworks like Travis\footnote{\url{https://travis-ci.org/}} that deal with automated integration testing.
-  The developer needs to define the unit test folder of the project, the URL that Travis should use for reporting its results, and the number of iterations for each unit test should be executed in order to generate a load for the application.
-
-  Each time a new build is created through Travis, the \tool automatically detects all available unit tests defined by the application developer and iterates through each one of them the number of times predefined by the developer while monitoring the response times for each request.
-The resulting measurements are persisted in a separate part of the tool database so as not to contaminate the ``live'' monitoring data.
-Beyond seeing the results of each Travis build in the \tool, this feature also allows for preemptive monitoring of the application, as we discuss in the following.
-
-  \ins{Besides functioning as an early warning system for performance, tracking the evolving performance of API tests serves a second purpose.
-To function as an anchor when the developer analyzes performance degradations.
-For example, when a developer sees that a given endpoint has become less performant after in a newly deployed version, they are able to investigate whether the performance degradation is visible also with the synthetic load.
-This would correspond to a performance degradation which is due to ``algorithmic performance degradation''.
-If on the other hand, the performance of the tests does not change between versions, the developer might conclude that the performance degradation could be due to the workload on the machines, or maybe to the workload mix of the users}
-
-  %By these means, an application developer can use unit tests as synthetic loads for driving the performance monitoring of the application before it reaches the production phase.
-As we will discuss further in Section~\ref{sec:testing}, this feature is integral in enabling preemptive performance monitoring of the application.
-
-\todo{here goes the discussion about our forecasting capabilities}
-
-\begin{figure}[h!]
-  \centering
-  \subfloat[The reported response time (in ms) per deployed version in the observation period]{\includegraphics[width=.5\columnwidth]{response_per_version_trunced_trend}}
-  \quad
-  \subfloat[The measured response time (in ms) using integration with Travis]{\includegraphics[width=.46\columnwidth]{travis_builds_no_outliers_trend}}
-  \caption{Comparison of the response times per endpoint: actual production system versus preemptive monitoring data}
-  \label{fig:preemptive}
-\end{figure}
-
-\begin{table}[h]
-  \begin{tabular}{lll}
-    \toprule
-    Iteration & \bfseries Live (median) & \bfseries Travis (median)\\
-    \midrule
-    1 & 1349.41 & 764.87\\ 
-    2 & 1466.13 & 992.87\\
-    3 & 1256.65 & 760.87\\
-    4 & 1266.42 & 813.89\\
-    5 & 1080.68 & 303.4\\
-    \bottomrule
-  \end{tabular}
-\end{table}
-
-Pearson correlation $r(3)=.93, p=.02$
-
-\section{Performance Perspectives}
-\label{sec:views}
-
-\ins{PERFORMANCE DISCUSSION: this introspection, which looks up endpoints, happens only once at the startup of the service, so it is going to affect the API startup performance, but not the actual endpoint performance.
-In our case, the overhead here was quite small: TO MEASURE}.
-
-  %In the remainder of the paper we present several of these perspectives.\footnote{We recommend obtaining a color version of this paper for better readability}
-
-  % The second endpoint conists of two parts, one of them being a table that shows for every monitored endpoint the number of hits it has gotten, the time it was last accessed and its average execution time.
-The second part is a view with four graphs which show:
-  % \begin{itemize} 
-  %   \item A heatmap of the total number of requests to the monitored endpoints
-  %   \item A stacked bar chart that shows the total number of requests to the monitored endpoints per endpoint per day
-  %   \item A boxplot graph showing the average execution time per version of the web service
-  %   \item A boxplot graph showing the average execution time for every monitored endpoint
-  % \end{itemize}
-
-\todo{evaluate which figures to keep and ideally take new screenshots for the ones we keep; consider adding two more views (like Fig. 3.2 of Patrick's thesis)}
-
-% In the Zeeguu case study, one of the slowest endpoints, and one with the highest variability as shown in \Fref{fig:ep} is \epFeedItems: it retrieves a list of recommended articles for a given user.
-However, since a user can be subscribed to anything from one to three dozen article sources, and since the computation of the difficulty is personalized and it is slow, the variability in time among users is likely to be very large.
-
-% \niceseparator
-
-%\section{Preemptive Monitoring}
-%\label{sec:testing}
-%
-%%\todo{discuss here idea of using unit testing as an early performance indicator for evolving services (`preemptive monitoring'?); add a side-by-side figure comparing unit testing vs live system performance per version, maybe calculate error in trend curves?
-find a better term than prediction for the section title (plus introduction)}
-%
-%\todo{Similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}, i.e.~testing service-based applications by using dedicated test input in parallel to its normal use and operation.
-The difference is that we take advantage of the capability of Travis to create an emulated live environment for integration testing purposes, and use unit testing as the dedicated test input in order to measure performance.
-What we are going provide evidence for in the following is that we can indeed use this preemptive monitoring concept as an early performance indicator for evolving services.}
-%
-%The developer needs to define the unit test folder of the project, the URL that Travis should use for reporting its results, and the number of iterations for each unit test should be executed in order to generate
-
-\section{Evaluation}
-\label{sec:evaluation}
-
-%\todo{evaluation of the \tool across the following dimensions: ease of use, effectiveness, overhead to the system; for the last part we need to do the overhead measurement experiment, for the other two we base it on the case study and bring over whatever is necessary from the discussion section}
-
-\section{Discussion}
-
-\todo{use this subsection to discuss limitations (from the previous evaluation section) and add the need to be able to handle the horizontal scaling of the application; remove sectioning and summarize briefly limitations that have already been discussed in the VISSOFT paper/move material to other sections; consider renaming as limitations or something similar}
-
-  \subsection{Distributed System Monitoring}
-  \todo{To think about}
-  Supporting situations where the API is deployed across multiple containers for example.
-
-  \subsubsection{Automatically Monitoring System Evolution}
-
-  The main goal of the \tool design was to allow analytics to be collected and insight to be gained by making the smallest possible changes to a running API.
-%To allow the collection of evolutionary information
-%
-  This technique assumes that the web application code which is the target of the monitoring is deployed using \git in the following way:
-
-  \begin{enumerate}
-    \item The deployment engineer pulls the latest version of the code from the integration server; this will result in a new commit being pointed at by the HEAD pointer. %than previously
-    \item The deployment engineer restarts the new version of the service.
-At this point, the \tool detects that a new HEAD is present in the local code base and consequently starts associating all the new data points with this new commit\footnote{The \tool detects the current version of the analyzed system the first time it is attached to the application object, and thus, assumes that the Flask application is restarted when a new version is deployed.
-This is in tune with the current version of Flask, but if the web server will support dynamic updates in the future, this might have to be taken into account}.
-  \end{enumerate}
-
-  The advantage of this approach is the need for minimal configuration effort, as discussed in the presentation of the tool.
-The disadvantage is that it will consider on equal ground the smallest of commits, even one that modifies a comment, and the shortest lived of commits, e.g.~a commit which was active only for a half an hour before a new version with a bug fix was deployed, with major and minor releases of the software.
-%as a distinct way of grouping the data points.
-  A mechanism to control which versions are important for monitoring purposes is therefore required to be added.
-%
-  A further possible extension point here is supporting other version control systems (e.g. Mercurial). However, this is a straightforward extension.
-
-  \subsubsection{NEW: API Renames}
-  \ml{just an idea}
-
-    The history would normally be lost in the case of an API rename. (also known as the {\em provenance issue} in source code evolution analysis)
-    In theory we could infer that an API was renamed if it is not to be found anymore, and another one with very similar timing characteristics would be found.
-Such a detection can never be completely sure, but it would help with meeting the needs of the API maintainer...
-
-  \subsubsection{User-Awareness }
-
-    For the situations in which the user information is not available, the \tool tracks by default information about different IPs and in some cases this might be a sufficiently good approximation of the user diversity and identity.
-    %
-
-    The visualizations for the user experiene perspectives as presented in Section \ref{sec:user} have been tested with several hundred users (of which about \activeUserCount were active during the course of the study), but the scalability of the visualizations must be further investigated for web services with tens of thousands of users.
-
-  \subsubsection{Other Possible Groupings}
-
-    There are other groupings of service utilization and performance that could be important to the maintainer, that we did not explore in this paper.
-For example, if the service is using OAuth, then together with every request, in the header of the request there is information about the application which is sending a request.
-Grouping the information by application that sends the request could be important in such a context.
-
-    In general, providing a mechanism that would allow very easy specification of groupings (either as code annotations, as normal code, or as configuration options) is an open problem that \tool and any other similar library will have to face.
-
-\section{Related Work}
-\label{sec:related}
-
-\todo{Refer the reader to~\cite{ghezzi2007run} and ~\cite{metzger2010analytical} for a more extensive discussion}
-
-There is a long tradition of using visualization for gaining insight into software performance.
-Tools like Jinsight \cite{Pauw02a} and Web Services Navigator \cite{Pauw05} pioneered such an approach for Java and for Web Services that communicate with SOAP messages.
-Both have an ``omniscient'' view of the services / objects and their interactions.
-As opposed to them, in our work we present an analytics platform which focuses on monitoring a single Python web service from its own point of view.
-
-From the perspective of service monitoring, our work falls within the server-side run-time monitoring of services ~\cite{ghezzi2007run}.
-While we don't implement the more advanced features of related monitoring solutions like QoS policies driving the monitoring, it presents nevertheless an easy to use approach support improving the performance of web applications.
-
-% \va{Mircea: Consider removing the rest for space...}
-% An existing monitoring tool is Pingdom \footnote{https://www.pingdom.com/company/why-pingdom}, which monitors the uptime of an existing web-service.
-This tool works by pinging the websites (up to 60 times) every minute automatically.
-Thus this creates a lot of overhead and is bound to be noisy since it will also be influenced by the speed of the network connection\footnote{Another problem is that such a tool would }
-
-% \todo{Runscope? Others?}
-
-\section{Conclusion and Future Work}
-\label{sec:conclusions}
-
-\todo{update as appropriate}
-
-In this paper we have shown that it is possible to create a monitoring solution which provides basic insight into web service utilization and performance  with very little effort from the developer.
-The user group that we are aiming for with this work is application developers using Flask and Python to build web applications with limited or no budget for implementing their own monitoring solutions.
-The emphasis is in allowing such users to gain insight into how the performance of the service evolves together with the application itself.
-We believe that the same architecture, and lessons can be applied to other frameworks and other languages.
-
-In the future, we plan to perform case studies with other sytstems, with the goal of discover other needs and to wean out the less useful visualizations in the \tool.
-We plan to also extend the tool towards supporting multiple deployments of the same applications across multiple nodes (e.g.
-for the situations where the application is deployed together with a load balancer).
-Finally, we plan to integrate \tool with unit testing as a complementary source of information about performance evolution.
-
-\newpage
-\appendix
-
-  \subsection{Changing the Default /dashboard Endpoint }
-
-  THIS STUFF SHOULD BE MOVED TO THE APPENDIX.
-And in a footnote we just mention that ...
-\ins{REQUIREMENT}: The system must provide an easy way to add more specific configurations if needed.
-A way of overwriting the common sense defaults.
-
-  Further configuration is not needed possible by adding additional statements before the binding definition, for example,
-
-  \begin{lstlisting}[style=custompython]
-  ...
-  dashboard.config.link = 'mydashboard'
-  dashboard.bind(flask_app)
-  \end{lstlisting}
-
-  allows for a custom route (\code{/mydashboard}) to the dashboard to be defined by the programmer. An external configuration file can be used instead, or in addition to this:
-
-  \begin{lstlisting}[style=custompython]
-  ...
-  dashboard.config.from_file('config.cfg')
-  ...
-  \end{lstlisting}
-
-  \subsection{For those who don't use GIT}
-
-    \textit{Manual} version control requires the developer to tag each new version of the application with an appropriate version identifier~\cite{papazoglou2011managing} using the \code{APP\_VERSION} configuration parameter, for example by adding to the configuration file:
-
-      \begin{lstlisting}[style=custompython]
-      [dashboard]
-      APP_VERSION=<versionID>
-      \end{lstlisting}
-
-% references section
-
-\bibliographystyle{abbrv}
-\bibliography{paper}
-
-% that's all folks
-\end{document}

### 2018-04-03 loc1.tex
split the huge file in smaller ones.

%!TEX root=restructured.tex

  \newpage
  \section{LOC \#1: Service Utilization and Performance}

  Since it is taylored for Flask applications, to deploy the \tool one simply needs to use one line of code\footnote{We don't count the import statement that enables that line...} to bind the dashboard to their Flask web application\footnote{\ins{ In this paper we present the integration with APIs written in Python and Flask hoping that this will not prevent the reader from seeing the more general idea; all the tools we show here for Flask can be applied to other API technologies (e.g. Django) by simply providing a few back-end adapters in the right places. }}:

  % caption=Configuring the \tool is straightforward,
  \begin{lstlisting}[style=custompython]
  import flask_monitoringdashboard as dashboard

  # app is main application object
  # every Flask web application has one
  dashboard.bind(app)

  \end{lstlisting}

  During binding, the \tool will search for all endpoints defined in the target application 
    \ins{ 
      and add function wrappers around all the corresponding endpoint functions
    }
% 
  % In order to monitor an endpoint, the \tool creates a function wrapper for the API function that corresponds to the endpoint. This way, the wrapper will be executed whenever that API call is made before the actual function is called. The wrapper contains the code that takes care of monitoring an endpoint. 
%

   \ins{
    The tool takes advantage of the fact that the monitored API already has a web presence, and makes available one extra endpoint (i.e. \code{/dashboard}), which serves the interactive visualization perspectives presented in the remainder of this paper. 
  }  
 The first perspective presents all the automatically discovered endpoints and lets the user select the ones that should be monitored. 
  \Fref{fig:sep} shows that the last access time of every endpoint is tracked irrespective of whether it is selected by the user to be monitored or not.

    \begin{figure}[h!]
      \centering
      \includegraphics[width=\linewidth]{selecting_endpoints.png}
      \caption{Once connected to an API the Dashboard presents the endpoints that are available for monitoring}
      \label{fig:sep}
    \end{figure}

  \ins{
    We have decided for an opt-in approach to monitoring to avoid any the performance penalties incurred by the dashboard to affect performance sensitive endpoints. We discuss performance issues later. Also, we discuss later one situation in which opt-out might be desirable. 
  }

  \ins{

    One alternative to allowing the user to use annotations would be to let them to annotate the code. However, this pollutes the code, and prevents deploying two versions which would monitor different endpoints.

  }

\niceseparator

  \ins
  {

      The remainder of this section presents several of the interactive
      visualizations that become available without any further configuration\footnote{We recommend obtaining a color version of this paper for better readability}.

  }

### 2018-04-03 performance.tex
split the huge file in smaller ones.

--- 2018-03-21 restructured.tex
@@
-\documentclass[conference]{IEEEtran}
-	\pdfpagewidth=8.5truein
-	\pdfpageheight=11truein
+%!TEX root=restructured.tex
 
-\usepackage{graphicx}
-\usepackage{subfig}
-\usepackage[bookmarks=false]{hyperref}
-\usepackage{xspace}
-\usepackage{listings}
-\usepackage[usenames, dvipsnames]{color}
+\subsection{Service Performance}
+\label{sec:perf}
 
-\usepackage{booktabs}
-
-\graphicspath{{./img/}}
-
-%\hyphenation{op-tical net-works semi-conduc-tor}
-
-\input{useful_defines}
-
-\begin{document}
-
-%
-\title{Agile Monitoring of Evolving Web Services}
-
-\author{
-\IEEEauthorblockN{Patrick Vogel, Thijs Klooster, Vasilios Andrikopoulos, Mircea Lungu}\\
-Johann Bernoulli Institute for Mathematics and Computer Science\\
-University of Groningen, Netherland\\
-Email: \{t.klooster.1,p.p.vogel\}@student.rug.nl, \{v.andrikopoulos,m.f.lungu\}@rug.nl 
-}
-
-% make the title area
-\maketitle
-
-\begin{abstract}
-\todo{Update abstract \& title}
-  Tens of thousands of web applications are written in Flask, a Python-based web framework.
-Despite a rich ecosystem of extensions, there is none that supports the developer in gaining insight into the evolving performance of their service.
-In this paper, we introduce \tool, a library that addresses this problem.
-We present the ease with which the library can be integrated in an already existing web application, discuss some of the visualization perspectives that the library provides and point to some future challenges for similar libraries.
-
-\end{abstract}
-
-\section{Introduction}
-
-\todo{Intro to be rewritten; refer to VISSOFT paper as previous work, explain focus is on integration with version control/CI for evolution support, and use of unit tests as early performance indicator}
-
-\todo{motivate with: monitoring in a world which is becoming more and more service-oriented is essential because it allows to perform three main types of actions: system adaptations to provide the request service at the desired level of quality, enabling flexibility in dealing with changing requirements and modes of operation, and enabling operational awareness through dashboards organizing the collected data~\cite{pernici2016monitoring}.
-In this work we focus on the last part and we discuss how a minimal effort probe-based solution for a particular type of web services can be used to facilitate the other two types of actions by involving the developer.}
-
-\todo{explain: in~\cite{vogel2017low} we introduced the \tool an a high level with a focus on presenting its performance visualization aspects; in this work we only summarize these features by means of introducing some of the provided functionalities of the \tool.
-We focus on discussing how the tool is implemented and operating, provide a deeper look at its capabilities for integration with version control and continuous integration environments, introduce a mechanism for performance prediction, and provide an evaluation of the proposed approach based on a case study.
-}
-
-%Every system is a distributed system nowadays \cite{cavage2013there}.
-Indeed a very large number of applications and web applications are nowadays implemented as two-tier architectures with a front-end implemented with web technologies and a service back-end.
-%\ml{I'm not completely happy with this paragraph}
-{\em There is no getting around it: you are building a distributed system} argues a recent article \cite{cavage2013there}.
-Indeed, even the simplest second-year student project is a web application implemented as two-tier architecture with a Javascript/HTML5 front-end a service backend, usually a REST API.
-
-% \hfill mds
-% Many contemporary programming languages are offering libraries, modules, or frameworks that facilitate the development of such architectures. 
-Python is one of the most popular programming language choices for implementing the back-end of web applications.
-GitHub contains more than 500K open source Python projects and the Tiobe Index\footnote{TIOBE programming community index is a measure of popularity of programming languages, created and maintained by the TIOBE Company based in Eindhoven, the Netherlands} ranks Python as the 4th most popular programming language as of June 2016.
-
-% possible flask summary
-Within the Python community, Flask\footnote{\url{http://flask.pocoo.org/}} is a very popular web framework\footnote{More than 25K projects on GitHub (5\% of all Python projects) are implemented with Flask (cf.
-a GitHub search for ``language:Python Flask'')}.
-It provides simplicity and flexibility by implementing a bare-minimum web server, and thus advertises as a micro-framework.
-The Flask tutorial shows how setting up a simple Flask {\em ``Hello World''} web-service requires no more than 5 lines of Python code \cite{ flask:tutorial}.
-% end of summary
-
-Despite their popularity, to the best of our knowledge, there is no simple solution for monitoring the evolving performance of Flask web applications.
-Thus, every one of the developers of these projects faces one of the following options when confronted with the need of gathering insight into the runtime behavior of their implemented services:
-
-  \begin{enumerate}
-
-    \item Use a commercial monitoring tool which treats the subject API as a black-box (e.g. Pingdom, Runscope). 
-    % , Graphite+Graphana+statd etc.
-
-    \item Implement their own ad-hoc analytics solution, having to reinvent basic visualization and interaction strategies.
-
-    \item Live without analytics insight into their services.
-
-  \end{enumerate}
-
-%\todo{For the first point in the list, we can also argue that analytics solutions like Google Analytics can be used, but they have no notion of versioning/integration with the development life cycle.
-Feel free to cite \cite{papazoglou2011managing} for service evolution purposes}
-
-For projects on a budget (e.g.
-research, startups) the first and the second options are often not available due to time and financial constraints.
-Even when using 3rd-party analytics solutions, a critical insight into the evolution of the exposed services of the web application, is missing because such solutions have no notion of versioning and no integration with the development life cycle.~\cite{papazoglou2011managing}
-
-To avoid projects ending up in the third situation, that of living without analytics, in this paper we present \tool~ --- a low-effort service monitoring library for Flask-based Python web services that is easy to integrate and enables the {\em agile assessment} of service evolution.
-\cite{Nier12b}
-
-As a case study, on which we will illustrate our solution, we are going to use an open source API which, for several years, was in the third of the above-presented situations.
-
-% In the next section, we will present a case study of an open source research API which was for a long time in the third situation presented above -- deployed without analytics insight.
-
-\todo{Add signposting paragraph}
-
-\section{Case Study: The API (Zeeguu)}
-\label{sec:case}
-
-%\todo{Either leave as is or add more info on usage by college if possible}
-
-  \zee\footnote{\url{https://github.com/zeeguu-ecosystem/}} is a platform and an ecosystem of applications for accelerating vocabulary acquisition in a foreign language \cite{Lungu16}. 
-%
-  The architecture of the ecosystem has at its core an API and a series of satellite applications that together offer to a learner three main inter-dependent features:
-
-  \begin{enumerate}
-
-    \item Reader applications that provide effortless translations for those texts which are too difficult for the readers.
-
-    \item Interactive exercises personally generated based on the preferences and past likes of the learner.
-
-    \item Article recommendations which are personalized for the interests of the learner and come with difficulty estimation that helps the learner find articles with the appropriate difficulty.
-
-  \end{enumerate}
-
-  The core API implemented with Flask and Python provides correspondingly three types of functionality: contextual translations, article recommendations, and personalized exercise suggestions.
-In total the API provides a bit less than 50 endpoints, out of which probably a dozen are very frequently used.
-  % The development of the core API itself is a research project.
-
-  At the time of writing this article, the ecosystem consists of a reader web application, a web based exercises platform, and a smartwatch application, which are used at the moment of writing this article by more than \activeUserCount active users.
-The users come from a highschool, a language school, and some users are using it on their own, without any educational context.
-The highest load we observed until now on the API consisted of 12K requests in one day.
-
-  We will use this \zee API as a case study for this paper. 
-  All the figures in this paper are captured from the actual deployment of \tool in the context of the \zee API.
-The figures are interactive offering basic data exploration capabilities: filter, zoom, and details on demand\cite{Shne99a}.
-The \tool deployment for the case study can be accessed publicly\footnote{\url{https://zeeguu.unibe.ch/api/dashboard}; username: {\em guest}, password: {\em soap-sac}}.
-
- % \todo{create new account, update URL if necessary, update activeUserCount, add a couple of sentences of the current state of the case study with the college/language center}
-% \ml{we should consider adding also one section in which the architecture/implementation and main features of the dashboard are presented before going on with discussing them in more depth in the following sections --- this should include a rundown on which views are provided from where (overview or per endpoint)}
-
-\newpage
-
-\section{Case Study: The Technology (Flask)}
-\label{sec:tool}
-
- Flask is a microframework for Python. It IS USED BY THOUSANDS OF PROJECTS. ELABORATE MORE ON THIS... 
- \vspace{2cm}
-
- Flask depends on two external libraries: the Jinja2 template engine and the Werkzeug WSGI toolkit.
-
- “Micro” does not mean that your whole web application has to fit into a single Python file (although it certainly can), nor does it mean that Flask is lacking in functionality.
-The “micro” in microframework means Flask aims to keep the core simple but extensible.
-Flask won’t make many decisions for you, such as what database to use.
-Those decisions that it does make, such as what templating engine to use, are easy to change.
-Everything else is up to you, so that Flask can be everything you need and nothing you don’t.
-
- Flask has many configuration values, with sensible defaults, and a few conventions when getting started.
-By convention, templates and static files are stored in subdirectories within the application’s Python source tree, with the names templates and static respectively.
-While this can be changed, you usually don’t have to, especially when getting started.
-
- A Python web application based on WSGI has to have one central callable object that implements the actual application.
-In Flask this is an instance of the Flask class.
-Each Flask application has to create an instance of this class itself and pass it the name of the module, but why can’t Flask do that itself?
-
-\begin{lstlisting}[style=custompython]
-
-from flask import Flask
-app = Flask(__name__)
-
-@app.route('/')
-def index():
-    return 'Hello World!'
-
-\end{lstlisting}
-
-\subsection*{NOTHING PREVENTS GENERALIZATION}
-In this article we take as an example the Flask case study.
-But there is nothing that prevents the generalization of 
-our approach.
-
-MAKE SURE THAT THIS IS CLEAR.
-
-    \begin{figure*}[ht!]
-      \centering
-      \includegraphics[width=0.8\linewidth]{interceptor}
-      \caption{The first thing that needs to be done, is to decorate the application object with an INTERCEPTOR}
-      \label{fig:sep}
-    \end{figure*}
-
-\newpage
-
-\section{Monitoring Service Utilization}
-
-REMEMBER THE GOAL: Our main goal with the \tool issues to allow the integration of the dashboard with an API with minimal effort.
-
-\ins{The first important question that such a 
-service monitoring system should provide is 
-information about what endpoints are being
-used and by whom}.
-
-TODO: Find references, arguments, related work
-that provide more details about why understanding
-who uses endpoints, and which are used are important.
-
-This is clearly the case in static analysis (see paper
-by haenni and lungu...) but should even more be the
-case in APIs.
-
-  \newpage
-TODO: TALK ABOUT THE DATABASE...
-
-  As discussed in the introductory section, in previous work~\cite{vogel2017low} we introduced \tool, a drop-in Python library that allows developers to monitor their Flask-based Python web applications with minimal effort.
-%
-  The \tool is implemented for Python 3.6 and is available on the Python Package Index repository\footnote{\url{https://pypi.python.org/pypi/flask-monitoring-dashboard/1.8}} from where it can be installed by running \install from the command line.
-%  
-  The source code of the \tool is published under a permissive MIT license and is available on GitHub\footnote{\url{https://github.com/flask-dashboard}}.
-
-  \newpage
-  \section{One Line Of Code Configuration With Technology Specific Interceptors}
-
-  The fact that the \tool itself is being developed in Python using Flask makes binding to the services of a to-be monitored application developed with the same technologies easy and intuitive.
-
-  \ins{In general, what matters is that with the right backend, we can do this minimal configuration for any technology... }
-
-  To start using the \tool, one simply needs to one lines of code\footnote{we don't count the import statement that enables that line...} to their Flask web service\footnote{\ins{ In this paper we present the integration with APIs written in Python and Flask hoping that this will not prevent the reader from seeing the more general idea; all the tools we show here for Flask can be applied to other API technologies (e.g.
-Django) by simply providing a few back-end adapters in the right places.
-}}:
-
-  % caption=Configuring the \tool is straightforward,
-  \begin{lstlisting}[style=custompython]
-  import dashboard
-
-  # flask_app is the Flask app object
-  dashboard.bind(flask_app)
-  \end{lstlisting}
-
-  \ins{The dashboard can take advantage of the fact that it is monitoring an API...
-so it can simply just add one extra endpoint there.} After (re)deployment of the application, the interactive visualizations of the \tool become available at the \code{/dashboard} route of the Flask application.
-
-  During binding, the \tool will search for all endpoints defined in the target application.
-
-  In order to monitor an endpoint, the \tool creates a function wrapper for the API function that corresponds to the endpoint.
-This way, the wrapper will be executed whenever that API call is made before the actual function is called.
-The wrapper contains the code that takes care of monitoring an endpoint.
-Data collected by the wrappers are persisted in a local database.
-The SQLAlchemy Object Relational Mapper\footnote{\url{https://www.sqlalchemy.org/}} on top of SQLite\footnote{\url{https://www.sqlite.org/}} is used for this purpose.
-  \ins{DISCUSSION about how do to this in other languages? Java: AspectJ probably, etc.}
-
-  These will be presented to the user in the tool web interface, where the user can select the ones that should be monitored, see \Fref{fig:sep}.
-
-    \begin{figure}[h!]
-      \centering
-      \includegraphics[width=\linewidth]{selecting_endpoints.png}
-      \caption{Once connected to an API the Dashboard presents the endpoints that are available for monitoring}
-      \label{fig:sep}
-    \end{figure}
-
-  This is the alternative to using annotations like other white-box tools (e.g. the other flask performance plugin).
-
-  \ins{PROMISE:} The last access time of the endpoint by users, irrespective of whether it is selected by the user to be monitored or not, is also displayed as the means of identifying potentially obsolete endpoints.
-
-  \ml{we must insert at least one dashboard view here...
-otherwise people will lose their patience...
-also it will help with explaining the outliers...
-also, one should see that a large variety of views are already available at this point...
-I know this risks being  }
+  The \tool also collects information regarding endpoint performance. The view in \Fref{fig:ep} summarizes the response times for various endpoints by using a box-and-whiskers plot.
 
@@
 However, since a user can be subscribed to anything from one to three dozen article sources, and since the computation of the difficulty is personalized and it is slow, the variability in time among users is likely to be very large.
-
-\begin{figure*}[h!]
-  \centering
-  \subfloat[number of requests per endpoint per day]{\includegraphics[width=.9\columnwidth]{number_of_requests_}\label{fig:aeu}}
-  \subfloat[requests per hour heatmap]{\includegraphics[width=\columnwidth]{daily_patterns_}\label{fig:dp}}
-  \caption{Some of the available views\label{fig:views}}
-\end{figure*}
-
-\newpage
-\section{Service Utilization}
-\label{sec:util}
-
-\ins{A discussion must be added here about the types of useful information and problems that can be solved with these views.}
-
-\textit{Service Utilization} presents information about the usage of all endpoints of interest to the developer,
-
-The most fundamental insight that a service maintainer needs regards service utilization. %\vspace{0.5cm}
-
-Simplest utilization perspective:
-
-\begin{figure}[h!]
-  \centering
-  \includegraphics[width=0.5\linewidth]{endpoint-call-count}
-  \caption{Caption here}
-  \label{fig:figure1}
-\end{figure}
-
-% REQUIREMENT: Adding a time interval is important. 
-% Otherwise the performance numbers in chart 1 are not good enough... also an outlier very far in the past, will make all the heatmap data points to look very pale in comparison forever after.
-
-Figure \ref{fig:aeu} shows a first perspective on endpoint utilization that \tool provides: a stacked bar chart of the number of hits to various endpoints grouped by day\footnote{Endpoint colors are the same in different views}.
-Figure~\ref{fig:aeu} in particular shows that at its peak the API has about 12.000 hits per day.
-
-The way users interact with the platform can also be inferred since the endpoints are indicators of different activity types, e.g.:
-
-\begin{enumerate}
-
-  \item {\color{myred} \epTranslations} is an indicator of the amount of foreign language reading the users are doing, and
-
-  \item {\color{mygreen} \epOutcome} is an indicator of the amount of foreign vocabulary practice the users are doing.
-
-\end{enumerate}
-
-%\begin{figure}[!ht]
-% \centering
-% \includegraphics[width=\linewidth]{number_of_requests_}
-% \caption{The number of requests per endpoint per day view shows the overall utilization of the monitored application}
-% \label{fig:aeu}
-%\end{figure}
-
-Besides showing the overall utilization, this endpoint provides the maintainer with information relevant for decisions regarding endpoint deprecation --- one of the most elementary ways of {\em understanding the needs of the downstream}\cite{Haen14a}.
-In our case study, the maintainer realized that one endpoint which they thought was not being used (i.e.
-\code{words\_to\_study}), contrary to their expectations, was actually being used\footnote{A complementary type of usage information can also be discovered in the view presented in Figure \ref{fig:sep} where seeing that an endpoint is never accessed can increase the confidence of the maintainer that a given endpoint is not used, although it can never be used a proof.}.
-
-\niceseparator
-
-%   \todo{Add the time series graph and discuss it before the heatmap? We can then sell the heatmap better} 
-%   \ml{Not sure about which graph you refer to here V}
-
-A second type of utilization question that the \tool can answer automatically regards {\em cyclic patterns of usage per hour of day} by means of a heatmap, as in \Fref{fig:dp}.
-
-% \mltp{ can we add vertical lines that highlight the beginning of a new week (e.g. before Sunday): }
-% Patrick: adding separators in the graph is unfortunately not supported by the library.
-
-%\begin{figure}[!ht]
-% \centering
-% \includegraphics[width=0.9\linewidth]{daily_patterns_}
-% \caption{Usage patterns become easy to spot in the requests per hour heatmap}
-% \label{fig:dp}
-%\end{figure}
-
-Figure \ref{fig:dp} shows the API not being used during the early morning hours, with most of the activity focused around working hours and some light activity during the evening.
-This is consistent with the fact that the current users are all in the central European timezone.
-Also, the figure shows that the spike in utilization that was visible also in the previous graph happended in on afternoon/evening.
-
-\newpage
-\section{Endpoint Performance}
-\label{sec:perf}
-
-  The \tool also collects information regarding endpoint performance. The view in \Fref{fig:ep} summarizes the response times for various endpoints by using a box-and-whiskers plot.
 
@@
 
-\begin{figure}[h!]
-  \centering
-  \subfloat[overview of the tracked endpoints]{\includegraphics[width=0.8\columnwidth]{overview}\label{fig:overview}}
-  \caption{Some of the available views\label{fig:views}}
-\end{figure}
-
   \newpage
-  \section{Evolution monitoring }
-
-  \begin{figure}[h!]
-  \centering
-  \includegraphics[width=0.8\linewidth]{translation_endpoint_evolution_}
-  \caption{The Performance Evolution of the \epTranslations endpoint}
-  \label{fig:tee}
-\end{figure}
-
-  When thinking about performance, 
-  one can not avoid thinking about the performance evolution.
-
-  Version control in \tool can be supported in two ways,
-  the developer explicitly states the current version, 
-  or the current version is automatically detected based
-  on some version control system.
-
-  However, with the current configuration of the tool, it would be impossible for the maintainer to see the improvements resulting from the optimization.
-
-  \ins{The latest STack Overflow Developer Survey: 88.4\% of **professional** developers use git.}  
-  If assume that the code that they run is deployed using .git, then with an extra line of configuration they can allow \tool to find the git\footnote{\url{https://git-scm.com/}} folder of the deployed service and automatically detect the version of the project that is running:
-
-    \begin{lstlisting}[style=custompython]
-
-      dashboard.config.git = 'path/to/.git'
-
-    \end{lstlisting}
-
-  If the \tool can automatically detect the current version of the project by reading the .git configuration as soon as the API is started and can then group measurements by version\footnote{Alternatively, the maintainer can add version identifiers manually for the web application through a configuration file if the system does not use git.}.
-\Fref{fig:tee} is a zoomed-in version of such a view for \epTranslations with versions increasing from top to bottom
-
-  This view confirms that the performance of the translation endpoint improved in the recent versions: the median of the last three versions is constantly moving towards the left, and progresses from 1.4 seconds (in the top-most box plot in \Fref{fig:tee}) to 0.8 in the latest version (bottom-most box plot).
-
-  CHALLENGE: Supporting other types of version control... not real challenge, as we said 88percent of devs use git. 
-  CHALLENGE 2: detecting minor versions which don't need to be tracked, since they didn't touch the performance of the system. E.g. a modification of the README, etc.
-
-\todo{TODO: endpoint evolution }
-
-  \begin{figure*}[h!]
-    \centering
-    \includegraphics[width=\linewidth]{outliers}
-    \caption{Automatically collected outlier information}
-    \label{fig:figure1}
-  \end{figure*}
-  \newpage
-  \section{Automated outlier detection and monitoring}
+  \subsection{Performance Insight: Automated outlier detection and monitoring}
 
@@
 
+  \begin{figure}[h!]
+    \centering
+    \includegraphics[width=\linewidth]{outlier-annotated}
+    \caption{Automatically collected outlier information}
+    \label{fig:figure1}
+  \end{figure}
+
 The bottom panel shows the stack trace. 
@@
 
-\subsection {A Special Type of Outlier: Exceptions}
-
-   TODO: probably something for the journal extension ...
-  statistics about which endpoints fail most often might be useful too.
-  same information as the outlier maybe?
-
-\begin{figure*}[!ht]
+\begin{figure*}[h!]
   \centering
-  \includegraphics[width=0.4\linewidth]{time_per_user}
-  \caption{The \epFeedItems shows a very high variability across users}
-  \label{fig:tpu}
+  \includegraphics[width=0.9\linewidth]{utilization-evolution}
+  \caption{The Evolution of Endpoint Utilization Across System Versions}
+  \label{fig:tee}
 \end{figure*}
-
-\newpage
-\section{Grouping Requests}
-\label{sec:user}
-
-For service endpoints which run computations in real time, the maintainer of a system might want to understand the endpoint performance on a per-user basis, especially for situations where the system response time is a function of some individual user load\footnote{E.g.
-in GMail some users have two emails while other have twenty thousand and this induces different response times for different users}.
-
-\todo{remove the configuration part and rephrase the following, the discussion has been partially subsumed by Section~\ref{sec:tool}.}
-
-To enable this, the \tool must be configured to associate an API call with a given user.
-The simplest way is to take advantage of the architecture of Flask applications in which a global \code{flask.request} object can be used to retrieve the session which can in turn lead to user identification:
-
-%\begin{lstlisting}[float,caption=Simply define a custom app-specific function for user retrieval and pass it to the \tool to group information by user,style=custompython]
-\begin{lstlisting}[style=custompython]  
-# app specific way of extracting the user
-# from a flask request object    
-def get_user_id():
-sid = int(flask.request.args['session'])
-session = User.find_for_session(sid)
-return user_id
-
-# attaching the get_user_id function
-dashboard.config.get_group_by = get_user_id
-
-\end{lstlisting}
-
-% \niceseparator
-
-\todo{consider losing one or both of the figures --- or at least updating them}
-
-In Zeeguu, \epFeedItems retrieves a list of recommended articles for a given user.
-Cf.
-\Fref{fig:ep} it is the endpoint with the slowest response time and highest variability.
-The reason for this is that a user can be subscribed to anything from one to three dozen article sources and for each of the sources the system must compute the personalized difficulty of each article at every request.
-
-A \perspective{Per-User Performance} perspective should show the different response times for different users.
-Figure \ref{fig:tpu} presents a subset of the corresponding view in the \tool.
-The figure shows that the response times for this endpoint can vary considerably for different users with some extreme cases where a user has to wait a full minute until their recommended articles are shown\footnote{After seeing this perspective, the maintainer refactored the architecture of the system to move part the difficulty computation out of the interactive loop}.
-
-The limitation of the previous view is that it does not present the information also on a per version basis.
-To address this, a different visual perspective entitled \perspective{Multi-Version per-User Performance} can be defined.
-Figure \ref{fig:tuv} presents such a perspective by mapping the average execution time for a given user (lines) and given version (columns) on the area of the corresponding circle.
-
-The colors represent users.
-The figure shows average performance varying  across users and versions with no clear trend: this is probably because varying user workload (i.e.
-number of sources to which the user is registered) is the reason for the variation in response times.
-\ins{one can see that for user 1 performance degrades over the }
-
-\begin{figure}[!ht]
-  \centering
-  \includegraphics[width=0.7\linewidth]{time_per_user_per_version}
-  \caption{This perspective shows that the evolution of response times for individual users (horizontal lines) across versions (the x-axis)}
-  \label{fig:tuv}
-\end{figure}
-
-  \newpage
-  \section{Preemptive monitoring}
-
-  \ins{INTRO - LINKING WITH PREVIOUS PARAGRAPH: the previous evolutionary view crucial for 
-  understanding the impact of software evolution on performance.
-  However, just by observing this one can't be sure whether 
-  performance changes because of the code or because of the payload.}
-
-  \todo{the following might need a bit of rewrite}
-
-  The concept of {\em preemptive monitoring} of the application performance by means of instrumenting integration unit testing as the synthetic load is similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}, i.e.~testing service-based applications by using dedicated test input in parallel to its normal use and operation.
-The difference is that we take advantage of the capability of the CI framework (i.e.
-Travis in this case) to create an emulated ``live'' environment for integration testing purposes, and use unit testing as the dedicated test input in order to measure performance.
-
-  For example, \Fref{fig:builds} is a screenshot from the dashboard showing the measured response times for 5 consequent builds, with 180 iterations of the unit tests executed in total across all endpoints.
-The outliers on the right of the figure are due to initial requests \ins{which must wait for a boot-up phase of the API}.
-
-    \begin{figure}[h!]
-        \centering
-        \includegraphics[width=\columnwidth]{travis_builds}
-        \caption{Response times in 5 consequent Travis builds in the \zee case study}
-        \label{fig:builds}
-      \end{figure}
-
-  While this integration testing environment is different from the actual production deployment one, and the load used is purely synthetic, as will show in Section~\ref{sec:evaluation}, it can actually be used successfully as an early performance indicator for the application developer.
-\ins{Although predicting the performance of the application would require advanced statistical work, it can be succesfully used as an advanced warning system for the overall performance trend for given endpoints}.
-
-  \paragraph{Integration with Travis}
-  \ml{added new section... also, this has to be moved after introducing the pre-emptive monitoring since it's about running tests. switched the order}
-
-  \ins{The pre-emptive monitoring system} can be configured to work together with Continuous Integration (CI) frameworks like Travis\footnote{\url{https://travis-ci.org/}} that deal with automated integration testing.
-  The developer needs to define the unit test folder of the project, the URL that Travis should use for reporting its results, and the number of iterations for each unit test should be executed in order to generate a load for the application.
-
-  Each time a new build is created through Travis, the \tool automatically detects all available unit tests defined by the application developer and iterates through each one of them the number of times predefined by the developer while monitoring the response times for each request.
-The resulting measurements are persisted in a separate part of the tool database so as not to contaminate the ``live'' monitoring data.
-Beyond seeing the results of each Travis build in the \tool, this feature also allows for preemptive monitoring of the application, as we discuss in the following.
-
-  \ins{Besides functioning as an early warning system for performance, tracking the evolving performance of API tests serves a second purpose.
-To function as an anchor when the developer analyzes performance degradations.
-For example, when a developer sees that a given endpoint has become less performant after in a newly deployed version, they are able to investigate whether the performance degradation is visible also with the synthetic load.
-This would correspond to a performance degradation which is due to ``algorithmic performance degradation''.
-If on the other hand, the performance of the tests does not change between versions, the developer might conclude that the performance degradation could be due to the workload on the machines, or maybe to the workload mix of the users}
-
-  %By these means, an application developer can use unit tests as synthetic loads for driving the performance monitoring of the application before it reaches the production phase.
-As we will discuss further in Section~\ref{sec:testing}, this feature is integral in enabling preemptive performance monitoring of the application.
-
-\todo{here goes the discussion about our forecasting capabilities}
-
-\begin{figure}[h!]
-  \centering
-  \subfloat[The reported response time (in ms) per deployed version in the observation period]{\includegraphics[width=.5\columnwidth]{response_per_version_trunced_trend}}
-  \quad
-  \subfloat[The measured response time (in ms) using integration with Travis]{\includegraphics[width=.46\columnwidth]{travis_builds_no_outliers_trend}}
-  \caption{Comparison of the response times per endpoint: actual production system versus preemptive monitoring data}
-  \label{fig:preemptive}
-\end{figure}
-
-\begin{table}[h]
-  \begin{tabular}{lll}
-    \toprule
-    Iteration & \bfseries Live (median) & \bfseries Travis (median)\\
-    \midrule
-    1 & 1349.41 & 764.87\\ 
-    2 & 1466.13 & 992.87\\
-    3 & 1256.65 & 760.87\\
-    4 & 1266.42 & 813.89\\
-    5 & 1080.68 & 303.4\\
-    \bottomrule
-  \end{tabular}
-\end{table}
-
-Pearson correlation $r(3)=.93, p=.02$
-
-\section{Performance Perspectives}
-\label{sec:views}
-
-\ins{PERFORMANCE DISCUSSION: this introspection, which looks up endpoints, happens only once at the startup of the service, so it is going to affect the API startup performance, but not the actual endpoint performance.
-In our case, the overhead here was quite small: TO MEASURE}.
-
-  %In the remainder of the paper we present several of these perspectives.\footnote{We recommend obtaining a color version of this paper for better readability}
-
-  % The second endpoint conists of two parts, one of them being a table that shows for every monitored endpoint the number of hits it has gotten, the time it was last accessed and its average execution time.
-The second part is a view with four graphs which show:
-  % \begin{itemize} 
-  %   \item A heatmap of the total number of requests to the monitored endpoints
-  %   \item A stacked bar chart that shows the total number of requests to the monitored endpoints per endpoint per day
-  %   \item A boxplot graph showing the average execution time per version of the web service
-  %   \item A boxplot graph showing the average execution time for every monitored endpoint
-  % \end{itemize}
-
-\todo{evaluate which figures to keep and ideally take new screenshots for the ones we keep; consider adding two more views (like Fig. 3.2 of Patrick's thesis)}
-
-% In the Zeeguu case study, one of the slowest endpoints, and one with the highest variability as shown in \Fref{fig:ep} is \epFeedItems: it retrieves a list of recommended articles for a given user.
-However, since a user can be subscribed to anything from one to three dozen article sources, and since the computation of the difficulty is personalized and it is slow, the variability in time among users is likely to be very large.
-
-% \niceseparator
-
-%\section{Preemptive Monitoring}
-%\label{sec:testing}
-%
-%%\todo{discuss here idea of using unit testing as an early performance indicator for evolving services (`preemptive monitoring'?); add a side-by-side figure comparing unit testing vs live system performance per version, maybe calculate error in trend curves?
-find a better term than prediction for the section title (plus introduction)}
-%
-%\todo{Similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}, i.e.~testing service-based applications by using dedicated test input in parallel to its normal use and operation.
-The difference is that we take advantage of the capability of Travis to create an emulated live environment for integration testing purposes, and use unit testing as the dedicated test input in order to measure performance.
-What we are going provide evidence for in the following is that we can indeed use this preemptive monitoring concept as an early performance indicator for evolving services.}
-%
-%The developer needs to define the unit test folder of the project, the URL that Travis should use for reporting its results, and the number of iterations for each unit test should be executed in order to generate
-
-\section{Evaluation}
-\label{sec:evaluation}
-
-%\todo{evaluation of the \tool across the following dimensions: ease of use, effectiveness, overhead to the system; for the last part we need to do the overhead measurement experiment, for the other two we base it on the case study and bring over whatever is necessary from the discussion section}
-
-\section{Discussion}
-
-\todo{use this subsection to discuss limitations (from the previous evaluation section) and add the need to be able to handle the horizontal scaling of the application; remove sectioning and summarize briefly limitations that have already been discussed in the VISSOFT paper/move material to other sections; consider renaming as limitations or something similar}
-
-  \subsection{Distributed System Monitoring}
-  \todo{To think about}
-  Supporting situations where the API is deployed across multiple containers for example.
-
-  \subsubsection{Automatically Monitoring System Evolution}
-
-  The main goal of the \tool design was to allow analytics to be collected and insight to be gained by making the smallest possible changes to a running API.
-%To allow the collection of evolutionary information
-%
-  This technique assumes that the web application code which is the target of the monitoring is deployed using \git in the following way:
-
-  \begin{enumerate}
-    \item The deployment engineer pulls the latest version of the code from the integration server; this will result in a new commit being pointed at by the HEAD pointer. %than previously
-    \item The deployment engineer restarts the new version of the service.
-At this point, the \tool detects that a new HEAD is present in the local code base and consequently starts associating all the new data points with this new commit\footnote{The \tool detects the current version of the analyzed system the first time it is attached to the application object, and thus, assumes that the Flask application is restarted when a new version is deployed.
-This is in tune with the current version of Flask, but if the web server will support dynamic updates in the future, this might have to be taken into account}.
-  \end{enumerate}
-
-  The advantage of this approach is the need for minimal configuration effort, as discussed in the presentation of the tool.
-The disadvantage is that it will consider on equal ground the smallest of commits, even one that modifies a comment, and the shortest lived of commits, e.g.~a commit which was active only for a half an hour before a new version with a bug fix was deployed, with major and minor releases of the software.
-%as a distinct way of grouping the data points.
-  A mechanism to control which versions are important for monitoring purposes is therefore required to be added.
-%
-  A further possible extension point here is supporting other version control systems (e.g. Mercurial). However, this is a straightforward extension.
-
-  \subsubsection{NEW: API Renames}
-  \ml{just an idea}
-
-    The history would normally be lost in the case of an API rename. (also known as the {\em provenance issue} in source code evolution analysis)
-    In theory we could infer that an API was renamed if it is not to be found anymore, and another one with very similar timing characteristics would be found.
-Such a detection can never be completely sure, but it would help with meeting the needs of the API maintainer...
-
-  \subsubsection{User-Awareness }
-
-    For the situations in which the user information is not available, the \tool tracks by default information about different IPs and in some cases this might be a sufficiently good approximation of the user diversity and identity.
-    %
-
-    The visualizations for the user experiene perspectives as presented in Section \ref{sec:user} have been tested with several hundred users (of which about \activeUserCount were active during the course of the study), but the scalability of the visualizations must be further investigated for web services with tens of thousands of users.
-
-  \subsubsection{Other Possible Groupings}
-
-    There are other groupings of service utilization and performance that could be important to the maintainer, that we did not explore in this paper.
-For example, if the service is using OAuth, then together with every request, in the header of the request there is information about the application which is sending a request.
-Grouping the information by application that sends the request could be important in such a context.
-
-    In general, providing a mechanism that would allow very easy specification of groupings (either as code annotations, as normal code, or as configuration options) is an open problem that \tool and any other similar library will have to face.
-
-\section{Related Work}
-\label{sec:related}
-
-\todo{Refer the reader to~\cite{ghezzi2007run} and ~\cite{metzger2010analytical} for a more extensive discussion}
-
-There is a long tradition of using visualization for gaining insight into software performance.
-Tools like Jinsight \cite{Pauw02a} and Web Services Navigator \cite{Pauw05} pioneered such an approach for Java and for Web Services that communicate with SOAP messages.
-Both have an ``omniscient'' view of the services / objects and their interactions.
-As opposed to them, in our work we present an analytics platform which focuses on monitoring a single Python web service from its own point of view.
-
-From the perspective of service monitoring, our work falls within the server-side run-time monitoring of services ~\cite{ghezzi2007run}.
-While we don't implement the more advanced features of related monitoring solutions like QoS policies driving the monitoring, it presents nevertheless an easy to use approach support improving the performance of web applications.
-
-% \va{Mircea: Consider removing the rest for space...}
-% An existing monitoring tool is Pingdom \footnote{https://www.pingdom.com/company/why-pingdom}, which monitors the uptime of an existing web-service.
-This tool works by pinging the websites (up to 60 times) every minute automatically.
-Thus this creates a lot of overhead and is bound to be noisy since it will also be influenced by the speed of the network connection\footnote{Another problem is that such a tool would }
-
-% \todo{Runscope? Others?}
-
-\section{Conclusion and Future Work}
-\label{sec:conclusions}
-
-\todo{update as appropriate}
-
-In this paper we have shown that it is possible to create a monitoring solution which provides basic insight into web service utilization and performance  with very little effort from the developer.
-The user group that we are aiming for with this work is application developers using Flask and Python to build web applications with limited or no budget for implementing their own monitoring solutions.
-The emphasis is in allowing such users to gain insight into how the performance of the service evolves together with the application itself.
-We believe that the same architecture, and lessons can be applied to other frameworks and other languages.
-
-In the future, we plan to perform case studies with other sytstems, with the goal of discover other needs and to wean out the less useful visualizations in the \tool.
-We plan to also extend the tool towards supporting multiple deployments of the same applications across multiple nodes (e.g.
-for the situations where the application is deployed together with a load balancer).
-Finally, we plan to integrate \tool with unit testing as a complementary source of information about performance evolution.
-
-\newpage
-\appendix
-
-  \subsection{Changing the Default /dashboard Endpoint }
-
-  THIS STUFF SHOULD BE MOVED TO THE APPENDIX.
-And in a footnote we just mention that ...
-\ins{REQUIREMENT}: The system must provide an easy way to add more specific configurations if needed.
-A way of overwriting the common sense defaults.
-
-  Further configuration is not needed possible by adding additional statements before the binding definition, for example,
-
-  \begin{lstlisting}[style=custompython]
-  ...
-  dashboard.config.link = 'mydashboard'
-  dashboard.bind(flask_app)
-  \end{lstlisting}
-
-  allows for a custom route (\code{/mydashboard}) to the dashboard to be defined by the programmer. An external configuration file can be used instead, or in addition to this:
-
-  \begin{lstlisting}[style=custompython]
-  ...
-  dashboard.config.from_file('config.cfg')
-  ...
-  \end{lstlisting}
-
-  \subsection{For those who don't use GIT}
-
-    \textit{Manual} version control requires the developer to tag each new version of the application with an appropriate version identifier~\cite{papazoglou2011managing} using the \code{APP\_VERSION} configuration parameter, for example by adding to the configuration file:
-
-      \begin{lstlisting}[style=custompython]
-      [dashboard]
-      APP_VERSION=<versionID>
-      \end{lstlisting}
-
-% references section
-
-\bibliographystyle{abbrv}
-\bibliography{paper}
-
-% that's all folks
-\end{document}

### 2018-04-03 related-work.tex
split the huge file in smaller ones.

--- 2018-03-21 restructured.tex
@@
-\documentclass[conference]{IEEEtran}
-	\pdfpagewidth=8.5truein
-	\pdfpageheight=11truein
-
-\usepackage{graphicx}
-\usepackage{subfig}
-\usepackage[bookmarks=false]{hyperref}
-\usepackage{xspace}
-\usepackage{listings}
-\usepackage[usenames, dvipsnames]{color}
-
-\usepackage{booktabs}
-
-\graphicspath{{./img/}}
-
-%\hyphenation{op-tical net-works semi-conduc-tor}
-
-\input{useful_defines}
-
-\begin{document}
-
-%
-\title{Agile Monitoring of Evolving Web Services}
-
-\author{
-\IEEEauthorblockN{Patrick Vogel, Thijs Klooster, Vasilios Andrikopoulos, Mircea Lungu}\\
-Johann Bernoulli Institute for Mathematics and Computer Science\\
-University of Groningen, Netherland\\
-Email: \{t.klooster.1,p.p.vogel\}@student.rug.nl, \{v.andrikopoulos,m.f.lungu\}@rug.nl 
-}
-
-% make the title area
-\maketitle
-
-\begin{abstract}
-\todo{Update abstract \& title}
-  Tens of thousands of web applications are written in Flask, a Python-based web framework.
-Despite a rich ecosystem of extensions, there is none that supports the developer in gaining insight into the evolving performance of their service.
-In this paper, we introduce \tool, a library that addresses this problem.
-We present the ease with which the library can be integrated in an already existing web application, discuss some of the visualization perspectives that the library provides and point to some future challenges for similar libraries.
-
-\end{abstract}
-
-\section{Introduction}
-
-\todo{Intro to be rewritten; refer to VISSOFT paper as previous work, explain focus is on integration with version control/CI for evolution support, and use of unit tests as early performance indicator}
-
-\todo{motivate with: monitoring in a world which is becoming more and more service-oriented is essential because it allows to perform three main types of actions: system adaptations to provide the request service at the desired level of quality, enabling flexibility in dealing with changing requirements and modes of operation, and enabling operational awareness through dashboards organizing the collected data~\cite{pernici2016monitoring}.
-In this work we focus on the last part and we discuss how a minimal effort probe-based solution for a particular type of web services can be used to facilitate the other two types of actions by involving the developer.}
-
-\todo{explain: in~\cite{vogel2017low} we introduced the \tool an a high level with a focus on presenting its performance visualization aspects; in this work we only summarize these features by means of introducing some of the provided functionalities of the \tool.
-We focus on discussing how the tool is implemented and operating, provide a deeper look at its capabilities for integration with version control and continuous integration environments, introduce a mechanism for performance prediction, and provide an evaluation of the proposed approach based on a case study.
-}
-
-%Every system is a distributed system nowadays \cite{cavage2013there}.
-Indeed a very large number of applications and web applications are nowadays implemented as two-tier architectures with a front-end implemented with web technologies and a service back-end.
-%\ml{I'm not completely happy with this paragraph}
-{\em There is no getting around it: you are building a distributed system} argues a recent article \cite{cavage2013there}.
-Indeed, even the simplest second-year student project is a web application implemented as two-tier architecture with a Javascript/HTML5 front-end a service backend, usually a REST API.
-
-% \hfill mds
-% Many contemporary programming languages are offering libraries, modules, or frameworks that facilitate the development of such architectures. 
-Python is one of the most popular programming language choices for implementing the back-end of web applications.
-GitHub contains more than 500K open source Python projects and the Tiobe Index\footnote{TIOBE programming community index is a measure of popularity of programming languages, created and maintained by the TIOBE Company based in Eindhoven, the Netherlands} ranks Python as the 4th most popular programming language as of June 2016.
-
-% possible flask summary
-Within the Python community, Flask\footnote{\url{http://flask.pocoo.org/}} is a very popular web framework\footnote{More than 25K projects on GitHub (5\% of all Python projects) are implemented with Flask (cf.
-a GitHub search for ``language:Python Flask'')}.
-It provides simplicity and flexibility by implementing a bare-minimum web server, and thus advertises as a micro-framework.
-The Flask tutorial shows how setting up a simple Flask {\em ``Hello World''} web-service requires no more than 5 lines of Python code \cite{ flask:tutorial}.
-% end of summary
-
-Despite their popularity, to the best of our knowledge, there is no simple solution for monitoring the evolving performance of Flask web applications.
-Thus, every one of the developers of these projects faces one of the following options when confronted with the need of gathering insight into the runtime behavior of their implemented services:
-
-  \begin{enumerate}
-
-    \item Use a commercial monitoring tool which treats the subject API as a black-box (e.g. Pingdom, Runscope). 
-    % , Graphite+Graphana+statd etc.
-
-    \item Implement their own ad-hoc analytics solution, having to reinvent basic visualization and interaction strategies.
-
-    \item Live without analytics insight into their services.
-
-  \end{enumerate}
-
-%\todo{For the first point in the list, we can also argue that analytics solutions like Google Analytics can be used, but they have no notion of versioning/integration with the development life cycle.
-Feel free to cite \cite{papazoglou2011managing} for service evolution purposes}
-
-For projects on a budget (e.g.
-research, startups) the first and the second options are often not available due to time and financial constraints.
-Even when using 3rd-party analytics solutions, a critical insight into the evolution of the exposed services of the web application, is missing because such solutions have no notion of versioning and no integration with the development life cycle.~\cite{papazoglou2011managing}
-
-To avoid projects ending up in the third situation, that of living without analytics, in this paper we present \tool~ --- a low-effort service monitoring library for Flask-based Python web services that is easy to integrate and enables the {\em agile assessment} of service evolution.
-\cite{Nier12b}
-
-As a case study, on which we will illustrate our solution, we are going to use an open source API which, for several years, was in the third of the above-presented situations.
-
-% In the next section, we will present a case study of an open source research API which was for a long time in the third situation presented above -- deployed without analytics insight.
-
-\todo{Add signposting paragraph}
-
-\section{Case Study: The API (Zeeguu)}
-\label{sec:case}
-
-%\todo{Either leave as is or add more info on usage by college if possible}
-
-  \zee\footnote{\url{https://github.com/zeeguu-ecosystem/}} is a platform and an ecosystem of applications for accelerating vocabulary acquisition in a foreign language \cite{Lungu16}. 
-%
-  The architecture of the ecosystem has at its core an API and a series of satellite applications that together offer to a learner three main inter-dependent features:
-
-  \begin{enumerate}
-
-    \item Reader applications that provide effortless translations for those texts which are too difficult for the readers.
-
-    \item Interactive exercises personally generated based on the preferences and past likes of the learner.
-
-    \item Article recommendations which are personalized for the interests of the learner and come with difficulty estimation that helps the learner find articles with the appropriate difficulty.
-
-  \end{enumerate}
-
-  The core API implemented with Flask and Python provides correspondingly three types of functionality: contextual translations, article recommendations, and personalized exercise suggestions.
-In total the API provides a bit less than 50 endpoints, out of which probably a dozen are very frequently used.
-  % The development of the core API itself is a research project.
-
-  At the time of writing this article, the ecosystem consists of a reader web application, a web based exercises platform, and a smartwatch application, which are used at the moment of writing this article by more than \activeUserCount active users.
-The users come from a highschool, a language school, and some users are using it on their own, without any educational context.
-The highest load we observed until now on the API consisted of 12K requests in one day.
-
-  We will use this \zee API as a case study for this paper. 
-  All the figures in this paper are captured from the actual deployment of \tool in the context of the \zee API.
-The figures are interactive offering basic data exploration capabilities: filter, zoom, and details on demand\cite{Shne99a}.
-The \tool deployment for the case study can be accessed publicly\footnote{\url{https://zeeguu.unibe.ch/api/dashboard}; username: {\em guest}, password: {\em soap-sac}}.
-
- % \todo{create new account, update URL if necessary, update activeUserCount, add a couple of sentences of the current state of the case study with the college/language center}
-% \ml{we should consider adding also one section in which the architecture/implementation and main features of the dashboard are presented before going on with discussing them in more depth in the following sections --- this should include a rundown on which views are provided from where (overview or per endpoint)}
-
-\newpage
-
-\section{Case Study: The Technology (Flask)}
-\label{sec:tool}
-
- Flask is a microframework for Python. It IS USED BY THOUSANDS OF PROJECTS. ELABORATE MORE ON THIS... 
- \vspace{2cm}
-
- Flask depends on two external libraries: the Jinja2 template engine and the Werkzeug WSGI toolkit.
-
- “Micro” does not mean that your whole web application has to fit into a single Python file (although it certainly can), nor does it mean that Flask is lacking in functionality.
-The “micro” in microframework means Flask aims to keep the core simple but extensible.
-Flask won’t make many decisions for you, such as what database to use.
-Those decisions that it does make, such as what templating engine to use, are easy to change.
-Everything else is up to you, so that Flask can be everything you need and nothing you don’t.
-
- Flask has many configuration values, with sensible defaults, and a few conventions when getting started.
-By convention, templates and static files are stored in subdirectories within the application’s Python source tree, with the names templates and static respectively.
-While this can be changed, you usually don’t have to, especially when getting started.
-
- A Python web application based on WSGI has to have one central callable object that implements the actual application.
-In Flask this is an instance of the Flask class.
-Each Flask application has to create an instance of this class itself and pass it the name of the module, but why can’t Flask do that itself?
-
-\begin{lstlisting}[style=custompython]
-
-from flask import Flask
-app = Flask(__name__)
-
-@app.route('/')
-def index():
-    return 'Hello World!'
-
-\end{lstlisting}
-
-\subsection*{NOTHING PREVENTS GENERALIZATION}
-In this article we take as an example the Flask case study.
-But there is nothing that prevents the generalization of 
-our approach.
-
-MAKE SURE THAT THIS IS CLEAR.
-
-    \begin{figure*}[ht!]
-      \centering
-      \includegraphics[width=0.8\linewidth]{interceptor}
-      \caption{The first thing that needs to be done, is to decorate the application object with an INTERCEPTOR}
-      \label{fig:sep}
-    \end{figure*}
-
-\newpage
-
-\section{Monitoring Service Utilization}
-
-REMEMBER THE GOAL: Our main goal with the \tool issues to allow the integration of the dashboard with an API with minimal effort.
-
-\ins{The first important question that such a 
-service monitoring system should provide is 
-information about what endpoints are being
-used and by whom}.
-
-TODO: Find references, arguments, related work
-that provide more details about why understanding
-who uses endpoints, and which are used are important.
-
-This is clearly the case in static analysis (see paper
-by haenni and lungu...) but should even more be the
-case in APIs.
-
-  \newpage
-TODO: TALK ABOUT THE DATABASE...
-
-  As discussed in the introductory section, in previous work~\cite{vogel2017low} we introduced \tool, a drop-in Python library that allows developers to monitor their Flask-based Python web applications with minimal effort.
-%
-  The \tool is implemented for Python 3.6 and is available on the Python Package Index repository\footnote{\url{https://pypi.python.org/pypi/flask-monitoring-dashboard/1.8}} from where it can be installed by running \install from the command line.
-%  
-  The source code of the \tool is published under a permissive MIT license and is available on GitHub\footnote{\url{https://github.com/flask-dashboard}}.
-
-  \newpage
-  \section{One Line Of Code Configuration With Technology Specific Interceptors}
-
-  The fact that the \tool itself is being developed in Python using Flask makes binding to the services of a to-be monitored application developed with the same technologies easy and intuitive.
-
-  \ins{In general, what matters is that with the right backend, we can do this minimal configuration for any technology... }
-
-  To start using the \tool, one simply needs to one lines of code\footnote{we don't count the import statement that enables that line...} to their Flask web service\footnote{\ins{ In this paper we present the integration with APIs written in Python and Flask hoping that this will not prevent the reader from seeing the more general idea; all the tools we show here for Flask can be applied to other API technologies (e.g.
-Django) by simply providing a few back-end adapters in the right places.
-}}:
-
-  % caption=Configuring the \tool is straightforward,
-  \begin{lstlisting}[style=custompython]
-  import dashboard
-
-  # flask_app is the Flask app object
-  dashboard.bind(flask_app)
-  \end{lstlisting}
-
-  \ins{The dashboard can take advantage of the fact that it is monitoring an API...
-so it can simply just add one extra endpoint there.} After (re)deployment of the application, the interactive visualizations of the \tool become available at the \code{/dashboard} route of the Flask application.
-
-  During binding, the \tool will search for all endpoints defined in the target application.
-
-  In order to monitor an endpoint, the \tool creates a function wrapper for the API function that corresponds to the endpoint.
-This way, the wrapper will be executed whenever that API call is made before the actual function is called.
-The wrapper contains the code that takes care of monitoring an endpoint.
-Data collected by the wrappers are persisted in a local database.
-The SQLAlchemy Object Relational Mapper\footnote{\url{https://www.sqlalchemy.org/}} on top of SQLite\footnote{\url{https://www.sqlite.org/}} is used for this purpose.
-  \ins{DISCUSSION about how do to this in other languages? Java: AspectJ probably, etc.}
-
-  These will be presented to the user in the tool web interface, where the user can select the ones that should be monitored, see \Fref{fig:sep}.
-
-    \begin{figure}[h!]
-      \centering
-      \includegraphics[width=\linewidth]{selecting_endpoints.png}
-      \caption{Once connected to an API the Dashboard presents the endpoints that are available for monitoring}
-      \label{fig:sep}
-    \end{figure}
-
-  This is the alternative to using annotations like other white-box tools (e.g. the other flask performance plugin).
-
-  \ins{PROMISE:} The last access time of the endpoint by users, irrespective of whether it is selected by the user to be monitored or not, is also displayed as the means of identifying potentially obsolete endpoints.
-
-  \ml{we must insert at least one dashboard view here...
-otherwise people will lose their patience...
-also it will help with explaining the outliers...
-also, one should see that a large variety of views are already available at this point...
-I know this risks being  }
-
-In the Zeeguu case study, one of the slowest endpoints, and one with the highest variability as shown in \Fref{fig:ep} is \epFeedItems: it retrieves a list of recommended articles for a given user.
-However, since a user can be subscribed to anything from one to three dozen article sources, and since the computation of the difficulty is personalized and it is slow, the variability in time among users is likely to be very large.
-
-\begin{figure*}[h!]
-  \centering
-  \subfloat[number of requests per endpoint per day]{\includegraphics[width=.9\columnwidth]{number_of_requests_}\label{fig:aeu}}
-  \subfloat[requests per hour heatmap]{\includegraphics[width=\columnwidth]{daily_patterns_}\label{fig:dp}}
-  \caption{Some of the available views\label{fig:views}}
-\end{figure*}
-
-\newpage
-\section{Service Utilization}
-\label{sec:util}
-
-\ins{A discussion must be added here about the types of useful information and problems that can be solved with these views.}
-
-\textit{Service Utilization} presents information about the usage of all endpoints of interest to the developer,
-
-The most fundamental insight that a service maintainer needs regards service utilization. %\vspace{0.5cm}
-
-Simplest utilization perspective:
-
-\begin{figure}[h!]
-  \centering
-  \includegraphics[width=0.5\linewidth]{endpoint-call-count}
-  \caption{Caption here}
-  \label{fig:figure1}
-\end{figure}
-
-% REQUIREMENT: Adding a time interval is important. 
-% Otherwise the performance numbers in chart 1 are not good enough... also an outlier very far in the past, will make all the heatmap data points to look very pale in comparison forever after.
-
-Figure \ref{fig:aeu} shows a first perspective on endpoint utilization that \tool provides: a stacked bar chart of the number of hits to various endpoints grouped by day\footnote{Endpoint colors are the same in different views}.
-Figure~\ref{fig:aeu} in particular shows that at its peak the API has about 12.000 hits per day.
-
-The way users interact with the platform can also be inferred since the endpoints are indicators of different activity types, e.g.:
-
-\begin{enumerate}
-
-  \item {\color{myred} \epTranslations} is an indicator of the amount of foreign language reading the users are doing, and
-
-  \item {\color{mygreen} \epOutcome} is an indicator of the amount of foreign vocabulary practice the users are doing.
-
-\end{enumerate}
-
-%\begin{figure}[!ht]
-% \centering
-% \includegraphics[width=\linewidth]{number_of_requests_}
-% \caption{The number of requests per endpoint per day view shows the overall utilization of the monitored application}
-% \label{fig:aeu}
-%\end{figure}
-
-Besides showing the overall utilization, this endpoint provides the maintainer with information relevant for decisions regarding endpoint deprecation --- one of the most elementary ways of {\em understanding the needs of the downstream}\cite{Haen14a}.
-In our case study, the maintainer realized that one endpoint which they thought was not being used (i.e.
-\code{words\_to\_study}), contrary to their expectations, was actually being used\footnote{A complementary type of usage information can also be discovered in the view presented in Figure \ref{fig:sep} where seeing that an endpoint is never accessed can increase the confidence of the maintainer that a given endpoint is not used, although it can never be used a proof.}.
-
-\niceseparator
-
-%   \todo{Add the time series graph and discuss it before the heatmap? We can then sell the heatmap better} 
-%   \ml{Not sure about which graph you refer to here V}
-
-A second type of utilization question that the \tool can answer automatically regards {\em cyclic patterns of usage per hour of day} by means of a heatmap, as in \Fref{fig:dp}.
-
-% \mltp{ can we add vertical lines that highlight the beginning of a new week (e.g. before Sunday): }
-% Patrick: adding separators in the graph is unfortunately not supported by the library.
-
-%\begin{figure}[!ht]
-% \centering
-% \includegraphics[width=0.9\linewidth]{daily_patterns_}
-% \caption{Usage patterns become easy to spot in the requests per hour heatmap}
-% \label{fig:dp}
-%\end{figure}
-
-Figure \ref{fig:dp} shows the API not being used during the early morning hours, with most of the activity focused around working hours and some light activity during the evening.
-This is consistent with the fact that the current users are all in the central European timezone.
-Also, the figure shows that the spike in utilization that was visible also in the previous graph happended in on afternoon/evening.
-
-\newpage
-\section{Endpoint Performance}
-\label{sec:perf}
-
-  The \tool also collects information regarding endpoint performance. The view in \Fref{fig:ep} summarizes the response times for various endpoints by using a box-and-whiskers plot.
-
- \begin{figure}[!ht]
-   \centering
-   \includegraphics[width=0.95\linewidth]{endpoint_performance_}
-   \caption{The response time (in ms) per monitored endpoint view allows for identifying performance variability and balancing issues}
-   \label{fig:ep}
- \end{figure}
-
-  From this view it became clear to the maintainer that four of the endpoints had very large variation in performance.
-The most critical for the application and consequently the one optimized first was the \epTranslations endpoint which was part of an interactive loop in the reader applications that relied on the Zeeguu API.
-Moreover, cf.
-\Fref{fig:aeu} this endpoint is one of the most used in the system.
-
-\begin{figure}[h!]
-  \centering
-  \subfloat[overview of the tracked endpoints]{\includegraphics[width=0.8\columnwidth]{overview}\label{fig:overview}}
-  \caption{Some of the available views\label{fig:views}}
-\end{figure}
-
-  \newpage
-  \section{Evolution monitoring }
-
-  \begin{figure}[h!]
-  \centering
-  \includegraphics[width=0.8\linewidth]{translation_endpoint_evolution_}
-  \caption{The Performance Evolution of the \epTranslations endpoint}
-  \label{fig:tee}
-\end{figure}
-
-  When thinking about performance, 
-  one can not avoid thinking about the performance evolution.
-
-  Version control in \tool can be supported in two ways,
-  the developer explicitly states the current version, 
-  or the current version is automatically detected based
-  on some version control system.
-
-  However, with the current configuration of the tool, it would be impossible for the maintainer to see the improvements resulting from the optimization.
-
-  \ins{The latest STack Overflow Developer Survey: 88.4\% of **professional** developers use git.}  
-  If assume that the code that they run is deployed using .git, then with an extra line of configuration they can allow \tool to find the git\footnote{\url{https://git-scm.com/}} folder of the deployed service and automatically detect the version of the project that is running:
-
-    \begin{lstlisting}[style=custompython]
-
-      dashboard.config.git = 'path/to/.git'
-
-    \end{lstlisting}
-
-  If the \tool can automatically detect the current version of the project by reading the .git configuration as soon as the API is started and can then group measurements by version\footnote{Alternatively, the maintainer can add version identifiers manually for the web application through a configuration file if the system does not use git.}.
-\Fref{fig:tee} is a zoomed-in version of such a view for \epTranslations with versions increasing from top to bottom
-
-  This view confirms that the performance of the translation endpoint improved in the recent versions: the median of the last three versions is constantly moving towards the left, and progresses from 1.4 seconds (in the top-most box plot in \Fref{fig:tee}) to 0.8 in the latest version (bottom-most box plot).
-
-  CHALLENGE: Supporting other types of version control... not real challenge, as we said 88percent of devs use git. 
-  CHALLENGE 2: detecting minor versions which don't need to be tracked, since they didn't touch the performance of the system. E.g. a modification of the README, etc.
-
-\todo{TODO: endpoint evolution }
-
-  \begin{figure*}[h!]
-    \centering
-    \includegraphics[width=\linewidth]{outliers}
-    \caption{Automatically collected outlier information}
-    \label{fig:figure1}
-  \end{figure*}
-  \newpage
-  \section{Automated outlier detection and monitoring}
-
-  When an API is called from within a highly interactive application (as it is the case with the case study in this paper) 
-  of particular interest to the API developers are performance {\em outliers}.
-
-  Indeed, a translation request that takes three times more than expected can seriously decrease the perceived quality of the application.
-Thus, identifying, collecting all appropriate data, and diagnosing the root causes of such outliers is especially critical in improving the quality of an application.
-
-  % In the context of the RESTful services support by Flask, this means request serving times that deviate from the average to an unexpected degree.
-
-  For this purpose the \tool tracks for every monitored endpoint a {\em running average} response time value\footnote{\ins{For performance reasons, we assume that the response times for the endpoints are normally distributed.
-Otherwise, more general density distribution information must be collected in real time.}}.
-When it detects that a given request is an outlier with respect to this past average running value, it triggers the {\em outlier data collection routine} which stores \ins{extra information} about the current execution environment.
-A configurable threshold with a default value of $2.5$ times the running average response time is used for this purpose.
-
-  For every detected outlier request, the \tool collects information about the current Python stack trace, CPU load, memory consumption, request parameters, etc.
-in order to allow the maintainer to investigate the causes of these exceptionally slow response times.
-In this way it is possible to get \ins{detailed insight into the operation of the application in the extreme cases without unnecessarily burdening it with logging this information for every request}.
-
-The bottom panel shows the stack trace. 
-In this particular case, it is revealing for the developer to learn that at the time of the stack trace snapshot, the code was in the microsoft\_translator.py: indeed, the system uses as back-end multiple translators, and it has been observed that many of the outliers happen to be waiting in the microsoft translator.
-
-This information has to be corroborated with the observations that neither the memory nor the processor are overloaded at the moment.
-Thus this functionality in microsoft\_translator is really slow in itself, and this is not a result of the machine being overloaded for example.
-
-\subsection {A Special Type of Outlier: Exceptions}
-
-   TODO: probably something for the journal extension ...
-  statistics about which endpoints fail most often might be useful too.
-  same information as the outlier maybe?
-
-\begin{figure*}[!ht]
-  \centering
-  \includegraphics[width=0.4\linewidth]{time_per_user}
-  \caption{The \epFeedItems shows a very high variability across users}
-  \label{fig:tpu}
-\end{figure*}
-
-\newpage
-\section{Grouping Requests}
-\label{sec:user}
-
-For service endpoints which run computations in real time, the maintainer of a system might want to understand the endpoint performance on a per-user basis, especially for situations where the system response time is a function of some individual user load\footnote{E.g.
-in GMail some users have two emails while other have twenty thousand and this induces different response times for different users}.
-
-\todo{remove the configuration part and rephrase the following, the discussion has been partially subsumed by Section~\ref{sec:tool}.}
-
-To enable this, the \tool must be configured to associate an API call with a given user.
-The simplest way is to take advantage of the architecture of Flask applications in which a global \code{flask.request} object can be used to retrieve the session which can in turn lead to user identification:
-
-%\begin{lstlisting}[float,caption=Simply define a custom app-specific function for user retrieval and pass it to the \tool to group information by user,style=custompython]
-\begin{lstlisting}[style=custompython]  
-# app specific way of extracting the user
-# from a flask request object    
-def get_user_id():
-sid = int(flask.request.args['session'])
-session = User.find_for_session(sid)
-return user_id
-
-# attaching the get_user_id function
-dashboard.config.get_group_by = get_user_id
-
-\end{lstlisting}
-
-% \niceseparator
-
-\todo{consider losing one or both of the figures --- or at least updating them}
-
-In Zeeguu, \epFeedItems retrieves a list of recommended articles for a given user.
-Cf.
-\Fref{fig:ep} it is the endpoint with the slowest response time and highest variability.
-The reason for this is that a user can be subscribed to anything from one to three dozen article sources and for each of the sources the system must compute the personalized difficulty of each article at every request.
-
-A \perspective{Per-User Performance} perspective should show the different response times for different users.
-Figure \ref{fig:tpu} presents a subset of the corresponding view in the \tool.
-The figure shows that the response times for this endpoint can vary considerably for different users with some extreme cases where a user has to wait a full minute until their recommended articles are shown\footnote{After seeing this perspective, the maintainer refactored the architecture of the system to move part the difficulty computation out of the interactive loop}.
-
-The limitation of the previous view is that it does not present the information also on a per version basis.
-To address this, a different visual perspective entitled \perspective{Multi-Version per-User Performance} can be defined.
-Figure \ref{fig:tuv} presents such a perspective by mapping the average execution time for a given user (lines) and given version (columns) on the area of the corresponding circle.
-
-The colors represent users.
-The figure shows average performance varying  across users and versions with no clear trend: this is probably because varying user workload (i.e.
-number of sources to which the user is registered) is the reason for the variation in response times.
-\ins{one can see that for user 1 performance degrades over the }
-
-\begin{figure}[!ht]
-  \centering
-  \includegraphics[width=0.7\linewidth]{time_per_user_per_version}
-  \caption{This perspective shows that the evolution of response times for individual users (horizontal lines) across versions (the x-axis)}
-  \label{fig:tuv}
-\end{figure}
-
-  \newpage
-  \section{Preemptive monitoring}
-
-  \ins{INTRO - LINKING WITH PREVIOUS PARAGRAPH: the previous evolutionary view crucial for 
-  understanding the impact of software evolution on performance.
-  However, just by observing this one can't be sure whether 
-  performance changes because of the code or because of the payload.}
-
-  \todo{the following might need a bit of rewrite}
-
-  The concept of {\em preemptive monitoring} of the application performance by means of instrumenting integration unit testing as the synthetic load is similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}, i.e.~testing service-based applications by using dedicated test input in parallel to its normal use and operation.
-The difference is that we take advantage of the capability of the CI framework (i.e.
-Travis in this case) to create an emulated ``live'' environment for integration testing purposes, and use unit testing as the dedicated test input in order to measure performance.
-
-  For example, \Fref{fig:builds} is a screenshot from the dashboard showing the measured response times for 5 consequent builds, with 180 iterations of the unit tests executed in total across all endpoints.
-The outliers on the right of the figure are due to initial requests \ins{which must wait for a boot-up phase of the API}.
-
-    \begin{figure}[h!]
-        \centering
-        \includegraphics[width=\columnwidth]{travis_builds}
-        \caption{Response times in 5 consequent Travis builds in the \zee case study}
-        \label{fig:builds}
-      \end{figure}
-
-  While this integration testing environment is different from the actual production deployment one, and the load used is purely synthetic, as will show in Section~\ref{sec:evaluation}, it can actually be used successfully as an early performance indicator for the application developer.
-\ins{Although predicting the performance of the application would require advanced statistical work, it can be succesfully used as an advanced warning system for the overall performance trend for given endpoints}.
-
-  \paragraph{Integration with Travis}
-  \ml{added new section... also, this has to be moved after introducing the pre-emptive monitoring since it's about running tests. switched the order}
-
-  \ins{The pre-emptive monitoring system} can be configured to work together with Continuous Integration (CI) frameworks like Travis\footnote{\url{https://travis-ci.org/}} that deal with automated integration testing.
-  The developer needs to define the unit test folder of the project, the URL that Travis should use for reporting its results, and the number of iterations for each unit test should be executed in order to generate a load for the application.
-
-  Each time a new build is created through Travis, the \tool automatically detects all available unit tests defined by the application developer and iterates through each one of them the number of times predefined by the developer while monitoring the response times for each request.
-The resulting measurements are persisted in a separate part of the tool database so as not to contaminate the ``live'' monitoring data.
-Beyond seeing the results of each Travis build in the \tool, this feature also allows for preemptive monitoring of the application, as we discuss in the following.
-
-  \ins{Besides functioning as an early warning system for performance, tracking the evolving performance of API tests serves a second purpose.
-To function as an anchor when the developer analyzes performance degradations.
-For example, when a developer sees that a given endpoint has become less performant after in a newly deployed version, they are able to investigate whether the performance degradation is visible also with the synthetic load.
-This would correspond to a performance degradation which is due to ``algorithmic performance degradation''.
-If on the other hand, the performance of the tests does not change between versions, the developer might conclude that the performance degradation could be due to the workload on the machines, or maybe to the workload mix of the users}
-
-  %By these means, an application developer can use unit tests as synthetic loads for driving the performance monitoring of the application before it reaches the production phase.
-As we will discuss further in Section~\ref{sec:testing}, this feature is integral in enabling preemptive performance monitoring of the application.
-
-\todo{here goes the discussion about our forecasting capabilities}
-
-\begin{figure}[h!]
-  \centering
-  \subfloat[The reported response time (in ms) per deployed version in the observation period]{\includegraphics[width=.5\columnwidth]{response_per_version_trunced_trend}}
-  \quad
-  \subfloat[The measured response time (in ms) using integration with Travis]{\includegraphics[width=.46\columnwidth]{travis_builds_no_outliers_trend}}
-  \caption{Comparison of the response times per endpoint: actual production system versus preemptive monitoring data}
-  \label{fig:preemptive}
-\end{figure}
-
-\begin{table}[h]
-  \begin{tabular}{lll}
-    \toprule
-    Iteration & \bfseries Live (median) & \bfseries Travis (median)\\
-    \midrule
-    1 & 1349.41 & 764.87\\ 
-    2 & 1466.13 & 992.87\\
-    3 & 1256.65 & 760.87\\
-    4 & 1266.42 & 813.89\\
-    5 & 1080.68 & 303.4\\
-    \bottomrule
-  \end{tabular}
-\end{table}
-
-Pearson correlation $r(3)=.93, p=.02$
-
-\section{Performance Perspectives}
-\label{sec:views}
-
-\ins{PERFORMANCE DISCUSSION: this introspection, which looks up endpoints, happens only once at the startup of the service, so it is going to affect the API startup performance, but not the actual endpoint performance.
-In our case, the overhead here was quite small: TO MEASURE}.
-
-  %In the remainder of the paper we present several of these perspectives.\footnote{We recommend obtaining a color version of this paper for better readability}
-
-  % The second endpoint conists of two parts, one of them being a table that shows for every monitored endpoint the number of hits it has gotten, the time it was last accessed and its average execution time.
-The second part is a view with four graphs which show:
-  % \begin{itemize} 
-  %   \item A heatmap of the total number of requests to the monitored endpoints
-  %   \item A stacked bar chart that shows the total number of requests to the monitored endpoints per endpoint per day
-  %   \item A boxplot graph showing the average execution time per version of the web service
-  %   \item A boxplot graph showing the average execution time for every monitored endpoint
-  % \end{itemize}
-
-\todo{evaluate which figures to keep and ideally take new screenshots for the ones we keep; consider adding two more views (like Fig. 3.2 of Patrick's thesis)}
-
-% In the Zeeguu case study, one of the slowest endpoints, and one with the highest variability as shown in \Fref{fig:ep} is \epFeedItems: it retrieves a list of recommended articles for a given user.
-However, since a user can be subscribed to anything from one to three dozen article sources, and since the computation of the difficulty is personalized and it is slow, the variability in time among users is likely to be very large.
-
-% \niceseparator
-
-%\section{Preemptive Monitoring}
-%\label{sec:testing}
-%
-%%\todo{discuss here idea of using unit testing as an early performance indicator for evolving services (`preemptive monitoring'?); add a side-by-side figure comparing unit testing vs live system performance per version, maybe calculate error in trend curves?
-find a better term than prediction for the section title (plus introduction)}
-%
-%\todo{Similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}, i.e.~testing service-based applications by using dedicated test input in parallel to its normal use and operation.
-The difference is that we take advantage of the capability of Travis to create an emulated live environment for integration testing purposes, and use unit testing as the dedicated test input in order to measure performance.
-What we are going provide evidence for in the following is that we can indeed use this preemptive monitoring concept as an early performance indicator for evolving services.}
-%
-%The developer needs to define the unit test folder of the project, the URL that Travis should use for reporting its results, and the number of iterations for each unit test should be executed in order to generate
-
-\section{Evaluation}
-\label{sec:evaluation}
-
-%\todo{evaluation of the \tool across the following dimensions: ease of use, effectiveness, overhead to the system; for the last part we need to do the overhead measurement experiment, for the other two we base it on the case study and bring over whatever is necessary from the discussion section}
-
-\section{Discussion}
-
-\todo{use this subsection to discuss limitations (from the previous evaluation section) and add the need to be able to handle the horizontal scaling of the application; remove sectioning and summarize briefly limitations that have already been discussed in the VISSOFT paper/move material to other sections; consider renaming as limitations or something similar}
-
-  \subsection{Distributed System Monitoring}
-  \todo{To think about}
-  Supporting situations where the API is deployed across multiple containers for example.
-
-  \subsubsection{Automatically Monitoring System Evolution}
-
-  The main goal of the \tool design was to allow analytics to be collected and insight to be gained by making the smallest possible changes to a running API.
-%To allow the collection of evolutionary information
-%
-  This technique assumes that the web application code which is the target of the monitoring is deployed using \git in the following way:
-
-  \begin{enumerate}
-    \item The deployment engineer pulls the latest version of the code from the integration server; this will result in a new commit being pointed at by the HEAD pointer. %than previously
-    \item The deployment engineer restarts the new version of the service.
-At this point, the \tool detects that a new HEAD is present in the local code base and consequently starts associating all the new data points with this new commit\footnote{The \tool detects the current version of the analyzed system the first time it is attached to the application object, and thus, assumes that the Flask application is restarted when a new version is deployed.
-This is in tune with the current version of Flask, but if the web server will support dynamic updates in the future, this might have to be taken into account}.
-  \end{enumerate}
-
-  The advantage of this approach is the need for minimal configuration effort, as discussed in the presentation of the tool.
-The disadvantage is that it will consider on equal ground the smallest of commits, even one that modifies a comment, and the shortest lived of commits, e.g.~a commit which was active only for a half an hour before a new version with a bug fix was deployed, with major and minor releases of the software.
-%as a distinct way of grouping the data points.
-  A mechanism to control which versions are important for monitoring purposes is therefore required to be added.
-%
-  A further possible extension point here is supporting other version control systems (e.g. Mercurial). However, this is a straightforward extension.
-
-  \subsubsection{NEW: API Renames}
-  \ml{just an idea}
-
-    The history would normally be lost in the case of an API rename. (also known as the {\em provenance issue} in source code evolution analysis)
-    In theory we could infer that an API was renamed if it is not to be found anymore, and another one with very similar timing characteristics would be found.
-Such a detection can never be completely sure, but it would help with meeting the needs of the API maintainer...
-
-  \subsubsection{User-Awareness }
-
-    For the situations in which the user information is not available, the \tool tracks by default information about different IPs and in some cases this might be a sufficiently good approximation of the user diversity and identity.
-    %
-
-    The visualizations for the user experiene perspectives as presented in Section \ref{sec:user} have been tested with several hundred users (of which about \activeUserCount were active during the course of the study), but the scalability of the visualizations must be further investigated for web services with tens of thousands of users.
-
-  \subsubsection{Other Possible Groupings}
-
-    There are other groupings of service utilization and performance that could be important to the maintainer, that we did not explore in this paper.
-For example, if the service is using OAuth, then together with every request, in the header of the request there is information about the application which is sending a request.
-Grouping the information by application that sends the request could be important in such a context.
-
-    In general, providing a mechanism that would allow very easy specification of groupings (either as code annotations, as normal code, or as configuration options) is an open problem that \tool and any other similar library will have to face.
+%!TEX root=restructured.tex
 
@@
 % \todo{Runscope? Others?}
-
-\section{Conclusion and Future Work}
-\label{sec:conclusions}
-
-\todo{update as appropriate}
-
-In this paper we have shown that it is possible to create a monitoring solution which provides basic insight into web service utilization and performance  with very little effort from the developer.
-The user group that we are aiming for with this work is application developers using Flask and Python to build web applications with limited or no budget for implementing their own monitoring solutions.
-The emphasis is in allowing such users to gain insight into how the performance of the service evolves together with the application itself.
-We believe that the same architecture, and lessons can be applied to other frameworks and other languages.
-
-In the future, we plan to perform case studies with other sytstems, with the goal of discover other needs and to wean out the less useful visualizations in the \tool.
-We plan to also extend the tool towards supporting multiple deployments of the same applications across multiple nodes (e.g.
-for the situations where the application is deployed together with a load balancer).
-Finally, we plan to integrate \tool with unit testing as a complementary source of information about performance evolution.
-
-\newpage
-\appendix
-
-  \subsection{Changing the Default /dashboard Endpoint }
-
-  THIS STUFF SHOULD BE MOVED TO THE APPENDIX.
-And in a footnote we just mention that ...
-\ins{REQUIREMENT}: The system must provide an easy way to add more specific configurations if needed.
-A way of overwriting the common sense defaults.
-
-  Further configuration is not needed possible by adding additional statements before the binding definition, for example,
-
-  \begin{lstlisting}[style=custompython]
-  ...
-  dashboard.config.link = 'mydashboard'
-  dashboard.bind(flask_app)
-  \end{lstlisting}
-
-  allows for a custom route (\code{/mydashboard}) to the dashboard to be defined by the programmer. An external configuration file can be used instead, or in addition to this:
-
-  \begin{lstlisting}[style=custompython]
-  ...
-  dashboard.config.from_file('config.cfg')
-  ...
-  \end{lstlisting}
-
-  \subsection{For those who don't use GIT}
-
-    \textit{Manual} version control requires the developer to tag each new version of the application with an appropriate version identifier~\cite{papazoglou2011managing} using the \code{APP\_VERSION} configuration parameter, for example by adding to the configuration file:
-
-      \begin{lstlisting}[style=custompython]
-      [dashboard]
-      APP_VERSION=<versionID>
-      \end{lstlisting}
-
-% references section
-
-\bibliographystyle{abbrv}
-\bibliography{paper}
-
-% that's all folks
-\end{document}

### 2018-04-03 restructured.tex
split the huge file in smaller ones.

--- 2018-03-21 restructured.tex
@@
 \author{
-\IEEEauthorblockN{Patrick Vogel, Thijs Klooster, Vasilios Andrikopoulos, Mircea Lungu}\\
-Johann Bernoulli Institute for Mathematics and Computer Science\\
-University of Groningen, Netherland\\
-Email: \{t.klooster.1,p.p.vogel\}@student.rug.nl, \{v.andrikopoulos,m.f.lungu\}@rug.nl 
+\IEEEauthorblockN{The Author, The Other Author}\\
+Institute\\
+University, Country\\
+Email: \{author\}@university.domain 
 }
@@
 
-\begin{abstract}
-\todo{Update abstract \& title}
-  Tens of thousands of web applications are written in Flask, a Python-based web framework.
-Despite a rich ecosystem of extensions, there is none that supports the developer in gaining insight into the evolving performance of their service.
-In this paper, we introduce \tool, a library that addresses this problem.
-We present the ease with which the library can be integrated in an already existing web application, discuss some of the visualization perspectives that the library provides and point to some future challenges for similar libraries.
+\input{introductory}
 
-\end{abstract}
+\input{loc1}
 
-\section{Introduction}
+\input{utilization}
 
-\todo{Intro to be rewritten; refer to VISSOFT paper as previous work, explain focus is on integration with version control/CI for evolution support, and use of unit tests as early performance indicator}
+\input{performance}
 
-\todo{motivate with: monitoring in a world which is becoming more and more service-oriented is essential because it allows to perform three main types of actions: system adaptations to provide the request service at the desired level of quality, enabling flexibility in dealing with changing requirements and modes of operation, and enabling operational awareness through dashboards organizing the collected data~\cite{pernici2016monitoring}.
-In this work we focus on the last part and we discuss how a minimal effort probe-based solution for a particular type of web services can be used to facilitate the other two types of actions by involving the developer.}
+\input{grouping}
 
-\todo{explain: in~\cite{vogel2017low} we introduced the \tool an a high level with a focus on presenting its performance visualization aspects; in this work we only summarize these features by means of introducing some of the provided functionalities of the \tool.
-We focus on discussing how the tool is implemented and operating, provide a deeper look at its capabilities for integration with version control and continuous integration environments, introduce a mechanism for performance prediction, and provide an evaluation of the proposed approach based on a case study.
-}
+\input{evolution}
 
-%Every system is a distributed system nowadays \cite{cavage2013there}.
-Indeed a very large number of applications and web applications are nowadays implemented as two-tier architectures with a front-end implemented with web technologies and a service back-end.
-%\ml{I'm not completely happy with this paragraph}
-{\em There is no getting around it: you are building a distributed system} argues a recent article \cite{cavage2013there}.
-Indeed, even the simplest second-year student project is a web application implemented as two-tier architecture with a Javascript/HTML5 front-end a service backend, usually a REST API.
-
-% \hfill mds
-% Many contemporary programming languages are offering libraries, modules, or frameworks that facilitate the development of such architectures. 
-Python is one of the most popular programming language choices for implementing the back-end of web applications.
-GitHub contains more than 500K open source Python projects and the Tiobe Index\footnote{TIOBE programming community index is a measure of popularity of programming languages, created and maintained by the TIOBE Company based in Eindhoven, the Netherlands} ranks Python as the 4th most popular programming language as of June 2016.
-
-% possible flask summary
-Within the Python community, Flask\footnote{\url{http://flask.pocoo.org/}} is a very popular web framework\footnote{More than 25K projects on GitHub (5\% of all Python projects) are implemented with Flask (cf.
-a GitHub search for ``language:Python Flask'')}.
-It provides simplicity and flexibility by implementing a bare-minimum web server, and thus advertises as a micro-framework.
-The Flask tutorial shows how setting up a simple Flask {\em ``Hello World''} web-service requires no more than 5 lines of Python code \cite{ flask:tutorial}.
-% end of summary
-
-Despite their popularity, to the best of our knowledge, there is no simple solution for monitoring the evolving performance of Flask web applications.
-Thus, every one of the developers of these projects faces one of the following options when confronted with the need of gathering insight into the runtime behavior of their implemented services:
-
-  \begin{enumerate}
-
-    \item Use a commercial monitoring tool which treats the subject API as a black-box (e.g. Pingdom, Runscope). 
-    % , Graphite+Graphana+statd etc.
-
-    \item Implement their own ad-hoc analytics solution, having to reinvent basic visualization and interaction strategies.
-
-    \item Live without analytics insight into their services.
-
-  \end{enumerate}
-
-%\todo{For the first point in the list, we can also argue that analytics solutions like Google Analytics can be used, but they have no notion of versioning/integration with the development life cycle.
-Feel free to cite \cite{papazoglou2011managing} for service evolution purposes}
-
-For projects on a budget (e.g.
-research, startups) the first and the second options are often not available due to time and financial constraints.
-Even when using 3rd-party analytics solutions, a critical insight into the evolution of the exposed services of the web application, is missing because such solutions have no notion of versioning and no integration with the development life cycle.~\cite{papazoglou2011managing}
-
-To avoid projects ending up in the third situation, that of living without analytics, in this paper we present \tool~ --- a low-effort service monitoring library for Flask-based Python web services that is easy to integrate and enables the {\em agile assessment} of service evolution.
-\cite{Nier12b}
-
-As a case study, on which we will illustrate our solution, we are going to use an open source API which, for several years, was in the third of the above-presented situations.
-
-% In the next section, we will present a case study of an open source research API which was for a long time in the third situation presented above -- deployed without analytics insight.
-
-\todo{Add signposting paragraph}
-
-\section{Case Study: The API (Zeeguu)}
-\label{sec:case}
-
-%\todo{Either leave as is or add more info on usage by college if possible}
-
-  \zee\footnote{\url{https://github.com/zeeguu-ecosystem/}} is a platform and an ecosystem of applications for accelerating vocabulary acquisition in a foreign language \cite{Lungu16}. 
-%
-  The architecture of the ecosystem has at its core an API and a series of satellite applications that together offer to a learner three main inter-dependent features:
-
-  \begin{enumerate}
-
-    \item Reader applications that provide effortless translations for those texts which are too difficult for the readers.
-
-    \item Interactive exercises personally generated based on the preferences and past likes of the learner.
-
-    \item Article recommendations which are personalized for the interests of the learner and come with difficulty estimation that helps the learner find articles with the appropriate difficulty.
-
-  \end{enumerate}
-
-  The core API implemented with Flask and Python provides correspondingly three types of functionality: contextual translations, article recommendations, and personalized exercise suggestions.
-In total the API provides a bit less than 50 endpoints, out of which probably a dozen are very frequently used.
-  % The development of the core API itself is a research project.
-
-  At the time of writing this article, the ecosystem consists of a reader web application, a web based exercises platform, and a smartwatch application, which are used at the moment of writing this article by more than \activeUserCount active users.
-The users come from a highschool, a language school, and some users are using it on their own, without any educational context.
-The highest load we observed until now on the API consisted of 12K requests in one day.
-
-  We will use this \zee API as a case study for this paper. 
-  All the figures in this paper are captured from the actual deployment of \tool in the context of the \zee API.
-The figures are interactive offering basic data exploration capabilities: filter, zoom, and details on demand\cite{Shne99a}.
-The \tool deployment for the case study can be accessed publicly\footnote{\url{https://zeeguu.unibe.ch/api/dashboard}; username: {\em guest}, password: {\em soap-sac}}.
-
- % \todo{create new account, update URL if necessary, update activeUserCount, add a couple of sentences of the current state of the case study with the college/language center}
-% \ml{we should consider adding also one section in which the architecture/implementation and main features of the dashboard are presented before going on with discussing them in more depth in the following sections --- this should include a rundown on which views are provided from where (overview or per endpoint)}
-
-\newpage
-
-\section{Case Study: The Technology (Flask)}
-\label{sec:tool}
-
- Flask is a microframework for Python. It IS USED BY THOUSANDS OF PROJECTS. ELABORATE MORE ON THIS... 
- \vspace{2cm}
-
- Flask depends on two external libraries: the Jinja2 template engine and the Werkzeug WSGI toolkit.
-
- “Micro” does not mean that your whole web application has to fit into a single Python file (although it certainly can), nor does it mean that Flask is lacking in functionality.
-The “micro” in microframework means Flask aims to keep the core simple but extensible.
-Flask won’t make many decisions for you, such as what database to use.
-Those decisions that it does make, such as what templating engine to use, are easy to change.
-Everything else is up to you, so that Flask can be everything you need and nothing you don’t.
-
- Flask has many configuration values, with sensible defaults, and a few conventions when getting started.
-By convention, templates and static files are stored in subdirectories within the application’s Python source tree, with the names templates and static respectively.
-While this can be changed, you usually don’t have to, especially when getting started.
-
- A Python web application based on WSGI has to have one central callable object that implements the actual application.
-In Flask this is an instance of the Flask class.
-Each Flask application has to create an instance of this class itself and pass it the name of the module, but why can’t Flask do that itself?
-
-\begin{lstlisting}[style=custompython]
-
-from flask import Flask
-app = Flask(__name__)
-
-@app.route('/')
-def index():
-    return 'Hello World!'
-
-\end{lstlisting}
-
-\subsection*{NOTHING PREVENTS GENERALIZATION}
-In this article we take as an example the Flask case study.
-But there is nothing that prevents the generalization of 
-our approach.
-
-MAKE SURE THAT THIS IS CLEAR.
-
-    \begin{figure*}[ht!]
-      \centering
-      \includegraphics[width=0.8\linewidth]{interceptor}
-      \caption{The first thing that needs to be done, is to decorate the application object with an INTERCEPTOR}
-      \label{fig:sep}
-    \end{figure*}
-
-\newpage
-
-\section{Monitoring Service Utilization}
-
-REMEMBER THE GOAL: Our main goal with the \tool issues to allow the integration of the dashboard with an API with minimal effort.
-
-\ins{The first important question that such a 
-service monitoring system should provide is 
-information about what endpoints are being
-used and by whom}.
-
-TODO: Find references, arguments, related work
-that provide more details about why understanding
-who uses endpoints, and which are used are important.
-
-This is clearly the case in static analysis (see paper
-by haenni and lungu...) but should even more be the
-case in APIs.
-
-  \newpage
-TODO: TALK ABOUT THE DATABASE...
-
-  As discussed in the introductory section, in previous work~\cite{vogel2017low} we introduced \tool, a drop-in Python library that allows developers to monitor their Flask-based Python web applications with minimal effort.
-%
-  The \tool is implemented for Python 3.6 and is available on the Python Package Index repository\footnote{\url{https://pypi.python.org/pypi/flask-monitoring-dashboard/1.8}} from where it can be installed by running \install from the command line.
-%  
-  The source code of the \tool is published under a permissive MIT license and is available on GitHub\footnote{\url{https://github.com/flask-dashboard}}.
-
-  \newpage
-  \section{One Line Of Code Configuration With Technology Specific Interceptors}
-
-  The fact that the \tool itself is being developed in Python using Flask makes binding to the services of a to-be monitored application developed with the same technologies easy and intuitive.
-
-  \ins{In general, what matters is that with the right backend, we can do this minimal configuration for any technology... }
-
-  To start using the \tool, one simply needs to one lines of code\footnote{we don't count the import statement that enables that line...} to their Flask web service\footnote{\ins{ In this paper we present the integration with APIs written in Python and Flask hoping that this will not prevent the reader from seeing the more general idea; all the tools we show here for Flask can be applied to other API technologies (e.g.
-Django) by simply providing a few back-end adapters in the right places.
-}}:
-
-  % caption=Configuring the \tool is straightforward,
-  \begin{lstlisting}[style=custompython]
-  import dashboard
-
-  # flask_app is the Flask app object
-  dashboard.bind(flask_app)
-  \end{lstlisting}
-
-  \ins{The dashboard can take advantage of the fact that it is monitoring an API...
-so it can simply just add one extra endpoint there.} After (re)deployment of the application, the interactive visualizations of the \tool become available at the \code{/dashboard} route of the Flask application.
-
-  During binding, the \tool will search for all endpoints defined in the target application.
-
-  In order to monitor an endpoint, the \tool creates a function wrapper for the API function that corresponds to the endpoint.
-This way, the wrapper will be executed whenever that API call is made before the actual function is called.
-The wrapper contains the code that takes care of monitoring an endpoint.
-Data collected by the wrappers are persisted in a local database.
-The SQLAlchemy Object Relational Mapper\footnote{\url{https://www.sqlalchemy.org/}} on top of SQLite\footnote{\url{https://www.sqlite.org/}} is used for this purpose.
-  \ins{DISCUSSION about how do to this in other languages? Java: AspectJ probably, etc.}
-
-  These will be presented to the user in the tool web interface, where the user can select the ones that should be monitored, see \Fref{fig:sep}.
-
-    \begin{figure}[h!]
-      \centering
-      \includegraphics[width=\linewidth]{selecting_endpoints.png}
-      \caption{Once connected to an API the Dashboard presents the endpoints that are available for monitoring}
-      \label{fig:sep}
-    \end{figure}
-
-  This is the alternative to using annotations like other white-box tools (e.g. the other flask performance plugin).
-
-  \ins{PROMISE:} The last access time of the endpoint by users, irrespective of whether it is selected by the user to be monitored or not, is also displayed as the means of identifying potentially obsolete endpoints.
-
-  \ml{we must insert at least one dashboard view here...
-otherwise people will lose their patience...
-also it will help with explaining the outliers...
-also, one should see that a large variety of views are already available at this point...
-I know this risks being  }
-
-In the Zeeguu case study, one of the slowest endpoints, and one with the highest variability as shown in \Fref{fig:ep} is \epFeedItems: it retrieves a list of recommended articles for a given user.
-However, since a user can be subscribed to anything from one to three dozen article sources, and since the computation of the difficulty is personalized and it is slow, the variability in time among users is likely to be very large.
-
-\begin{figure*}[h!]
-  \centering
-  \subfloat[number of requests per endpoint per day]{\includegraphics[width=.9\columnwidth]{number_of_requests_}\label{fig:aeu}}
-  \subfloat[requests per hour heatmap]{\includegraphics[width=\columnwidth]{daily_patterns_}\label{fig:dp}}
-  \caption{Some of the available views\label{fig:views}}
-\end{figure*}
-
-\newpage
-\section{Service Utilization}
-\label{sec:util}
-
-\ins{A discussion must be added here about the types of useful information and problems that can be solved with these views.}
-
-\textit{Service Utilization} presents information about the usage of all endpoints of interest to the developer,
-
-The most fundamental insight that a service maintainer needs regards service utilization. %\vspace{0.5cm}
-
-Simplest utilization perspective:
-
-\begin{figure}[h!]
-  \centering
-  \includegraphics[width=0.5\linewidth]{endpoint-call-count}
-  \caption{Caption here}
-  \label{fig:figure1}
-\end{figure}
-
-% REQUIREMENT: Adding a time interval is important. 
-% Otherwise the performance numbers in chart 1 are not good enough... also an outlier very far in the past, will make all the heatmap data points to look very pale in comparison forever after.
-
-Figure \ref{fig:aeu} shows a first perspective on endpoint utilization that \tool provides: a stacked bar chart of the number of hits to various endpoints grouped by day\footnote{Endpoint colors are the same in different views}.
-Figure~\ref{fig:aeu} in particular shows that at its peak the API has about 12.000 hits per day.
-
-The way users interact with the platform can also be inferred since the endpoints are indicators of different activity types, e.g.:
-
-\begin{enumerate}
-
-  \item {\color{myred} \epTranslations} is an indicator of the amount of foreign language reading the users are doing, and
-
-  \item {\color{mygreen} \epOutcome} is an indicator of the amount of foreign vocabulary practice the users are doing.
-
-\end{enumerate}
-
-%\begin{figure}[!ht]
-% \centering
-% \includegraphics[width=\linewidth]{number_of_requests_}
-% \caption{The number of requests per endpoint per day view shows the overall utilization of the monitored application}
-% \label{fig:aeu}
-%\end{figure}
-
-Besides showing the overall utilization, this endpoint provides the maintainer with information relevant for decisions regarding endpoint deprecation --- one of the most elementary ways of {\em understanding the needs of the downstream}\cite{Haen14a}.
-In our case study, the maintainer realized that one endpoint which they thought was not being used (i.e.
-\code{words\_to\_study}), contrary to their expectations, was actually being used\footnote{A complementary type of usage information can also be discovered in the view presented in Figure \ref{fig:sep} where seeing that an endpoint is never accessed can increase the confidence of the maintainer that a given endpoint is not used, although it can never be used a proof.}.
-
-\niceseparator
-
-%   \todo{Add the time series graph and discuss it before the heatmap? We can then sell the heatmap better} 
-%   \ml{Not sure about which graph you refer to here V}
-
-A second type of utilization question that the \tool can answer automatically regards {\em cyclic patterns of usage per hour of day} by means of a heatmap, as in \Fref{fig:dp}.
-
-% \mltp{ can we add vertical lines that highlight the beginning of a new week (e.g. before Sunday): }
-% Patrick: adding separators in the graph is unfortunately not supported by the library.
-
-%\begin{figure}[!ht]
-% \centering
-% \includegraphics[width=0.9\linewidth]{daily_patterns_}
-% \caption{Usage patterns become easy to spot in the requests per hour heatmap}
-% \label{fig:dp}
-%\end{figure}
-
-Figure \ref{fig:dp} shows the API not being used during the early morning hours, with most of the activity focused around working hours and some light activity during the evening.
-This is consistent with the fact that the current users are all in the central European timezone.
-Also, the figure shows that the spike in utilization that was visible also in the previous graph happended in on afternoon/evening.
-
-\newpage
-\section{Endpoint Performance}
-\label{sec:perf}
-
-  The \tool also collects information regarding endpoint performance. The view in \Fref{fig:ep} summarizes the response times for various endpoints by using a box-and-whiskers plot.
-
- \begin{figure}[!ht]
-   \centering
-   \includegraphics[width=0.95\linewidth]{endpoint_performance_}
-   \caption{The response time (in ms) per monitored endpoint view allows for identifying performance variability and balancing issues}
-   \label{fig:ep}
- \end{figure}
-
-  From this view it became clear to the maintainer that four of the endpoints had very large variation in performance.
-The most critical for the application and consequently the one optimized first was the \epTranslations endpoint which was part of an interactive loop in the reader applications that relied on the Zeeguu API.
-Moreover, cf.
-\Fref{fig:aeu} this endpoint is one of the most used in the system.
-
-\begin{figure}[h!]
-  \centering
-  \subfloat[overview of the tracked endpoints]{\includegraphics[width=0.8\columnwidth]{overview}\label{fig:overview}}
-  \caption{Some of the available views\label{fig:views}}
-\end{figure}
-
-  \newpage
-  \section{Evolution monitoring }
-
-  \begin{figure}[h!]
-  \centering
-  \includegraphics[width=0.8\linewidth]{translation_endpoint_evolution_}
-  \caption{The Performance Evolution of the \epTranslations endpoint}
-  \label{fig:tee}
-\end{figure}
-
-  When thinking about performance, 
-  one can not avoid thinking about the performance evolution.
-
-  Version control in \tool can be supported in two ways,
-  the developer explicitly states the current version, 
-  or the current version is automatically detected based
-  on some version control system.
-
-  However, with the current configuration of the tool, it would be impossible for the maintainer to see the improvements resulting from the optimization.
-
-  \ins{The latest STack Overflow Developer Survey: 88.4\% of **professional** developers use git.}  
-  If assume that the code that they run is deployed using .git, then with an extra line of configuration they can allow \tool to find the git\footnote{\url{https://git-scm.com/}} folder of the deployed service and automatically detect the version of the project that is running:
-
-    \begin{lstlisting}[style=custompython]
-
-      dashboard.config.git = 'path/to/.git'
-
-    \end{lstlisting}
-
-  If the \tool can automatically detect the current version of the project by reading the .git configuration as soon as the API is started and can then group measurements by version\footnote{Alternatively, the maintainer can add version identifiers manually for the web application through a configuration file if the system does not use git.}.
-\Fref{fig:tee} is a zoomed-in version of such a view for \epTranslations with versions increasing from top to bottom
-
-  This view confirms that the performance of the translation endpoint improved in the recent versions: the median of the last three versions is constantly moving towards the left, and progresses from 1.4 seconds (in the top-most box plot in \Fref{fig:tee}) to 0.8 in the latest version (bottom-most box plot).
-
-  CHALLENGE: Supporting other types of version control... not real challenge, as we said 88percent of devs use git. 
-  CHALLENGE 2: detecting minor versions which don't need to be tracked, since they didn't touch the performance of the system. E.g. a modification of the README, etc.
-
-\todo{TODO: endpoint evolution }
-
-  \begin{figure*}[h!]
-    \centering
-    \includegraphics[width=\linewidth]{outliers}
-    \caption{Automatically collected outlier information}
-    \label{fig:figure1}
-  \end{figure*}
-  \newpage
-  \section{Automated outlier detection and monitoring}
-
-  When an API is called from within a highly interactive application (as it is the case with the case study in this paper) 
-  of particular interest to the API developers are performance {\em outliers}.
-
-  Indeed, a translation request that takes three times more than expected can seriously decrease the perceived quality of the application.
-Thus, identifying, collecting all appropriate data, and diagnosing the root causes of such outliers is especially critical in improving the quality of an application.
-
-  % In the context of the RESTful services support by Flask, this means request serving times that deviate from the average to an unexpected degree.
-
-  For this purpose the \tool tracks for every monitored endpoint a {\em running average} response time value\footnote{\ins{For performance reasons, we assume that the response times for the endpoints are normally distributed.
-Otherwise, more general density distribution information must be collected in real time.}}.
-When it detects that a given request is an outlier with respect to this past average running value, it triggers the {\em outlier data collection routine} which stores \ins{extra information} about the current execution environment.
-A configurable threshold with a default value of $2.5$ times the running average response time is used for this purpose.
-
-  For every detected outlier request, the \tool collects information about the current Python stack trace, CPU load, memory consumption, request parameters, etc.
-in order to allow the maintainer to investigate the causes of these exceptionally slow response times.
-In this way it is possible to get \ins{detailed insight into the operation of the application in the extreme cases without unnecessarily burdening it with logging this information for every request}.
-
-The bottom panel shows the stack trace. 
-In this particular case, it is revealing for the developer to learn that at the time of the stack trace snapshot, the code was in the microsoft\_translator.py: indeed, the system uses as back-end multiple translators, and it has been observed that many of the outliers happen to be waiting in the microsoft translator.
-
-This information has to be corroborated with the observations that neither the memory nor the processor are overloaded at the moment.
-Thus this functionality in microsoft\_translator is really slow in itself, and this is not a result of the machine being overloaded for example.
-
-\subsection {A Special Type of Outlier: Exceptions}
-
-   TODO: probably something for the journal extension ...
-  statistics about which endpoints fail most often might be useful too.
-  same information as the outlier maybe?
-
-\begin{figure*}[!ht]
-  \centering
-  \includegraphics[width=0.4\linewidth]{time_per_user}
-  \caption{The \epFeedItems shows a very high variability across users}
-  \label{fig:tpu}
-\end{figure*}
-
-\newpage
-\section{Grouping Requests}
-\label{sec:user}
-
-For service endpoints which run computations in real time, the maintainer of a system might want to understand the endpoint performance on a per-user basis, especially for situations where the system response time is a function of some individual user load\footnote{E.g.
-in GMail some users have two emails while other have twenty thousand and this induces different response times for different users}.
-
-\todo{remove the configuration part and rephrase the following, the discussion has been partially subsumed by Section~\ref{sec:tool}.}
-
-To enable this, the \tool must be configured to associate an API call with a given user.
-The simplest way is to take advantage of the architecture of Flask applications in which a global \code{flask.request} object can be used to retrieve the session which can in turn lead to user identification:
-
-%\begin{lstlisting}[float,caption=Simply define a custom app-specific function for user retrieval and pass it to the \tool to group information by user,style=custompython]
-\begin{lstlisting}[style=custompython]  
-# app specific way of extracting the user
-# from a flask request object    
-def get_user_id():
-sid = int(flask.request.args['session'])
-session = User.find_for_session(sid)
-return user_id
-
-# attaching the get_user_id function
-dashboard.config.get_group_by = get_user_id
-
-\end{lstlisting}
-
-% \niceseparator
-
-\todo{consider losing one or both of the figures --- or at least updating them}
-
-In Zeeguu, \epFeedItems retrieves a list of recommended articles for a given user.
-Cf.
-\Fref{fig:ep} it is the endpoint with the slowest response time and highest variability.
-The reason for this is that a user can be subscribed to anything from one to three dozen article sources and for each of the sources the system must compute the personalized difficulty of each article at every request.
-
-A \perspective{Per-User Performance} perspective should show the different response times for different users.
-Figure \ref{fig:tpu} presents a subset of the corresponding view in the \tool.
-The figure shows that the response times for this endpoint can vary considerably for different users with some extreme cases where a user has to wait a full minute until their recommended articles are shown\footnote{After seeing this perspective, the maintainer refactored the architecture of the system to move part the difficulty computation out of the interactive loop}.
-
-The limitation of the previous view is that it does not present the information also on a per version basis.
-To address this, a different visual perspective entitled \perspective{Multi-Version per-User Performance} can be defined.
-Figure \ref{fig:tuv} presents such a perspective by mapping the average execution time for a given user (lines) and given version (columns) on the area of the corresponding circle.
-
-The colors represent users.
-The figure shows average performance varying  across users and versions with no clear trend: this is probably because varying user workload (i.e.
-number of sources to which the user is registered) is the reason for the variation in response times.
-\ins{one can see that for user 1 performance degrades over the }
-
-\begin{figure}[!ht]
-  \centering
-  \includegraphics[width=0.7\linewidth]{time_per_user_per_version}
-  \caption{This perspective shows that the evolution of response times for individual users (horizontal lines) across versions (the x-axis)}
-  \label{fig:tuv}
-\end{figure}
-
-  \newpage
-  \section{Preemptive monitoring}
-
-  \ins{INTRO - LINKING WITH PREVIOUS PARAGRAPH: the previous evolutionary view crucial for 
-  understanding the impact of software evolution on performance.
-  However, just by observing this one can't be sure whether 
-  performance changes because of the code or because of the payload.}
-
-  \todo{the following might need a bit of rewrite}
-
-  The concept of {\em preemptive monitoring} of the application performance by means of instrumenting integration unit testing as the synthetic load is similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}, i.e.~testing service-based applications by using dedicated test input in parallel to its normal use and operation.
-The difference is that we take advantage of the capability of the CI framework (i.e.
-Travis in this case) to create an emulated ``live'' environment for integration testing purposes, and use unit testing as the dedicated test input in order to measure performance.
-
-  For example, \Fref{fig:builds} is a screenshot from the dashboard showing the measured response times for 5 consequent builds, with 180 iterations of the unit tests executed in total across all endpoints.
-The outliers on the right of the figure are due to initial requests \ins{which must wait for a boot-up phase of the API}.
-
-    \begin{figure}[h!]
-        \centering
-        \includegraphics[width=\columnwidth]{travis_builds}
-        \caption{Response times in 5 consequent Travis builds in the \zee case study}
-        \label{fig:builds}
-      \end{figure}
-
-  While this integration testing environment is different from the actual production deployment one, and the load used is purely synthetic, as will show in Section~\ref{sec:evaluation}, it can actually be used successfully as an early performance indicator for the application developer.
-\ins{Although predicting the performance of the application would require advanced statistical work, it can be succesfully used as an advanced warning system for the overall performance trend for given endpoints}.
-
-  \paragraph{Integration with Travis}
-  \ml{added new section... also, this has to be moved after introducing the pre-emptive monitoring since it's about running tests. switched the order}
-
-  \ins{The pre-emptive monitoring system} can be configured to work together with Continuous Integration (CI) frameworks like Travis\footnote{\url{https://travis-ci.org/}} that deal with automated integration testing.
-  The developer needs to define the unit test folder of the project, the URL that Travis should use for reporting its results, and the number of iterations for each unit test should be executed in order to generate a load for the application.
-
-  Each time a new build is created through Travis, the \tool automatically detects all available unit tests defined by the application developer and iterates through each one of them the number of times predefined by the developer while monitoring the response times for each request.
-The resulting measurements are persisted in a separate part of the tool database so as not to contaminate the ``live'' monitoring data.
-Beyond seeing the results of each Travis build in the \tool, this feature also allows for preemptive monitoring of the application, as we discuss in the following.
-
-  \ins{Besides functioning as an early warning system for performance, tracking the evolving performance of API tests serves a second purpose.
-To function as an anchor when the developer analyzes performance degradations.
-For example, when a developer sees that a given endpoint has become less performant after in a newly deployed version, they are able to investigate whether the performance degradation is visible also with the synthetic load.
-This would correspond to a performance degradation which is due to ``algorithmic performance degradation''.
-If on the other hand, the performance of the tests does not change between versions, the developer might conclude that the performance degradation could be due to the workload on the machines, or maybe to the workload mix of the users}
-
-  %By these means, an application developer can use unit tests as synthetic loads for driving the performance monitoring of the application before it reaches the production phase.
-As we will discuss further in Section~\ref{sec:testing}, this feature is integral in enabling preemptive performance monitoring of the application.
-
-\todo{here goes the discussion about our forecasting capabilities}
-
-\begin{figure}[h!]
-  \centering
-  \subfloat[The reported response time (in ms) per deployed version in the observation period]{\includegraphics[width=.5\columnwidth]{response_per_version_trunced_trend}}
-  \quad
-  \subfloat[The measured response time (in ms) using integration with Travis]{\includegraphics[width=.46\columnwidth]{travis_builds_no_outliers_trend}}
-  \caption{Comparison of the response times per endpoint: actual production system versus preemptive monitoring data}
-  \label{fig:preemptive}
-\end{figure}
-
-\begin{table}[h]
-  \begin{tabular}{lll}
-    \toprule
-    Iteration & \bfseries Live (median) & \bfseries Travis (median)\\
-    \midrule
-    1 & 1349.41 & 764.87\\ 
-    2 & 1466.13 & 992.87\\
-    3 & 1256.65 & 760.87\\
-    4 & 1266.42 & 813.89\\
-    5 & 1080.68 & 303.4\\
-    \bottomrule
-  \end{tabular}
-\end{table}
-
-Pearson correlation $r(3)=.93, p=.02$
+\input{testing}
 
@@
 
-  %In the remainder of the paper we present several of these perspectives.\footnote{We recommend obtaining a color version of this paper for better readability}
+  %In the remainder of the paper we present several of these perspectives.
 
@@
 
-\section{Discussion}
+\input{discussion}
 
-\todo{use this subsection to discuss limitations (from the previous evaluation section) and add the need to be able to handle the horizontal scaling of the application; remove sectioning and summarize briefly limitations that have already been discussed in the VISSOFT paper/move material to other sections; consider renaming as limitations or something similar}
-
-  \subsection{Distributed System Monitoring}
-  \todo{To think about}
-  Supporting situations where the API is deployed across multiple containers for example.
-
-  \subsubsection{Automatically Monitoring System Evolution}
-
-  The main goal of the \tool design was to allow analytics to be collected and insight to be gained by making the smallest possible changes to a running API.
-%To allow the collection of evolutionary information
-%
-  This technique assumes that the web application code which is the target of the monitoring is deployed using \git in the following way:
-
-  \begin{enumerate}
-    \item The deployment engineer pulls the latest version of the code from the integration server; this will result in a new commit being pointed at by the HEAD pointer. %than previously
-    \item The deployment engineer restarts the new version of the service.
-At this point, the \tool detects that a new HEAD is present in the local code base and consequently starts associating all the new data points with this new commit\footnote{The \tool detects the current version of the analyzed system the first time it is attached to the application object, and thus, assumes that the Flask application is restarted when a new version is deployed.
-This is in tune with the current version of Flask, but if the web server will support dynamic updates in the future, this might have to be taken into account}.
-  \end{enumerate}
-
-  The advantage of this approach is the need for minimal configuration effort, as discussed in the presentation of the tool.
-The disadvantage is that it will consider on equal ground the smallest of commits, even one that modifies a comment, and the shortest lived of commits, e.g.~a commit which was active only for a half an hour before a new version with a bug fix was deployed, with major and minor releases of the software.
-%as a distinct way of grouping the data points.
-  A mechanism to control which versions are important for monitoring purposes is therefore required to be added.
-%
-  A further possible extension point here is supporting other version control systems (e.g. Mercurial). However, this is a straightforward extension.
-
-  \subsubsection{NEW: API Renames}
-  \ml{just an idea}
-
-    The history would normally be lost in the case of an API rename. (also known as the {\em provenance issue} in source code evolution analysis)
-    In theory we could infer that an API was renamed if it is not to be found anymore, and another one with very similar timing characteristics would be found.
-Such a detection can never be completely sure, but it would help with meeting the needs of the API maintainer...
-
-  \subsubsection{User-Awareness }
-
-    For the situations in which the user information is not available, the \tool tracks by default information about different IPs and in some cases this might be a sufficiently good approximation of the user diversity and identity.
-    %
-
-    The visualizations for the user experiene perspectives as presented in Section \ref{sec:user} have been tested with several hundred users (of which about \activeUserCount were active during the course of the study), but the scalability of the visualizations must be further investigated for web services with tens of thousands of users.
-
-  \subsubsection{Other Possible Groupings}
-
-    There are other groupings of service utilization and performance that could be important to the maintainer, that we did not explore in this paper.
-For example, if the service is using OAuth, then together with every request, in the header of the request there is information about the application which is sending a request.
-Grouping the information by application that sends the request could be important in such a context.
-
-    In general, providing a mechanism that would allow very easy specification of groupings (either as code annotations, as normal code, or as configuration options) is an open problem that \tool and any other similar library will have to face.
-
-\section{Related Work}
-\label{sec:related}
-
-\todo{Refer the reader to~\cite{ghezzi2007run} and ~\cite{metzger2010analytical} for a more extensive discussion}
-
-There is a long tradition of using visualization for gaining insight into software performance.
-Tools like Jinsight \cite{Pauw02a} and Web Services Navigator \cite{Pauw05} pioneered such an approach for Java and for Web Services that communicate with SOAP messages.
-Both have an ``omniscient'' view of the services / objects and their interactions.
-As opposed to them, in our work we present an analytics platform which focuses on monitoring a single Python web service from its own point of view.
-
-From the perspective of service monitoring, our work falls within the server-side run-time monitoring of services ~\cite{ghezzi2007run}.
-While we don't implement the more advanced features of related monitoring solutions like QoS policies driving the monitoring, it presents nevertheless an easy to use approach support improving the performance of web applications.
-
-% \va{Mircea: Consider removing the rest for space...}
-% An existing monitoring tool is Pingdom \footnote{https://www.pingdom.com/company/why-pingdom}, which monitors the uptime of an existing web-service.
-This tool works by pinging the websites (up to 60 times) every minute automatically.
-Thus this creates a lot of overhead and is bound to be noisy since it will also be influenced by the speed of the network connection\footnote{Another problem is that such a tool would }
-
-% \todo{Runscope? Others?}
+\input{related-work}
 

### 2018-04-03 testing.tex
split the huge file in smaller ones.

--- 2018-03-20 paper.tex
@@
-\documentclass[conference]{IEEEtran}
-	\pdfpagewidth=8.5truein
-	\pdfpageheight=11truein
+%!TEX root=restructured.tex
 
-\usepackage{graphicx}
-\usepackage{subfig}
-\usepackage[bookmarks=false]{hyperref}
-\usepackage{xspace}
-\usepackage{listings}
-\usepackage[usenames, dvipsnames]{color}
+  \newpage
+  \section{Preemptive monitoring}
 
-\usepackage{booktabs}
-
-\graphicspath{{./img/}}
-
-%\hyphenation{op-tical net-works semi-conduc-tor}
-
-\input{useful_defines}
-
-\begin{document}
-
-%
-\title{A Low-Effort Analytics Platform for Visualizing Evolving Flask-Based Python Web Services}
-% Alternative Titles: The Importance of Visualization in the Performance Monitoring of Python Web Services
-
-\author{
-\IEEEauthorblockN{Patrick Vogel, Thijs Klooster, Vasilios Andrikopoulos, Mircea Lungu}\\
-Johann Bernoulli Institute for Mathematics and Computer Science\\
-University of Groningen, Netherland\\
-Email: \{t.klooster.1,p.p.vogel\}@student.rug.nl, \{v.andrikopoulos,m.f.lungu\}@rug.nl 
-}
-
-% make the title area
-\maketitle
-
-\begin{abstract}
-\todo{Update abstract \& title}
-  Tens of thousands of web applications are written in Flask, a Python-based web framework.
-Despite a rich ecosystem of extensions, there is none that supports the developer in gaining insight into the evolving performance of their service.
-In this paper, we introduce \tool, a library that addresses this problem.
-We present the ease with which the library can be integrated in an already existing web application, discuss some of the visualization perspectives that the library provides and point to some future challenges for similar libraries.
-
-\end{abstract}
-
-\section{Introduction}
-
-\todo{Intro to be rewritten; refer to VISSOFT paper as previous work, explain focus is on integration with version control/CI for evolution support, and use of unit tests as early performance indicator}
-
-\todo{motivate with: monitoring in a world which is becoming more and more service-oriented is essential because it allows to perform three main types of actions: system adaptations to provide the request service at the desired level of quality, enabling flexibility in dealing with changing requirements and modes of operation, and enabling operational awareness through dashboards organizing the collected data~\cite{pernici2016monitoring}.
-In this work we focus on the last part and we discuss how a minimal effort probe-based solution for a particular type of web services can be used to facilitate the other two types of actions by involving the developer.}
-
-\todo{explain: in~\cite{vogel2017low} we introduced the \tool an a high level with a focus on presenting its performance visualization aspects; in this work we only summarize these features by means of introducing some of the provided functionalities of the \tool.
-We focus on discussing how the tool is implemented and operating, provide a deeper look at its capabilities for integration with version control and continuous integration environments, introduce a mechanism for performance prediction, and provide an evaluation of the proposed approach based on a case study.
-}
-
-%Every system is a distributed system nowadays \cite{cavage2013there}.
-Indeed a very large number of applications and web applications are nowadays implemented as two-tier architectures with a front-end implemented with web technologies and a service back-end.
-%\ml{I'm not completely happy with this paragraph}
-{\em There is no getting around it: you are building a distributed system} argues a recent article \cite{cavage2013there}.
-Indeed, even the simplest second-year student project is a web application implemented as two-tier architecture with a Javascript/HTML5 front-end a service backend, usually a REST API.
-
-% \hfill mds
-% Many contemporary programming languages are offering libraries, modules, or frameworks that facilitate the development of such architectures. 
-Python is one of the most popular programming language choices for implementing the back-end of web applications.
-GitHub contains more than 500K open source Python projects and the Tiobe Index\footnote{TIOBE programming community index is a measure of popularity of programming languages, created and maintained by the TIOBE Company based in Eindhoven, the Netherlands} ranks Python as the 4th most popular programming language as of June 2016.
-
-% possible flask summary
-Within the Python community, Flask\footnote{\url{http://flask.pocoo.org/}} is a very popular web framework\footnote{More than 25K projects on GitHub (5\% of all Python projects) are implemented with Flask (cf.
-a GitHub search for ``language:Python Flask'')}.
-It provides simplicity and flexibility by implementing a bare-minimum web server, and thus advertises as a micro-framework.
-The Flask tutorial shows how setting up a simple Flask {\em ``Hello World''} web-service requires no more than 5 lines of Python code \cite{ flask:tutorial}.
-% end of summary
-
-Despite their popularity, to the best of our knowledge, there is no simple solution for monitoring the evolving performance of Flask web applications.
-Thus, every one of the developers of these projects faces one of the following options when confronted with the need of gathering insight into the runtime behavior of their implemented services:
-
-  \begin{enumerate}
-
-    \item Use a commercial monitoring tool which treats the subject API as a black-box (e.g. Pingdom, Runscope). 
-    % , Graphite+Graphana+statd etc.
-
-    \item Implement their own ad-hoc analytics solution, having to reinvent basic visualization and interaction strategies.
-
-    \item Live without analytics insight into their services.
-
-  \end{enumerate}
-
-%\todo{For the first point in the list, we can also argue that analytics solutions like Google Analytics can be used, but they have no notion of versioning/integration with the development life cycle.
-Feel free to cite \cite{papazoglou2011managing} for service evolution purposes}
-
-For projects on a budget (e.g.
-research, startups) the first and the second options are often not available due to time and financial constraints.
-Even when using 3rd-party analytics solutions, a critical insight into the evolution of the exposed services of the web application, is missing because such solutions have no notion of versioning and no integration with the development life cycle.~\cite{papazoglou2011managing}
-
-To avoid projects ending up in the third situation, that of living without analytics, in this paper we present \tool~ --- a low-effort service monitoring library for Flask-based Python web services that is easy to integrate and enables the {\em agile assessment} of service evolution.
-\cite{Nier12b}
-
-As a case study, on which we will illustrate our solution, we are going to use an open source API which, for several years, was in the third of the above-presented situations.
-
-% In the next section, we will present a case study of an open source research API which was for a long time in the third situation presented above -- deployed without analytics insight.
-
-\todo{Add signposting paragraph}
-
-\section{Case Study}
-\label{sec:case}
-
-%\todo{Either leave as is or add more info on usage by college if possible}
-
-  \zee\footnote{\url{https://github.com/zeeguu-ecosystem/}} is a platform and an ecosystem of applications for accelerating vocabulary acquisition in a foreign language \cite{Lungu16}. 
-%
-  The architecture of the ecosystem has at its core an API and a series of satellite applications that together offer to a learner three main inter-dependent features:
-
-  \begin{enumerate}
-
-    \item Reader applications that provide effortless translations for those texts which are too difficult for the readers.
-
-    \item Interactive exercises personally generated based on the preferences and past likes of the learner.
-
-    \item Article recommendations which are personalized for the interests of the learner and come with difficulty estimation that helps the learner find articles with the appropriate difficulty.
-
-  \end{enumerate}
-
-  The core API implemented with Flask and Python provides correspondingly three types of functionality: contextual translations, article recommendations, and personalized exercise suggestions.
-In total the API provides a bit less than 50 endpoints, out of which probably a dozen are very frequently used.
-  % The development of the core API itself is a research project.
-
-  At the time of writing this article, the ecosystem consists of a reader web application, a web based exercises platform, and a smartwatch application, which are used at the moment of writing this article by more than \activeUserCount active users.
-The users come from a highschool, a language school, and some users are using it on their own, without any educational context.
-The highest load we observed until now on the API consisted of 12K requests in one day.
-
-  We will use this \zee API as a case study for this paper. 
-  All the figures in this paper are captured from the actual deployment of \tool in the context of the \zee API.
-The figures are interactive offering basic data exploration capabilities: filter, zoom, and details on demand\cite{Shne99a}.
-The \tool deployment for the case study can be accessed publicly\footnote{\url{https://zeeguu.unibe.ch/api/dashboard}; username: {\em guest}, password: {\em soap-sac}}.
-
- % \todo{create new account, update URL if necessary, update activeUserCount, add a couple of sentences of the current state of the case study with the college/language center}
-% \ml{we should consider adding also one section in which the architecture/implementation and main features of the dashboard are presented before going on with discussing them in more depth in the following sections --- this should include a rundown on which views are provided from where (overview or per endpoint)}
-
-\section{The Flask Dashboard}
-\label{sec:tool}
-
-%\todo{Refer to VISSOFT paper~\cite{vogel2017low} and rephrase where possible; add more info on implementation, e.g. annotation mechanism, database design, session handling etc.}
-
-  As discussed in the introductory section, in previous work~\cite{vogel2017low} we introduced \tool, a drop-in Python library that allows developers to monitor their Flask-based Python web applications with minimal effort.
-%
-  The \tool is implemented for Python 3.6 and is available on the Python Package Index repository\footnote{\url{https://pypi.python.org/pypi/flask-monitoring-dashboard/1.8}} from where it can be installed by running \install from the command line.
-%  
-  The source code of the \tool is published under a permissive MIT license and is available on GitHub\footnote{\url{https://github.com/flask-dashboard}}.
-
-  \paragraph{Usage \& Configuration}
-
-  The fact that the \tool itself is being developed in Python using Flask makes binding to the services of a to-be monitored application developed with the same technologies easy and intuitive.
-\ml{this is not what matters...
-rather, what matters is that with the right backend, we can do this minimal configuration for any technology...
-do you see what i mean?
-let me see if i can put it in words...}
-
-  Our main goal with the \tool was to allow the integration of the dashboard with an API with minimal effort.
-  To start using the \tool, one simply needs to add two lines of code to their Flask web service\footnote{\ins{ In this paper we present the integration with APIs written in Python and Flask hoping that this will not prevent the reader from seeing the more general idea; all the tools we show here for Flask can be applied to other API technologies (e.g.
-Django) by simply providing a few back-end adapters in the right places.
-}}:
-
-  % caption=Configuring the \tool is straightforward,
-  \begin{lstlisting}[style=custompython]
-  import dashboard
-
-  # flask_app is the Flask app object
-  dashboard.bind(flask_app)
-  \end{lstlisting}
-
-  After (re)deployment of the application, the \tool becomes available at the \code{/dashboard} route of the Flask application.
-Further configuration is possible by adding additional statements before the binding definition, for example,
-
-  \begin{lstlisting}[style=custompython]
-  ...
-  dashboard.config.link = 'mydashboard'
-  dashboard.bind(flask_app)
-  \end{lstlisting}
-
-  allows for a custom route (\code{/mydashboard}) to the dashboard to be defined by the programmer. An external configuration file can be used instead, or in addition to this:
-
-  \begin{lstlisting}[style=custompython]
-  ...
-  dashboard.config.from_file('config.cfg')
-  ...
-  \end{lstlisting}
-
-  During binding, the \tool will search for all endpoints defined in the target application.
-These will be presented to the user in the tool web interface, where the user can select the ones that should be monitored, see \Fref{fig:sep}.
-The last access time of the endpoint by users, irrespective of whether it is selected by the user to be monitored or not, is also displayed as the means of identifying potentially obsolete endpoints.
-
-    % \mltp{a small screenshot of how the dashboard allows one to select the interesting }
-    \begin{figure}
-      \centering
-      \includegraphics[width=\linewidth]{selecting_endpoints.png}
-      \caption{Endpoints that are available for monitoring in the \zee API}
-      \label{fig:sep}
-    \end{figure}
-
-  In order to monitor an endpoint, the \tool creates a function wrapper for the API function that corresponds to the endpoint.
-This way, the wrapper will be executed whenever that API call is made.
-The wrapper contains the code that takes care of monitoring an endpoint.
-Data collected by the wrappers are persisted in a local database.
-The SQLAlchemy Object Relational Mapper\footnote{\url{https://www.sqlalchemy.org/}} on top of SQLite\footnote{\url{https://www.sqlite.org/}} is used for this purpose.
-
-  \ml{we must insert at least one dashboard view here...
-otherwise people will lose their patience...
-also it will help with explaining the outliers...
-also, one should see that a large variety of views are already available at this point...
-I know this risks being  }
-
-  \paragraph{Automated outlier detection and monitoring}
-
-  \ins{When an API is called from within a highly interactive application (as it is the case with the Zeeguu ecosystem) 
-  of particular interest to the API developers are performance {\em outliers}.
-Indeed, a translation request that takes three times more than expected can seriously decrease the perceived quality of the application.
-Thus,} identifying, collecting all appropriate data, and diagnosing the root causes of such outliers is especially critical in improving the quality of an application.
-
-  % In the context of the RESTful services support by Flask, this means request serving times that deviate from the average to an unexpected degree.
-
-  For this purpose the \tool tracks for every monitored endpoint a {\em running average} response time value\footnote{\ins{For performance reasons, we assume that the response times for the endpoints are normally distributed.
-Otherwise, more general density distribution information must be collected in real time.}}.
-When it detects that a given request is an outlier with respect to this past average running value, it triggers the {\em outlier data collection routine} which stores \ins{extra information} about the current execution environment.
-A configurable threshold with a default value of $2.5$ times the running average response time is used for this purpose.
-
-  For every detected outlier request, the \tool collects information about the current Python stack trace, CPU load, memory consumption, request parameters, etc.
-in order to allow the maintainer to investigate the causes of these exceptionally slow response times.
-In this way it is possible to get \ins{detailed insight into the operation of the application in the extreme cases without unnecessarily burdening it with logging this information for every request}.
-
-  \paragraph{User-level monitoring}
-
-  \ins{Sometimes, the same service endpoint will have different execution times for different users, based on the amount of data that a user has in their account.
-Gmail engineers admitted in 2010 that the performance of their service was dependent on the amount of emails a user would have in their Inbox \footnote{https://techcrunch.com/2010/03/14/gmail-slow/, visited September 23 2017}.
-}
-
-  \ins{If the API provider needs to understand the performance of their service for individual users, or for different user groups, the dashboard should have a way of distilling a user or a user group from the API request.
-For a large number of Flask applications that use the global \code{flask.request} object, the object} can be used to retrieve session information, \ins{as in the following snippet}:
-  % .
-This ability can be used in order to allow, for example, the identification of application users from the \tool, and the consequent aggregation of monitoring information on a per user basis as follows:
-
-  \begin{lstlisting}[style=custompython]  
-  # app specific way of extracting the user
-  # from a flask request object    
-  def get_user_id():
-  sid = int(flask.request.args['session'])
-  session = User.find_for_session(sid)
-  return user_id
-
-  # attaching the get_user_id function
-  dashboard.config.get_group_by = 
-      get_user_id
-
-  \end{lstlisting}
-  This feature, as discussed further in the following section, is important for the clustering of users depending on their application usage.
-It also provides a view on the distribution of the work load across the application endpoints on a per user basis that allows the identification of more or less popular endpoints for further performance improvements.
-\ml{this previous paragraph seems a bit reduntant.
-must revisit}
-
-  \paragraph{Evolution monitoring for Git-versioned projects }
-
-  Version control in \tool is supported in two ways.
-
-\begin{enumerate}
-
-  \item  \textit{Manual} version control requires the developer to tag each new version of the application with an appropriate version identifier~\cite{papazoglou2011managing} using the \code{APP\_VERSION} configuration parameter, for example by adding to the configuration file:
-
-    \begin{lstlisting}[style=custompython]
-    [dashboard]
-    APP_VERSION=<versionID>
-    \end{lstlisting}
-
-  \item \textit{Automatic} version control, aimed for use in conjunction with a version control system and currently only supporting Git\footnote{\url{https://git-scm.com/}}, is enabled by linking to the Git folder in the application source code:
-
-      \begin{lstlisting}[style=custompython]
-      ....
-      dashboard.config.git = '.git'
-      ....
-      \end{lstlisting}
-
-    With this extra configuration parameter, the \tool automatically detects the current version of the project by reading the hash code of the deployed version as soon as the application is deployed and started.
-Information about the version history of the application is also persisted in the tool database and is used during the collection of measurements for the application.
-
-\end{enumerate}
-
-  \paragraph{Preemptive monitoring}
+  \ins{INTRO - LINKING WITH PREVIOUS PARAGRAPH: the previous evolutionary view crucial for 
+  understanding the impact of software evolution on performance.
+  However, just by observing this one can't be sure whether 
+  performance changes because of the code or because of the payload.}
 
@@
 
-\section{Performance Perspectives}
-\label{sec:views}
-
-  There are three main categories of visual perspectives that are available using \tool:
-  \begin{enumerate}
-    \item \textit{Service Utilization} presents information about the usage of all endpoints of interest to the developer,
-    \item \textit{Endpoint Performance} presents response times of the various service endpoints, and
-    \item \textit{User Experience} presents information about the user-perceived performance of monitored service endpoints.
-  \end{enumerate}
-
-  %In the remainder of the paper we present several of these perspectives.\footnote{We recommend obtaining a color version of this paper for better readability}
-
-  % The second endpoint conists of two parts, one of them being a table that shows for every monitored endpoint the number of hits it has gotten, the time it was last accessed and its average execution time.
-The second part is a view with four graphs which show:
-  % \begin{itemize} 
-  %   \item A heatmap of the total number of requests to the monitored endpoints
-  %   \item A stacked bar chart that shows the total number of requests to the monitored endpoints per endpoint per day
-  %   \item A boxplot graph showing the average execution time per version of the web service
-  %   \item A boxplot graph showing the average execution time for every monitored endpoint
-  % \end{itemize}
-
-\todo{Summarize the subsections extensively; evaluate which figures to keep and ideally take new screenshots for the ones we keep; consider adding two more views (like Fig. 3.2 of Patrick's thesis)}
-
-\begin{figure*}[t]
-	\centering
-	\subfloat[number of requests per endpoint per day]{\includegraphics[width=.9\columnwidth]{number_of_requests_}\label{fig:aeu}}
-	\subfloat[requests per hour heatmap]{\includegraphics[width=\columnwidth]{daily_patterns_}\label{fig:dp}}
-	\qquad
-	\subfloat[response time (in ms) per monitored endpoint view]{\includegraphics[width=.8\columnwidth]{endpoint_performance_}\label{fig:ep}}\qquad
-	\subfloat[The Performance Evolution of the \epTranslations endpoint]{\includegraphics[width=.95\columnwidth]{translation_endpoint_evolution_}\label{fig:tee}}
-	\caption{Some of the available views\label{fig:views}}
-\end{figure*}
-
-\paragraph{Service Utilization}
-\label{sec:util}
-
-The most fundamental insight that a service maintainer needs regards service utilization. %\vspace{0.5cm}
-
-Figure \ref{fig:aeu} shows a first perspective on endpoint utilization that \tool provides: a stacked bar chart of the number of hits to various endpoints grouped by day\footnote{Endpoint colors are the same in different views}.
-Figure~\ref{fig:aeu} in particular shows that at its peak the API has about 12.000 hits per day.
-The way users interact with the platform can also be inferred since the endpoints are indicators of different activity types, e.g.:
-
-\begin{enumerate}
-
-	\item {\color{myred} \epTranslations} is an indicator of the amount of foreign language reading the users are doing, and
-
-	\item {\color{mygreen} \epOutcome} is an indicator of the amount of foreign vocabulary practice the users are doing.
-
-\end{enumerate}
-
-%\begin{figure}[!ht]
-%	\centering
-%	\includegraphics[width=\linewidth]{number_of_requests_}
-%	\caption{The number of requests per endpoint per day view shows the overall utilization of the monitored application}
-%	\label{fig:aeu}
-%\end{figure}
-
-Besides showing the overall utilization, this endpoint provides the maintainer with information relevant for decisions regarding endpoint deprecation --- one of the most elementary ways of {\em understanding the needs of the downstream}\cite{Haen14a}.
-In our case study, the maintainer realized that one endpoint which they thought was not being used (i.e.
-\code{words\_to\_study}), contrary to their expectations, was actually being used\footnote{A complementary type of usage information can also be discovered in the view presented in Figure \ref{fig:sep} where seeing that an endpoint is never accessed can increase the confidence of the maintainer that a given endpoint is not used, although it can never be used a proof.}.
-
-\niceseparator
-
-%   \todo{Add the time series graph and discuss it before the heatmap? We can then sell the heatmap better} 
-%   \ml{Not sure about which graph you refer to here V}
-
-A second type of utilization question that the \tool can answer automatically regards {\em cyclic patterns of usage per hour of day} by means of a heatmap, as in \Fref{fig:dp}.
-
-% \mltp{ can we add vertical lines that highlight the beginning of a new week (e.g. before Sunday): }
-% Patrick: adding separators in the graph is unfortunately not supported by the library.
-
-%\begin{figure}[!ht]
-%	\centering
-%	\includegraphics[width=0.9\linewidth]{daily_patterns_}
-%	\caption{Usage patterns become easy to spot in the requests per hour heatmap}
-%	\label{fig:dp}
-%\end{figure}
-
-Figure \ref{fig:dp} shows the API not being used during the early morning hours, with most of the activity focused around working hours and some light activity during the evening.
-This is consistent with the fact that the current users are all in the central European timezone.
-Also, the figure shows that the spike in utilization that was visible also in the previous graph happended in on afternoon/evening.
-
-\paragraph{Endpoint Performance}
-\label{sec:perf}
-
-  The \tool also collects information regarding endpoint performance. The view in \Fref{fig:ep} summarizes the response times for various endpoints by using a box-and-whiskers plot.
-
-  % \ml{Thijs and Patrick...
-the visualization people will complain when they see that we have different colors for the same endpoint in different graphs.
-Can we insure that there is consistency in colors?
-Simplest trick would be to obtain the color by hashing the name of the endpoint...
-in that case the same endpoint would have the same color in various graphs.}
-  % This is supported in the latest version of the dashboard
-
-%  \begin{figure}[!ht]
-%    \centering
-%    \includegraphics[width=0.95\linewidth]{endpoint_performance_}
-%    \caption{The response time (in ms) per monitored endpoint view allows for identifying performance variability and balancing issues}
-%    \label{fig:ep}
-%  \end{figure}
-
-  From this view it became clear to the maintainer that four of the endpoints had very large variation in performance.
-The most critical for the application and consequently the one optimized first was the \epTranslations endpoint which was part of an interactive loop in the reader applications that relied on the Zeeguu API.
-Moreover, cf.
-\Fref{fig:aeu} this endpoint is one of the most used in the system.
-
-  \niceseparator
-
-  \todo{remove the part about how to configure git, the discussion has been subsumed by Section~\ref{sec:tool}.}
-
-  However, with the current configuration of the tool, it would be impossible for the maintainer to see the improvements resulting from the optimization.
-One way to do this is to add an extra line of configuration to allow \tool to find the git\footnote{\url{https://git-scm.com/}} folder of the deployed service:
-
-%  \begin{lstlisting}[caption=Configuring the \tool with the path to the .git folder enables the generation of evolutionary performance graphs, style=custompython]
-%
-%dashboard.config.git = 'path/to/.git'
-%
-%
-%    \end{lstlisting}
-
-    \begin{lstlisting}[style=custompython]
-
-  dashboard.config.git = 'path/to/.git'
-
-  \end{lstlisting}
-
-  With this extra configuration, the \tool can now automatically detect the current version of the project, and group measurements by version\footnote{Alternatively, the maintainer can add version identifiers manually for the web application through a configuration file if the system does not use git.}.
-\Fref{fig:tee} is a zoomed-in version of such a view for \epTranslations with versions increasing from top to bottom
-
-%    \begin{figure}[h!]
-%      \centering
-%      \includegraphics[width=0.95\linewidth]{translation_endpoint_evolution_}
-%      \caption{Visualizing The Performance Evolution of the \epTranslations endpoint}
-%      \label{fig:tee}
-%    \end{figure}
-
-  This view confirms that the performance of the translation endpoint improved in the recent versions: the median of the last three versions is constantly moving towards the left, and progresses from 1.4 seconds (in the top-most box plot in \Fref{fig:tee}) to 0.8 in the latest version (bottom-most box plot).
-
-\paragraph{User Experience}
-\label{sec:user}
-
-For service endpoints which run computations in real time, the maintainer of a system might want to understand the endpoint performance on a per-user basis, especially for situations where the system response time is a function of some individual user load\footnote{E.g.
-in GMail some users have two emails while other have twenty thousand and this induces different response times for different users}.
-
-\todo{remove the configuration part and rephrase the following, the discussion has been partially subsumed by Section~\ref{sec:tool}.}
-
-To enable this, the \tool must be configured to associate an API call with a given user.
-The simplest way is to take advantage of the architecture of Flask applications in which a global \code{flask.request} object can be used to retrieve the session which can in turn lead to user identification:
-
-%\begin{lstlisting}[float,caption=Simply define a custom app-specific function for user retrieval and pass it to the \tool to group information by user,style=custompython]
-\begin{lstlisting}[style=custompython]  
-# app specific way of extracting the user
-# from a flask request object    
-def get_user_id():
-sid = int(flask.request.args['session'])
-session = User.find_for_session(sid)
-return user_id
-
-# attaching the get_user_id function
-dashboard.config.get_group_by = get_user_id
-
-\end{lstlisting}
-
-% \niceseparator
-
-\todo{consider losing one or both of the figures --- or at least updating them}
-
-In Zeeguu, \epFeedItems retrieves a list of recommended articles for a given user.
-Cf.
-\Fref{fig:ep} it is the endpoint with the slowest response time and highest variability.
-The reason for this is that a user can be subscribed to anything from one to three dozen article sources and for each of the sources the system must compute the personalized difficulty of each article at every request.
-
-% In the Zeeguu case study, one of the slowest endpoints, and one with the highest variability as shown in \Fref{fig:ep} is \epFeedItems: it retrieves a list of recommended articles for a given user.
-However, since a user can be subscribed to anything from one to three dozen article sources, and since the computation of the difficulty is personalized and it is slow, the variability in time among users is likely to be very large.
-
-\begin{figure}[!ht]
-	\centering
-	\includegraphics[width=0.85\linewidth]{time_per_user}
-	\caption{The \epFeedItems shows a very high variability across users}
-	\label{fig:tpu}
-\end{figure}
-
-A \perspective{Per-User Performance} perspective should show the different response times for different users.
-Figure \ref{fig:tpu} presents a subset of the corresponding view in the \tool.
-The figure shows that the response times for this endpoint can vary considerably for different users with some extreme cases where a user has to wait a full minute until their recommended articles are shown\footnote{After seeing this perspective, the maintainer refactored the architecture of the system to move part the difficulty computation out of the interactive loop}.
-
-% \niceseparator
-
-The limitation of the previous view is that it does not present the information also on a per version basis.
-To address this, a different visual perspective entitled \perspective{Multi-Version per-User Performance} can be defined.
-Figure \ref{fig:tuv} presents such a perspective by mapping the average execution time for a given user (lines) and given version (columns) on the area of the corresponding circle.
-The colors represent users.
-The figure shows average performance varying  across users and versions with no clear trend: this is probably because varying user workload (i.e.
-number of sources to which the user is registered) is the reason for the variation in response times.
-
-\begin{figure}[!ht]
-	\centering
-	\includegraphics[width=0.9\linewidth]{time_per_user_per_version}
-	\caption{This perspective shows that the evolution of response times for individual users (horizontal lines) across versions (the x-axis)}
-	\label{fig:tuv}
-\end{figure}
-
-%\section{Preemptive Monitoring}
-%\label{sec:testing}
-%
-%%\todo{discuss here idea of using unit testing as an early performance indicator for evolving services (`preemptive monitoring'?); add a side-by-side figure comparing unit testing vs live system performance per version, maybe calculate error in trend curves?
-find a better term than prediction for the section title (plus introduction)}
-%
-%\todo{Similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}, i.e.~testing service-based applications by using dedicated test input in parallel to its normal use and operation.
-The difference is that we take advantage of the capability of Travis to create an emulated live environment for integration testing purposes, and use unit testing as the dedicated test input in order to measure performance.
-What we are going provide evidence for in the following is that we can indeed use this preemptive monitoring concept as an early performance indicator for evolving services.}
-%
-%The developer needs to define the unit test folder of the project, the URL that Travis should use for reporting its results, and the number of iterations for each unit test should be executed in order to generate
-
-\section{Evaluation}
-\label{sec:evaluation}
-
-%\todo{evaluation of the \tool across the following dimensions: ease of use, effectiveness, overhead to the system; for the last part we need to do the overhead measurement experiment, for the other two we base it on the case study and bring over whatever is necessary from the discussion section}
-
-\subsection{Effectiveness \& Ease of Use}
-
-\todo{@M: move here the war stories, and discuss how useful the whole thing was to the application maintainer}
-
-\subsection{Efficiency of Preemptive Monitoring}
-
 \todo{here goes the discussion about our forecasting capabilities}
 
-\begin{figure}
+\begin{figure}[h!]
   \centering
@@
 
-\subsection{Discussion}
+    \begin{figure*}[ht!]
+      \centering
+      \includegraphics[width=0.8\linewidth]{interceptor}
+      \caption{The first thing that needs to be done, is to decorate the application object with an INTERCEPTOR}
+      \label{fig:sep}
+    \end{figure*}
 
-\todo{use this subsection to discuss limitations (from the previous evaluation section) and add the need to be able to handle the horizontal scaling of the application; remove sectioning and summarize briefly limitations that have already been discussed in the VISSOFT paper/move material to other sections; consider renaming as limitations or something similar}
-
-  \subsubsection{Automatically Monitoring System Evolution}
-
-  The main goal of the \tool design was to allow analytics to be collected and insight to be gained by making the smallest possible changes to a running API.
-%To allow the collection of evolutionary information
-%
-  This technique assumes that the web application code which is the target of the monitoring is deployed using \git in the following way:
-
-  \begin{enumerate}
-    \item The deployment engineer pulls the latest version of the code from the integration server; this will result in a new commit being pointed at by the HEAD pointer. %than previously
-    \item The deployment engineer restarts the new version of the service.
-At this point, the \tool detects that a new HEAD is present in the local code base and consequently starts associating all the new data points with this new commit\footnote{The \tool detects the current version of the analyzed system the first time it is attached to the application object, and thus, assumes that the Flask application is restarted when a new version is deployed.
-This is in tune with the current version of Flask, but if the web server will support dynamic updates in the future, this might have to be taken into account}.
-  \end{enumerate}
-
-  The advantage of this approach is the need for minimal configuration effort, as discussed in the presentation of the tool.
-The disadvantage is that it will consider on equal ground the smallest of commits, even one that modifies a comment, and the shortest lived of commits, e.g.~a commit which was active only for a half an hour before a new version with a bug fix was deployed, with major and minor releases of the software.
-%as a distinct way of grouping the data points.
-  A mechanism to control which versions are important for monitoring purposes is therefore required to be added.
-%
-  A further possible extension point here is supporting other version control systems (e.g. Mercurial). However, this is a straightforward extension.
-
-  \subsubsection{NEW: API Renames}
-  \ml{just an idea}
-
-    The history would normally be lost in the case of an API rename. (also known as the {\em provenance issue} in source code evolution analysis)
-    In theory we could infer that an API was renamed if it is not to be found anymore, and another one with very similar timing characteristics would be found.
-Such a detection can never be completely sure, but it would help with meeting the needs of the API maintainer...
-
-  \subsubsection{User-Awareness }
-
-    For the situations in which the user information is not available, the \tool tracks by default information about different IPs and in some cases this might be a sufficiently good approximation of the user diversity and identity.
-    %
-
-    The visualizations for the user experiene perspectives as presented in Section \ref{sec:user} have been tested with several hundred users (of which about \activeUserCount were active during the course of the study), but the scalability of the visualizations must be further investigated for web services with tens of thousands of users.
-
-  \subsubsection{Other Possible Groupings}
-
-    There are other groupings of service utilization and performance that could be important to the maintainer, that we did not explore in this paper.
-For example, if the service is using OAuth, then together with every request, in the header of the request there is information about the application which is sending a request.
-Grouping the information by application that sends the request could be important in such a context.
-
-    In general, providing a mechanism that would allow very easy specification of groupings (either as code annotations, as normal code, or as configuration options) is an open problem that \tool and any other similar library will have to face.
-
-\section{Related Work}
-\label{sec:related}
-
-\todo{Refer the reader to~\cite{ghezzi2007run} and ~\cite{metzger2010analytical} for a more extensive discussion}
-
-There is a long tradition of using visualization for gaining insight into software performance.
-Tools like Jinsight \cite{Pauw02a} and Web Services Navigator \cite{Pauw05} pioneered such an approach for Java and for Web Services that communicate with SOAP messages.
-Both have an ``omniscient'' view of the services / objects and their interactions.
-As opposed to them, in our work we present an analytics platform which focuses on monitoring a single Python web service from its own point of view.
-
-From the perspective of service monitoring, our work falls within the server-side run-time monitoring of services ~\cite{ghezzi2007run}.
-While we don't implement the more advanced features of related monitoring solutions like QoS policies driving the monitoring, it presents nevertheless an easy to use approach support improving the performance of web applications.
-
-% \va{Mircea: Consider removing the rest for space...}
-% An existing monitoring tool is Pingdom \footnote{https://www.pingdom.com/company/why-pingdom}, which monitors the uptime of an existing web-service.
-This tool works by pinging the websites (up to 60 times) every minute automatically.
-Thus this creates a lot of overhead and is bound to be noisy since it will also be influenced by the speed of the network connection\footnote{Another problem is that such a tool would }
-
-% \todo{Runscope? Others?}
-
-\section{Conclusion and Future Work}
-\label{sec:conclusions}
-
-\todo{update as appropriate}
-
-In this paper we have shown that it is possible to create a monitoring solution which provides basic insight into web service utilization and performance  with very little effort from the developer.
-The user group that we are aiming for with this work is application developers using Flask and Python to build web applications with limited or no budget for implementing their own monitoring solutions.
-The emphasis is in allowing such users to gain insight into how the performance of the service evolves together with the application itself.
-We believe that the same architecture, and lessons can be applied to other frameworks and other languages.
-
-In the future, we plan to perform case studies with other sytstems, with the goal of discover other needs and to wean out the less useful visualizations in the \tool.
-We plan to also extend the tool towards supporting multiple deployments of the same applications across multiple nodes (e.g.
-for the situations where the application is deployed together with a load balancer).
-Finally, we plan to integrate \tool with unit testing as a complementary source of information about performance evolution.
-
-% references section
-
-\bibliographystyle{abbrv}
-\bibliography{paper}
-
-% that's all folks
-\end{document}
+\newpage

### 2018-04-03 utilization.tex
split the huge file in smaller ones.

%!TEX root=restructured.tex

\newpage
\subsection{Service Utilization}
\label{sec:util}

\ins{Knowing how third parties use one's API is difficult even in the case of static dependencies. In the case of services, there is no other way but monitoring service utilization. \tool introduces a series of perspectives on utilization. }

\ins{

  The most basic possible view shows the cummulative information about all the endpoints and the number of calls to that endpoint (over the lifetime of its tracking as well as in the current day and the last seven days). HERE? ELSEWHERE? Actually, probably better for performance. The importance of the last day and last seven days became clear only after having used the dashboard for several months. 
}

  \begin{figure}[h!]
  \centering
  \includegraphics[width=\linewidth]{basicest-utilization}
  \caption{....}
  \label{fig:basicest}
  \end{figure}

This kind of information is already very useful, and can provide the API maintainer with insight into the evolution of their system. By looking at \Fref{fig:basicest}, which presents the top 7 endpoints (for lack of space) one can already see several patterns: 
- the two most used endpoints stand for the two main activities in the system (doing exercises and asking for translations while reading personalized materials )
- \epUserActivity has been disproportionately been used in the last day and last week
- \epFeedItems has not been used in the last 7 days

  % \begin{figure}[h!]
  % \centering
  % \includegraphics[width=0.5\linewidth]{number_of_requests_}
  % \caption{The number of requests per endpoint per day view shows the overall utilization of the monitored application}
  % \label{fig:aeu}
  % \end{figure}

\begin{figure}[h!]
  \centering
  \subfloat[Daily Utilization]{\includegraphics[width=.42\columnwidth]{number_of_requests_}\label{fig:aeu}}
  \subfloat[Hourly Utilization]{\includegraphics[width=.5\columnwidth]{daily_patterns_}\label{fig:dp}}
  \caption{Some of the available views\label{fig:views}}
\end{figure}

Figure \ref{fig:aeu} shows a first perspective on endpoint utilization that \tool provides: a stacked bar chart of the number of hits to various endpoints grouped by day\footnote{Endpoint colors are the same in different views}. Figure~\ref{fig:aeu} in particular shows that at its peak the API has about 12.000 hits per day.

% The way users interact with the platform can also be inferred since the endpoints are indicators of different activity types, e.g.:

% \begin{enumerate}

%   \item {\color{myred} \epTranslations} is an indicator of the amount of foreign language reading the users are doing, and

%   \item {\color{mygreen} \epOutcome} is an indicator of the amount of foreign vocabulary practice the users are doing.

% \end{enumerate}

% Besides showing the overall utilization, this endpoint provides the maintainer with information relevant for decisions regarding endpoint deprecation --- one of the most elementary ways of {\em understanding the needs of the downstream}\cite{Haen14a}. In our case study, the maintainer realized that one endpoint which they thought was not being used (i.e. \code{words\_to\_study}), contrary to their expectations, was actually being used\footnote{A complementary type of usage information can also be discovered in the view presented in Figure \ref{fig:sep} where seeing that an endpoint is never accessed can increase the confidence of the maintainer that a given endpoint is not used, although it can never be used a proof.}.

% \niceseparator

%   \todo{Add the time series graph and discuss it before the heatmap? We can then sell the heatmap better} 
%   \ml{Not sure about which graph you refer to here V}

A second type of utilization question that the \tool can answer automatically regards {\em cyclic patterns of usage per hour of day} by means of a heatmap, as in \Fref{fig:dp}.

% \mltp{ can we add vertical lines that highlight the beginning of a new week (e.g. before Sunday): }
% Patrick: adding separators in the graph is unfortunately not supported by the library.

%\begin{figure}[!ht]
% \centering
% \includegraphics[width=0.9\linewidth]{daily_patterns_}
% \caption{Usage patterns become easy to spot in the requests per hour heatmap}
% \label{fig:dp}
%\end{figure}

Figure \ref{fig:dp} shows the API not being used during the early morning hours, with most of the activity focused around working hours and some light activity during the evening. This is consistent with the fact that the current users are all in the central European timezone. Also, the figure shows that the spike in utilization that was visible also in the previous graph happended in on afternoon/evening.

### 2018-04-03 evolution.tex
cleaning up

@@
 
-\begin{figure*}[h!]
-  \centering
-  \subfloat[overview of the tracked endpoints]{\includegraphics[width=1.5\columnwidth]{overview}\label{fig:overview}}
-  \caption{Some of the available views\label{fig:views}}
-\end{figure*}
-
-  \newpage
   \section{Evolution monitoring }
@@
   If the \tool can automatically detect the current version of the project by reading the .git configuration as soon as the API is started and can then group measurements by version\footnote{Alternatively, the maintainer can add version identifiers manually for the web application through a configuration file if the system does not use git.}.
+
+  \subsubsection*{Evolving Utilization}
+
+  \Fref{fig:mv-util} shows the \perspective{Multi-Version Utilization} perspective. 
+  It shows the utilization of all the tracked endpoints across versions.
+
+    \begin{figure}[h!]
+      \centering
+      \includegraphics[width=\linewidth]{utilization-evolution}
+      \caption{The Evolution of Endpoint Utilization Across System Versions}
+      \label{fig:mv-util}
+    \end{figure}
+
+    \subsubsection*{Evolving Performance}
+
     \Fref{fig:tee} is a zoomed-in version of such a view for \epTranslations with versions increasing from top to bottom
@@
 
-\begin{figure*}[!ht]
+\subsubsection*{Evolving Groups}
+The limitation of the previous view is that it does not present the information also on a per version basis.
+To address this, a different visual perspective entitled \perspective{Multi-Version per-User Performance} can be defined.
+Figure \ref{fig:tuv} presents such a perspective by mapping the average execution time for a given user (lines) and given version (columns) on the area of the corresponding circle.
+
+\begin{figure}[h!]
   \centering
-  \includegraphics[width=0.4\linewidth]{time_per_user}
-  \caption{The \epFeedItems shows a very high variability across users}
-  \label{fig:tpu}
-\end{figure*}
+  \includegraphics[width=0.7\linewidth]{time_per_user_per_version}
+  \caption{This perspective shows that the evolution of response times for individual users (horizontal lines) across versions (the x-axis)}
+  \label{fig:tuv}
+\end{figure}
+
+The colors represent users.
+The figure shows average performance varying  across users and versions with no clear trend: this is probably because varying user workload (i.e.
+number of sources to which the user is registered) is the reason for the variation in response times.
+\ins{one can see that for user 1 performance degrades over the }

### 2018-04-03 grouping.tex
cleaning up

@@
 
-\newpage
 \section{Grouping Requests}
@@
 in GMail some users have two emails while other have twenty thousand and this induces different response times for different users}.
-
 \todo{remove the configuration part and rephrase the following, the discussion has been partially subsumed by Section~\ref{sec:tool}.}
@@
 
-%\begin{lstlisting}[float,caption=Simply define a custom app-specific function for user retrieval and pass it to the \tool to group information by user,style=custompython]
 \begin{lstlisting}[style=custompython]
-# app specific way of extracting the user
-# from a flask request object    
-def get_user_id():
-sid = int(flask.request.args['session'])
-session = User.find_for_session(sid)
-return user_id
 
-# attaching the get_user_id function
-dashboard.config.get_group_by = get_user_id
+# attaching a group by function 
+dashboard.config.group_by = 'User',
+	lambda: Session.find(flask.request).userid
 
@@
 
-The limitation of the previous view is that it does not present the information also on a per version basis.
-To address this, a different visual perspective entitled \perspective{Multi-Version per-User Performance} can be defined.
-Figure \ref{fig:tuv} presents such a perspective by mapping the average execution time for a given user (lines) and given version (columns) on the area of the corresponding circle.
-
-The colors represent users.
-The figure shows average performance varying  across users and versions with no clear trend: this is probably because varying user workload (i.e.
-number of sources to which the user is registered) is the reason for the variation in response times.
-\ins{one can see that for user 1 performance degrades over the }
-
-\begin{figure}[!ht]
+\begin{figure}[h!]
   \centering
-  \includegraphics[width=0.7\linewidth]{time_per_user_per_version}
-  \caption{This perspective shows that the evolution of response times for individual users (horizontal lines) across versions (the x-axis)}
-  \label{fig:tuv}
+  \includegraphics[width=.8\linewidth]{time_per_user}
+  \caption{The \epFeedItems shows a very high variability across users}
+  \label{fig:tpu}
 \end{figure}

### 2018-04-03 performance.tex
cleaning up

@@
 
+Knowing the performance of one’s API is critical for the quality of a service API.
+Flask Dashboard introduces a series of perspectives on performance, which focus on the response times of the various endpoints.
+
+\ins{HERE? ELSEWHERE? Actually, probably better for performance. The importance of the last day and last seven days became clear only after having used the dashboard for several months. }
+
 The \tool also collects information regarding endpoint performance. The view in \Fref{fig:ep} summarizes the response times for various endpoints by using a box-and-whiskers plot.
-
-In the Zeeguu case study, one of the slowest endpoints, and one with the highest variability as shown in \Fref{fig:ep} is \epFeedItems: it retrieves a list of recommended articles for a given user.
-However, since a user can be subscribed to anything from one to three dozen article sources, and since the computation of the difficulty is personalized and it is slow, the variability in time among users is likely to be very large.
 
@@
 
+In the Zeeguu case study, one of the slowest endpoints, and one with the highest variability as shown in \Fref{fig:ep} is \epFeedItems: it retrieves a list of recommended articles for a given user.
+However, since a user can be subscribed to anything from one to three dozen article sources, and since the computation of the difficulty is personalized and it is slow, the variability in time among users is likely to be very large.
+
   From this view it became clear to the maintainer that four of the endpoints had very large variation in performance.
@@
   \newpage
-  \subsection{Performance Insight: Automated outlier detection and monitoring}
+  \subsection{Automated Outlier Detection and Monitoring}
 
@@
 The bottom panel shows the stack trace. 
-In this particular case, it is revealing for the developer to learn that at the time of the stack trace snapshot, the code was in the microsoft\_translator.py: indeed, the system uses as back-end multiple translators, and it has been observed that many of the outliers happen to be waiting in the microsoft translator.
+In this particular case, it is revealing for the developer to learn that at the time of the stack trace snapshot, the code was in the google\_translator.py: indeed, the system uses as back-end multiple translators, and it has been observed that many of the outliers happen to be waiting in the google translator.
 
@@
 Thus this functionality in microsoft\_translator is really slow in itself, and this is not a result of the machine being overloaded for example.
-
-\begin{figure*}[h!]
-  \centering
-  \includegraphics[width=0.9\linewidth]{utilization-evolution}
-  \caption{The Evolution of Endpoint Utilization Across System Versions}
-  \label{fig:tee}
-\end{figure*}

### 2018-04-03 testing.tex
cleaning up

@@
 
-  \newpage
+  \begin{figure*}[h!]
+  \centering
+  \subfloat[The reported response time (in ms) per deployed version in the observation period]{\includegraphics[width=.6\columnwidth]{response_per_version_trunced_trend}}
+  \quad
+  \subfloat[The measured response time (in ms) using integration with Travis]{\includegraphics[width=.56\columnwidth]{travis_builds_no_outliers_trend}}
+  \caption{Comparison of the response times per endpoint: actual production system versus preemptive monitoring data}
+  \label{fig:preemptive}
+\end{figure*}
+
   \section{Preemptive monitoring}
@@
 
-  For example, \Fref{fig:builds} is a screenshot from the dashboard showing the measured response times for 5 consequent builds, with 180 iterations of the unit tests executed in total across all endpoints.
-The outliers on the right of the figure are due to initial requests \ins{which must wait for a boot-up phase of the API}.
-
       \begin{figure}[h!]
         \centering
-        \includegraphics[width=\columnwidth]{travis_builds}
+        \includegraphics[width=0.7\columnwidth]{travis_builds}
         \caption{Response times in 5 consequent Travis builds in the \zee case study}
@@
       \end{figure}
+
+  For example, \Fref{fig:builds} is a screenshot from the dashboard showing the measured response times for 5 consequent builds, with 180 iterations of the unit tests executed in total across all endpoints.
+The outliers on the right of the figure are due to initial requests \ins{which must wait for a boot-up phase of the API}.
 
@@
 
-\begin{figure}[h!]
-  \centering
-  \subfloat[The reported response time (in ms) per deployed version in the observation period]{\includegraphics[width=.5\columnwidth]{response_per_version_trunced_trend}}
-  \quad
-  \subfloat[The measured response time (in ms) using integration with Travis]{\includegraphics[width=.46\columnwidth]{travis_builds_no_outliers_trend}}
-  \caption{Comparison of the response times per endpoint: actual production system versus preemptive monitoring data}
-  \label{fig:preemptive}
-\end{figure}
-
 \begin{table}[h]

### 2018-04-03 utilization.tex
cleaning up

@@
 
-  The most basic possible view shows the cummulative information about all the endpoints and the number of calls to that endpoint (over the lifetime of its tracking as well as in the current day and the last seven days).
-HERE?
-ELSEWHERE?
-Actually, probably better for performance.
-The importance of the last day and last seven days became clear only after having used the dashboard for several months.
+  The most basic possible view shows the cummulative information about all the endpoints and the number of calls to that endpoint over the lifetime of its tracking as well as in the current day and the last seven days.
 }
@@
 By looking at \Fref{fig:basicest}, which presents the top 7 endpoints (for lack of space) one can already see several patterns:
-- the two most used endpoints stand for the two main activities in the system (doing exercises and asking for translations while reading personalized materials )
-- \epUserActivity has been disproportionately been used in the last day and last week
-- \epFeedItems has not been used in the last 7 days
+
+\begin{itemize}
+
+  \item {\bf Frequent Patterns of System Usage}. The two most used endpoints stand for the two main activities in the system:
+
+  \begin{itemize}
+
+    \item \epTranslations is an indicator of the amount of foreign language reading the users are doing
+
+    \epOutcome is an indicator of the amount of foreign vocabulary practice the users are doing.
+
+  \end{itemize}
+
+  \item {\bf Sudden Increase of Endpoint Usage}. One endpoint (\epUserActivity) has been disproportionately been used in the last day and last week; much more than before.
+
+  \item {\bf Possibly Discontinued Endpoint Usage}. One endpoint (\epFeedItems) has not been used in the last 7 days
+
+\end{itemize}
+
+\lesson{Although very useful, it was only after several months of using the system that the client requested the extra columsn for one day and seven days.}
 
@@
 
+The table view is limited, and the one day and one week periods are conveniently but arbitrarily selected.
+For a more detailed evolution of utilization, Figure \ref{fig:aeu} shows the \perspective{Daily Utilization} perspective on endpoint utilization that \tool provides: a stacked bar chart of the number of hits to various endpoints grouped by day\footnote{Endpoint colors are the same in different views}.
+Figure~\ref{fig:aeu} in particular shows a peak utilization, a day when the API had more than 12.000 hits.
+
 \begin{figure}[h!]
@@
 
-Figure \ref{fig:aeu} shows a first perspective on endpoint utilization that \tool provides: a stacked bar chart of the number of hits to various endpoints grouped by day\footnote{Endpoint colors are the same in different views}.
-Figure~\ref{fig:aeu} in particular shows that at its peak the API has about 12.000 hits per day.
-
 % The way users interact with the platform can also be inferred since the endpoints are indicators of different activity types, e.g.:
-
-% \begin{enumerate}
-
-%   \item {\color{myred} \epTranslations} is an indicator of the amount of foreign language reading the users are doing, and
-
-%   \item {\color{mygreen} \epOutcome} is an indicator of the amount of foreign vocabulary practice the users are doing.
-
-% \end{enumerate}
 
@@
 
-A second type of utilization question that the \tool can answer automatically regards {\em cyclic patterns of usage per hour of day} by means of a heatmap, as in \Fref{fig:dp}.
-
-% \mltp{ can we add vertical lines that highlight the beginning of a new week (e.g. before Sunday): }
-% Patrick: adding separators in the graph is unfortunately not supported by the library.
+Another utilization perspective is the \perspective{Hourly Utilization} in which  the \tool can highlight {\em cyclic patterns of usage per hour of day} by means of a heatmap, as in \Fref{fig:dp}.
 

### 2018-04-03 appendix.tex
done or the day

--- 2018-04-03 restructured.tex
@@
-\documentclass[conference]{IEEEtran}
-	\pdfpagewidth=8.5truein
-	\pdfpageheight=11truein
-
-\usepackage{graphicx}
-\usepackage{subfig}
-\usepackage[bookmarks=false]{hyperref}
-\usepackage{xspace}
-\usepackage{listings}
-\usepackage[usenames, dvipsnames]{color}
-
-\usepackage{booktabs}
-
-\graphicspath{{./img/}}
-
-%\hyphenation{op-tical net-works semi-conduc-tor}
-
-\input{useful_defines}
-
-\begin{document}
-
-%
-\title{Agile Monitoring of Evolving Web Services}
-
-\author{
-\IEEEauthorblockN{The Author, The Other Author}\\
-Institute\\
-University, Country\\
-Email: \{author\}@university.domain 
-}
-
-% make the title area
-\maketitle
-
-\input{introductory}
-
-\input{loc1}
-
-\input{utilization}
-
-\input{performance}
-
-\input{grouping}
-
-\input{evolution}
-
-\input{testing}
-
-\section{Performance Perspectives}
-\label{sec:views}
-
-\ins{PERFORMANCE DISCUSSION: this introspection, which looks up endpoints, happens only once at the startup of the service, so it is going to affect the API startup performance, but not the actual endpoint performance.
-In our case, the overhead here was quite small: TO MEASURE}.
-
-  %In the remainder of the paper we present several of these perspectives.
-
-  % The second endpoint conists of two parts, one of them being a table that shows for every monitored endpoint the number of hits it has gotten, the time it was last accessed and its average execution time.
-The second part is a view with four graphs which show:
-  % \begin{itemize} 
-  %   \item A heatmap of the total number of requests to the monitored endpoints
-  %   \item A stacked bar chart that shows the total number of requests to the monitored endpoints per endpoint per day
-  %   \item A boxplot graph showing the average execution time per version of the web service
-  %   \item A boxplot graph showing the average execution time for every monitored endpoint
-  % \end{itemize}
-
-\todo{evaluate which figures to keep and ideally take new screenshots for the ones we keep; consider adding two more views (like Fig. 3.2 of Patrick's thesis)}
-
-% In the Zeeguu case study, one of the slowest endpoints, and one with the highest variability as shown in \Fref{fig:ep} is \epFeedItems: it retrieves a list of recommended articles for a given user.
-However, since a user can be subscribed to anything from one to three dozen article sources, and since the computation of the difficulty is personalized and it is slow, the variability in time among users is likely to be very large.
-
-% \niceseparator
-
-%\section{Preemptive Monitoring}
-%\label{sec:testing}
-%
-%%\todo{discuss here idea of using unit testing as an early performance indicator for evolving services (`preemptive monitoring'?); add a side-by-side figure comparing unit testing vs live system performance per version, maybe calculate error in trend curves?
-find a better term than prediction for the section title (plus introduction)}
-%
-%\todo{Similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}, i.e.~testing service-based applications by using dedicated test input in parallel to its normal use and operation.
-The difference is that we take advantage of the capability of Travis to create an emulated live environment for integration testing purposes, and use unit testing as the dedicated test input in order to measure performance.
-What we are going provide evidence for in the following is that we can indeed use this preemptive monitoring concept as an early performance indicator for evolving services.}
-%
-%The developer needs to define the unit test folder of the project, the URL that Travis should use for reporting its results, and the number of iterations for each unit test should be executed in order to generate
-
-\section{Evaluation}
-\label{sec:evaluation}
-
-%\todo{evaluation of the \tool across the following dimensions: ease of use, effectiveness, overhead to the system; for the last part we need to do the overhead measurement experiment, for the other two we base it on the case study and bring over whatever is necessary from the discussion section}
-
-\input{discussion}
-
-\input{related-work}
-
-\section{Conclusion and Future Work}
-\label{sec:conclusions}
-
-\todo{update as appropriate}
-
-In this paper we have shown that it is possible to create a monitoring solution which provides basic insight into web service utilization and performance  with very little effort from the developer.
-The user group that we are aiming for with this work is application developers using Flask and Python to build web applications with limited or no budget for implementing their own monitoring solutions.
-The emphasis is in allowing such users to gain insight into how the performance of the service evolves together with the application itself.
-We believe that the same architecture, and lessons can be applied to other frameworks and other languages.
-
-In the future, we plan to perform case studies with other sytstems, with the goal of discover other needs and to wean out the less useful visualizations in the \tool.
-We plan to also extend the tool towards supporting multiple deployments of the same applications across multiple nodes (e.g.
-for the situations where the application is deployed together with a load balancer).
-Finally, we plan to integrate \tool with unit testing as a complementary source of information about performance evolution.
-
-\newpage
 \appendix
@@
       \end{lstlisting}
-
-% references section
-
-\bibliographystyle{abbrv}
-\bibliography{paper}
-
-% that's all folks
-\end{document}

### 2018-04-03 paper.tex
done or the day

--- 2018-03-03 paper.tex
@@
 \usepackage[usenames, dvipsnames]{color}
-\usepackage{cite}
-\usepackage{caption}
+\usepackage{amssymb}
+
 \usepackage{booktabs}
@@
 \begin{document}
-\title{The Importance of Visualization in Performance Monitoring of Python Web Services}
-% Alternative Titles: A Low-Effort Analytics Platform for Visualizing Evolving Flask-Based Python Web Services
+
+%
+\title{Agile Monitoring of Evolving Web Services}
 
 \author{
-\IEEEauthorblockN{Patrick Vogel, Thijs Klooster, Vasilios Andrikopoulos, Mircea Lungu}\\
-Johann Bernoulli Institute for Mathematics and Computer Science\\
-University of Groningen, Netherlands\\
-Email: \{t.klooster.1,p.p.vogel\}@student.rug.nl, 
-\{v.andrikopoulos,m.f.lungu\}@rug.nl 
+\IEEEauthorblockN{The Author, The Other Author}\\
+Institute\\
+University, Country\\
+Email: \{author\}@university.domain 
 }
@@
 
-\begin{abstract}
-% TODO: Update abstract
-  Tens of thousands of web applications are written in Flask, a Python-based web framework.
-Despite a rich ecosystem of extensions, there is none that supports the developer in gaining insight into the evolving performance of their service.
-In this paper, we introduce \tool, a library that addresses this problem.
-We present the ease with which the library can be integrated in an already existing web application, discuss some of the visualization perspectives that the library provides and point to some future challenges for similar libraries.
-\end{abstract}
+\input{introductory}
 
-\section{Introduction}
+\input{loc1}
 
-% \todo{Intro to be rewritten; refer to VISSOFT paper as previous work, explain focus is on integration with version control/CI for evolution support, and use of unit tests as early performance indicator}
+\input{utilization}
 
-Networked computer systems play important roles in the operation of many businesses and organizations \cite{davenport1998working}. 
-The performance of a computer system providing services to a business or customers of a business is integral to the successful operation of the business.
-And, as every system is nowadays a distributed system \cite{cavage2013there}, it is often hard to achieve the desired quality of those systems. 
-Therefore, monitoring those services is essential because it allows to perform three main types of actions: system adaptations to provide the request service at the desired level of quality, enabling flexibility in dealing with changing requirements and modes of operation, and enabling operational awareness through dashboards organizing the collected data~\cite{pernici2016monitoring}.
+\input{performance}
 
-In this work we focus on the last part and we discuss how a minimal effort probe-based solution for a particular type of web services can be used to facilitate the other two types of actions by involving the developer.
+\input{grouping}
 
-In \cite{vogel2017low} we introduced the \tool on a high level with a focus on presenting its performance visualization aspects; in this work we only summarize these features by means of introducing some of the provided functionalities of the \tool.
-We focus on discussing how the tool is implemented and operating, provide a deeper look at its capabilities for integration with version control and continuous integration environments, introduce a mechanism for performance prediction, and provide an evaluation of the proposed approach based on a case study.
+\input{evolution}
 
-%{\em There is no getting around it: you are building a distributed system} argues a recent article \cite{cavage2013there}.
-Indeed, even the simplest second-year student project is a web application implemented as two-tier architecture with a Javascript/HTML5 front-end a service backend, usually a REST API.
+\input{testing}
 
-% \hfill mds
-% Many contemporary programming languages are offering libraries, modules, or frameworks that facilitate the development of such architectures. 
-Python is one of the most popular programming language choices for implementing the back-end of web applications.
-GitHub contains more than 500K open source Python projects and the Tiobe Index\footnote{TIOBE programming community index is a measure of popularity of programming languages, created and maintained by the TIOBE Company based in Eindhoven, the Netherlands} ranks Python as the 4th most popular programming language as of February 2018 \cite{tiobe}.
-
-Within the Python community, Flask\footnote{\url{http://flask.pocoo.org/}} is a very popular web framework\footnote{More than 35K projects on GitHub are implemented with Flask (cf.
-a GitHub search for ``language:Python Flask'')}.
-It provides simplicity and flexibility by implementing a bare-minimum web server, and thus advertises as a micro-framework.
-The Flask tutorial shows how setting up a simple Flask {\em ``Hello World''} web-service requires no more than 5 lines of Python code \cite{ flask:tutorial}.
-
-Despite their popularity, to the best of our knowledge, there is no simple solution for monitoring the evolving performance of Flask web applications.
-Thus, every one of the developers of these projects faces one of the following options when confronted with the need of gathering insight into the runtime behavior of their implemented services:
-
-  \begin{enumerate}
-
-    \item Use a commercial monitoring tool which treats the subject API as a black-box (e.g. Pingdom, Runscope). 
-    % , Graphite+Graphana+statd etc.
-
-    \item Implement their own ad-hoc analytics solution, having to reinvent basic visualization and interaction strategies.
-
-    \item Live without analytics insight into their services.
-
-  \end{enumerate}
-
-%\todo{For the first point in the list, we can also argue that analytics solutions like Google Analytics can be used, but they have no notion of versioning/integration with the development life cycle.
-Feel free to cite \cite{papazoglou2011managing} for service evolution purposes}
-
-For projects on a budget (e.g.
-research, startups) the first and the second options are often not available due to time and financial constraints.
-Even when using 3rd-party analytics solutions, a critical insight into the evolution of the exposed services of the web application, is missing because such solutions have no notion of versioning and no integration with the development life cycle.~\cite{papazoglou2011managing}
-
-To avoid projects ending up in the third situation, that of living without analytics, in this paper we present \tool~ --- a low-effort service monitoring library for Flask-based Python web services that is easy to integrate and enables the {\em agile assessment} of service evolution.
-\cite{Nier12b}
-
-As a case study, on which we will illustrate our solution, we are going to use an open source API which, for several years, was in the third of the above-presented situations.
-
-% In the next section, we will present a case study of an open source research API which was for a long time in the third situation presented above -- deployed without analytics insight.
-
-\todo{Add signposting paragraph}
-
-\section{Case Study}
-\label{sec:case}
-  \zee\footnote{\url{https://github.com/zeeguu-ecosystem/}} is a platform and an ecosystem of applications for accelerating vocabulary acquisition in a foreign language \cite{Lungu16}. 
-The architecture of the ecosystem has at its core an API and a series of satellite applications that together offer to a learner three main inter-dependent features:
-
-  \begin{enumerate}
-
-    \item Reader applications that provide effortless translations for those texts which are too difficult for the readers.
-
-    \item Interactive exercises personally generated based on the preferences and past likes of the learner.
-
-    \item Article recommendations which are personalized for the interests of the learner and come with difficulty estimation that helps the learner find articles with the appropriate difficulty.
-
-  \end{enumerate}
-
-  The core API implemented with Flask and Python provides correspondingly three types of functionality: contextual translations, article recommendations, and personalized exercise suggestions.
-In total the API provides a bit less than 50 endpoints, out of which probably a dozen are very frequently used.
-  % The development of the core API itself is a research project.
-
-  At the time of writing this article, the ecosystem consists of a reader web application, a web based exercises platform, and a smartwatch application, which are used at the moment of writing this article by more than \activeUserCount active users.
-The users come from a highschool, a language school, and some users are using it on their own, without any educational context.
-The highest load we observed until now on the API consisted of 12K requests in one day.
-
-  We will use this \zee API as a case study for this paper. 
-  All the figures in this paper are captured from the actual deployment of \tool in the context of the \zee API.
-The figures are interactive offering basic data exploration capabilities: filter, zoom, and details on demand\cite{Shne99a}.
-The \tool deployment for the case study can be accessed publicly\footnote{\url{https://zeeguu.unibe.ch/api/dashboard}; username: {\em guest}, password: {\em soap-sac}}.
-
- % \todo{create new account, update URL if necessary, update activeUserCount, add a couple of sentences of the current state of the case study with the college/language center}
-% \ml{we should consider adding also one section in which the architecture/implementation and main features of the dashboard are presented before going on with discussing them in more depth in the following sections --- this should include a rundown on which views are provided from where (overview or per endpoint)}
-
-\section{The Flask Dashboard}
-\label{sec:tool}
-
-%\todo{Refer to VISSOFT paper~\cite{vogel2017low} and rephrase where possible; add more info on implementation, e.g. annotation mechanism, database design, session handling etc.}
-
-  As discussed in the introductory section, in previous work~\cite{vogel2017low} we introduced \tool, a drop-in Flask Extension in Python that allows developers to monitor their web applications with minimal effort.
-%
-  The \tool is implemented in Python and works in Python 2.7 up to Python 3.6 and is available on the Python Package Index repository\footnote{\url{https://pypi.org/project/Flask-MonitoringDashboard/}} from where it can be installed by running \install from the command line.
-%  
-  The source code of the \tool is published under a permissive MIT license and is available on GitHub\footnote{\url{https://github.com/flask-dashboard/flask_monitoringdashboard/}}.
-
-  \paragraph{Usage \& Configuration}
-
-  The fact that the \tool itself is being developed in Python using Flask makes binding to the services of a to-be monitored application developed with the same technologies easy and intuitive.
-Thus, there are no external dependencies needed (apart from the \tool itself).
-This leads to minimal effort, as the configuration is the only thing that has be done.
-
-  To start using the \tool, one simply needs to add two lines of code to their Flask web service:
-
-  \begin{lstlisting}[style=custompython, caption=Configuring the \tool is straightforward]
-  import flask_monitoringdashboard as dashboard
-
-  # flask_app is the Flask app object
-  dashboard.bind(flask_app)
-  \end{lstlisting}
-
-  After (re)deployment of the application, the \tool becomes available at the \code{/dashboard} route of the Flask application.
-Further configuration is possible by adding additional statements before the binding definition, for example:
-
-  \begin{lstlisting}[style=custompython]
-  ...
-  dashboard.config.link = 'mydashboard'
-  dashboard.bind(flask_app)
-  \end{lstlisting}
-
-  allows for a custom route (\code{/mydashboard}) to the dashboard to be defined by the programmer. An external configuration file can be used instead, or in addition to this:
-
-  \begin{lstlisting}[style=custompython]
-  ...
-  dashboard.config.init_form(file='config.cfg')
-  ...
-  \end{lstlisting}
-
-  During binding, the \tool will iterate through all endpoints defined in the target application.
-These will be presented to the user in the tool web interface, where the user can select the ones that should be monitored, see \Fref{fig:sep}.
-
-    \begin{figure}
-      \centering
-      \includegraphics[width=\linewidth]{selecting_endpoints.png}
-      \caption{Endpoints that are available for monitoring in the \zee API}
-      \label{fig:sep}
-    \end{figure}
-
-  In order to monitor an endpoint, the \tool attaches a function wrapper for the API function that corresponds to the endpoint.
-An alternative for monitoring endpoints would be to wrap the entire application into a middleware \footnote{This is presented in \cite{Zeng}.}.
-However, this alternative results in an overhead for the entire application, while using a function wrapper only provides overhead for the endpoints that are wrapped.
-Thus, the wrapper will be executed only whenever the corresponding API call is made.
-The wrapper contains the code that takes care of monitoring an endpoint.
-Data collected by the wrappers are persisted in a local database.
-The SQLAlchemy Object Relational Mapper\footnote{\url{https://www.sqlalchemy.org/}} on top of SQLite\footnote{\url{https://www.sqlite.org/}} is used for this purpose.
-
-  In \Fref{fig:sep} the last access time of the endpoint by users, irrespective of whether it is selected by the user to be monitored or not, is also displayed.
-This has the means of identifying potentially obsolete endpoints.
-This functionality is implemented using a different wrapper.
-But, since this wrapper is attached to every endpoint, it introduces some overhead for the entire application.
-
-% PV: Maybe remove paragraph and figure, as it doesn't make really sense in this place.
-Apart form the monitoring-API, a visualization-API is provided.
-The functionality of this API is to present the results that are collected in the former API to the user of the \tool.
-The visualization-API is further explained in Section \ref{sec:views}, but can already be seen in Figure \ref{fig:overview}.
-
-    \begin{figure}
-      \centering
-      \includegraphics[width=\linewidth]{overview.png}
-      \caption{Overview of the endpoints that are monitored in \zee API}
-      \label{fig:overview}
-    \end{figure}
-
-\paragraph{Automated outlier detection and monitoring}
-  When an API is called from within a highly interactive application (as it is the case with the Zeeguu ecosystem) of particular interest to the API developers are performance {\em outliers}.
-We consider a request an outlier, if processing that requests takes three times more than the {\em running average} for that endpoint\footnote{It is also possible to configure a different number for detecting outliers.}.
-Identifying and collecting all appropriate data, and diagnosing the root causes of such outliers is especially critical in improving the quality of an application.
-
-  An assumption that have to be made is that the response time values are normally distributed, otherwise more general density distribution information must be collected in real time.
-When it detects that a given request is an outlier with respect to this past average running value, it triggers the {\em outlier data collection routine} which stores extra information, such as the current stack trace, all request variables, memory consumption, and CPU load at the moment of processing the request.
-This allow the maintainer to investigate the causes of these exceptionally slow response times.
-In this way it is possible to get detailed insight into the operation of the application in the extreme cases without unnecessarily burdening it with logging this information for every request.
-
-  \paragraph{User-level monitoring}
-  \label{sec:user-monitoring}
-  Another possibility for the occurrence of outliers is due to differences in response time per user.
-For example, Gmail engineers admitted in 2010 that the performance of their service was dependent on the amount of emails a user would have in their Inbox \cite{gmail}.
-Therefore, it is important to keep track which request belongs to which user.
-This provides the functionality for the developer to understand the performance of their service for individual users, or for different user groups.
-However, since Flask itself support no functionality for user-management, it is not possible for the \tool to retrieve the corresponding user for a certain request.
-Neither it is an option for storing this in the configuration file, due to retrieving the user-information is only possible in the code itself.
-
-This leads to a functionality in which the API developer can pass a function that retrieves the information per user.
-This implements a way of distilling a user or a user group from the API request.
-Another advantage of specifying the function by the API developer is that the results can be clustered on more than users only.
-In the example above, the developers might group on email-count, which allows them to find a relation between this and the execution time.
-For a large number of Flask applications that use the global \code{flask.request} object, the object can be used to retrieve session information, as in the following snippet:
-  % .
-This ability can be used in order to allow, for example, the identification of application users from the \tool, and the consequent aggregation of monitoring information on a per user basis as follows:
-
-  \begin{lstlisting}[style=custompython,caption=App specific way of extracting the user from a Flask request object]
-  def get_user_id():
-    sid = int(flask.request.args['session'])
-    session = User.find_for_session(sid)
-    return user_id
-
-  # attaching the get_user_id function
-  dashboard.config.get_group_by = 
-      get_user_id
-
-  \end{lstlisting}
-  %This feature, as discussed further in the following section, is important for the clustering of users depending on their application usage.
-It also provides a view on the distribution of the work load across the application endpoints on a per user basis that allows the identification of more or less popular endpoints for further performance improvements.
-\ml{this previous paragraph seems a bit reduntant.
-must revisit}
-
-  \paragraph{Evolution monitoring for Git-versioned projects}
-  It is not only useful to find relations in processing endpoints across users, but it is also important to detect whether the overall performance of the web-service decreases during its evolution, which is likely to happen according to \cite{Kaminski:2006:DAW:1137677.1137694}.
-There are two ways in \tool to support version control:
-
-\begin{enumerate}
-  \item  \textit{Manual} version control requires the developer to tag each new version of the application with an appropriate version identifier~\cite{papazoglou2011managing} using the \code{APP\_VERSION} configuration parameter, for example by adding to the configuration file:
-
-    \begin{lstlisting}[style=custompython,caption=Manually adding version control]
-    [dashboard]
-    APP_VERSION=1.2.3
-    \end{lstlisting}
-
-  \item \textit{Automatic} version control, aimed for use in conjunction with a version control system and currently only supporting Git\footnote{\url{https://git-scm.com/}}, is enabled by linking to the Git folder in the application source code:
-
-      \begin{lstlisting}[style=custompython]
-      ....
-      dashboard.config.git = '/path/to/.git'
-      ....
-      \end{lstlisting}
-
-    With this extra configuration parameter, the \tool automatically detects the current version of the project by reading the hash code of the deployed version as soon as the application is deployed and started.
-Information about the version history of the application is also persisted in the tool database and is used for clustering the requests result per endpoint.
-
-\end{enumerate}
-
-  \paragraph{Preemptive monitoring}
-
-  The concept of {\em preemptive monitoring} of the application performance by means of instrumenting integration unit testing as the synthetic load is similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}.
-Testing service-based application can be done in parallel to its normal use and operation.
-By using the capability of a Continuous Integration (CI) framework, a ``live'' environment for integration testing purposes can be simulated without affecting the performance of the deployed application.
-After the results on the testing framework have been collected, they are send to the deployed application, on which they are visualized in the \tool.
-
-  For example, \Fref{fig:builds} is a screenshot from the dashboard showing the measured response times for 5 consequent builds, with 180 iterations of the unit tests executed in total across all endpoints.
-The outliers on the right of the figure are due to initial requests, which must wait for a boot-up phase of the API.
-
-    \begin{figure}[h!]
-        \centering
-        \includegraphics[width=\columnwidth]{travis_builds}
-        \caption{Response times in 5 consequent Travis builds in the \zee case study}
-        \label{fig:builds}
-      \end{figure}
-
-  While this integration testing environment is different from the actual production deployment one, and the load used is purely synthetic, as will show in Section~\ref{sec:evaluation}, it can actually be used successfully as an early performance indicator for the application developer.
-Although predicting the performance of the application would require advanced statistical work, it can be successfully used as an advanced warning system for the overall performance trend for given endpoints.
-
-%  \paragraph{Integration with Travis}
-%  \ml{added new section... also, this has to be moved after introducing the pre-emptive monitoring since it's about running tests. switched the order}
-
-  The {\em pre-emptive monitoring system} can be configured to work together with CI frameworks like Travis\footnote{\url{https://travis-ci.org/}} that deal with automated integration testing.
-The developer needs to define the unit test folder of the project, the URL that Travis should use for reporting its results, and the number of iterations for each unit test should be executed in order to generate a load for the application.
-
-  Each time a new build is created through Travis, the \tool automatically detects all available unit tests defined by the application developer and iterates through each one of them the number of times predefined by the developer while monitoring the response times for each request.
-The resulting measurements are persisted in a separate part of the tool database so as not to contaminate the ``live'' monitoring data.
-Beyond seeing the results of each Travis build in the \tool, this feature also allows for preemptive monitoring of the application, as we discuss in the following.
-
-  Besides functioning as an early warning system for performance, tracking the evolving performance of API tests serves a second purpose.
-To function as an anchor when the developer analyzes performance degradations.
-For example, when a developer sees that a given endpoint has become less performant after in a newly deployed version, they are able to investigate whether the performance degradation is visible also with the synthetic load.
-This would correspond to a performance degradation which is due to ``algorithmic performance degradation''.
-If on the other hand, the performance of the tests does not change between versions, the developer might conclude that the performance degradation could be due to the workload on the machines, or maybe to the workload mix of the users.
-
-  %By these means, an application developer can use unit tests as synthetic loads for driving the performance monitoring of the application before it reaches the production phase.
-As we will discuss further in Section~\ref{sec:testing}, this feature is integral in enabling preemptive performance monitoring of the application.
+\input{architecture}
 
@@
 
-  There are three main categories of visual perspectives that are available using \tool:
-  \begin{enumerate}
-    \item \textit{Service Utilization} presents information about the usage of all endpoints of interest to the developer,
-    \item \textit{Endpoint Performance} presents response times of the various service endpoints, and
-    \item \textit{User Experience} presents information about the user-perceived performance of monitored service endpoints.
-  \end{enumerate}
+\ins{PERFORMANCE DISCUSSION: this introspection, which looks up endpoints, happens only once at the startup of the service, so it is going to affect the API startup performance, but not the actual endpoint performance.
+In our case, the overhead here was quite small: TO MEASURE}.
 
-In the remainder of this section we present the views into more depth\footnote{We recommend obtaining a color version of this paper for better readability.}.
-% TODO: Elaborate on the introduction, as it is quite short
+  %In the remainder of the paper we present several of these perspectives.
 
@@
 
-% \todo{Summarize the subsections extensively; evaluate which figures to keep and ideally take new screenshots for the ones we keep; consider adding two more views (like Fig. 3.2 of Patrick's thesis)}
+\todo{evaluate which figures to keep and ideally take new screenshots for the ones we keep; consider adding two more views (like Fig. 3.2 of Patrick's thesis)}
 
-\begin{figure*}[t]
-	\centering
-	\subfloat[Number of requests per endpoint per day]{\includegraphics[width=.9\columnwidth]{number_of_requests_}\label{fig:aeu}}
-	\subfloat[Daily patterns for the \epOutcome endpoint]{\includegraphics[width=\columnwidth]{heatmap_requests}\label{fig:dp}}
-	\qquad
-\subfloat[The Performance Evolution of the \epTranslations endpoint]{\includegraphics[width=.95\columnwidth]{exec_time_reo}\label{fig:tee}}\qquad	
-\subfloat[response time (in ms) per monitored endpoint view]{\includegraphics[width=.8\columnwidth]{endpoint_performance_}\label{fig:ep}}
-%	\subfloat[The Performance Evolution of the \epTranslations endpoint]{\includegraphics[width=.95\columnwidth]{translation_endpoint_evolution_}\label{fig:tee}}
+% In the Zeeguu case study, one of the slowest endpoints, and one with the highest variability as shown in \Fref{fig:ep} is \epFeedItems: it retrieves a list of recommended articles for a given user.
+However, since a user can be subscribed to anything from one to three dozen article sources, and since the computation of the difficulty is personalized and it is slow, the variability in time among users is likely to be very large.
 
-	\caption{Some of the available views\label{fig:views}}
-\end{figure*}
-
-\paragraph{Service Utilization}
-\label{sec:util}
-
-The most fundamental insight that a service maintainer needs regards service utilization. 
-\Fref{fig:aeu} shows a first perspective on endpoint utilization that \tool provides: a stacked bar chart of the number of hits to various endpoints grouped by day\footnote{Endpoint colors are the same in different views.}.
-This figure is interesting for two reasons.
-First, it provides a quick overview of the number of requests that the API processes.
-It shows that at its peak the API has about 12.000 requests per day.
-Second, it is easy to spot what the most popular endpoints are, which are in our case study the following three endpoints:
-
-\begin{enumerate}
-
-	\item \textbf{\color{myred} \epTranslations} is an indicator of the amount of foreign language reading the users are doing, and
-
-	\item \textbf{\color{myblue} \epLearnedLanguage} is an indicator of ... % TODO for Mircea
-
-	\item \textbf{\color{mygreen} \epOutcome} is an indicator of the amount of foreign vocabulary practice the users are doing.
-
-\end{enumerate}
-
-%\begin{figure}[!ht]
-%	\centering
-%	\includegraphics[width=\linewidth]{number_of_requests_}
-%	\caption{The number of requests per endpoint per day view shows the overall utilization of the monitored application}
-%	\label{fig:aeu}
-%\end{figure}
-
-Besides showing the overall utilization, this endpoint provides the maintainer with information relevant for decisions regarding endpoint deprecation --- one of the most elementary ways of {\em understanding the needs of the downstream}\cite{Haen14a}.
-In our case study, the maintainer realized that one endpoint which they thought was not being used (i.e.
-\code{words\_to\_study}), contrary to their expectations, was actually being used\footnote{A complementary type of usage information can also be discovered in the view presented in Figure \ref{fig:sep} where seeing that an endpoint is never accessed can increase the confidence of the maintainer that a given endpoint is not used, although it can never be used a proof.}.
-
-%   \todo{Add the time series graph and discuss it before the heatmap? We can then sell the heatmap better} 
-%   \ml{Not sure about which graph you refer to here V}
-
-As \epOutcome is one of the endpoints that are most used in the API, it is interesting to have investigate the usage patterns into more depth.
-For this purpose, \Fref{fig:dp} has been created.
-This graph shows the usage of the endpoint in the number of requests per hour, which is presented in a color-scale from low usage (blue) to highly active (red).
-This figure shows the API is not being used during the early morning hours, with most of the activity focused around working hours and some light activity during the evening.
-This is consistent with the fact that the current users are all in the central European timezone.
-Also, the figure shows that the spike in utilization that was visible also in \Fref{fig:aeu} happened in the afternoon/evening.
-Another utilization that is visible in the \Fref{fig:dp} regards {\em cyclic patterns of usage per hour of day}, which is of interest to attack possible service overloads in the future.
-% PV I'm not sure whether this is a good argument, but the following source might be of interest: http://static.usenix.org/events/usits03/tech/full_papers/welsh/welsh_html/
-
-%A second type of utilization question that the \tool can answer automatically regards {\em cyclic patterns of usage per hour of day} by means of a heatmap, as in \Fref{fig:dp}.
-
-%\begin{figure}[!ht]
-%	\centering
-%	\includegraphics[width=0.9\linewidth]{daily_patterns_}
-%	\caption{Usage patterns become easy to spot in the requests per hour heatmap}
-%	\label{fig:dp}
-%\end{figure}
-
-\paragraph{Endpoint Performance}
-\label{sec:perf}
-The \tool also collects information regarding endpoint performance, and as discussed in the previous section, \epTranslations is one of the most used endpoints in the API.
-Therefore, this endpoint is used for further investigating the overall response time.
-In order to do so, one might ask himself whether the overall execution time remains constant during its evolution.
-In order to answer this question, the \tool creates a view which is presented in \Fref{fig:tee}.
-This Figure creates a horizontal boxplot for every version using the response times.
-It is easy to see that there is an increase in the performance times since version {\em 257e1a}.
-The maintainer has confirmed that this decrease in performance is due to a new algorithm that was deployed.
-
-Another useful view that the \tool generates is the overall response time clustered per endpoint.
-This is of interest to the maintainer of the API since it provides information about what endpoints might need optimization.
-The results of this view for the case study are presented in \Fref{fig:ep}.
-From this figure it is visible that the response time for \epLearnedLanguage and \epOutcome are in most cases almost instant, while the response time for \epTranslations is far from instant.
-Another observation that can be made is the fact that some endpoints show a huge variation in the response time, this is explained below:
-
-\begin{figure*}[t]
-	\centering
-	\subfloat[User specific performance for the \epLearnedLanguage endpoint]{\includegraphics[width=\linewidth]{exec_time_user}\label{fig:etu}}\qquad
-	\subfloat[User specific performance per version for the \epLearnedLanguage endpoint]{\includegraphics[width=\linewidth]{exec_time_user_version}\label{fig:etuv}}
-	% TODO: Change to order of one of the two figures, such that the users are listed in the right order (they are now different listed).
-	\caption{Some of the available views\label{fig:user_experience}}
-\end{figure*}
-
-\paragraph{User Experience}
-\label{sec:user}
-According to \cite{gmail}, the performance might be user-specific.
-In such case, the maintainer of a system might want to understand the impact on the resulting response time.
-The \tool allows several ways of doing this, some of them are presented in \Fref{fig:user_experience}\footnote{Note that both figures show a subset of the corresponding view in the \tool.}.
-First of all, the view presented in \Fref{fig:etu} can be used to detect the impact per version across the entire evolution of the API.
-It shows a number of horizontal boxplots, one for each user (a user is represented by a user-id).
-From this view, it can be concluded that \epLearnedLanguage has a user-specific processing time.
-
-The limitation of this view is that it is not possible to detect whether the performance increases per user throughout the evolution of the API.
-To address this, a different view can be defined, which is shown in \Fref{fig:etuv}.
-This view presents a {\em circle-plot}, which shows a circle for the average execution time for every user per version.
-A larger circle represents a larger execution time.
-While it is impossible to infer the actual value of this average execution time, this plot is useful for determining for which users the performance increases throughout its evolution.
-However, for the \epLearnedLanguage endpoint, there is no clear trend visible: this is probably because varying user workload (i.e.
-number of sources to which the user is registered) is the reason for the variation in response times.
-
-% TODO: Add paragraph about Preemptive Monitoring
+% \niceseparator
 
@@
 
-\subsection{Effectiveness \& Ease of Use}
+\input{discussion}
 
-\todo{@M: move here the war stories, and discuss how useful the whole thing was to the application maintainer}
-
-\subsection{Efficiency of Preemptive Monitoring}
-
-\todo{here goes the discussion about our forecasting capabilities}
-
-\begin{figure}
-	\centering
-	\subfloat[The reported response time (in ms) per deployed version in the observation period]{\includegraphics[width=.5\columnwidth]{response_per_version_trunced_trend}}
-	\quad
-	\subfloat[The measured response time (in ms) using integration with Travis]{\includegraphics[width=.46\columnwidth]{travis_builds_no_outliers_trend}}
-	\caption{Comparison of the response times per endpoint: actual production system versus preemptive monitoring data}
-	\label{fig:preemptive}
-\end{figure}
-
-\begin{table}[h]
-	\begin{tabular}{lll}
-		\toprule
-		Iteration & \bfseries Live (median) & \bfseries Travis (median)\\
-		\midrule
-		1 & 1349.41 & 764.87\\ 
-		2 & 1466.13 & 992.87\\
-		3 & 1256.65 & 760.87\\
-		4 & 1266.42 & 813.89\\
-		5 & 1080.68 & 303.4\\
-		\bottomrule
-	\end{tabular}
-\end{table}
-
-Pearson correlation $r(3)=.93, p=.02$
-
-\subsection{Discussion}
-
-\todo{use this subsection to discuss limitations (from the previous evaluation section) and add the need to be able to handle the horizontal scaling of the application; remove sectioning and summarize briefly limitations that have already been discussed in the VISSOFT paper/move material to other sections; consider renaming as limitations or something similar}
-
-  \subsubsection{Automatically Monitoring System Evolution}
-
-  The main goal of the \tool design was to allow analytics to be collected and insight to be gained by making the smallest possible changes to a running API.
-%To allow the collection of evolutionary information
-%
-  This technique assumes that the web application code which is the target of the monitoring is deployed using \git in the following way:
-
-  \begin{enumerate}
-    \item The deployment engineer pulls the latest version of the code from the integration server; this will result in a new commit being pointed at by the HEAD pointer. %than previously
-    \item The deployment engineer restarts the new version of the service.
-At this point, the \tool detects that a new HEAD is present in the local code base and consequently starts associating all the new data points with this new commit\footnote{The \tool detects the current version of the analyzed system the first time it is attached to the application object, and thus, assumes that the Flask application is restarted when a new version is deployed.
-This is in tune with the current version of Flask, but if the web server will support dynamic updates in the future, this might have to be taken into account}.
-  \end{enumerate}
-
-  The advantage of this approach is the need for minimal configuration effort, as discussed in the presentation of the tool.
-The disadvantage is that it will consider on equal ground the smallest of commits, even one that modifies a comment, and the shortest lived of commits, e.g.~a commit which was active only for a half an hour before a new version with a bug fix was deployed, with major and minor releases of the software.
-%as a distinct way of grouping the data points.
-  A mechanism to control which versions are important for monitoring purposes is therefore required to be added.
-%
-  A further possible extension point here is supporting other version control systems (e.g. Mercurial). However, this is a straightforward extension.
-
-  \subsubsection{User-Awareness }
-
-    For the situations in which the user information is not available, the \tool tracks by default information about different IPs and in some cases this might be a sufficiently good approximation of the user diversity and identity.
-    %
-
-    The visualizations for the user experiene perspectives as presented in Section \ref{sec:user} have been tested with several hundred users (of which about \activeUserCount were active during the course of the study), but the scalability of the visualizations must be further investigated for web services with tens of thousands of users.
-
-  \subsubsection{Other Possible Groupings}
-
-    There are other groupings of service utilization and performance that could be important to the maintainer, that we did not explore in this paper.
-For example, if the service is using OAuth, then together with every request, in the header of the request there is information about the application which is sending a request.
-Grouping the information by application that sends the request could be important in such a context.
-
-    In general, providing a mechanism that would allow very easy specification of groupings (either as code annotations, as normal code, or as configuration options) is an open problem that \tool and any other similar library will have to face.
-
-\section{Related Work}
-\label{sec:related}
-
-\todo{Refer the reader to~\cite{ghezzi2007run} and ~\cite{metzger2010analytical} for a more extensive discussion}
-
-There is a long tradition of using visualization for gaining insight into software performance.
-Tools like Jinsight \cite{Pauw02a} and Web Services Navigator \cite{Pauw05} pioneered such an approach for Java and for Web Services that communicate with SOAP messages.
-Both have an ``omniscient'' view of the services / objects and their interactions.
-As opposed to them, in our work we present an analytics platform which focuses on monitoring a single Python web service from its own point of view.
-
-From the perspective of service monitoring, our work falls within the server-side run-time monitoring of services ~\cite{ghezzi2007run}.
-While we don't implement the more advanced features of related monitoring solutions like QoS policies driving the monitoring, it presents nevertheless an easy to use approach support improving the performance of web applications.
-
-% \va{Mircea: Consider removing the rest for space...}
-% An existing monitoring tool is Pingdom \footnote{https://www.pingdom.com/company/why-pingdom}, which monitors the uptime of an existing web-service.
-This tool works by pinging the websites (up to 60 times) every minute automatically.
-Thus this creates a lot of overhead and is bound to be noisy since it will also be influenced by the speed of the network connection\footnote{Another problem is that such a tool would }
-
-% \todo{Runscope? Others?}
+\input{related-work}
 
@@
 
+% \input{appendix}
+
 % references section
+
 \bibliographystyle{abbrv}

### 2018-04-03 utilization.tex
done or the day

@@
-%!TEX root=restructured.tex
+%!TEX root=paper.tex
 
@@
 
-\begin{figure}[h!]
+\begin{figure}[!ht]
 \centering
-  \subfloat[Daily Utilization]{\includegraphics[width=.42\columnwidth]{number_of_requests_}\label{fig:aeu}}
-  \subfloat[Hourly Utilization]{\includegraphics[width=.5\columnwidth]{daily_patterns_}\label{fig:dp}}
-  \caption{Some of the available views\label{fig:views}}
+\includegraphics[width=\linewidth]{number_of_requests_}
+\caption{Usage patterns become easy to spot in the requests per hour heatmap}
+\label{fig:aeu}
 \end{figure}
+
+% \begin{figure}[h!]
+%   \centering
+%   \subfloat[Daily Utilization]{\includegraphics[width=.42\columnwidth]{number_of_requests_}\label{fig:aeu}}
+%   \subfloat[Hourly Utilization]{\includegraphics[width=.5\columnwidth]{daily_patterns_}\label{fig:dp}}
+%   \caption{Some of the available views\label{fig:views}}
+% \end{figure}
 
@@
 
-%\begin{figure}[!ht]
-% \centering
-% \includegraphics[width=0.9\linewidth]{daily_patterns_}
-% \caption{Usage patterns become easy to spot in the requests per hour heatmap}
-% \label{fig:dp}
-%\end{figure}
+\begin{figure}[!ht]
+\centering
+\includegraphics[width=\linewidth]{daily_patterns_}
+\caption{Usage patterns become easy to spot in the requests per hour heatmap}
+\label{fig:dp}
+\end{figure}
 

### 2018-04-04 discussion.tex
done or the day
cleaned up the performance testing part

@@
   same information as the outlier maybe?
+
+- time as measured in days and commits are two sides of the same coin.
+
+- currently the system supports only one grouping. this need not be so, and there are many scenarios where one could envision multiple groupins in parallel.
+
+- threats to the validity of the constant-load performance testing. performance regressions? 
+- the endpoint where the unit-testing results can be uploaded is prone to being tampered with by Bob who would want to upload junk information to mess up with the dashboard.
+to solve this, a UUID can also be added that is generated by the server, and thus prevents anybody else but the rightful owner of the dashboard
+
+- user management. although not explained in the paper, the dashboard also comes with a user management system. and two classes of users. admins and guests.
 

### 2018-04-04 evolution.tex
done or the day
cleaned up the performance testing part

@@
-%!TEX root=restructured.tex
+%!TEX root=paper.tex
 
-  \section{Evolution monitoring }
+  \section{LOC\#3: Monitoring Evolution}
 
@@
 
+# LOC #3: provide the dashboard with 
+# information on where to find the 
+# git information 
 dashboard.config.git = 'path/to/.git'
@@
       \centering
-      \includegraphics[width=\linewidth]{utilization-evolution}
+      \includegraphics[width=1\linewidth]{utilization-evolution}
       \caption{The Evolution of Endpoint Utilization Across System Versions}
@@
       \centering
-      \includegraphics[width=0.9\linewidth]{translation_endpoint_evolution_}
+      \includegraphics[width=\linewidth]{translation_endpoint_evolution_}
       \caption{The Performance Evolution of the \epTranslations endpoint}
@@
   \centering
-  \includegraphics[width=0.7\linewidth]{time_per_user_per_version}
-  \caption{This perspective shows that the evolution of response times for individual users (horizontal lines) across versions (the x-axis)}
+  \includegraphics[width=0.97\linewidth]{time_per_user_per_version}
+  \caption{This perspective shows that the evolution of response times for individual users (horizontal lines) across versions (the x-axis) for a given endpoint}
   \label{fig:tuv}
@@
 number of sources to which the user is registered) is the reason for the variation in response times.
-\ins{one can see that for user 1 performance degrades over the }
+\ins{one can see that for user 1 performance degrades over the versions.
+Given that for other users the performance does not degrade in the same way, it is safe to assume that the problem might lay elsewhere.}

### 2018-04-04 grouping.tex
done or the day
cleaned up the performance testing part

@@
 in GMail some users have two emails while other have twenty thousand and this induces different response times for different users}.
-\todo{remove the configuration part and rephrase the following, the discussion has been partially subsumed by Section~\ref{sec:tool}.}
 
@@
 
-# attaching a group by function 
+# LOC #2: configure the dashboard
+# to group requests by the user id
 dashboard.config.group_by = 'User',
@@
 
+Explain the code: the group by is assigned a tuple:
+
+\begin{itemize}
+	\item 'User' stands for the name of the grouping
+	\item the ``lambda'' stands for a function that takes as argument
+	the global flask.request. Every application must have a way of
+	associating a user with a request. Or an app with a request. 
+	A user is usually retrieved from some session variable. 
+\end{itemize}
+
 % \niceseparator
-
-\todo{consider losing one or both of the figures --- or at least updating them}
 
@@
 
+\begin{figure}[h!]
+  \centering
+  \includegraphics[width=.6\linewidth]{time_per_user}
+  \caption{The \epFeedItems shows a very high variability across users}
+  \label{fig:tpu}
+\end{figure}
+
 A \perspective{Per-User Performance} perspective should show the different response times for different users.
@@
 The figure shows that the response times for this endpoint can vary considerably for different users with some extreme cases where a user has to wait a full minute until their recommended articles are shown\footnote{After seeing this perspective, the maintainer refactored the architecture of the system to move part the difficulty computation out of the interactive loop}.
-
-\begin{figure}[h!]
-  \centering
-  \includegraphics[width=.8\linewidth]{time_per_user}
-  \caption{The \epFeedItems shows a very high variability across users}
-  \label{fig:tpu}
-\end{figure}

### 2018-04-04 introductory.tex
done or the day
cleaned up the performance testing part

@@
 GitHub contains more than 500K open source Python projects and the Tiobe Index\footnote{TIOBE programming community index is a measure of popularity of programming languages, created and maintained by the TIOBE Company based in Eindhoven, the Netherlands} ranks Python as the 4th most popular programming language as of June 2016.
+An analysis of  StackOverflow from September 2017 argues that {\em ``Python has a solid claim to being the fastest-growing major programming language''}\footnote{https://stackoverflow.blog/2017/09/06/incredible-growth-python/}.
 
@@
 
-\subsection*{NOTHING PREVENTS GENERALIZATION}
-In this article we take as an example the Flask case study.
+\subsection*{ENSURE CLARITY: NOTHING PREVENTS GENERALIZATION}
+In this article we take as an example the Flask case study
+and of the \tool as a dashboard. 
 But there is nothing that prevents the generalization of 
-our approach.
-
-MAKE SURE THAT THIS IS CLEAR.
+our approach to other languages and technologies.

### 2018-04-04 testing.tex
done or the day
cleaned up the performance testing part

--- 2018-04-03 testing.tex
@@
-%!TEX root=restructured.tex
+%!TEX root=paper.tex
 
 \newpage
-  \section{Preemptive monitoring}
+  \section{Performance Regression Testing}
 
-  \ins{INTRO - LINKING WITH PREVIOUS PARAGRAPH: the previous evolutionary view crucial for 
-  understanding the impact of software evolution on performance.
-  However, just by observing this one can't be sure whether 
-  performance changes because of the code or because of the payload.}
+  What happens when the developer observes a performance degradation of a given API endpoint?
+How does he know whether the reason is the degradation of the performance of the code, the load on the machines, or maybe the change in the magnitude of the user data?
 
-  \todo{the following might need a bit of rewrite}
+  \vspace{0.1cm}
 
+  One way of solving the problem is to monitor the evolution of the performance of the endpoints measured with a constant load.
+Thus, if the performance decreases when an endpoint is tested with a constant load the problem must be the endpoint implementation; otherwise, it must be found elsewhere.
+
+  \subsection*{Integrating with CI Servers}
+
+  To monitor endpoint performance with a constant load \tool takes advantage of two best practices in software evolution: (1) and API must have unit tests for its endpoints, and (2) these unit tests will be run within a CI server.
+
+  This raises the opportunity of piggybacking on the CI process to track the timing of these unit tests as a proxy of the endpoint performance evolution.
+
+  \tool can be configured to work together with Continuous Integration (CI) frameworks like Travis\footnote{\url{https://travis-ci.org/}, one of the most popular CI solutions at the moment} that deal with automated integration testing.
+To do this one needs to simply add one extra line in the script that runs their continuous integration testing.
+In the case of Travis CI, a developer has to add the following line in `.travis.yml':
+
+  \begin{lstlisting}[style=custompython]  
+# LOC #4: to be added to the .travis.yml file
+python -m flask_monitoringdashboard.collect
+   --test_folder=./tests_zeeguu_api --times=5 
+   --url=https://zeeguu.unibe.ch/api/dashboard
+
+  \end{lstlisting}
+
+  Each time a new build is created through Travis, the \tool automatically detects all available unit tests defined by the application developer (in (-{}-test\_folder)) and iterates through each one of them a number of times (-{}-times) while monitoring the response times for each test.
+The resulting measurements are uploaded to a specific endpoint that is created by the tool at the API url (-{}-url).
+The data uploaded to this endpoint is   persisted in a separate part of the tool database so as not to contaminate the ``live'' monitoring data.
+
+      \begin{figure}[h!]
+        \centering
+        \includegraphics[width=0.45\columnwidth]{travis_builds}
+        \caption{Response times in 5 consequent Travis builds}
+        \label{fig:builds}
+      \end{figure}
+
+  \Fref{fig:builds} is a screenshot from the dashboard showing the measured response times for 5 consequent builds, with 180 iterations of the unit tests executed in total across all endpoints.
+The outliers on the right of the figure are due to initial requests \ins{which must wait for a boot-up phase of the API}.
+
+  \subsection*{Preemptive Monitoring}
   The concept of {\em preemptive monitoring} of the application performance by means of instrumenting integration unit testing as the synthetic load is similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}, i.e.~testing service-based applications by using dedicated test input in parallel to its normal use and operation.
@@
 
-  For example, \Fref{fig:builds} is a screenshot from the dashboard showing the measured response times for 5 consequent builds, with 180 iterations of the unit tests executed in total across all endpoints.
-The outliers on the right of the figure are due to initial requests \ins{which must wait for a boot-up phase of the API}.
+  Figure \ref{fig:response_times_preemptive} shows one example of correlation between the actual endpoint response times across five versions of the system (above) and the corresponding testing times for the synthetic load in the same five versions (below).
+As suggested by the green line that unites the medians there seems to be a correlation between the two datasets as they are moving in the same.
 
@@
         \centering
-        \includegraphics[width=\columnwidth]{travis_builds}
-        \caption{Response times in 5 consequent Travis builds in the \zee case study}
-        \label{fig:builds}
+        \includegraphics[width=0.75\columnwidth]{response_per_version_trunced_trend}
+
+        \advance\leftskip-0.2cm
+        \includegraphics[width=0.5\columnwidth]{travis_builds_no_outliers_trend}
+        \caption{The reported response time (in ms) per deployed version in the observation period: 
+        in the actual deployment (above) and using integration with Travis (below)}
+
+        \label{fig:response_times_preemptive}
       \end{figure}
+
+  Computing Pearson correlation ($r(3)=.93, p=.02$) between the median values of the two datasets (cf. Table \ref{tab:correlations}) shows that there is a correlation between the two.
+
+    \begin{table}[h]
+
+      \centering
+      \begin{tabular}{lll}
+        \toprule
+        Iteration & \bfseries Live (median) & \bfseries Travis (median)\\
+        \midrule
+        1 & 1349.41 & 764.87\\ 
+        2 & 1466.13 & 992.87\\
+        3 & 1256.65 & 760.87\\
+        4 & 1266.42 & 813.89\\
+        5 & 1080.68 & 303.4\\
+        \bottomrule
+
+      \end{tabular}
+      \caption{The measured response time (in ms) using integration with Travis}
+      \label{tab:correlations}
+    \end{table}
+
+%   \begin{figure*}[hb!]
+%   \centering
+%   \subfloat[The reported response time (in ms) per deployed version in the observation period]{\includegraphics[width=.6\columnwidth]{response_per_version_trunced_trend}}
+%   \quad
+%   \subfloat[The measured response time (in ms) using integration with Travis]{\includegraphics[width=.56\columnwidth]{travis_builds_no_outliers_trend}}
+%   \caption{Comparison of the response times per endpoint: actual production system versus preemptive monitoring data}
+%   \label{fig:preemptive}
+% \end{figure*}
 
@@
   \paragraph{Integration with Travis}
-  \ml{added new section... also, this has to be moved after introducing the pre-emptive monitoring since it's about running tests. switched the order}
-
-  \ins{The pre-emptive monitoring system} can be configured to work together with Continuous Integration (CI) frameworks like Travis\footnote{\url{https://travis-ci.org/}} that deal with automated integration testing.
-  The developer needs to define the unit test folder of the project, the URL that Travis should use for reporting its results, and the number of iterations for each unit test should be executed in order to generate a load for the application.
-
-  Each time a new build is created through Travis, the \tool automatically detects all available unit tests defined by the application developer and iterates through each one of them the number of times predefined by the developer while monitoring the response times for each request.
-The resulting measurements are persisted in a separate part of the tool database so as not to contaminate the ``live'' monitoring data.
-Beyond seeing the results of each Travis build in the \tool, this feature also allows for preemptive monitoring of the application, as we discuss in the following.
 
@@
 
-\begin{figure}[h!]
-  \centering
-  \subfloat[The reported response time (in ms) per deployed version in the observation period]{\includegraphics[width=.5\columnwidth]{response_per_version_trunced_trend}}
-  \quad
-  \subfloat[The measured response time (in ms) using integration with Travis]{\includegraphics[width=.46\columnwidth]{travis_builds_no_outliers_trend}}
-  \caption{Comparison of the response times per endpoint: actual production system versus preemptive monitoring data}
-  \label{fig:preemptive}
-\end{figure}
-
-\begin{table}[h]
-  \begin{tabular}{lll}
-    \toprule
-    Iteration & \bfseries Live (median) & \bfseries Travis (median)\\
-    \midrule
-    1 & 1349.41 & 764.87\\ 
-    2 & 1466.13 & 992.87\\
-    3 & 1256.65 & 760.87\\
-    4 & 1266.42 & 813.89\\
-    5 & 1080.68 & 303.4\\
-    \bottomrule
-  \end{tabular}
-\end{table}
-
-Pearson correlation $r(3)=.93, p=.02$
-
     \begin{figure*}[ht!]

### 2018-04-04 evolution.tex
work in progress

@@
 
-  \section{LOC\#3: Monitoring Evolution}
+  \section{Evolution Monitoring}
 
@@
       \centering
-      \includegraphics[width=1\linewidth]{utilization-evolution}
+      \includegraphics[width=0.9\linewidth]{utilization-evolution}
       \caption{The Evolution of Endpoint Utilization Across System Versions}
@@
       \centering
-      \includegraphics[width=\linewidth]{translation_endpoint_evolution_}
+      \includegraphics[width=0.9\linewidth]{translation_endpoint_evolution_}
       \caption{The Performance Evolution of the \epTranslations endpoint}
@@
 \ins{one can see that for user 1 performance degrades over the versions.
-Given that for other users the performance does not degrade in the same way, it is safe to assume that the problem might lay elsewhere.}
+Given that for other users the performance does not degrade in the same way, it is probable that the problem might lay elsewhere: the server was overloaded or the endpoint is ``algorithmically slower''.}

### 2018-04-04 grouping.tex
work in progress

@@
 \begin{lstlisting}[style=custompython]  
-
 # LOC #2: configure the dashboard
@@
 dashboard.config.group_by = 'User',
-	lambda: Session.find(flask.request).userid
+	lambda: Session.find(flask.request).user.id
 
@@
 	A user is usually retrieved from some session variable. 
+	If we wanted to group by the user group for example, the code
+	would change slightly by replacing ``user.id'' with ``user.group.id''
 \end{itemize}
@@
   \centering
-  \includegraphics[width=.6\linewidth]{time_per_user}
+  \includegraphics[width=.5\linewidth]{time_per_user}
   \caption{The \epFeedItems shows a very high variability across users}

### 2018-04-04 testing.tex
work in progress

@@
 \newpage
-  \section{Performance Regression Testing}
+  \section{Performance Regression Monitoring}
 
-  What happens when the developer observes a performance degradation of a given API endpoint?
-How does he know whether the reason is the degradation of the performance of the code, the load on the machines, or maybe the change in the magnitude of the user data?
+  Suppose the developer observes a performance degradation of a given API endpoint in the latest version of the system.
+How do they know whether the degradation is due to the performance of the code, the server load, or the workload mix of the user?
 
@@
 
-  One way of solving the problem is to monitor the evolution of the performance of the endpoints measured with a constant load.
-Thus, if the performance decreases when an endpoint is tested with a constant load the problem must be the endpoint implementation; otherwise, it must be found elsewhere.
+  One way of triangulating such an observation is by monitoring the evolution of the performance of the endpoints when tested with a constant load across versions.
+Thus, if the performance decreases in a new version, and the  when an endpoint is tested with a constant load the problem must be the endpoint implementation.
 
@@
   To monitor endpoint performance with a constant load \tool takes advantage of two best practices in software evolution: (1) and API must have unit tests for its endpoints, and (2) these unit tests will be run within a CI server.
-
-  This raises the opportunity of piggybacking on the CI process to track the timing of these unit tests as a proxy of the endpoint performance evolution.
+By tracking the timing of these unit tests one obtains a perspective on endpoint performance evolution with a constant load (as long as the unit tests do not change).
 
@@
   The concept of {\em preemptive monitoring} of the application performance by means of instrumenting integration unit testing as the synthetic load is similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}, i.e.~testing service-based applications by using dedicated test input in parallel to its normal use and operation.
-The difference is that we take advantage of the capability of the CI framework (i.e.
-Travis in this case) to create an emulated ``live'' environment for integration testing purposes, and use unit testing as the dedicated test input in order to measure performance.
+The difference is that we take advantage of the capability of the CI framework to create an emulated ``live'' environment for integration testing purposes, and use unit testing as the dedicated test input in order to measure performance.
 
-  Figure \ref{fig:response_times_preemptive} shows one example of correlation between the actual endpoint response times across five versions of the system (above) and the corresponding testing times for the synthetic load in the same five versions (below).
-As suggested by the green line that unites the medians there seems to be a correlation between the two datasets as they are moving in the same.
+  While this integration testing environment is different from the production one, and the load used is purely synthetic, it can serve as an early performance indicator for the developer.
+Figure \ref{fig:response_times_preemptive} shows the actual endpoint response times across five versions of the system (above) and the corresponding testing times for the synthetic load in the same five versions (below).
+The green line joining the medians hints to a correlation between the two sets of measurements.
 
@@
       \end{tabular}
-      \caption{The measured response time (in ms) using integration with Travis}
+      \caption{Median response times in Figure \ref{fig:response_times_preemptive}}
       \label{tab:correlations}
@@
 
-  While this integration testing environment is different from the actual production deployment one, and the load used is purely synthetic, as will show in Section~\ref{sec:evaluation}, it can actually be used successfully as an early performance indicator for the application developer.
-\ins{Although predicting the performance of the application would require advanced statistical work, it can be succesfully used as an advanced warning system for the overall performance trend for given endpoints}.
-
-  \paragraph{Integration with Travis}
-
+  \ml{This following paragraph is beautifully written, but it's a bit redundant with what it's already said.
+However maybe after some more refactorings we can put it to use.
+Worst case, we move it to the conclusion}
   \ins{Besides functioning as an early warning system for performance, tracking the evolving performance of API tests serves a second purpose.
@@
 If on the other hand, the performance of the tests does not change between versions, the developer might conclude that the performance degradation could be due to the workload on the machines, or maybe to the workload mix of the users}
-
-  %By these means, an application developer can use unit tests as synthetic loads for driving the performance monitoring of the application before it reaches the production phase.
-As we will discuss further in Section~\ref{sec:testing}, this feature is integral in enabling preemptive performance monitoring of the application.
-
-\todo{here goes the discussion about our forecasting capabilities}
-
-    \begin{figure*}[ht!]
-      \centering
-      \includegraphics[width=0.8\linewidth]{interceptor}
-      \caption{The first thing that needs to be done, is to decorate the application object with an INTERCEPTOR}
-      \label{fig:sep}
-    \end{figure*}
-
-\newpage

### 2018-04-04 architecture.tex
done or the day
work in progress
cleaning up stuff

@@
-%!TEX root=restructured.tex
+%!TEX root=paper.tex
+
+    \begin{figure*}[ht!]
+      \centering
+      \includegraphics[width=0.8\linewidth]{interceptor}
+      \caption{The first thing that needs to be done, is to decorate the application object with an INTERCEPTOR}
+      \label{fig:sep}
+    \end{figure*}
 
@@
 REMEMBER THE GOAL: Our main goal with the \tool issues to allow the integration of the dashboard with an API with minimal effort.
+
+The fact that the \tool itself is being developed in Python using Flask makes binding to the services of a to-be monitored application developed with the same technologies easy and intuitive.
+However, by writing different backends that do the measurement and monitoring, the frontend can be reused.
 

### 2018-04-04 discussion.tex
work in progress
cleaning up stuff

@@
 We showed here how to do it for Python and Flask because they are some of the most popular web development technologies at the moment.}
-
--   The fact that the \tool itself is being developed in Python using Flask makes binding to the services of a to-be monitored application developed with the same technologies easy and intuitive.
-However, by writing different backends that do the measurement and monitoring, the frontend can be reused.
 
@@
 - user management. although not explained in the paper, the dashboard also comes with a user management system. and two classes of users. admins and guests.
+
+- Although predicting the performance of the application would require advanced statistical work, it can be succesfully used as an advanced warning system for the overall performance trend for given endpoints.
 

### 2018-04-04 introductory.tex
work in progress
cleaning up stuff

@@
 \todo{Update abstract \& title}
-  Tens of thousands of web applications are written in Flask, a Python-based web framework.
-Despite a rich ecosystem of extensions, there is none that supports the developer in gaining insight into the evolving performance of their service.
-In this paper, we introduce \tool, a library that addresses this problem.
+  Python is one of the fastest growing programming languages of the moment.
+Flask, a Python-based web framework is the technology chosen by tens of thousands of web applications.
+Many of these Flask-based projects are small websites which can not afford to deploy an advanced service monitoring solution because they are too complicated to install.
+
+  In this paper, investigate how simple can it be to start monitoring a Flask API by developing and presenting \tool, a library that aims to provide insight into the utilization and performance of evolving services.
 We present the ease with which the library can be integrated in an already existing web application, discuss some of the visualization perspectives that the library provides and point to some future challenges for similar libraries.
@@
 
+  \ml{add more numbers that show how impressive is our API}
+
   We will use this \zee API as a case study for this paper. 

### 2018-04-04 overhead.tex
cleaning up stuff

--- 2018-04-03 restructured.tex
@@
-\documentclass[conference]{IEEEtran}
-	\pdfpagewidth=8.5truein
-	\pdfpageheight=11truein
+%!TEX root=paper.tex
 
-\usepackage{graphicx}
-\usepackage{subfig}
-\usepackage[bookmarks=false]{hyperref}
-\usepackage{xspace}
-\usepackage{listings}
-\usepackage[usenames, dvipsnames]{color}
-
-\usepackage{booktabs}
-
-\graphicspath{{./img/}}
-
-%\hyphenation{op-tical net-works semi-conduc-tor}
-
-\input{useful_defines}
-
-\begin{document}
-
-%
-\title{Agile Monitoring of Evolving Web Services}
-
-\author{
-\IEEEauthorblockN{The Author, The Other Author}\\
-Institute\\
-University, Country\\
-Email: \{author\}@university.domain 
-}
-
-% make the title area
-\maketitle
-
-\input{introductory}
-
-\input{loc1}
-
-\input{utilization}
-
-\input{performance}
-
-\input{grouping}
-
-\input{evolution}
-
-\input{testing}
-
-\section{Performance Perspectives}
-\label{sec:views}
+\section{Overhead of the \tool}
+\label{sec:overhead}
 
@@
 In our case, the overhead here was quite small: TO MEASURE}.
-
-  %In the remainder of the paper we present several of these perspectives.
-
-  % The second endpoint conists of two parts, one of them being a table that shows for every monitored endpoint the number of hits it has gotten, the time it was last accessed and its average execution time.
-The second part is a view with four graphs which show:
-  % \begin{itemize} 
-  %   \item A heatmap of the total number of requests to the monitored endpoints
-  %   \item A stacked bar chart that shows the total number of requests to the monitored endpoints per endpoint per day
-  %   \item A boxplot graph showing the average execution time per version of the web service
-  %   \item A boxplot graph showing the average execution time for every monitored endpoint
-  % \end{itemize}
-
-\todo{evaluate which figures to keep and ideally take new screenshots for the ones we keep; consider adding two more views (like Fig. 3.2 of Patrick's thesis)}
-
-% In the Zeeguu case study, one of the slowest endpoints, and one with the highest variability as shown in \Fref{fig:ep} is \epFeedItems: it retrieves a list of recommended articles for a given user.
-However, since a user can be subscribed to anything from one to three dozen article sources, and since the computation of the difficulty is personalized and it is slow, the variability in time among users is likely to be very large.
-
-% \niceseparator
-
-%\section{Preemptive Monitoring}
-%\label{sec:testing}
-%
-%%\todo{discuss here idea of using unit testing as an early performance indicator for evolving services (`preemptive monitoring'?); add a side-by-side figure comparing unit testing vs live system performance per version, maybe calculate error in trend curves?
-find a better term than prediction for the section title (plus introduction)}
-%
-%\todo{Similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}, i.e.~testing service-based applications by using dedicated test input in parallel to its normal use and operation.
-The difference is that we take advantage of the capability of Travis to create an emulated live environment for integration testing purposes, and use unit testing as the dedicated test input in order to measure performance.
-What we are going provide evidence for in the following is that we can indeed use this preemptive monitoring concept as an early performance indicator for evolving services.}
-%
-%The developer needs to define the unit test folder of the project, the URL that Travis should use for reporting its results, and the number of iterations for each unit test should be executed in order to generate
-
-\section{Evaluation}
-\label{sec:evaluation}
-
-%\todo{evaluation of the \tool across the following dimensions: ease of use, effectiveness, overhead to the system; for the last part we need to do the overhead measurement experiment, for the other two we base it on the case study and bring over whatever is necessary from the discussion section}
-
-\input{discussion}
-
-\input{related-work}
-
-\section{Conclusion and Future Work}
-\label{sec:conclusions}
-
-\todo{update as appropriate}
-
-In this paper we have shown that it is possible to create a monitoring solution which provides basic insight into web service utilization and performance  with very little effort from the developer.
-The user group that we are aiming for with this work is application developers using Flask and Python to build web applications with limited or no budget for implementing their own monitoring solutions.
-The emphasis is in allowing such users to gain insight into how the performance of the service evolves together with the application itself.
-We believe that the same architecture, and lessons can be applied to other frameworks and other languages.
-
-In the future, we plan to perform case studies with other sytstems, with the goal of discover other needs and to wean out the less useful visualizations in the \tool.
-We plan to also extend the tool towards supporting multiple deployments of the same applications across multiple nodes (e.g.
-for the situations where the application is deployed together with a load balancer).
-Finally, we plan to integrate \tool with unit testing as a complementary source of information about performance evolution.
-
-\newpage
-\appendix
-
-  \subsection{Changing the Default /dashboard Endpoint }
-
-  THIS STUFF SHOULD BE MOVED TO THE APPENDIX.
-And in a footnote we just mention that ...
-\ins{REQUIREMENT}: The system must provide an easy way to add more specific configurations if needed.
-A way of overwriting the common sense defaults.
-
-  Further configuration is not needed possible by adding additional statements before the binding definition, for example,
-
-  \begin{lstlisting}[style=custompython]
-  ...
-  dashboard.config.link = 'mydashboard'
-  dashboard.bind(flask_app)
-  \end{lstlisting}
-
-  allows for a custom route (\code{/mydashboard}) to the dashboard to be defined by the programmer. An external configuration file can be used instead, or in addition to this:
-
-  \begin{lstlisting}[style=custompython]
-  ...
-  dashboard.config.from_file('config.cfg')
-  ...
-  \end{lstlisting}
-
-  \subsection{For those who don't use GIT}
-
-    \textit{Manual} version control requires the developer to tag each new version of the application with an appropriate version identifier~\cite{papazoglou2011managing} using the \code{APP\_VERSION} configuration parameter, for example by adding to the configuration file:
-
-      \begin{lstlisting}[style=custompython]
-      [dashboard]
-      APP_VERSION=<versionID>
-      \end{lstlisting}
-
-% references section
-
-\bibliographystyle{abbrv}
-\bibliography{paper}
-
-% that's all folks
-\end{document}

### 2018-04-04 paper.tex
cleaned up the performance testing part
cleaning up stuff

@@
 %
-\title{Agile Monitoring of Evolving Web Services}
+\title{Agile Monitoring of Evolving Web Services: For the Little Man}
 
@@
 
+\input{overhead}
+
 \input{architecture}
-
-\section{Performance Perspectives}
-\label{sec:views}
-
-\ins{PERFORMANCE DISCUSSION: this introspection, which looks up endpoints, happens only once at the startup of the service, so it is going to affect the API startup performance, but not the actual endpoint performance.
-In our case, the overhead here was quite small: TO MEASURE}.
-
-  %In the remainder of the paper we present several of these perspectives.
-
-  % The second endpoint conists of two parts, one of them being a table that shows for every monitored endpoint the number of hits it has gotten, the time it was last accessed and its average execution time.
-The second part is a view with four graphs which show:
-  % \begin{itemize} 
-  %   \item A heatmap of the total number of requests to the monitored endpoints
-  %   \item A stacked bar chart that shows the total number of requests to the monitored endpoints per endpoint per day
-  %   \item A boxplot graph showing the average execution time per version of the web service
-  %   \item A boxplot graph showing the average execution time for every monitored endpoint
-  % \end{itemize}
-
-\todo{evaluate which figures to keep and ideally take new screenshots for the ones we keep; consider adding two more views (like Fig. 3.2 of Patrick's thesis)}
-
-% In the Zeeguu case study, one of the slowest endpoints, and one with the highest variability as shown in \Fref{fig:ep} is \epFeedItems: it retrieves a list of recommended articles for a given user.
-However, since a user can be subscribed to anything from one to three dozen article sources, and since the computation of the difficulty is personalized and it is slow, the variability in time among users is likely to be very large.
-
-% \niceseparator
-
-%\section{Preemptive Monitoring}
-%\label{sec:testing}
-%
-%%\todo{discuss here idea of using unit testing as an early performance indicator for evolving services (`preemptive monitoring'?); add a side-by-side figure comparing unit testing vs live system performance per version, maybe calculate error in trend curves?
-find a better term than prediction for the section title (plus introduction)}
-%
-%\todo{Similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}, i.e.~testing service-based applications by using dedicated test input in parallel to its normal use and operation.
-The difference is that we take advantage of the capability of Travis to create an emulated live environment for integration testing purposes, and use unit testing as the dedicated test input in order to measure performance.
-What we are going provide evidence for in the following is that we can indeed use this preemptive monitoring concept as an early performance indicator for evolving services.}
-%
-%The developer needs to define the unit test folder of the project, the URL that Travis should use for reporting its results, and the number of iterations for each unit test should be executed in order to generate
-
-\section{Evaluation}
-\label{sec:evaluation}
-
-%\todo{evaluation of the \tool across the following dimensions: ease of use, effectiveness, overhead to the system; for the last part we need to do the overhead measurement experiment, for the other two we base it on the case study and bring over whatever is necessary from the discussion section}
 

### 2018-04-04 overhead.tex
.

@@
 
+To know the performance overhead of the \tool, we have implemented an automated benchmark system.
+It is open source and available online and can be tested by the reader.
+The benchmark does the following steps:
+
+\begin{itemize}
+	\item downloads the latest version of the Zeeguu API and installs it in a Docker container
+	\item enables the dasboard and calls several endpoints tracking the response times
+	\item disables the dashboard and calls the same endpoitns tracking the response times
+	\item reports the response times
+\end{itemize}
+
+Based on running the benchmarking script we can report
+
 \ins{PERFORMANCE DISCUSSION: this introspection, which looks up endpoints, happens only once at the startup of the service, so it is going to affect the API startup performance, but not the actual endpoint performance.

### 2018-04-04 evolution.tex
cleaning up stuff
small improvement

@@
 
-  \section{Evolution Monitoring}
+  \section{Version-Aware Monitoring}
 
-  When thinking about performance, 
-  one can not avoid thinking about the performance evolution.
+  The Stack Overflow Developer Survey from 2018 revealed the fact that 88.4\% of professional developers use git for version control. 
+  Given that versioning using git is the main way in which the source code of services evolves, it is natural to monitor their progress also across versions.
 
@@
 
-  However, with the current configuration of the tool, it would be impossible for the maintainer to see the improvements resulting from the optimization.
+  % However, with the current configuration of the tool, it would be impossible for the maintainer to see the improvements resulting from the optimization.
 
-  \ins{The latest STack Overflow Developer Survey: 88.4\% of **professional** developers use git.}  
   If assume that the code that they run is deployed using .git, then with an extra line of configuration they can allow \tool to find the git\footnote{\url{https://git-scm.com/}} folder of the deployed service and automatically detect the version of the project that is running:
@@
 
-  \subsubsection*{Evolving Utilization}
+  \subsection*{Evolving Utilization}
 
-  \Fref{fig:mv-util} shows the \perspective{Multi-Version Utilization} perspective. 
-  It shows the utilization of all the tracked endpoints across versions.
+  \Fref{fig:mv-util} shows the \perspective{Multi-Version API Utilization} perspective which presents the utilization of the tracked endpoints across versions.
+Because the different versions might be deployed for very different periods of time, at the intersection of an endpoint and a version the chart does not plot the absolute utilization of that endpoint in that version but rather the percentage of all the API calls that go to that endpoint.
+Otherwise a version that is deployed for many weeks would make all those deployed for several days invisible.
 
@@
       \includegraphics[width=0.9\linewidth]{utilization-evolution}
-      \caption{The Evolution of Endpoint Utilization Across System Versions}
+      \caption{The Evolution of All the API Endpoints Utilization Across System Versions}
       \label{fig:mv-util}
@@
 
-    \subsubsection*{Evolving Performance}
+  With this view, several patterns are visible:
+  \begin{itemize}
+
+    \item \epFeedItems is discontinued a little bit after \epTopArticles is being introduced
+
+  \end{itemize}
+
+  \subsection*{Evolving Performance}
 
@@
 
-  CHALLENGE: Supporting other types of version control... not real challenge, as we said 88percent of devs use git. 
-  CHALLENGE 2: detecting minor versions which don't need to be tracked, since they didn't touch the performance of the system. E.g. a modification of the README, etc.
-
-\todo{TODO: endpoint evolution }
-
-\subsubsection*{Evolving Groups}
+\subsection*{Evolving Groups}
   The limitation of the previous view is that it does not present the information also on a per version basis.

### 2018-04-04 performance.tex
done or the day
work in progress
cleaning up stuff
small improvement

@@
-%!TEX root=restructured.tex
+%!TEX root=paper.tex
 
-\subsection{Service Performance}
+\newpage
+\subsection{Performance}
 \label{sec:perf}
@@
    \centering
-   \includegraphics[width=0.95\linewidth]{endpoint_performance_}
+   \includegraphics[width=0.85\linewidth]{endpoint_performance_}
    \caption{The response time (in ms) per monitored endpoint view allows for identifying performance variability and balancing issues}
@@
 
-In the Zeeguu case study, one of the slowest endpoints, and one with the highest variability as shown in \Fref{fig:ep} is \epFeedItems: it retrieves a list of recommended articles for a given user.
+In the Zeeguu case study, one of the slowest endpoints, and one with the highest variability as shown in \Fref{fig:ep} is \epFeedItemsColor which retrieves a list of recommended articles for a given user.
 However, since a user can be subscribed to anything from one to three dozen article sources, and since the computation of the difficulty is personalized and it is slow, the variability in time among users is likely to be very large.
@@
 From this view it became clear to the maintainer that four of the endpoints had very large variation in performance.
-The most critical for the application and consequently the one optimized first was the \epTranslations endpoint which was part of an interactive loop in the reader applications that relied on the Zeeguu API.
+The most critical for the application and consequently the one optimized first was \epTranslationsColor endpoint which was part of an interactive loop in the reader applications that relied on the Zeeguu API.
 Moreover, cf.
@@
 
-  \newpage
   \subsection{Automated Outlier Detection and Monitoring}
@@
     \centering
-    \includegraphics[width=\linewidth]{outlier-annotated}
+    \includegraphics[width=0.9\linewidth]{outlier-annotated}
     \caption{Automatically collected outlier information}
@@
 Thus this functionality in microsoft\_translator is really slow in itself, and this is not a result of the machine being overloaded for example.
+
+\lesson{We need more advanced outlier analysis tools... Instead of manually searching in the page to build stats about Google vs. Microsoft}

### 2018-04-04 utilization.tex
work in progress
.
smarter table
small improvement

@@
 \newpage
-\subsection{Service Utilization}
+\subsection{Utilization}
 \label{sec:util}
@@
   \centering
-  \includegraphics[width=\linewidth]{basicest-utilization}
+  \includegraphics[width=\linewidth]{utilization-table}
   \caption{....}
@@
 
-  \item {\bf Frequent Patterns of System Usage}. The two most used endpoints stand for the two main activities in the system:
+  \item {\bf Application Specific Patterns of System Usage}. The two most used endpoints in the figure stand for the two main activities in the system:
 
@@
 
-    \item \epTranslations is an indicator of the amount of foreign language reading the users are doing
+    \item \epTranslationsColor is an indicator of the amount of foreign language reading the users are doing
 
-    \epOutcome is an indicator of the amount of foreign vocabulary practice the users are doing.
+    \item \epOutcomeColor is an indicator of the amount of foreign vocabulary practice the users are doing.
 
@@
 
-  \item {\bf Sudden Increase of Endpoint Usage}. One endpoint (\epUserActivity) has been disproportionately been used in the last day and last week; much more than before.
+  \item {\bf Sudden Increase of Endpoint Usage}. One endpoint (\epUserActivityColor) has disproportionately been used in the last day and last week.
 
-  \item {\bf Possibly Discontinued Endpoint Usage}. One endpoint (\epFeedItems) has not been used in the last 7 days
+  \item {\bf Possibly Discontinued Endpoint Usage}. One endpoint (\epFeedItemsColor) has not been used in the last 7 days. This might be an endpoint that is deprecated.
 
@@
 The table view is limited, and the one day and one week periods are conveniently but arbitrarily selected.
-For a more detailed evolution of utilization, Figure \ref{fig:aeu} shows the \perspective{Daily Utilization} perspective on endpoint utilization that \tool provides: a stacked bar chart of the number of hits to various endpoints grouped by day\footnote{Endpoint colors are the same in different views}.
+For a more detailed evolution of utilization, Figure \ref{fig:aeu} shows the \perspective{Daily API Utilization} perspective on endpoint utilization that \tool provides: a stacked bar chart of the number of hits to various endpoints grouped by day\footnote{Endpoint colors are the same in different views}.
 Figure~\ref{fig:aeu} in particular shows a peak utilization, a day when the API had more than 12.000 hits.
@@
 
-Another utilization perspective is the \perspective{Hourly Utilization} in which  the \tool can highlight {\em cyclic patterns of usage per hour of day} by means of a heatmap, as in \Fref{fig:dp}.
+Another utilization perspective is the \perspective{Hourly API Utilization} in which  the \tool can highlight {\em cyclic patterns of usage per hour of day} by means of a heatmap, as in \Fref{fig:dp}.
 

### 2018-04-04 evolution.tex
...

@@
 
-    \Fref{fig:tee} is a zoomed-in version of such a view for \epTranslations with versions increasing from top to bottom
+    \Fref{fig:tee} is a zoomed-in version of such a view for \epTranslationsColor with versions increasing from top to bottom
 
@@
 
-\subsection*{Evolving Groups}
+\subsection*{Evolving Grouped Performance}
   The limitation of the previous view is that it does not present the information also on a per version basis.
-To address this, a different visual perspective entitled \perspective{Multi-Version per-User Performance} can be defined.
+To address this, a different visual perspective entitled \perspective{User-Focused Multi-Version Endpoint Performance} can be defined.
 Figure \ref{fig:tuv} presents such a perspective by mapping the average execution time for a given user (lines) and given version (columns) on the area of the corresponding circle.
@@
   \centering
-  \includegraphics[width=0.97\linewidth]{time_per_user_per_version}
+  \includegraphics[width=0.77\linewidth]{time_per_user_per_version}
   \caption{This perspective shows that the evolution of response times for individual users (horizontal lines) across versions (the x-axis) for a given endpoint}

### 2018-04-05 architecture.tex
smarter table
cleaning up...

@@
 However, by writing different backends that do the measurement and monitoring, the frontend can be reused.
+
+The viewpoints we showed are only a subset of all the viewpoints of the \tool. In practice, the tool provides other perspectives as long as they are a combination of:
+
+\begin{itemize}
+  \item Cardinality: one endpoint vs multiple
+  \item Objective: utilization vs. performance (response time)
+  \item Evolution Axis: time vs. versions
+  \item Grouping: Grouped (e.g. by user) vs. All
+\end{itemize}
 
@@
   % Maybe it fits better localized here: 
-  Data collected by the wrappers are persisted in a local database.
-The SQLAlchemy Object Relational Mapper\footnote{\url{https://www.sqlalchemy.org/}} on top of SQLite\footnote{\url{https://www.sqlite.org/}} is used for this purpose.
+  Data collected by the wrappers are persisted in a local database. The SQLAlchemy Object Relational Mapper\footnote{\url{https://www.sqlalchemy.org/}} allows for a DB system independent solution.
+
+  By default the system is deployed with an SQLite database \footnote{\url{https://www.sqlite.org/}} is used for this purpose.
 

### 2018-04-05 ease of use.tex
cleaning up...

ease of use:

statsd
collectd

Graphite used to be fairly difficult to install.

https://www.digitalocean.com/community/tutorials/how-to-install-and-use-graphite-on-an-ubuntu-14-04-server

25 pages document

https://community.netapp.com/t5/OnCommand-Storage-Management-Software-Articles-and-Resources/How-to-install-Graphite-and-Grafana/ta-p/109456

recipe for installing: 
-- tested recipe on RHEL 6 & 7, and Ubuntu, to get you up and running fast

how to install graphite:

https://www.digitalocean.com/community/tutorials/how-to-install-and-use-graphite-on-an-ubuntu-14-04-server

immense...

how to install collectd: 
https://www.digitalocean.com/community/tutorials/how-to-configure-collectd-to-gather-system-metrics-for-graphite-on-ubuntu-14-04

immense...

### 2018-04-05 introductory.tex
cleaning up...

@@
 \begin{abstract}
-\todo{Update abstract \& title}
   Python is one of the fastest growing programming languages of the moment.
@@
 Many of these Flask-based projects are small websites which can not afford to deploy an advanced service monitoring solution because they are too complicated to install.
+Many of these projects might offer a service API.
+Given the complexity of professional monitoring tools it is very likely that
 
-  In this paper, investigate how simple can it be to start monitoring a Flask API by developing and presenting \tool, a library that aims to provide insight into the utilization and performance of evolving services.
+  In this paper, we set out to show how to design a dashboard for a Flask API which is useful in understanding and guiding service evolution while at the same time, requiring minimal configuration and installation.
+We present \tool, a tool that provide insight into the utilization and performance of evolving services.
 We present the ease with which the library can be integrated in an already existing web application, discuss some of the visualization perspectives that the library provides and point to some future challenges for similar libraries.

### 2018-04-05 loc1.tex
done or the day
cleaned up the performance testing part
cleaning up...

@@
-%!TEX root=restructured.tex
+%!TEX root=paper.tex
 
   \newpage
-  \section{LOC \#1: Service Utilization and Performance}
+  \section{Service Utilization and Performance}
 
-  Since it is taylored for Flask applications, to deploy the \tool one simply needs to use one line of code\footnote{We don't count the import statement that enables that line...} to bind the dashboard to their Flask web application\footnote{\ins{ In this paper we present the integration with APIs written in Python and Flask hoping that this will not prevent the reader from seeing the more general idea; all the tools we show here for Flask can be applied to other API technologies (e.g.
+  % Since it is tailored for Flask applications, the deployment of \tool is trivial.
+
+  The tool is designed to take advantage of the fact that it runs in the same web application context and that the monitored API already has a web presence.
+It thus simply makes available one extra endpoint (i.e.
+\code{/dashboard}) at the same address as the API, endpoint which serves the interactive application presented in the remainder of this paper.
+
+  A dashboard can be attached to an already existing Flask application with a single line of code as the code in the following snippet shows
+  \footnote{We do not count the import statement that enables that line}:
+
+  % which binds the dashboard to a Flask application\footnote{\ins{ In this paper we present the integration with APIs written in Python and Flask hoping that this will not prevent the reader from seeing the more general idea; all the tools we show here for Flask can be applied to other API technologies (e.g.
 Django) by simply providing a few back-end adapters in the right places.
@@
 
-  # app is main application object
-  # every Flask web application has one
+  # LOC #1: associate the main Flask application 
+  # object with the dashboard
   dashboard.bind(app)
@@
 
-  During binding, the \tool will search for all endpoints defined in the target application 
-    \ins{ 
-      and add function wrappers around all the corresponding endpoint functions
-    }
-% 
+  During binding, the \tool will search for all endpoints defined in the target application and make them available in the dashboard in an interactive configuration panel (\Fref{fig:sep}).
+The panel presents all the automatically discovered endpoints and lets the user select the ones that should be monitored (the checkboxes in the last column).
+When the user enables the monitoring of an endpoint a function wrapper is added to the implementation of that endpoint (which is a function in Flask).
+The information is also saved in a database such that it is persistent across application restarts.
+
   % In order to monitor an endpoint, the \tool creates a function wrapper for the API function that corresponds to the endpoint.
@@
 The wrapper contains the code that takes care of monitoring an endpoint.
-%
-
-   \ins{
-    The tool takes advantage of the fact that the monitored API already has a web presence, and makes available one extra endpoint (i.e.
-\code{/dashboard}), which serves the interactive visualization perspectives presented in the remainder of this paper.
-  }  
- The first perspective presents all the automatically discovered endpoints and lets the user select the ones that should be monitored. 
-  \Fref{fig:sep} shows that the last access time of every endpoint is tracked irrespective of whether it is selected by the user to be monitored or not.
 
@@
 
-  \ins{
-    We have decided for an opt-in approach to monitoring to avoid any the performance penalties incurred by the dashboard to affect performance sensitive endpoints.
-We discuss performance issues later.
-Also, we discuss later one situation in which opt-out might be desirable.
-  }
+  By default the dashboard takes an {\em opt-in} approach to monitoring: to prevent the performance penalties incurred by the dashboard\footnote{We discuss performance issues later} to affect performance sensitive endpoints, the service developer is asked to manually add the endpoints they want to monitor.
 
-  \ins{
-
-    One alternative to allowing the user to use annotations would be to let them to annotate the code.
-However, this pollutes the code, and prevents deploying two versions which would monitor different endpoints.
-
-  }
+  One alternative to configuring the monitored endpoints from the user interface is to allow the service developer to annotate the code.
+This would however pollute the code, and prevent deploying two versions which would monitor different endpoints.
+Also, would pose problems with endpoint evolution in the future.
 
@@
 
-  \ins
-  {
-
-      The remainder of this section presents several of the interactive
-      visualizations that become available without any further configuration\footnote{We recommend obtaining a color version of this paper for better readability}.
-
-  }
+  The remainder of this section presents several interactive
+  visualizations that are available in the dashboared without any further configuration\footnote{We recommend obtaining a color version of this paper for better readability}.

### 2018-04-05 discussion.tex
small improvement
cleaning up...
cleaning up till performance

@@
 - Although predicting the performance of the application would require advanced statistical work, it can be succesfully used as an advanced warning system for the overall performance trend for given endpoints.
+
+- detecting minor versions which don't need to be tracked, since they didn't touch the performance of the system. E.g. a modification of the README, etc.
+
+- discuss later one situation in which opt-out might be desirable.
+
+- tracking other types of information: exceptions, etc. especially since logs usually get lost... nobody has time to look in them...
 

### 2018-04-05 introductory.tex
cleaning up till performance

@@
 
-\section{Case Study: The Technology (Flask)}
+\subsection*{Case Study: The Technology (Flask)}
 \label{sec:tool}
@@
 
-\subsection*{ENSURE CLARITY: NOTHING PREVENTS GENERALIZATION}
+\subsection*{This is a LONGITUDINAL STUDY}
+
+The dashboard has been deployed with the case stuy API for 
+ten months, time in which more than 112.000 hits of the API have 
+been recorded.
+
+\subsection*{TO ARGUE: NOTHING PREVENTS GENERALIZATION}
 In this article we take as an example the Flask case study

### 2018-04-05 loc1.tex
cleaning up till performance

@@
   visualizations that are available in the dashboared without any further configuration\footnote{We recommend obtaining a color version of this paper for better readability}.
+They are focused on two types of information that is important to the API maintainer:
+
+  \begin{itemize}
+
+    \item {\bf Utilization} -- refers to how the third parties use the API, which parts are most used, which are little used, etc.
+This is known to be critical information for upstream developer in general\cite{Haen14a}.
+In the case of source code dependencies one can detect such information by analyzing software repositories.
+However, in the case of services, there is no other way but monitoring service utilization.
+
+    \item {\bf Performance} as measured in response times. This is important, especially for APIs that are supposed to be integrated in live systems.
+
+  \end{itemize}

### 2018-04-05 utilization.tex
cleaning up...
cleaning up till performance

@@
 
-\newpage
 \subsection{Utilization}
@@
 
-\ins{Knowing how third parties use one's API is difficult even in the case of static dependencies.
-In the case of services, there is no other way but monitoring service utilization.
-\tool introduces a series of perspectives on utilization.
-}
-
-\ins{
-
-  The most basic possible view shows the cummulative information about all the endpoints and the number of calls to that endpoint over the lifetime of its tracking as well as in the current day and the last seven days.
-}
+  The default landing page of the dashboard presents an overview of all the endpoints, 
+  overview which presents some basic utilization information as:
+  (1) the number of calls to that endpoint since the begining of tracking
+  (2) the number of calls in the last seven days, and
+  (3) the number of calls in the last day\footnote{Although very useful, it was only after several months of using the system that the client requested the extra columsn for one day and seven days.
+An argument for slow research}s
 
@@
       \includegraphics[width=\linewidth]{utilization-table}
-  \caption{....}
+      \caption{The API Overview presents synthetic endpoint utilization information}
       \label{fig:basicest}
@@
 
-This kind of information is already very useful, and can provide the API maintainer with insight into the evolution of their system.
-By looking at \Fref{fig:basicest}, which presents the top 7 endpoints (for lack of space) one can already see several patterns:
+  This kind of information is already very useful, and can provide the API maintainer with insight into the evolution of their service API.
+The seven endpoints present in \Fref{fig:basicest} illustrate three types of insight that the maintainer can obtain by looking at such a view:
 
@@
 
-  \item {\bf Application Specific Patterns of System Usage}. The two most used endpoints in the figure stand for the two main activities in the system:
+    \item {\bf Service Specific Usage Patterns}. The two most used endpoints in the figure stand for the two main activities in the system:
 
@@
 
-\lesson{Although very useful, it was only after several months of using the system that the client requested the extra columsn for one day and seven days.}
+\niceseparator
 
-  % \begin{figure}[h!]
-  % \centering
-  % \includegraphics[width=0.5\linewidth]{number_of_requests_}
-  % \caption{The number of requests per endpoint per day view shows the overall utilization of the monitored application}
-  % \label{fig:aeu}
-  % \end{figure}
-
-The table view is limited, and the one day and one week periods are conveniently but arbitrarily selected.
-For a more detailed evolution of utilization, Figure \ref{fig:aeu} shows the \perspective{Daily API Utilization} perspective on endpoint utilization that \tool provides: a stacked bar chart of the number of hits to various endpoints grouped by day\footnote{Endpoint colors are the same in different views}.
-Figure~\ref{fig:aeu} in particular shows a peak utilization, a day when the API had more than 12.000 hits.
+  For a more detailed evolution of utilization, Figure \ref{fig:aeu} shows the \perspective{Daily API Utilization} perspective on API utilization that \tool provides: a stacked bar chart of the number of hits to various endpoints grouped by day\footnote{Endpoint colors are the same in different views}.
+Figure~\ref{fig:aeu} in particular shows a peak utilization, on June 28 when the API had more than 12.000 hits\footnote{A highschool of students that were using the system for their class had a deadline on that date}.
 
@@
     \centering
-\includegraphics[width=\linewidth]{number_of_requests_}
+    \includegraphics[width=0.8\linewidth]{number_of_requests_}
     \caption{Usage patterns become easy to spot in the requests per hour heatmap}
@@
 
-% \begin{figure}[h!]
-%   \centering
-%   \subfloat[Daily Utilization]{\includegraphics[width=.42\columnwidth]{number_of_requests_}\label{fig:aeu}}
-%   \subfloat[Hourly Utilization]{\includegraphics[width=.5\columnwidth]{daily_patterns_}\label{fig:dp}}
-%   \caption{Some of the available views\label{fig:views}}
-% \end{figure}
-
-% The way users interact with the platform can also be inferred since the endpoints are indicators of different activity types, e.g.:
-
-% Besides showing the overall utilization, this endpoint provides the maintainer with information relevant for decisions regarding endpoint deprecation --- one of the most elementary ways of {\em understanding the needs of the downstream}\cite{Haen14a}.
-In our case study, the maintainer realized that one endpoint which they thought was not being used (i.e.
-\code{words\_to\_study}), contrary to their expectations, was actually being used\footnote{A complementary type of usage information can also be discovered in the view presented in Figure \ref{fig:sep} where seeing that an endpoint is never accessed can increase the confidence of the maintainer that a given endpoint is not used, although it can never be used a proof.}.
-
-% \niceseparator
-
-%   \todo{Add the time series graph and discuss it before the heatmap? We can then sell the heatmap better} 
-%   \ml{Not sure about which graph you refer to here V}
+\niceseparator
 
@@
       \centering
-\includegraphics[width=\linewidth]{daily_patterns_}
+      \includegraphics[width=0.8\linewidth]{daily_patterns_}
       \caption{Usage patterns become easy to spot in the requests per hour heatmap}

### 2018-04-05 discussion.tex
...

@@
 - tracking other types of information: exceptions, etc. especially since logs usually get lost... nobody has time to look in them...
+
+-  In this paper we present the integration with APIs written in Python and Flask hoping that this will not prevent the reader from seeing the more general idea; all the tools we show here for Flask can be applied to other API technologies (e.g.
+Django) by simply providing a few back-end adapters in the right places.
+
+- We need more advanced outlier analysis tools... Instead of manually searching in the page to build stats about Google vs. Microsoft
 

### 2018-04-05 performance.tex
cleaning up till performance
...

@@
 
-\newpage
 \subsection{Performance}
@@
 
-Knowing the performance of one’s API is critical for the quality of a service API.
-Flask Dashboard introduces a series of perspectives on performance, which focus on the response times of the various endpoints.
+% The \tool also collects information regarding endpoint performance. 
+The \perspective{API Performance} visualization perspective presents
+an overview of the response times for all the tracked endpoints 
+by using a box-and-whiskers plot.
 
-\ins{HERE? ELSEWHERE? Actually, probably better for performance. The importance of the last day and last seven days became clear only after having used the dashboard for several months. }
+\Fref{fig:ep} presents such a perspective from our case study which 
+was taken in the last quarter of 2017. Several observatrions are 
+straightforward:
 
-The \tool also collects information regarding endpoint performance. The view in \Fref{fig:ep} summarizes the response times for various endpoints by using a box-and-whiskers plot.
+\begin{itemize}
+
+  \item The slowest endpoint and the one with the highest variability is \epFeedItemsColor: it retrieves a list of recommended articles for a given user.
+Since a user can be subscribed to anything from one to three dozen article sources, and since the computation of the difficulty was done in real time, the variability in time among users is likely to be very large.
+
+  \item In the top three slowest endpoints, but the second most used endpoint in the system (cf.
+\Fref{fig:aeu}) is \epTranslationsColor.
+The endpoint is critical being embedded in the interactive loop in the reader applications that rely on the Zeeguu API: a user would tap on a word and receive a translation from this endpoint.
+
+\end{itemize}
 
@@
 
-In the Zeeguu case study, one of the slowest endpoints, and one with the highest variability as shown in \Fref{fig:ep} is \epFeedItemsColor which retrieves a list of recommended articles for a given user.
-However, since a user can be subscribed to anything from one to three dozen article sources, and since the computation of the difficulty is personalized and it is slow, the variability in time among users is likely to be very large.
-
-From this view it became clear to the maintainer that four of the endpoints had very large variation in performance.
-The most critical for the application and consequently the one optimized first was \epTranslationsColor endpoint which was part of an interactive loop in the reader applications that relied on the Zeeguu API.
-Moreover, cf.
-\Fref{fig:aeu} this endpoint is one of the most used in the system.
+Such a view is important for allocating optimization efforts.
+In our case study, the API developers started by improving the performance of \epTranslationsColor and then later optimized the performance of \epFeedItemsColor.
+However, in order to understand the problems of the two endpoints, a big component represents understanding the outliers: because they are many.
 
@@
 The bottom panel shows the stack trace. 
-In this particular case, it is revealing for the developer to learn that at the time of the stack trace snapshot, the code was in the google\_translator.py: indeed, the system uses as back-end multiple translators, and it has been observed that many of the outliers happen to be waiting in the google translator.
+In this particular case, it is revealing for the developer to learn that at the time of the stack trace snapshot, the code was in the google\_translator.py: indeed, the system uses as back-end multiple translators, and it has been observed that many of the outliers happen while the system is waiting for the Google translator.
 
@@
 Thus this functionality in microsoft\_translator is really slow in itself, and this is not a result of the machine being overloaded for example.
-
-\lesson{We need more advanced outlier analysis tools... Instead of manually searching in the page to build stats about Google vs. Microsoft}

### 2018-04-05 utilization.tex
...

@@
       \centering
-      \includegraphics[width=\linewidth]{utilization-table}
+      \includegraphics[width=0.8\linewidth]{utilization-table}
       \caption{The API Overview presents synthetic endpoint utilization information}
@@
   For a more detailed evolution of utilization, Figure \ref{fig:aeu} shows the \perspective{Daily API Utilization} perspective on API utilization that \tool provides: a stacked bar chart of the number of hits to various endpoints grouped by day\footnote{Endpoint colors are the same in different views}.
-Figure~\ref{fig:aeu} in particular shows a peak utilization, on June 28 when the API had more than 12.000 hits\footnote{A highschool of students that were using the system for their class had a deadline on that date}.
+Figure~\ref{fig:aeu} in particular shows a peak utilization, on June 28 when the API had more than 12.000 hits\footnote{Turns out that the highschool students that were using the system for their class had a deadline on that date}.
 

### 2018-04-06 evolution.tex
intro to evolution

@@
   The Stack Overflow Developer Survey from 2018 revealed the fact that 88.4\% of professional developers use git for version control. 
-  Given that versioning using git is the main way in which the source code of services evolves, it is natural to monitor their progress also across versions.
+  Given that versioning using git is the main way in which the source code of services evolves and is deployed, it is natural to monitor their progress also across versions.
 
@@
   on some version control system. 
-
-  % However, with the current configuration of the tool, it would be impossible for the maintainer to see the improvements resulting from the optimization.
-
-  If assume that the code that they run is deployed using .git, then with an extra line of configuration they can allow \tool to find the git\footnote{\url{https://git-scm.com/}} folder of the deployed service and automatically detect the version of the project that is running:
+  The second option is the most interesting one.
+  For the systems deployed using Git, an extra configuration line enables \tool to automatically\footnote{Alternatively, the maintainer can add version identifiers manually for the web application through a configuration file if the system does not use git.} associate API traffic with the currently deployed version based on information in the .git folder \footnote{\url{https://git-scm.com/}}:
 
@@
 \end{lstlisting}
-
-  If the \tool can automatically detect the current version of the project by reading the .git configuration as soon as the API is started and can then group measurements by version\footnote{Alternatively, the maintainer can add version identifiers manually for the web application through a configuration file if the system does not use git.}.
 

### 2018-04-06 grouping.tex
...

@@
 
-For service endpoints which run computations in real time, the maintainer of a system might want to understand the endpoint performance on a per-user basis, especially for situations where the system response time is a function of some individual user load\footnote{E.g.
-in GMail some users have two emails while other have twenty thousand and this induces different response times for different users}.
+It is often the case that a service has diferent types of users: the paying and the free-for-evaluation ones, or the in-house and those in a different organization.
 
-To enable this, the \tool must be configured to associate an API call with a given user.
-The simplest way is to take advantage of the architecture of Flask applications in which a global \code{flask.request} object can be used to retrieve the session which can in turn lead to user identification:
+In some other cases, it might be important for the maintainer of the API to understand the performance on a per-user basis, especially for situations where the system response time is a function of some individual user load\footnote{E.g.
+in GMail some users have a few dozen emails while other have tens of thousands: this difference in user loads will eventually induce a difference in the response times for different users}.
+
+To support narrowing down the analysis to groups of users, or even individual users, the \tool must log together with every request grouping information for that request.
+The simplest way of achieving this is to take advantage of the architecture of Flask applications in which a global \code{flask.request} is used to retrieve session information which can in turn is normally used to identify the user sending the request.
+From the user one can obtain the group.
+
+The following code snippet shows how to configure the dashboard to enable user-by-user analysis\footnote{	If we wanted to group by the user group for example, the code would change slightly by replacing ``user.id'' with ``user.group.id''}:
 
@@
 
-Explain the code: the group by is assigned a tuple:
+The \code{dashboard.config.group\_by} is assigned a tuple with two elements, in which:
 
 \begin{itemize}
-	\item 'User' stands for the name of the grouping
-	\item the ``lambda'' stands for a function that takes as argument
-	the global flask.request. Every application must have a way of
-	associating a user with a request. Or an app with a request. 
-	A user is usually retrieved from some session variable. 
-	If we wanted to group by the user group for example, the code
-	would change slightly by replacing ``user.id'' with ``user.group.id''
+	\item \code{'User'} stands for the name of this particulart grouping strategy
+	\item \code{lambda: Session.find(... } is a callback with no arguments, 
+	that makes use of the global flask.request and extracts from 
+	the request the current session, and in turn the user id\footnote{
+		Note that every web service or application must have a way of 
+		associating a request with a user. In fact, the Session.find(request) 
+		was already an existing function in the analyzed system}
 \end{itemize}

### 2018-04-06 evolution.tex
...

@@
 
-  The Stack Overflow Developer Survey from 2018 revealed the fact that 88.4\% of professional developers use git for version control. 
+  The Stack Overflow Developer Survey from 2018 revealed the fact that 88.4\% of professional developers use Git. 
   Given that versioning using git is the main way in which the source code of services evolves and is deployed, it is natural to monitor their progress also across versions.
 
-  Version control in \tool can be supported in two ways,
-  the developer explicitly states the current version, 
-  or the current version is automatically detected based
-  on some version control system. 
-  The second option is the most interesting one.
-  For the systems deployed using Git, an extra configuration line enables \tool to automatically\footnote{Alternatively, the maintainer can add version identifiers manually for the web application through a configuration file if the system does not use git.} associate API traffic with the currently deployed version based on information in the .git folder \footnote{\url{https://git-scm.com/}}:
+  Version control in a dashboard can be supported in two ways: 
+  (1) the developer explicitly states the current version, or 
+  (2) the current version is automatically detected based on some version control system.
+
+  For \tool to start using Git, an extra configuration line enables it to automatically\footnote{Alternatively, the maintainer can add version identifiers manually for the web application through a configuration file if the system does not use Git.} associate API traffic with the currently deployed version based on information in the .git folder \footnote{\url{https://git-scm.com/}}:
 
@@
   \Fref{fig:mv-util} shows the \perspective{Multi-Version API Utilization} perspective which presents the utilization of the tracked endpoints across versions.
-Because the different versions might be deployed for very different periods of time, at the intersection of an endpoint and a version the chart does not plot the absolute utilization of that endpoint in that version but rather the percentage of all the API calls that go to that endpoint.
-Otherwise a version that is deployed for many weeks would make all those deployed for several days invisible.
+The view is a matrix, in which, the intensity of the color at the intersection of an endpoint line and a version column is proportional to the percentage of requests served by that endpoint in that particular version\footnote{The chart does not plot the absolute utilization of that endpoint in that version but rather the percentage of all the API calls that go to that endpoint.
+Otherwise a version that is deployed for many weeks would make all those deployed for a few days invisible}.
 
@@
 
-  With this view, several patterns are visible:
+  With this view, several patterns could be visible on such a graph:
   \begin{itemize}
 
-    \item \epFeedItems is discontinued a little bit after \epTopArticles is being introduced
+    \item Endpoints being deprecateds. In the case study \epFeedItems stops being used somewhere after the half of the presented timeline.
+
+    \item Endpoints being introduced. In the case study, \epTopArticles appears soon after the half o the presented timeline.
+
+    \item Endpoint renames. Although not visible in the figure, the candidates for rename would be visible as one would discontinued at the same time when the other would appear.
 

### 2018-04-06 grouping.tex
...

@@
 
-It is often the case that a service has diferent types of users: the paying and the free-for-evaluation ones, or the in-house and those in a different organization.
+It is sometimes the case that the utilization and performance of an API must be understood not only per endpoint, but also by grouping the requests by another criterion. Two such examples are: 
+\begin{itemize}
+	\item When a service has diferent types of users (e.g.
+the paying vs.
+the free-for-evaluation ones, or the in-house vs.
+those in a different organization) the maintainer must understand the performance for the different groups
+	\item When the load mix o the users can vary dramatically and the system response time is a function of the individual user load\footnote{E.g.
+in GMail some users have a few dozen emails while other have tens of thousands: this difference in user loads will eventually induce a difference in the response times for different users} the maintainer must understand the variation of performance with the users (and implicitly the user load).
+s
+\end{itemize}
 
-In some other cases, it might be important for the maintainer of the API to understand the performance on a per-user basis, especially for situations where the system response time is a function of some individual user load\footnote{E.g.
-in GMail some users have a few dozen emails while other have tens of thousands: this difference in user loads will eventually induce a difference in the response times for different users}.
-
-To support narrowing down the analysis to groups of users, or even individual users, the \tool must log together with every request grouping information for that request.
+To support narrowing down the analysis to groups of users, or even individual users, the \tool can log together with every request grouping information for that request.
 The simplest way of achieving this is to take advantage of the architecture of Flask applications in which a global \code{flask.request} is used to retrieve session information which can in turn is normally used to identify the user sending the request.
@@
 
-The following code snippet shows how to configure the dashboard to enable user-by-user analysis\footnote{	If we wanted to group by the user group for example, the code would change slightly by replacing ``user.id'' with ``user.group.id''}:
+\niceseparator
+
+The following code snippet shows how we enabled \tool to enable user-by-user analysis\footnote{If we wanted to group by the user group for example, the code would change slightly by replacing ``user.id'' with ``user.group.id''}:
 
@@
 
-% \niceseparator
+In our case study, we wanted to understand the per-user performance of \epFeedItems 
+--  an endpoint that retrieves a list of recommended articles for a given user.
+It is the endpoint with the slowest response time and highest variability (see \Fref{fig:ep}).
 
-In Zeeguu, \epFeedItems retrieves a list of recommended articles for a given user.
-Cf.
-\Fref{fig:ep} it is the endpoint with the slowest response time and highest variability.
-The reason for this is that a user can be subscribed to anything from one to three dozen article sources and for each of the sources the system must compute the personalized difficulty of each article at every request.
+% Once the \tool is configured with the line of code aforementioned.
 
@@
 
-A \perspective{Per-User Performance} perspective should show the different response times for different users.
+After configuring the \tool as shown earlier, the \perspective{Per-User Performance} perspective becomes available to  present the different response times for different users.
 Figure \ref{fig:tpu} presents a subset of the corresponding view in the \tool.
-The figure shows that the response times for this endpoint can vary considerably for different users with some extreme cases where a user has to wait a full minute until their recommended articles are shown\footnote{After seeing this perspective, the maintainer refactored the architecture of the system to move part the difficulty computation out of the interactive loop}.
+The figure shows that the response times for this endpoint varied considerably for different users with some extreme cases where a user has to wait a full minute until their recommended articles were shown.
+
+The reason for this behavior is that a user can be subscribed to anything from one to three dozen 
+article sources and for each of the sources the system computed the personalized difficulty 
+of each article at every request.
+Only after seeing the two perspectives presented in Figures \ref{fig:ep} and \ref{fig:tpu}, the maintainer refactored the architecture of the system to move the difficulty computation out of the interactive loop.

### 2018-04-06 evolution.tex
Edits to Sections 2 to 6

--- 2018-04-06 evolution.tex
@@
   \section{Version-Aware Monitoring}
+  \label{sec:version}
 
@@
 
-  Version control in \tool can be supported in two ways,
+  Version control in \tool can be supported in two ways:
   the developer explicitly states the current version, 
@@
   The second option is the most interesting one.
-  For the systems deployed using Git, an extra configuration line enables \tool to automatically\footnote{Alternatively, the maintainer can add version identifiers manually for the web application through a configuration file if the system does not use git.} associate API traffic with the currently deployed version based on information in the .git folder \footnote{\url{https://git-scm.com/}}:
+  For the systems deployed using Git, an extra configuration line enables \tool to automatically\footnote{Alternatively, the maintainer can add version identifiers manually for the web application through a configuration file if the system does not use git.} associate API traffic with the currently deployed version based on information in the .git folder\footnote{\url{https://git-scm.com/}}:
 
@@
 
-    \begin{figure}[h!]
+    \begin{figure}
       \centering
@@
     \item \epFeedItems is discontinued a little bit after \epTopArticles is being introduced
+	\item TBA
 
@@
 \subsection*{Evolving Grouped Performance}
-  The limitation of the previous view is that it does not present the information also on a per version basis.
+  The limitation of the previous view is that it does not present the information also on a per user basis.
 To address this, a different visual perspective entitled \perspective{User-Focused Multi-Version Endpoint Performance} can be defined.
-Figure \ref{fig:tuv} presents such a perspective by mapping the average execution time for a given user (lines) and given version (columns) on the area of the corresponding circle.
+\Fref{fig:tuv} presents such a perspective by mapping the average execution time for a given user (lines) and given version (columns) on the area of the corresponding circle.
 
@@
 number of sources to which the user is registered) is the reason for the variation in response times.
-\ins{one can see that for user 1 performance degrades over the versions.
-Given that for other users the performance does not degrade in the same way, it is probable that the problem might lay elsewhere: the server was overloaded or the endpoint is ``algorithmically slower''.}
+\ins{One can see that for user with id=1, performance degrades over consecutive versions.
+Given that for other users the performance does not degrade in the same way, it is probable that the problem might lay elsewhere: the server was overloaded or the endpoint is ``algorithmically slower''.} \va{could it be that the user kept on adding sources to their profile?}

### 2018-04-06 grouping.tex
Edits to Sections 2 to 6

--- 2018-04-06 grouping.tex
@@
 
-\section{Grouping Requests}
+\section{Per-User Performance Monitoring}
 \label{sec:user}
@@
 \begin{itemize}
-	\item \code{'User'} stands for the name of this particulart grouping strategy
-	\item \code{lambda: Session.find(... } is a callback with no arguments, 
-	that makes use of the global flask.request and extracts from 
+	\item \code{'User'} stands for the name of this particular grouping strategy
+	\item \code{lambda: Session.find... } is a callback with no arguments, 
+	that makes use of the global \code{flask.request} and extracts from 
 	the request the current session, and in turn the user id\footnote{
@@
 
-In Zeeguu, \epFeedItems retrieves a list of recommended articles for a given user.
-Cf.
-\Fref{fig:ep} it is the endpoint with the slowest response time and highest variability.
+In Zeeguu, the \epFeedItems endpoint retrieves a list of recommended articles for a given user.
+As shown in \Fref{fig:ep}, this is the endpoint with the slowest response time and highest variability.
 The reason for this is that a user can be subscribed to anything from one to three dozen article sources and for each of the sources the system must compute the personalized difficulty of each article at every request.
@@
   \centering
-  \includegraphics[width=.5\linewidth]{time_per_user}
+  \includegraphics[width=.7\linewidth]{time_per_user}
   \caption{The \epFeedItems shows a very high variability across users}
@@
 A \perspective{Per-User Performance} perspective should show the different response times for different users.
-Figure \ref{fig:tpu} presents a subset of the corresponding view in the \tool.
+\Fref{fig:tpu} presents a subset of the corresponding view in the \tool.
 The figure shows that the response times for this endpoint can vary considerably for different users with some extreme cases where a user has to wait a full minute until their recommended articles are shown\footnote{After seeing this perspective, the maintainer refactored the architecture of the system to move part the difficulty computation out of the interactive loop}.

### 2018-04-06 introductory.tex
...
Edits to Sections 2 to 6

--- 2018-04-05 introductory.tex
@@
 
-\section{Case Study: The API (Zeeguu)}
+\section{Case Study}
 \label{sec:case}
@@
 %\todo{Either leave as is or add more info on usage by college if possible}
+\subsection*{Case Study: The API (Zeeguu)}
+\label{sec:api}
 
@@
 
-    \item Reader applications that provide effortless translations for those texts which are too difficult for the readers.
+    \item \textit{Reader applications} that provide effortless translations for those texts which are too difficult for the readers.
 
-    \item Interactive exercises personally generated based on the preferences and past likes of the learner.
+    \item \textit{Interactive exercises} that are personally generated based on the preferences and past likes of the learner.
 
-    \item Article recommendations which are personalized for the interests of the learner and come with difficulty estimation that helps the learner find articles with the appropriate difficulty.
+    \item \textit{Article recommendations} which are also personalized for the interests of each learner and come with a difficulty estimation that helps the learner find articles with appropriate difficulty.
 
@@
 
-  The core API implemented with Flask and Python provides correspondingly three types of functionality: contextual translations, article recommendations, and personalized exercise suggestions.
-In total the API provides a bit less than 50 endpoints, out of which probably a dozen are very frequently used.
+  The core API is implemented with Flask and Python provides and correspondingly three types of functionality: contextual translations, personalized exercise suggestions, and article recommendations.
+In total, the API provides a bit under 50 endpoints, out of which around a dozen are very frequently used.
   % The development of the core API itself is a research project. 
-
-  At the time of writing this article, the ecosystem consists of a reader web application, a web based exercises platform, and a smartwatch application, which are used at the moment of writing this article by more than \activeUserCount active users.
-The users come from a highschool, a language school, and some users are using it on their own, without any educational context.
-The highest load we observed until now on the API consisted of 12K requests in one day.
+  %
+  At the time of writing, the ecosystem consists of a reader web application, a web-based exercises platform, and a smartwatch application, which are used by more than \activeUserCount active users on a regular basis.
+  %The majority of the users come from a high school and a language school in the city of Groningen.
+However, some users are using it on their own, outside of any formal educational context.
+The highest load we observed until now on the API consisted of 12K requests in a single day.
 
   \ml{add more numbers that show how impressive is our API}
+  \va{I don't think this is necessary - I am also moving the user characteristics in the respective subsection below}
 
   We will use this \zee API as a case study for this paper. 
-  All the figures in this paper are captured from the actual deployment of \tool in the context of the \zee API.
+  All the figures and measurements in this paper are captured from the actual deployment of \tool in the context of the \zee API.
 The figures are interactive offering basic data exploration capabilities: filter, zoom, and details on demand\cite{Shne99a}.
@@
 
-\section{Case Study: The Technology (Flask)}
-\label{sec:tool}
+\subsection*{Case Study: The Technology (Flask)}
+\label{sec:flask}
+
+ \va{Consider this subsection moving before the introduction of Zeeguu; it would make a smoother reading}
 
  Flask is a microframework for Python. It IS USED BY THOUSANDS OF PROJECTS. ELABORATE MORE ON THIS... 
- \vspace{2cm}
-
- Flask depends on two external libraries: the Jinja2 template engine and the Werkzeug WSGI toolkit.
-
- “Micro” does not mean that your whole web application has to fit into a single Python file (although it certainly can), nor does it mean that Flask is lacking in functionality.
-The “micro” in microframework means Flask aims to keep the core simple but extensible.
-Flask won’t make many decisions for you, such as what database to use.
+ %\vspace{2cm}
+ \va{Actually, you probably have already introduced Flask in the first section, so you just need the first sentence: As discussed in the previous section, Flask is a microframework for Python.}
+%
+% Flask depends on two external libraries: the Jinja2 template engine and the Werkzeug WSGI toolkit. 
+%
+ ``Micro'' does not mean that a whole web application has to fit into a single Python file (although it certainly can), nor does it mean that Flask is lacking in functionality.
+The ``micro'' in microframework means Flask aims to keep the core simple but extensible.
+Flask won’t make many decisions for the user, such as what database to use.
 Those decisions that it does make, such as what templating engine to use, are easy to change.
-Everything else is up to you, so that Flask can be everything you need and nothing you don’t.
+Everything else is up to the user.%, so that Flask can be everything you need and nothing you don’t.
 
  Flask has many configuration values, with sensible defaults, and a few conventions when getting started.
-By convention, templates and static files are stored in subdirectories within the application’s Python source tree, with the names templates and static respectively.
-While this can be changed, you usually don’t have to, especially when getting started.
-
+By convention, templates and static files are stored in subdirectories within the application's Python source tree, with the names templates and static respectively.
+While this can be changed, it usually don’t have to, especially when getting started.
+%
  A Python web application based on WSGI has to have one central callable object that implements the actual application.
@@
 Each Flask application has to create an instance of this class itself and pass it the name of the module, but why can’t Flask do that itself?
+ \va{Do you really need the part in the previous sentence after the comma? Just ref the listing}
 
@@
 
-\subsection*{ENSURE CLARITY: NOTHING PREVENTS GENERALIZATION}
+\subsection*{Study characteristics}
+
+The \tool has been deployed with the case study API for 
+ten months, time in which more than 112.000 hits of the API have 
+been recorded. \va{which ten months?}
+The majority of the users come from a high school and a language school in the city of Groningen using \zee to study foreign languages.
+However, some users are using it on their own, outside of any formal educational context.
+The highest load we observed until now on the API consisted of 12K requests in a single day.
+\va{not sure what can be added here; maybe that's enough}
+
+\subsection*{Role of Authors in the Case Study}
+\va{not sure what else you would like to add here...}
+One of the authors leads the developer team that develops the \zee API.
+
+Another one of the authors is the main developer in a team 
+that develops the \tool.
+
+\subsection*{TO ARGUE: NOTHING PREVENTS GENERALIZATION}
+\va{Move this to the discussion section - it's premature to discuss this here} 
 In this article we take as an example the Flask case study

### 2018-04-06 loc1.tex
...
Edits to Sections 2 to 6

@@
 
-  The tool is designed to take advantage of the fact that it runs in the same web application context and that the monitored API already has a web presence.
+  The \tool is designed to take advantage of the fact that it runs in the same context as the web application it monitors, and that the monitored API already has a web presence.
 It thus simply makes available one extra endpoint (i.e.
-\code{/dashboard}) at the same address as the API, endpoint which serves the interactive application presented in the remainder of this paper.
+\code{/dashboard}) at the same address as the API, which serves the interactive application presented in the remainder of this paper.
 
-  A dashboard can be attached to an already existing Flask application with a single line of code as the code in the following snippet shows
-  \footnote{We do not count the import statement that enables that line}:
+  A dashboard can be attached to an already existing Flask application with a single line of code~\cite{vogel2017low} as the following snippet shows\footnote{We do not count the import statement that enables that line.}:
 
-  % which binds the dashboard to a Flask application\footnote{\ins{ In this paper we present the integration with APIs written in Python and Flask hoping that this will not prevent the reader from seeing the more general idea; all the tools we show here for Flask can be applied to other API technologies (e.g.
-Django) by simply providing a few back-end adapters in the right places.
-}}:
-
-  % caption=Configuring the \tool is straightforward,
   \begin{lstlisting}[style=custompython]
@@
 
+  \va{\Fref{fig:sep} needs to be updated --- it still has 2017 dates}
+
   % In order to monitor an endpoint, the \tool creates a function wrapper for the API function that corresponds to the endpoint.
@@
 
-    \begin{figure}[h!]
+    \begin{figure}[h]
       \centering
@@
 
-  By default the dashboard takes an {\em opt-in} approach to monitoring: to prevent the performance penalties incurred by the dashboard\footnote{We discuss performance issues later} to affect performance sensitive endpoints, the service developer is asked to manually add the endpoints they want to monitor.
+  By default the dashboard takes an {\em opt-in} approach to monitoring: to prevent the performance penalties incurred by the dashboard\footnote{We discuss performance issues later.} to affect performance-sensitive endpoints, the service developer is asked to manually define the endpoints they want to monitor.
 
@@
     \item {\bf Utilization} -- refers to how the third parties use the API, which parts are most used, which are little used, etc.
-This is known to be critical information for upstream developer in general\cite{Haen14a}.
+This is known to be critical information for upstream developers in general\cite{Haen14a}.
 In the case of source code dependencies one can detect such information by analyzing software repositories.
@@
 
-    \item {\bf Performance} as measured in response times. This is important, especially for APIs that are supposed to be integrated in live systems.
+    \item {\bf Performance} as measured in response times.
+Knowing the performance of one's API is critical for the quality of a service API, and especially so for APIs that are supposed to be integrated in live systems.
 

### 2018-04-06 performance.tex
intro to evolution
Edits to Sections 2 to 6

@@
 \Fref{fig:aeu}) is \epTranslationsColor.
-The endpoint is critical being embedded in the interactive loop in the reader applications that rely on the Zeeguu API: a user would tap on a word and receive a translation from this endpoint.
+The endpoint is critical for the reader applications that rely on the Zeeguu API: each time a user taps on a word on their screen they receive a translation from this endpoint.
 
@@
 
-\begin{figure}[!ht]
+\begin{figure}
  \centering
@@
 However, in order to understand the problems of the two endpoints, a big component represents understanding the outliers: because they are many.
+\va{the last sentence does not parse}
 
   \subsection{Automated Outlier Detection and Monitoring}
+  \va{Shouldn't this be promoted to a section?
+It is neither utilization- nor performance- related.
+Either that, or Section~\ref{sec:user} should be folded into this section too, and then Sections~\ref{sec:version} and~\ref{sec:regression} merged into another section.
+Otherwise we have a very scattered paper}
 
@@
   of particular interest to the API developers are performance {\em outliers}.
-
 Indeed, a translation request that takes three times more than expected can seriously decrease the perceived quality of the application.
@@
     \caption{Automatically collected outlier information}
-    \label{fig:figure1}
+    \label{fig:stack}
   \end{figure}
 
-The bottom panel shows the stack trace. 
-In this particular case, it is revealing for the developer to learn that at the time of the stack trace snapshot, the code was in the google\_translator.py: indeed, the system uses as back-end multiple translators, and it has been observed that many of the outliers happen while the system is waiting for the Google translator.
-
+The bottom panel in \Fref{fig:stack} shows the stack trace. 
+In this particular case, it is revealing for the developer to learn that at the time of the stack trace snapshot, the code was in the \code{google\_translator}: indeed, the system uses as back-end multiple translators, and it has been observed that many of the outliers happen while the system is waiting for the Google translator.
+%
 This information has to be corroborated with the observations that neither the memory nor the processor are overloaded at the moment.
-Thus this functionality in microsoft\_translator is really slow in itself, and this is not a result of the machine being overloaded for example.
+Thus this functionality in \code{microsoft\_translator} is really slow in itself, and this is not a result of the machine being overloaded for example.
+\va{last sentence does not parse}

### 2018-04-06 grouping.tex

--- 2018-04-06 grouping.tex
@@
 
-In our case study, we wanted to understand the per-user performance of \epFeedItems 
+In Zeeguu, it was decided to further understand the per-user performance of \epFeedItems
 --  an endpoint that retrieves a list of recommended articles for a given user.
@@
 
-% Once the \tool is configured with the line of code aforementioned.
-
 \begin{figure}[h!]
   \centering
-  \includegraphics[width=.5\linewidth]{time_per_user}
+  \includegraphics[width=.7\linewidth]{time_per_user}
   \caption{The \epFeedItems shows a very high variability across users}

### 2018-04-06 paper.tex
intro to evolution

@@
 \IEEEauthorblockN{The Author, The Other Author}\\
-Institute\\
-University, Country\\
+Johann Bernoulli Institute for Mathematics and Computer Science\\
+University of Groningen, Groningen, the Netherlands\\
 Email: \{author\}@university.domain 
@@
 
-\input{architecture}
+% \input{architecture}
 

### 2018-04-06 testing.tex
cleaning up stuff
...

@@
 
-\newpage
-  \section{Performance Regression Monitoring}
+  \section{Performance Triangulation With Regression Monitoring}
+  \label{sec:regression}
 
@@
 
-  Each time a new build is created through Travis, the \tool automatically detects all available unit tests defined by the application developer (in (-{}-test\_folder)) and iterates through each one of them a number of times (-{}-times) while monitoring the response times for each test.
-The resulting measurements are uploaded to a specific endpoint that is created by the tool at the API url (-{}-url).
+  Each time a new build is created through Travis, the \tool automatically detects all available unit tests defined by the application developer (in (\code{-{}-test\_folder)}) and iterates through each one of them a number of times (\code{-{}-times}) while monitoring the response times for each test.
+The resulting measurements are uploaded to a specific endpoint that is created by the tool at the API url (\code{-{}-url}).
 The data uploaded to this endpoint is   persisted in a separate part of the tool database so as not to contaminate the ``live'' monitoring data.
@@
         \centering
-        \includegraphics[width=0.45\columnwidth]{travis_builds}
+        \includegraphics[width=0.65\columnwidth]{travis_builds}
         \caption{Response times in 5 consequent Travis builds}
@@
 The outliers on the right of the figure are due to initial requests \ins{which must wait for a boot-up phase of the API}.
+\va{Remember to make greyscale for consistency}
 
@@
   While this integration testing environment is different from the production one, and the load used is purely synthetic, it can serve as an early performance indicator for the developer.
-Figure \ref{fig:response_times_preemptive} shows the actual endpoint response times across five versions of the system (above) and the corresponding testing times for the synthetic load in the same five versions (below).
+\Fref{fig:response_times_preemptive} shows the actual endpoint response times across five versions of the system (above) and the corresponding testing times for the synthetic load in the same five versions (below).
 The green line joining the medians hints to a correlation between the two sets of measurements.
@@
 
-  Computing Pearson correlation ($r(3)=.93, p=.02$) between the median values of the two datasets (cf. Table \ref{tab:correlations}) shows that there is a correlation between the two.
+  Computing Pearson correlation ($r(3)=.93, p=.02$) between the median values of the two datasets (cf. Table \ref{tab:correlations}) shows that there is indeed a correlation between the two.
 
@@
 Worst case, we move it to the conclusion}
+  \va{I concur}
   \ins{Besides functioning as an early warning system for performance, tracking the evolving performance of API tests serves a second purpose.

### 2018-04-06 utilization.tex

@@
   The default landing page of the dashboard presents an overview of all the endpoints, 
-  overview which presents some basic utilization information as:
+  which presents some basic utilization information as:
   (1) the number of calls to that endpoint since the begining of tracking
@@
 An argument for slow research}s
+  \va{Please remove the last footnote - it's the equivalent of shooting yourself in the leg.}
 
@@
 
-        \item \epTranslationsColor is an indicator of the amount of foreign language reading the users are doing
+        \item \epTranslationsColor is an indicator of the amount of foreign language reading the users are doing.
 
@@
 
-  For a more detailed evolution of utilization, Figure \ref{fig:aeu} shows the \perspective{Daily API Utilization} perspective on API utilization that \tool provides: a stacked bar chart of the number of hits to various endpoints grouped by day\footnote{Endpoint colors are the same in different views}.
-Figure~\ref{fig:aeu} in particular shows a peak utilization, on June 28 when the API had more than 12.000 hits\footnote{Turns out that the highschool students that were using the system for their class had a deadline on that date}.
+  For a more detailed view of utilization evolution, \Fref{fig:aeu} shows the \perspective{Daily API Utilization} perspective on API utilization that \tool provides: a stacked bar chart of the number of hits to various endpoints grouped by day\footnote{Endpoint colors are the same in different views}.
+\Fref{fig:aeu} in particular shows a peak utilization, on June 28, 2017 when the API had more than 12.000 hits\footnote{Turns out that the high school students that were using the system for their French-learning class had a deadline on that date}.
 
@@
 
-    \begin{figure}[!ht]
+    \begin{figure}
       \centering

### 2018-04-07 introductory.tex
working on the case study section

@@
 \section{Case Study}
-\label{sec:case}
 
-%\todo{Either leave as is or add more info on usage by college if possible}
-\subsection*{Case Study: The API (Zeeguu)}
-\label{sec:api}
-
-  \zee\footnote{\url{https://github.com/zeeguu-ecosystem/}} is a platform and an ecosystem of applications for accelerating vocabulary acquisition in a foreign language \cite{Lungu16}. 
-%
-  The architecture of the ecosystem has at its core an API and a series of satellite applications that together offer to a learner three main inter-dependent features:
-
-  \begin{enumerate}
-
-    \item \textit{Reader applications} that provide effortless translations for those texts which are too difficult for the readers.
-
-    \item \textit{Interactive exercises} that are personally generated based on the preferences and past likes of the learner.
-
-    \item \textit{Article recommendations} which are also personalized for the interests of each learner and come with a difficulty estimation that helps the learner find articles with appropriate difficulty.
-
-  \end{enumerate}
-
-  The core API is implemented with Flask and Python provides and correspondingly three types of functionality: contextual translations, personalized exercise suggestions, and article recommendations.
-In total, the API provides a bit under 50 endpoints, out of which around a dozen are very frequently used.
-  % The development of the core API itself is a research project. 
-  %
-  At the time of writing, the ecosystem consists of a reader web application, a web-based exercises platform, and a smartwatch application, which are used by more than \activeUserCount active users on a regular basis.
-  %The majority of the users come from a high school and a language school in the city of Groningen.
-However, some users are using it on their own, outside of any formal educational context.
-The highest load we observed until now on the API consisted of 12K requests in a single day.
-
-  \ml{add more numbers that show how impressive is our API}
-  \va{I don't think this is necessary - I am also moving the user characteristics in the respective subsection below}
-
-  We will use this \zee API as a case study for this paper. 
-  All the figures and measurements in this paper are captured from the actual deployment of \tool in the context of the \zee API.
-The figures are interactive offering basic data exploration capabilities: filter, zoom, and details on demand\cite{Shne99a}.
-The \tool deployment for the case study can be accessed publicly\footnote{\url{https://zeeguu.unibe.ch/api/dashboard}; username: {\em guest}, password: {\em soap-sac}}.
-
- % \todo{create new account, update URL if necessary, update activeUserCount, add a couple of sentences of the current state of the case study with the college/language center}
-% \ml{we should consider adding also one section in which the architecture/implementation and main features of the dashboard are presented before going on with discussing them in more depth in the following sections --- this should include a rundown on which views are provided from where (overview or per endpoint)}
-
-\newpage
-
-\subsection*{Case Study: The Technology (Flask)}
+\subsection{The Technology Stack}
 \label{sec:flask}
 
- \va{Consider this subsection moving before the introduction of Zeeguu; it would make a smoother reading}
-
- Flask is a microframework for Python. It IS USED BY THOUSANDS OF PROJECTS. ELABORATE MORE ON THIS... 
- %\vspace{2cm}
- \va{Actually, you probably have already introduced Flask in the first section, so you just need the first sentence: As discussed in the previous section, Flask is a microframework for Python.}
-%
-% Flask depends on two external libraries: the Jinja2 template engine and the Werkzeug WSGI toolkit. 
-%
+ As discussed in the previous section, Flask is a microframework for Python.
 ``Micro'' does not mean that a whole web application has to fit into a single Python file (although it certainly can), nor does it mean that Flask is lacking in functionality.
@@
 Those decisions that it does make, such as what templating engine to use, are easy to change.
-Everything else is up to the user.%, so that Flask can be everything you need and nothing you don’t.
+Everything else is up to the user.
 
@@
 In Flask this is an instance of the Flask class.
-Each Flask application has to create an instance of this class itself and pass it the name of the module, but why can’t Flask do that itself?
- \va{Do you really need the part in the previous sentence after the comma? Just ref the listing}
+Each Flask application has to create an instance of this class itself and pass it the name of the module.
+The following code snippet presents a simple ``Hello World'' web service written in Python with the help of the Flask framework.
 
@@
 
+%\todo{Either leave as is or add more info on usage by college if possible}
+\subsection{An Ecosystem With a Service API at its Core}
+\label{sec:api}
+\label{sec:case}
+
+  \zee\footnote{\url{https://github.com/zeeguu-ecosystem/}} is a platform and an ecosystem of applications for accelerating vocabulary acquisition in a foreign language \cite{Lungu16}. 
+%
+  The architecture of the ecosystem has at its core an API and a series of ecosystem applications that together offer to a learner three main inter-dependent features:
+
+  \begin{enumerate}
+
+    \item \textit{Reader applications} that provide one-click (or one touch, on touch-enabled devices) translations for the words and expressions in a text that the reader does not understand.
+Besides facilitating reading, the translations serve as input to a user knowledge model that is used further to generate personalized exercises and further reading recommendations.
+
+    \item \textit{Interactive exercises} are generated based on the texts that the reader has studied in the past and are scheduled in such a way as to optimize the memory retention and to prioritize the most important unknown vocabulary items.
+
+    \item \textit{Article recommendations} are also personalized for the interests and language capabilities of each learner.
+The recommendations come with a difficulty estimation that helps the learner find articles with appropriate difficulty.
+
+  \end{enumerate}
+
+  \Fref{fig:zeeguuarch} presents a high-level view of the case study ecosystem.
+The core API is implemented with Flask and Python provides and correspondingly three types of functionality: contextual translations, personalized exercise suggestions, and article recommendations.
+In total, the API provides a bit under 50 endpoints, out of which around a dozen are very frequently used.
+The development of the core API itself is a research project.
+One of the is part of the developer team that maintains and evolves the \zee API.
+
+    \begin{figure}[h!]
+      \centering
+      \includegraphics[width=0.7\linewidth]{zeeguu-architecture}
+      \caption{The architecture of the Zeeguu ecosystem}
+      \label{fig:zeeguuarch}
+    \end{figure}
+
+  At the time of writing, the ecosystem consists of a reader web application, a web-based exercises platform, and a smartwatch application, which are used by more than \activeUserCount active users on a regular basis.
+The highest load we observed until now on the API consisted of 12K requests in a single day.
+
+  We will use this \zee API as a case study for this paper. 
+  All the figures and measurements in this paper are captured from the actual deployment of \tool in the context of the \zee API.
+The figures are interactive offering basic data exploration capabilities: filter, zoom, and details on demand\cite{Shne99a}.
+The \tool deployment for the case study can be accessed publicly\footnote{\url{https://zeeguu.unibe.ch/api/dashboard}; username: {\em guest}, password: {\em soap-sac}}.
+
+ % \todo{create new account, update URL if necessary, update activeUserCount, add a couple of sentences of the current state of the case study with the college/language center}
+% \ml{we should consider adding also one section in which the architecture/implementation and main features of the dashboard are presented before going on with discussing them in more depth in the following sections --- this should include a rundown on which views are provided from where (overview or per endpoint)}
+
 \subsection*{Study characteristics}
@@
 been recorded. \va{which ten months?}
-The majority of the users come from a high school and a language school in the city of Groningen using \zee to study foreign languages.
+
+The users come from several Dutch highschools and a language center associated with a DutchUniversity.
 However, some users are using it on their own, outside of any formal educational context.
-The highest load we observed until now on the API consisted of 12K requests in a single day.
-\va{not sure what can be added here; maybe that's enough}
+The learners use \zee to study a variety of of foreign languages including German, Dutch, French.
+They have also diverse native languages, including: Chinese, Dutch, and English.
+
+The highest load we observed until now on the API consisted of 12K requests in a single day. \va{not sure what can be added here; maybe that's enough}
 
@@
 \va{not sure what else you would like to add here...}
-One of the authors leads the developer team that develops the \zee API.
 

### 2018-04-07 discussion.tex
intro to evolution
cleaned up quite nicely section 2

@@
 \section{Discussion}
+
+- could even start with this: NOTHING PREVENTS GENERALIZATION
+In this article we take as an example the Flask case study
+and of the \tool as a dashboard. 
+But there is nothing that prevents the generalization of 
+our approach to other languages and technologies.
 
@@
 - We need more advanced outlier analysis tools... Instead of manually searching in the page to build stats about Google vs. Microsoft
+
+- branches. the system does not care. it links to whatever commit is the current one. it could be that visualizations could be developed to comapre branches, but this optiona. future work.
 

### 2018-04-07 introductory.tex
cleaned up quite nicely section 2

@@
 
-%\todo{Either leave as is or add more info on usage by college if possible}
 \subsection{An Ecosystem With a Service API at its Core}
@@
 The development of the core API itself is a research project.
-One of the is part of the developer team that maintains and evolves the \zee API.
+One of the authors of this paper is part of the developer team that maintains and evolves the \zee API.
 
@@
       \includegraphics[width=0.7\linewidth]{zeeguu-architecture}
-      \caption{The architecture of the Zeeguu ecosystem}
+      \caption{The architecture of the \zee ecosystem}
       \label{fig:zeeguuarch}
@@
   At the time of writing, the ecosystem consists of a reader web application, a web-based exercises platform, and a smartwatch application, which are used by more than \activeUserCount active users on a regular basis.
-The highest load we observed until now on the API consisted of 12K requests in a single day.
-
-  We will use this \zee API as a case study for this paper. 
-  All the figures and measurements in this paper are captured from the actual deployment of \tool in the context of the \zee API.
-The figures are interactive offering basic data exploration capabilities: filter, zoom, and details on demand\cite{Shne99a}.
-The \tool deployment for the case study can be accessed publicly\footnote{\url{https://zeeguu.unibe.ch/api/dashboard}; username: {\em guest}, password: {\em soap-sac}}.
-
- % \todo{create new account, update URL if necessary, update activeUserCount, add a couple of sentences of the current state of the case study with the college/language center}
-% \ml{we should consider adding also one section in which the architecture/implementation and main features of the dashboard are presented before going on with discussing them in more depth in the following sections --- this should include a rundown on which views are provided from where (overview or per endpoint)}
+The users come from several Dutch highschools and a language center associated with a Dutch University.
+Several users are using it on their own, outside of any formal educational context.
+The learners use \zee to study a variety of of foreign languages including German, Dutch, French.
+They have also diverse native languages, including: Chinese, Dutch, and English.
+The system can scale to many languages by delegating the machine translation tasks to third party APIs.
 
@@
   The \tool has been deployed with the case study API for 
-ten months, time in which more than 112.000 hits of the API have 
-been recorded. \va{which ten months?}
+  ten months (between June 2017 and the time of writing this
+  article, April 2018), time in which more than 112.000 hits 
+  of the API have been recorded. The highest load observed 
+  until now on the API consisted of 12K requests in a single day.
 
-The users come from several Dutch highschools and a language center associated with a DutchUniversity.
-However, some users are using it on their own, outside of any formal educational context.
-The learners use \zee to study a variety of of foreign languages including German, Dutch, French.
-They have also diverse native languages, including: Chinese, Dutch, and English.
+  All the figures and measurements in this paper are captured from the actual deployment of \tool in the context of the \zee API.
+The figures are interactive offering basic data exploration capabilities: filter, zoom, and details on demand\cite{Shne99a}.
+The \tool deployment for the case study can be accessed publicly\footnote{\url{https://zeeguu.unibe.ch/api/dashboard}; username: {\em guest}, password: {\em icsme}}.
 
-The highest load we observed until now on the API consisted of 12K requests in a single day. \va{not sure what can be added here; maybe that's enough}
+  We will use this \zee API as a case study for this paper.
 
-\subsection*{Role of Authors in the Case Study}
-\va{not sure what else you would like to add here...}
-
-Another one of the authors is the main developer in a team 
-that develops the \tool.
-
-\subsection*{TO ARGUE: NOTHING PREVENTS GENERALIZATION}
-\va{Move this to the discussion section - it's premature to discuss this here} 
-In this article we take as an example the Flask case study
-and of the \tool as a dashboard. 
-But there is nothing that prevents the generalization of 
-our approach to other languages and technologies.
+ % \todo{create new account, update URL if necessary, update activeUserCount, add a couple of sentences of the current state of the case study with the college/language center}
+% \ml{we should consider adding also one section in which the architecture/implementation and main features of the dashboard are presented before going on with discussing them in more depth in the following sections --- this should include a rundown on which views are provided from where (overview or per endpoint)}

### 2018-04-07 introductory.tex
...

@@
 
+  We will use this \zee API as a case study for this paper.
+
   The \tool has been deployed with the case study API for 
   ten months (between June 2017 and the time of writing this
-  article, April 2018), time in which more than 112.000 hits 
-  of the API have been recorded. The highest load observed 
+  article, April 2018), time in which more than 112.000 requests
+  to the API have been recorded \footnote{This is a lower bounds
+  on usage, as only a subset of the API endpoints have been 
+  actually instrumented} . The highest load observed 
   until now on the API consisted of 12K requests in a single day. 
+  With the help of the applications in the ecoystem the users have 
+  performed about 30K vocabulary exercises and have translated 
+  about 28K words and experssions while reading foreign language 
+  texts.
 
@@
 
-  We will use this \zee API as a case study for this paper.
-
  % \todo{create new account, update URL if necessary, update activeUserCount, add a couple of sentences of the current state of the case study with the college/language center}

### 2018-04-07 utilization.tex
stupid footnote removed.

@@
   (2) the number of calls in the last seven days, and
-  (3) the number of calls in the last day\footnote{Although very useful, it was only after several months of using the system that the client requested the extra columsn for one day and seven days.
-An argument for slow research}s
-  \va{Please remove the last footnote - it's the equivalent of shooting yourself in the leg.}
+  (3) the number of calls in the last day
 

### 2018-04-07 introductory.tex
cleaning up the introduction.

--- 2018-04-07 introductory.tex
@@
 \begin{abstract}
+
   Python is one of the fastest growing programming languages of the moment.
-Flask, a Python-based web framework is the technology chosen by tens of thousands of web applications.
-Many of these Flask-based projects are small websites which can not afford to deploy an advanced service monitoring solution because they are too complicated to install.
-Many of these projects might offer a service API.
-Given the complexity of professional monitoring tools it is very likely that
+Flask, a Python-based web framework is the technology used by tens of thousands of web applications.
+These applications, from interactive websites to service APIs, lack a technology specific service monitoring solution.
 
-  In this paper, we set out to show how to design a dashboard for a Flask API which is useful in understanding and guiding service evolution while at the same time, requiring minimal configuration and installation.
-We present \tool, a tool that provide insight into the utilization and performance of evolving services.
-We present the ease with which the library can be integrated in an already existing web application, discuss some of the visualization perspectives that the library provides and point to some future challenges for similar libraries.
+  In this paper, we present \tool, a tool that provide insight into the utilization and performance of evolving Flask-based web services.
+We present the ease with which the tool can be integrated in an already existing web application, discuss some of the visualization perspectives that the library provides and point to some future challenges for similar libraries.
 
@@
 
-\todo{Intro to be rewritten; refer to VISSOFT paper as previous work, explain focus is on integration with version control/CI for evolution support, and use of unit tests as early performance indicator}
-
-\todo{motivate with: monitoring in a world which is becoming more and more service-oriented is essential because it allows to perform three main types of actions: system adaptations to provide the request service at the desired level of quality, enabling flexibility in dealing with changing requirements and modes of operation, and enabling operational awareness through dashboards organizing the collected data~\cite{pernici2016monitoring}.
+% \todo{motivate with: monitoring in a world which is becoming more and more service-oriented is essential because it allows to perform three main types of actions: system adaptations to provide the request service at the desired level of quality, enabling flexibility in dealing with changing requirements and modes of operation, and enabling operational awareness through dashboards organizing the collected data~\cite{pernici2016monitoring}.
 In this work we focus on the last part and we discuss how a minimal effort probe-based solution for a particular type of web services can be used to facilitate the other two types of actions by involving the developer.}
 
-\todo{explain: in~\cite{vogel2017low} we introduced the \tool an a high level with a focus on presenting its performance visualization aspects; in this work we only summarize these features by means of introducing some of the provided functionalities of the \tool.
+% \todo{explain: in~\cite{vogel2017low} we introduced the \tool an a high level with a focus on presenting its performance visualization aspects; in this work we only summarize these features by means of introducing some of the provided functionalities of the \tool.
 We focus on discussing how the tool is implemented and operating, provide a deeper look at its capabilities for integration with version control and continuous integration environments, introduce a mechanism for performance prediction, and provide an evaluation of the proposed approach based on a case study.
@@
 %\ml{I'm not completely happy with this paragraph}
-{\em There is no getting around it: you are building a distributed system} argues a recent article \cite{cavage2013there}.
+{\em There is no getting around it: you are building a distributed system} argues Cavage  in a recent article written for the {\em Communications of the ACM} \cite{cavage2013there}.
 Indeed, even the simplest second-year student project is a web application implemented as two-tier architecture with a Javascript/HTML5 front-end a service backend, usually a REST API.
@@
 
-\todo{Add signposting paragraph}
+\subsection*{Extending Previous Work}
+
+This paper is an extension of [Anonymous Paper by Anonymous Auhthors] which was published as a NIER paper in [Anonymous Conference]. The current paper extends that paper in the following ways: 
+\begin{itemize}
+  \item Providing more details about the case study (in Section \ref{sec:case}) and extending the data collection period with half year
+  \item A better categorization of the dashboard perspectives presented in Sections \ref{sec:util}, \ref{sec:grouping}, \ref{sec:evolution} including several new perspectives
+  \item A detailed discussion on automated outlier detection and monitoring (Section \ref{sec:outliers})
+  \item An advanced discussion integration continuous integration environments in order to triangulate the live collected data with information from testing the endpoints with a constant load and introduce a mechanism for performance prediction (Section \ref{sec:testing})
+  \item A discussion on the overhead incurred by the presented tool \ml{(... if we make it... )}
+\end{itemize}
 
@@
 
-\subsection{The Technology Stack}
-\label{sec:flask}
-
- As discussed in the previous section, Flask is a microframework for Python.
-``Micro'' does not mean that a whole web application has to fit into a single Python file (although it certainly can), nor does it mean that Flask is lacking in functionality.
-The ``micro'' in microframework means Flask aims to keep the core simple but extensible.
-Flask won’t make many decisions for the user, such as what database to use.
-Those decisions that it does make, such as what templating engine to use, are easy to change.
-Everything else is up to the user.
-
- Flask has many configuration values, with sensible defaults, and a few conventions when getting started.
-By convention, templates and static files are stored in subdirectories within the application's Python source tree, with the names templates and static respectively.
-While this can be changed, it usually don’t have to, especially when getting started.
-%
- A Python web application based on WSGI has to have one central callable object that implements the actual application.
-In Flask this is an instance of the Flask class.
-Each Flask application has to create an instance of this class itself and pass it the name of the module.
-The following code snippet presents a simple ``Hello World'' web service written in Python with the help of the Flask framework.
-
-\begin{lstlisting}[style=custompython]
-
-from flask import Flask
-app = Flask(__name__)
-
-@app.route('/')
-def index():
-    return 'Hello World!'
-
-\end{lstlisting}
-
-\subsection{An Ecosystem With a Service API at its Core}
+% \subsection{An Ecosystem With a Service API at its Core}
 \label{sec:api}
@@
 
-  The \tool has been deployed with the case study API for 
-  ten months (between June 2017 and the time of writing this
-  article, April 2018), time in which more than 112.000 hits 
-  of the API have been recorded. The highest load observed 
+  We will use this \zee service API as a case study for this paper. 
+  The API has been deployed for several years without any monitoring
+  solution until June 2017, when the \tool has been deployed with it. 
+  During the ten month deployment (between June 2017 and the time of writing this
+  article, April 2018) more than 112.000 requests
+  to the API have been recorded by the API \footnote{Note that the 
+  recorded requests are a lower bound on the usage of the API, 
+  as only a subset of the API endpoints have been actually instrumented} . The highest load observed 
   until now on the API consisted of 12K requests in a single day. 
+  With the help of the applications in the ecoystem, in the aforementioned
+  time period the users have 
+  performed about 30K vocabulary exercises and have translated 
+  about 28K words and experssions while reading foreign language 
+  texts.
 
@@
 
-  We will use this \zee API as a case study for this paper.
-
  % \todo{create new account, update URL if necessary, update activeUserCount, add a couple of sentences of the current state of the case study with the college/language center}
 % \ml{we should consider adding also one section in which the architecture/implementation and main features of the dashboard are presented before going on with discussing them in more depth in the following sections --- this should include a rundown on which views are provided from where (overview or per endpoint)}
+
+\subsection{The Technology Stack}
+\label{sec:flask}
+
+ As discussed in the previous section, Flask is a microframework for Python. 
+ The ``micro'' in microframework means Flask aims to keep the core simple but extensible. 
+ The framework does not make many decisions for the user, such as what database to use and
+ those decisions that it does make, such as what templating engine to use, can be changed.
+
+ Flask is built on top of WSGI -- The Web Server Gateway Interface -- a calling convention for web servers to forward requests to web applications or frameworks written in Python.
+A Python web application based on WSGI has to have one central callable object that implements the actual application.
+In Flask this is an instance of the Flask class.
+Each Flask application has to create an instance of this class itself and pass it the name of the module.
+The following code snippet presents a simple ``Hello World'' web service written in Python with the help of the Flask framework.
+
+\begin{lstlisting}[style=custompython]
+
+from flask import Flask
+app = Flask(__name__)
+
+@app.route('/')
+def index():
+    return 'Hello World!'
+
+\end{lstlisting}

### 2018-04-07 paper.tex
cleaning up the introduction.

--- 2018-04-04 paper.tex
@@
 \author{
-\IEEEauthorblockN{The Author, The Other Author}\\
-Institute\\
-University, Country\\
-Email: \{author\}@university.domain 
+\IEEEauthorblockN{Author One, Author Two, Author Three}\\
+Famous Institute for Software Engineering and Computer Science\\
+University of Great City, Some Country\\
+Email: \{first.last\}@university.to 
 }

### 2018-04-07 performance.tex
cleaning up the introduction.

@@
 \Fref{fig:ep} presents such a perspective from our case study which 
-was taken in the last quarter of 2017. Several observatrions are 
+was taken in the last quarter of 2017. Several observations are 
 straightforward:
@@
 
-  \subsection{Automated Outlier Detection and Monitoring}
-  \va{Shouldn't this be promoted to a section?
+  \section{Automated Outlier Detection and Monitoring}
+  \label{sec:outliers}
+  % \va{Shouldn't this be promoted to a section?
 It is neither utilization- nor performance- related.

### 2018-04-07 introductory.tex
cleaned up the beginning bits a bit more.

@@
 \begin{itemize}
-  \item Providing more details about the case study (in Section \ref{sec:case}) and extending the data collection period with half year
-  \item A better categorization of the dashboard perspectives presented in Sections \ref{sec:util}, \ref{sec:grouping}, \ref{sec:evolution} including several new perspectives
+  \item Providing more details about the case study in Section \ref{sec:case} and extending the data collection period with half year
+  \item Novel perspectives and a better categorization of the perspectives presented in Sections \ref{sec:utilandperf} -- \ref{sec:evolution}
   \item A detailed discussion on automated outlier detection and monitoring (Section \ref{sec:outliers})
   \item An advanced discussion integration continuous integration environments in order to triangulate the live collected data with information from testing the endpoints with a constant load and introduce a mechanism for performance prediction (Section \ref{sec:testing})
-  \item A discussion on the overhead incurred by the presented tool \ml{(... if we make it... )}
+  \item A more extended discussion about lessons learned while developing the tool including measurements on the overhead incurred by the presented tool \ml{(... if we make it... )}
 \end{itemize}

### 2018-04-07 paper.tex
cleaned up the beginning bits a bit more.

--- 2018-04-04 paper.tex
@@
 \author{
-\IEEEauthorblockN{The Author, The Other Author}\\
-Institute\\
-University, Country\\
-Email: \{author\}@university.domain 
+\IEEEauthorblockN{Author One, Segundo Autor, Troisi\`eme Auteur}\\
+Famous Institute for Software Engineering and Computer Science\\
+University of Great City, Some Country\\
+Email: \{first.last\}@university.to 
 }

### 2018-04-07 outliers.tex
all iz well till section IV

--- 2018-04-07 performance.tex
@@
 %!TEX root=paper.tex
-
-\subsection{Performance}
-\label{sec:perf}
-
-% The \tool also collects information regarding endpoint performance. 
-The \perspective{API Performance} visualization perspective presents
-an overview of the response times for all the tracked endpoints 
-by using a box-and-whiskers plot.
-
-\Fref{fig:ep} presents such a perspective from our case study which 
-was taken in the last quarter of 2017. Several observations are 
-straightforward:
-
-\begin{itemize}
-
-  \item The slowest endpoint and the one with the highest variability is \epFeedItemsColor: it retrieves a list of recommended articles for a given user.
-Since a user can be subscribed to anything from one to three dozen article sources, and since the computation of the difficulty was done in real time, the variability in time among users is likely to be very large.
-
-  \item In the top three slowest endpoints, but the second most used endpoint in the system (cf.
-\Fref{fig:aeu}) is \epTranslationsColor.
-The endpoint is critical for the reader applications that rely on the Zeeguu API: each time a user taps on a word on their screen they receive a translation from this endpoint.
-
-\end{itemize}
-
-\begin{figure}
- \centering
- \includegraphics[width=0.85\linewidth]{endpoint_performance_}
- \caption{The response time (in ms) per monitored endpoint view allows for identifying performance variability and balancing issues}
- \label{fig:ep}
-\end{figure}
-
-Such a view is important for allocating optimization efforts.
-In our case study, the API developers started by improving the performance of \epTranslationsColor and then later optimized the performance of \epFeedItemsColor.
-However, in order to understand the problems of the two endpoints, a big component represents understanding the outliers: because they are many.
-\va{the last sentence does not parse}
 
@@
 
-  When an API is called from within a highly interactive application (as it is the case with the case study in this paper) 
-  of particular interest to the API developers are performance {\em outliers}.
+  When an API endpoint is called from within a highly interactive application (as it is the case with the case with \epTranslationsColor in the previous section) of particular interest to the API developers are performance {\em outliers}.
 Indeed, a translation request that takes three times more than expected can seriously decrease the perceived quality of the application.

### 2018-04-07 performance.tex
all iz well till section IV

--- 2018-04-05 performance.tex
@@
 
-  \item In the top three slowest endpoints, but the second most used endpoint in the system (cf.
-\Fref{fig:aeu}) is \epTranslationsColor.
-The endpoint is critical being embedded in the interactive loop in the reader applications that rely on the Zeeguu API: a user would tap on a word and receive a translation from this endpoint.
+  \item \epTranslationsColor and \epBookmarksToStudyColor, two of the most used endpoints (as seen in \Fref{fig:aeu}) are also some of the slowest and with most outliers.
+The first one is very critical for the usability of the Reader applications since it called when a user taps on a word to receive a translation.
 
@@
 
-\begin{figure}[!ht]
+\begin{figure}
  \centering
- \includegraphics[width=0.85\linewidth]{endpoint_performance_}
+ \includegraphics[width=0.95\linewidth]{endpoint_performance_}
  \caption{The response time (in ms) per monitored endpoint view allows for identifying performance variability and balancing issues}
@@
 
-Such a view is important for allocating optimization efforts.
-In our case study, the API developers started by improving the performance of \epTranslationsColor and then later optimized the performance of \epFeedItemsColor.
-However, in order to understand the problems of the two endpoints, a big component represents understanding the outliers: because they are many.
+The \perspective{API Performance} perspective is important for allocating optimization efforts.
+In our case study, after seeing this perspecgive, the API developers decided to rearchitect the system in order to improve the performance of \epTranslations and \epFeedItems.
 
-  \subsection{Automated Outlier Detection and Monitoring}
-
-  When an API is called from within a highly interactive application (as it is the case with the case study in this paper) 
-  of particular interest to the API developers are performance {\em outliers}.
-
-  Indeed, a translation request that takes three times more than expected can seriously decrease the perceived quality of the application.
-Thus, identifying, collecting all appropriate data, and diagnosing the root causes of such outliers is especially critical in improving the quality of an application.
-
-  % In the context of the RESTful services support by Flask, this means request serving times that deviate from the average to an unexpected degree.
-
-  For this purpose the \tool tracks for every monitored endpoint a {\em running average} response time value\footnote{\ins{For performance reasons, we assume that the response times for the endpoints are normally distributed.
-Otherwise, more general density distribution information must be collected in real time.}}.
-When it detects that a given request is an outlier with respect to this past average running value, it triggers the {\em outlier data collection routine} which stores \ins{extra information} about the current execution environment.
-A configurable threshold with a default value of $2.5$ times the running average response time is used for this purpose.
-
-  For every detected outlier request, the \tool collects information about the current Python stack trace, CPU load, memory consumption, request parameters, etc.
-in order to allow the maintainer to investigate the causes of these exceptionally slow response times.
-In this way it is possible to get \ins{detailed insight into the operation of the application in the extreme cases without unnecessarily burdening it with logging this information for every request}.
-
-  \begin{figure}[h!]
-    \centering
-    \includegraphics[width=0.9\linewidth]{outlier-annotated}
-    \caption{Automatically collected outlier information}
-    \label{fig:figure1}
-  \end{figure}
-
-The bottom panel shows the stack trace. 
-In this particular case, it is revealing for the developer to learn that at the time of the stack trace snapshot, the code was in the google\_translator.py: indeed, the system uses as back-end multiple translators, and it has been observed that many of the outliers happen while the system is waiting for the Google translator.
-
-This information has to be corroborated with the observations that neither the memory nor the processor are overloaded at the moment.
-Thus this functionality in microsoft\_translator is really slow in itself, and this is not a result of the machine being overloaded for example.
+% However, in order to understand the problems of the two endpoints, a big component represents understanding the outliers: because they are many. \va{the last sentence does not parse}

### 2018-04-07 grouping.tex
cleaning up the introduction.
all is nice up to V

@@
 
-\section{Per-User Performance Monitoring}
-\label{sec:user}
+\section{Identity-Based Grouping of Requests}
+\label{sec:grouping}
 
@@
 in GMail some users have a few dozen emails while other have tens of thousands: this difference in user loads will eventually induce a difference in the response times for different users} the maintainer must understand the variation of performance with the users (and implicitly the user load).
-s
 \end{itemize}
@@
 
-After configuring the \tool as shown earlier, the \perspective{Per-User Performance} perspective becomes available to  present the different response times for different users.
+After configuring the \tool as shown earlier, the \perspective{Per-User Endpoint Performance} perspective becomes available to  present the different response times for different users.
 Figure \ref{fig:tpu} presents a subset of the corresponding view in the \tool.

### 2018-04-07 outliers.tex
all is nice up to V

@@
 
-  When an API endpoint is called from within a highly interactive application (as it is the case with the case with \epTranslationsColor in the previous section) of particular interest to the API developers are performance {\em outliers}.
-Indeed, a translation request that takes three times more than expected can seriously decrease the perceived quality of the application.
+  When an API endpoint is called from within a highly interactive application 
+    (as it is the case with \epTranslationsColor in the previous section) 
+  of particular interest to the API developers are {\em performance outliers}.   
+  Indeed, a translation request that takes three times more than expected can   decrease the perceived quality of the application. 
   Thus, identifying, collecting all appropriate data, and diagnosing the root   causes of such outliers is especially critical in improving the quality of an application.
 
-  % In the context of the RESTful services support by Flask, this means request serving times that deviate from the average to an unexpected degree.
-
-  For this purpose the \tool tracks for every monitored endpoint a {\em running average} response time value\footnote{\ins{For performance reasons, we assume that the response times for the endpoints are normally distributed.
-Otherwise, more general density distribution information must be collected in real time.}}.
-When it detects that a given request is an outlier with respect to this past average running value, it triggers the {\em outlier data collection routine} which stores \ins{extra information} about the current execution environment.
-A configurable threshold with a default value of $2.5$ times the running average response time is used for this purpose.
-
-  For every detected outlier request, the \tool collects information about the current Python stack trace, CPU load, memory consumption, request parameters, etc.
-in order to allow the maintainer to investigate the causes of these exceptionally slow response times.
-In this way it is possible to get \ins{detailed insight into the operation of the application in the extreme cases without unnecessarily burdening it with logging this information for every request}.
+  For this purpose the \tool tracks for every monitored endpoint a {\em running average} response time value.\footnote{
+    For performance reasons, we assume that the response times for the endpoints are normally distributed. 
+    Otherwise, more general density distribution information must be collected in real time.}
+  When it detects that a given request is an outlier with respect to this past average running value\footnote{A configurable threshold with a default value of $2.5$ times the running average response time is used for this purpose}, it triggers the {\em outlier data collection routine} which stores extra information about the current execution environment.
+This extra data includes: the current Python stack trace, CPU load, memory consumption, request parameters, etc.
 
@@
     \centering
-    \includegraphics[width=0.9\linewidth]{outlier-annotated}
-    \caption{Automatically collected outlier information}
+    \includegraphics[width=\linewidth]{outlier-annotated}
+    \caption{Automatically collected outlier information for the \epTranslationsColor endpoint}
     \label{fig:stack}
@@
 
-The bottom panel in \Fref{fig:stack} shows the stack trace. 
-In this particular case, it is revealing for the developer to learn that at the time of the stack trace snapshot, the code was in the \code{google\_translator}: indeed, the system uses as back-end multiple translators, and it has been observed that many of the outliers happen while the system is waiting for the Google translator.
-%
-This information has to be corroborated with the observations that neither the memory nor the processor are overloaded at the moment.
-Thus this functionality in \code{microsoft\_translator} is really slow in itself, and this is not a result of the machine being overloaded for example.
-\va{last sentence does not parse}
+\Fref{fig:stack} shows an example of a \perspective{Performance Outlier Detail}  perspective on one of the outliers in the \epTranslationsColor endpoints.
+In this particular case, the analyst could corroborate several bits of informations:
+
+\begin{itemize}
+
+  \item Neither the memory (1) nor the processor (2) were overloaded at the moment of the request. Thus the slow response is not due to the machine being overloaded.
+
+  \item The stack trace snapshot shows that code was in the \code{google\_translator.py}.
+Since the system uses as back-end multiple translators it is revealing to see that in this particular case it was waiting for the Google Translator.
+By manually investigating multiple outliers, the developer discovered that in a large percentage of the cases the Google Translator was indeed the reason for the slow response time.
+
+\end{itemize}
+
+  \niceseparator
+
+  In this way it is possible to get detailed insight into the operation of the application in the extreme cases without unnecessarily burdening it with logging this information for every request.
+In the discussion section we will come back and discuss some of the limitations of this approach.\ml{memento!}

### 2018-04-07 beautiful-stuff.tex
all iz well up to section VIII

--- 2018-04-04 testing.tex
@@
-%!TEX root=paper.tex
-
-\newpage
-  \section{Performance Regression Monitoring}
-
-  Suppose the developer observes a performance degradation of a given API endpoint in the latest version of the system.
-How do they know whether the degradation is due to the performance of the code, the server load, or the workload mix of the user?
-
-  \vspace{0.1cm}
-
-  One way of triangulating such an observation is by monitoring the evolution of the performance of the endpoints when tested with a constant load across versions.
-Thus, if the performance decreases in a new version, and the  when an endpoint is tested with a constant load the problem must be the endpoint implementation.
-
-  \subsection*{Integrating with CI Servers}
-
-  To monitor endpoint performance with a constant load \tool takes advantage of two best practices in software evolution: (1) and API must have unit tests for its endpoints, and (2) these unit tests will be run within a CI server.
-By tracking the timing of these unit tests one obtains a perspective on endpoint performance evolution with a constant load (as long as the unit tests do not change).
-
-  \tool can be configured to work together with Continuous Integration (CI) frameworks like Travis\footnote{\url{https://travis-ci.org/}, one of the most popular CI solutions at the moment} that deal with automated integration testing.
-To do this one needs to simply add one extra line in the script that runs their continuous integration testing.
-In the case of Travis CI, a developer has to add the following line in `.travis.yml':
-
-  \begin{lstlisting}[style=custompython]  
-# LOC #4: to be added to the .travis.yml file
-python -m flask_monitoringdashboard.collect
-   --test_folder=./tests_zeeguu_api --times=5 
-   --url=https://zeeguu.unibe.ch/api/dashboard
-
-  \end{lstlisting}
-
-  Each time a new build is created through Travis, the \tool automatically detects all available unit tests defined by the application developer (in (-{}-test\_folder)) and iterates through each one of them a number of times (-{}-times) while monitoring the response times for each test.
-The resulting measurements are uploaded to a specific endpoint that is created by the tool at the API url (-{}-url).
-The data uploaded to this endpoint is   persisted in a separate part of the tool database so as not to contaminate the ``live'' monitoring data.
-
-      \begin{figure}[h!]
-        \centering
-        \includegraphics[width=0.45\columnwidth]{travis_builds}
-        \caption{Response times in 5 consequent Travis builds}
-        \label{fig:builds}
-      \end{figure}
-
-  \Fref{fig:builds} is a screenshot from the dashboard showing the measured response times for 5 consequent builds, with 180 iterations of the unit tests executed in total across all endpoints.
-The outliers on the right of the figure are due to initial requests \ins{which must wait for a boot-up phase of the API}.
-
-  \subsection*{Preemptive Monitoring}
-  The concept of {\em preemptive monitoring} of the application performance by means of instrumenting integration unit testing as the synthetic load is similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}, i.e.~testing service-based applications by using dedicated test input in parallel to its normal use and operation.
-The difference is that we take advantage of the capability of the CI framework to create an emulated ``live'' environment for integration testing purposes, and use unit testing as the dedicated test input in order to measure performance.
-
-  While this integration testing environment is different from the production one, and the load used is purely synthetic, it can serve as an early performance indicator for the developer.
-Figure \ref{fig:response_times_preemptive} shows the actual endpoint response times across five versions of the system (above) and the corresponding testing times for the synthetic load in the same five versions (below).
-The green line joining the medians hints to a correlation between the two sets of measurements.
-
-      \begin{figure}[h!]
-        \centering
-        \includegraphics[width=0.75\columnwidth]{response_per_version_trunced_trend}
-
-        \advance\leftskip-0.2cm
-        \includegraphics[width=0.5\columnwidth]{travis_builds_no_outliers_trend}
-        \caption{The reported response time (in ms) per deployed version in the observation period: 
-        in the actual deployment (above) and using integration with Travis (below)}
-
-        \label{fig:response_times_preemptive}
-      \end{figure}
-
-  Computing Pearson correlation ($r(3)=.93, p=.02$) between the median values of the two datasets (cf. Table \ref{tab:correlations}) shows that there is a correlation between the two.
-
-    \begin{table}[h]
-
-      \centering
-      \begin{tabular}{lll}
-        \toprule
-        Iteration & \bfseries Live (median) & \bfseries Travis (median)\\
-        \midrule
-        1 & 1349.41 & 764.87\\ 
-        2 & 1466.13 & 992.87\\
-        3 & 1256.65 & 760.87\\
-        4 & 1266.42 & 813.89\\
-        5 & 1080.68 & 303.4\\
-        \bottomrule
-
-      \end{tabular}
-      \caption{Median response times in Figure \ref{fig:response_times_preemptive}}
-      \label{tab:correlations}
-    \end{table}
-
-%   \begin{figure*}[hb!]
-%   \centering
-%   \subfloat[The reported response time (in ms) per deployed version in the observation period]{\includegraphics[width=.6\columnwidth]{response_per_version_trunced_trend}}
-%   \quad
-%   \subfloat[The measured response time (in ms) using integration with Travis]{\includegraphics[width=.56\columnwidth]{travis_builds_no_outliers_trend}}
-%   \caption{Comparison of the response times per endpoint: actual production system versus preemptive monitoring data}
-%   \label{fig:preemptive}
-% \end{figure*}
+% Let's keep here stuff that's nicely said, but doesn't yet have a place in the paper.
 

### 2018-04-07 discussion.tex
all is nice up to V
all iz well up to section VIII

@@
 
--   \ins{We can do this minimal configuration for any technology.
-We showed here how to do it for Python and Flask because they are some of the most popular web development technologies at the moment.}
+- We can do this minimal configuration for any technology. We showed here how to do it for Python and Flask because they are some of the most popular web development technologies at the moment.
 
@@
 - branches. the system does not care. it links to whatever commit is the current one. it could be that visualizations could be developed to comapre branches, but this optiona. future work.
+
+- outlier detection and monitoring is important.
+manual exploration is critical.
+however, one needs a performant way of analyzing multiple logs.
+in the example we saw we had to use the search facility to investigate our hypothesis that Google Translate was slower than Microsoft Translate.
+Moreover, these trends might change too...
 

### 2018-04-07 evolution.tex
cleaning up the introduction.
all iz well up to section VIII

@@
   \section{Version-Aware Monitoring}
-  \label{sec:version}
+    \label{sec:evolution}
 
@@
 number of sources to which the user is registered) is the reason for the variation in response times.
-\ins{One can see that for user with id=1, performance degrades over consecutive versions.
-Given that for other users the performance does not degrade in the same way, it is probable that the problem might lay elsewhere: the server was overloaded or the endpoint is ``algorithmically slower''.} \va{could it be that the user kept on adding sources to their profile?}
+One can see that for user with id=1, performance degrades over consecutive versions.
+Given that for other users the performance does not degrade in the same way, it is probable that the problem might lay elsewhere: the server was overloaded or the endpoint is ``algorithmically slower''.

### 2018-04-07 overhead.tex
working on the case study section
all iz well up to section VIII

@@
 \label{sec:overhead}
+
+\ml{To Do. Or to Drop.}
 
@@
 
-\ins{PERFORMANCE DISCUSSION: this introspection, which looks up endpoints, happens only once at the startup of the service, so it is going to affect the API startup performance, but not the actual endpoint performance.
-In our case, the overhead here was quite small: TO MEASURE}.
+Observation -- not easy to benchmark this... it's flask on top of python communicating with
+
+- this introspection, which looks up endpoints, happens only once at the startup of the service, so it is going to affect the API startup performance, but not the actual endpoint performance.
+In our case, the overhead here was quite small: TO MEASURE

### 2018-04-07 testing.tex
cleaning up the introduction.
all iz well up to section VIII

@@
   \section{Performance Triangulation With Regression Monitoring}
-  \label{sec:regression}
+  \label{sec:testing}
 
@@
         \centering
-        \includegraphics[width=0.65\columnwidth]{travis_builds}
+        \includegraphics[width=0.65\columnwidth]{travis_builds-bw}
         \caption{Response times in 5 consequent Travis builds}
@@
   \Fref{fig:builds} is a screenshot from the dashboard showing the measured response times for 5 consequent builds, with 180 iterations of the unit tests executed in total across all endpoints.
-The outliers on the right of the figure are due to initial requests \ins{which must wait for a boot-up phase of the API}.
-\va{Remember to make greyscale for consistency}
+The outliers on the right of the figure are due to initial requests which must wait for a boot-up phase of the API.
 
@@
     \end{table}
-
-%   \begin{figure*}[hb!]
-%   \centering
-%   \subfloat[The reported response time (in ms) per deployed version in the observation period]{\includegraphics[width=.6\columnwidth]{response_per_version_trunced_trend}}
-%   \quad
-%   \subfloat[The measured response time (in ms) using integration with Travis]{\includegraphics[width=.56\columnwidth]{travis_builds_no_outliers_trend}}
-%   \caption{Comparison of the response times per endpoint: actual production system versus preemptive monitoring data}
-%   \label{fig:preemptive}
-% \end{figure*}
-
-  \ml{This following paragraph is beautifully written, but it's a bit redundant with what it's already said.
-However maybe after some more refactorings we can put it to use.
-Worst case, we move it to the conclusion}
-  \va{I concur}
-  \ins{Besides functioning as an early warning system for performance, tracking the evolving performance of API tests serves a second purpose.
-To function as an anchor when the developer analyzes performance degradations.
-For example, when a developer sees that a given endpoint has become less performant after in a newly deployed version, they are able to investigate whether the performance degradation is visible also with the synthetic load.
-This would correspond to a performance degradation which is due to ``algorithmic performance degradation''.
-If on the other hand, the performance of the tests does not change between versions, the developer might conclude that the performance degradation could be due to the workload on the machines, or maybe to the workload mix of the users}

### 2018-04-08 introductory.tex
all iz well till section IV
Edited from top to Section II (incl.)

@@
 
-  Python is one of the fastest growing programming languages of the moment.
+  Python is one of the fastest growing programming languages at the moment.
 Flask, a Python-based web framework is the technology used by tens of thousands of web applications.
-These applications, from interactive websites to service APIs, lack a technology specific service monitoring solution.
+These applications, from interactive websites to service APIs, lack a technology-specific service monitoring solution.
 
-  In this paper, we present \tool, a tool that provide insight into the utilization and performance of evolving Flask-based web services.
-We present the ease with which the tool can be integrated in an already existing web application, discuss some of the visualization perspectives that the library provides and point to some future challenges for similar libraries.
+  In this paper, we present \tool, a tool that provides insight into the utilization and performance of evolving Flask-based web services.
+We present the ease with which the tool can be integrated in an already existing web application, discuss some of the visualization perspectives that the library provides, and point to some future challenges for similar libraries.
 
@@
 %\ml{I'm not completely happy with this paragraph}
-{\em There is no getting around it: you are building a distributed system} argues Cavage  in a recent article written for the {\em Communications of the ACM} \cite{cavage2013there}.
-Indeed, even the simplest second-year student project is a web application implemented as two-tier architecture with a Javascript/HTML5 front-end a service backend, usually a REST API.
+{\em There is no getting around it: you are building a distributed system} argues Mark Cavage  in a recent article written for the {\em Communications of the ACM} \cite{cavage2013there}.
+Indeed, even the simplest second-year student project these days is a web application implemented as two-tier architecture with a Javascript/HTML5 front-end a service backend, usually a REST API.
 
@@
 
-Despite their popularity, to the best of our knowledge, there is no simple solution for monitoring the evolving performance of Flask web applications.
+Despite their popularity however, and to the best of our knowledge, there is no simple solution for monitoring the evolving performance of Flask web applications.
 Thus, every one of the developers of these projects faces one of the following options when confronted with the need of gathering insight into the runtime behavior of their implemented services:
@@
 research, startups) the first and the second options are often not available due to time and financial constraints.
-Even when using 3rd-party analytics solutions, a critical insight into the evolution of the exposed services of the web application, is missing because such solutions have no notion of versioning and no integration with the development life cycle.~\cite{papazoglou2011managing}
+Even when using 3rd-party analytics solutions, a critical insight into the evolution of the exposed services of the web application is missing because such solutions have no notion of versioning and no integration with the development life cycle~\cite{papazoglou2011managing}.
 
-To avoid projects ending up in the third situation, that of living without analytics, in this paper we present \tool~ --- a low-effort service monitoring library for Flask-based Python web services that is easy to integrate and enables the {\em agile assessment} of service evolution.
-\cite{Nier12b}
+To avoid projects ending up in the third situation, that of living without analytics, in this paper we present \tool~ --- a low-effort service monitoring library for Flask-based Python web services that is easy to integrate and enables the {\em agile assessment} of service evolution~\cite{Nier12b}.
 
-As a case study, on which we will illustrate our solution, we are going to use an open source API which, for several years, was in the third of the above-presented situations.
+As a case study, on which we will illustrate our solution, we are going to use an open source API which, for several years, was in the third of the above-presented situations, i.e.
+it was being operated without any monitoring solution.
 
@@
 \begin{itemize}
-  \item Providing more details about the case study in Section \ref{sec:case} and extending the data collection period with half year
-  \item Novel perspectives and a better categorization of the perspectives presented in Sections \ref{sec:utilandperf} -- \ref{sec:evolution}
-  \item A detailed discussion on automated outlier detection and monitoring (Section \ref{sec:outliers})
-  \item An advanced discussion integration continuous integration environments in order to triangulate the live collected data with information from testing the endpoints with a constant load and introduce a mechanism for performance prediction (Section \ref{sec:testing})
+  \item Providing more details about the case study in Section \ref{sec:case} and extending the data collection period by half a year.
+  \item New perspectives and a better categorization of the perspectives are presented in Sections \ref{sec:utilandperf} -- \ref{sec:evolution}.
+  \item A detailed discussion on automated outlier detection and monitoring is provided (Section \ref{sec:outliers}).
+  \item An advanced discussion on the integration with continuous integration environments in order to triangulate the live collected data with information from testing the endpoints with a constant load and introduce a mechanism for performance prediction is added (Section \ref{sec:testing})
   \item A more extended discussion about lessons learned while developing the tool including measurements on the overhead incurred by the presented tool \ml{(... if we make it... )}
@@
   \Fref{fig:zeeguuarch} presents a high-level view of the case study ecosystem.
-The core API is implemented with Flask and Python provides and correspondingly three types of functionality: contextual translations, personalized exercise suggestions, and article recommendations.
-In total, the API provides a bit under 50 endpoints, out of which around a dozen are very frequently used.
+The core API is implemented with Flask and Python and provides three types of functionality, respectively: contextual translations, personalized exercise suggestions, and article recommendations.
+In total, the API provides a bit under 50 endpoints in total, out of which around a dozen are very frequently used.
 The development of the core API itself is a research project.
-One of the authors of this paper is part of the developer team that maintains and evolves the \zee API.
+  % One of the authors of this paper is part of the developer team that maintains and evolves the \zee API.
 
@@
 Several users are using it on their own, outside of any formal educational context.
-The learners use \zee to study a variety of of foreign languages including German, Dutch, French.
+The learners use \zee to study a variety of foreign languages including German, Dutch, and French.
 They have also diverse native languages, including: Chinese, Dutch, and English.
@@
 
-\subsection*{Study characteristics}
+\subsection{Study characteristics}
 
@@
   All the figures and measurements in this paper are captured from the actual deployment of \tool in the context of the \zee API.
-The figures are interactive offering basic data exploration capabilities: filter, zoom, and details on demand\cite{Shne99a}.
+The figures are interactively offering basic data exploration capabilities: filter, zoom, and details on demand~\cite{Shne99a}.
 The \tool deployment for the case study can be accessed publicly\footnote{\url{https://zeeguu.unibe.ch/api/dashboard}; username: {\em guest}, password: {\em icsme}}.
@@
 
- Flask is built on top of WSGI -- The Web Server Gateway Interface -- a calling convention for web servers to forward requests to web applications or frameworks written in Python.
+ Flask is built on top of WSGI\footnote{\url{http://wsgi.org}} --- The Web Server Gateway Interface --- a calling convention for web servers to forward requests to web applications or frameworks written in Python.
 A Python web application based on WSGI has to have one central callable object that implements the actual application.
@@
 \end{lstlisting}
+
+In the following we discuss the capabilities of the \tool in more depth.

### 2018-04-08 discussion.tex
.
started cleaning up discussion a bit

@@
 \section{Discussion}
+
+\todo{use this subsection to discuss limitations (from the previous evaluation section) and add the need to be able to handle the horizontal scaling of the application; remove sectioning and summarize briefly limitations that have already been discussed in the VISSOFT paper/move material to other sections; consider renaming as limitations or something similar}
+
+- The main goal of the \tool design was to allow analytics to be collected and insight to be gained by making the smallest possible changes to a running API.
+%To allow the collection of evolutionary information
+
+- We can do this minimal configuration for any technology.
+We showed here how to do it for Python and Flask because they are some of the most popular web development technologies at the moment.
+however, nothing prevents generalization.
+sure, for different technologies, several components have to be adapted:
+-- endpoint discovery -- this will be done differently for different technologies
+-- deploying the dashboard as a new endpoint -- this will be also done differently for different technologies 
+--   A further possible extension point here is supporting other version control systems (e.g. Mercurial). However, this is a straightforward extension.
 
@@
 
-- We can do this minimal configuration for any technology. We showed here how to do it for Python and Flask because they are some of the most popular web development technologies at the moment.
+- Lanza has done nice stuff for source code evolution patterns. we need more work on automatic detection of service / endpoint evolution patterns. as exemplified by the utilization evolution.
 
@@
 
-\todo{use this subsection to discuss limitations (from the previous evaluation section) and add the need to be able to handle the horizontal scaling of the application; remove sectioning and summarize briefly limitations that have already been discussed in the VISSOFT paper/move material to other sections; consider renaming as limitations or something similar}
+-   Supporting situations where the API is deployed across multiple containers for example.
 
-  \subsection{Distributed System Monitoring}
-  \todo{To think about}
-  Supporting situations where the API is deployed across multiple containers for example.
+-   DETECTING ENDPOINT RENAMES.  The history would normally be lost in the case of an API rename. (also known as the {\em provenance issue} in source code evolution analysis)
+    In theory we could infer that an API endpoint was renamed if it is not to be found anymore, and another one with very similar timing characteristics would be found.
+Such a detection can never be completely sure, but it would help with meeting the needs of the API maintainer...
 
-  \subsubsection{Automatically Monitoring System Evolution}
-
-  The main goal of the \tool design was to allow analytics to be collected and insight to be gained by making the smallest possible changes to a running API.
-%To allow the collection of evolutionary information
-%
+  \subsection*{Limitations of the Current Evolution Approach}
   This technique assumes that the web application code which is the target of the monitoring is deployed using \git in the following way:
@@
 %
-  A further possible extension point here is supporting other version control systems (e.g. Mercurial). However, this is a straightforward extension.
 
-  \subsubsection{NEW: API Renames}
-  \ml{just an idea}
-
-    The history would normally be lost in the case of an API rename. (also known as the {\em provenance issue} in source code evolution analysis)
-    In theory we could infer that an API was renamed if it is not to be found anymore, and another one with very similar timing characteristics would be found.
-Such a detection can never be completely sure, but it would help with meeting the needs of the API maintainer...
-
-  \subsubsection{User-Awareness }
-
-    For the situations in which the user information is not available, the \tool tracks by default information about different IPs and in some cases this might be a sufficiently good approximation of the user diversity and identity.
-    %
+  \subsection*{Limitation of Visualizations }
 
@@
 
-  \subsubsection{Other Possible Groupings}
+  \subsection*{Allowing Multiple Groupings}
 
-    There are other groupings of service utilization and performance that could be important to the maintainer, that we did not explore in this paper.
-For example, if the service is using OAuth, then together with every request, in the header of the request there is information about the application which is sending a request.
-Grouping the information by application that sends the request could be important in such a context.
+  One of the current limitations is that one can only specify one type of grouping. 
+  However, there could be cases where multiple groupins would be desirable.
 
-    In general, providing a mechanism that would allow very easy specification of groupings (either as code annotations, as normal code, or as configuration options) is an open problem that \tool and any other similar library will have to face.
+  Grouping the information by client application that sends the request could be important as a complementary grouping even in the current \zee case study.

### 2018-04-08 grouping.tex
Rel work + sections 3 to 5 incl. edited

@@
 those in a different organization) the maintainer must understand the performance for the different groups
-	\item When the load mix o the users can vary dramatically and the system response time is a function of the individual user load\footnote{E.g.
+	\item When the load mix of the users can vary dramatically and the system response time is a function of the individual user load\footnote{E.g.
 in GMail some users have a few dozen emails while other have tens of thousands: this difference in user loads will eventually induce a difference in the response times for different users} the maintainer must understand the variation of performance with the users (and implicitly the user load).
@@
 
-\niceseparator
+%\niceseparator
 
@@
 In Zeeguu, it was decided to further understand the per-user performance of \epFeedItems
---  an endpoint that retrieves a list of recommended articles for a given user.
+~---  an endpoint that retrieves a list of recommended articles for a given user.
 It is the endpoint with the slowest response time and highest variability (see \Fref{fig:ep}).
@@
 of each article at every request.
-Only after seeing the two perspectives presented in Figures \ref{fig:ep} and \ref{fig:tpu}, the maintainer refactored the architecture of the system to move the difficulty computation out of the interactive loop.
+After seeing the two perspectives presented in Figures \ref{fig:ep} and \ref{fig:tpu}, the \zee maintainer refactored the architecture of the system to move this difficult computation out of the interactive loop.

### 2018-04-08 related-work.tex
done or the day
added baresi13 to rel work
Rel work + sections 3 to 5 incl. edited

@@
 
-\todo{Refer the reader to~\cite{ghezzi2007run} and ~\cite{metzger2010analytical} for a more extensive discussion}
+Service monitoring data are increasingly being used and more across a variety of systems in which adaptability, flexibility, and environmental awareness are essential~\cite{pernici2016monitoring}.
+Monitoring can be performed on different types of system components and for several purposes.
+Service-oriented literature on the subject discussed the multitude of aspects related to monitoring for services; the interested reader is referred to ~\cite{ghezzi2007run},~\cite{metzger2010analytical}, or more recently~\cite{pernici2016monitoring} for an extensive presentation of the subject.
+
+In this context, our work falls within the server-side run-time monitoring of services ~\cite{ghezzi2007run}.
+While we don't implement the more advanced features of related monitoring solutions like QoS policies driving the monitoring, it presents nevertheless an easy to use approach towards improving the performance of web applications.
+The emphasis in this work is given in identifying the appropriate way to visualize the monitoring data.
 
@@
 As opposed to them, in our work we present an analytics platform which focuses on monitoring a single Python web service from its own point of view.
-
-From the perspective of service monitoring, our work falls within the server-side run-time monitoring of services ~\cite{ghezzi2007run}.
-While we don't implement the more advanced features of related monitoring solutions like QoS policies driving the monitoring, it presents nevertheless an easy to use approach support improving the performance of web applications.
 

### 2018-04-08 evolution.tex
Sec. 6 & 7 edited

@@
 
-  Version control in a dashboard can be supported in two ways: 
+  Version control in the dashboard can be supported in two ways: 
   (1) the developer explicitly states the current version, or 
@@
 # LOC #3: provide the dashboard with 
-# information on where to find the 
+# information on where to find 
 # git information 
@@
 
-    \item Endpoints being deprecateds. In the case study \epFeedItems stops being used somewhere after the half of the presented timeline.
+    \item Endpoints being deprecated. In the case study, \epFeedItems stops being used somewhere after the half of the presented timeline.
 
@@
 
-    \item Endpoint renames. Although not visible in the figure, the candidates for rename would be visible as one would discontinued at the same time when the other would appear.
+    \item Endpoint renames. Although not visible in the figure, the candidates for rename would be made easily visible as one would be discontinued at the same time when the other would appear.
 
@@
 
-    \Fref{fig:tee} is a zoomed-in version of such a view for \epTranslationsColor with versions increasing from top to bottom
+    \Fref{fig:tee} is a zoomed-in version of such a view for \epTranslationsColor with versions increasing from top to bottom.
 
@@
 number of sources to which the user is registered) is the reason for the variation in response times.
-One can see that for user with id=1, performance degrades over consecutive versions.
-Given that for other users the performance does not degrade in the same way, it is probable that the problem might lay elsewhere: the server was overloaded or the endpoint is ``algorithmically slower''.
+One can see that for user with \code{id=1}, performance degrades over consecutive versions.
+Given that for other users the performance does not degrade in the same way, it is probable that the problem might lay elsewhere: the server was overloaded or the endpoint is inherently ``algorithmically slower''.

### 2018-04-08 testing.tex
Sec. 6 & 7 edited

@@
   One way of triangulating such an observation is by monitoring the evolution of the performance of the endpoints when tested with a constant load across versions.
-Thus, if the performance decreases in a new version, and the  when an endpoint is tested with a constant load the problem must be the endpoint implementation.
+Thus, if the performance decreases in a new version, and the endpoint in question is tested with a constant load the problem must be the endpoint implementation.
 
@@
 
-  To monitor endpoint performance with a constant load \tool takes advantage of two best practices in software evolution: (1) and API must have unit tests for its endpoints, and (2) these unit tests will be run within a CI server.
-By tracking the timing of these unit tests one obtains a perspective on endpoint performance evolution with a constant load (as long as the unit tests do not change).
+  To monitor endpoint performance with a constant load \tool takes advantage of two best practices in software evolution: (1) an API must have unit tests for its endpoints, and (2) these unit tests will be run within a Continuous Integration (CI) server.
+By tracking the timing of these unit tests one can obtain a perspective on endpoint performance evolution with a constant load (as long as the unit tests do not change).
 
-  \tool can be configured to work together with Continuous Integration (CI) frameworks like Travis\footnote{\url{https://travis-ci.org/}, one of the most popular CI solutions at the moment} that deal with automated integration testing.
+  \tool can be configured to work together with CI frameworks like Travis\footnote{\url{https://travis-ci.org/}, one of the most popular CI solutions at the moment} that deal with automated integration testing.
 To do this one needs to simply add one extra line in the script that runs their continuous integration testing.
-In the case of Travis CI, a developer has to add the following line in `.travis.yml':
+In the case of Travis CI, a developer has to add the following line in the `.travis.yml' configuration file:
 
@@
 The resulting measurements are uploaded to a specific endpoint that is created by the tool at the API url (\code{-{}-url}).
-The data uploaded to this endpoint is   persisted in a separate part of the tool database so as not to contaminate the ``live'' monitoring data.
+The data uploaded to this endpoint are persisted in a separate part of the \tool database so as not to contaminate the ``live'' monitoring data.
 
@@
 
-  Computing Pearson correlation ($r(3)=.93, p=.02$) between the median values of the two datasets (cf. Table \ref{tab:correlations}) shows that there is indeed a correlation between the two.
+  Computing Pearson correlation ($r(3)=.93, p=.02$) between the median values of the two datasets (cf.
+Table \ref{tab:correlations}) shows that there is indeed a correlation between the two.
+Further investigation on the characteristics of this observation is currently ongoing and it will be elaborated further in future work.
 

### 2018-04-08 related-work.tex
rel

--- 2018-04-03 related-work.tex
@@
 
+In their work Baresi and Guinea have proposed the Multi-layer Collection and Constraint Language (mlCCL) which allows them to define how to collect, aggregate, and analyze runtime data in a multi-layered system.
+They also present ECoWare, a framework for event correlation and aggregation that supports mlCCL, and provides a dashboard for on-line and off-line drill-down analyses of collected data \cite{Bare13-monitoring}.
+
 % \va{Mircea: Consider removing the rest for space...}

### 2018-04-08 loc1.tex
cleaned up the beginning bits a bit more.

@@
   \section{Service Utilization and Performance}
+  \label{sec:utilandperf}
 
@@
 This would however pollute the code, and prevent deploying two versions which would monitor different endpoints.
-Also, would pose problems with endpoint evolution in the future.
+In addition, it would pose problems with endpoint evolution in the future.
 
@@
 
-    \item {\bf Utilization} -- refers to how the third parties use the API, which parts are most used, which are little used, etc.
+    \item {\bf Utilization} -- refers to how the third parties use the API, which parts are most used, which are less used, etc.
 This is known to be critical information for upstream developers in general\cite{Haen14a}.

### 2018-04-08 outliers.tex

@@
   of particular interest to the API developers are {\em performance outliers}.   
-  Indeed, a translation request that takes three times more than expected can   decrease the perceived quality of the application. 
-  Thus, identifying, collecting all appropriate data, and diagnosing the root   causes of such outliers is especially critical in improving the quality of an application.
+  Indeed, a translation request that takes three times more than expected can   decrease the perceived quality of the application significantly. 
+  Thus, identifying, collecting all appropriate data, and diagnosing the root   causes of such outliers is especially critical in improving application quality.
 

### 2018-04-08 overhead.tex
benchmark

@@
 \label{sec:overhead}
-
-\ml{To Do. Or to Drop.}
 
@@
 	\item downloads the latest version of the Zeeguu API and installs it in a Docker container
-	\item enables the dasboard and calls several endpoints tracking the response times
-	\item disables the dashboard and calls the same endpoitns tracking the response times
-	\item reports the response times
+	\item calls multiple endpoints for 500 times tracking the response times. the endpoints are called in three configurations
+	\begin{itemize}
+		\item with the dashboard disabled
+		\item with the dashboard enabled but with the outlier detection deactivated
+		\item with the dashboard enabled and the outlier tracking enabled always
 	\end{itemize}
+\end{itemize}
+
+\Fref{fig:bench} shows the results of calling every one of the three endpoints for 500 times.
+
+\begin{figure}[h!]
+	\centering
+	\includegraphics[width=\linewidth]{benchmark}
+	\caption{The distribution of response times when calling the three endpoints for 500 times in three conditions: no dashboard, dashboard but no outliers, dashboard and always outlier routine}
+	\label{fig:bench}
+\end{figure}
 

### 2018-04-09 overhead-table.tex
a nice table with the benchmark numbers is in

%!TEX root=paper.tex

\newcommand{\yes}{YES}

\begin{table}[tb]

	\centering

	\begin{tabular}{rllrr}
		\toprule
		\bfseries Endpoint & \bfseries Dashboard & \bfseries Outliers & \bfseries Mean (ms) & \bfseries SD (ms) \\

		\midrule

	         available\_languages  &    &	    &   6.3 &  2.3 \\ 
	         available\_languages  &  \yes &    &  19.9 &  5.6 \\ 
	         available\_languages  &  \yes &  \yes &  29.4 &  5.4 \\ \\

	                user\_article  &    &	    &  13.9 &  2.6 \\
	                user\_article  &  \yes &    &  26.9 &  3.2 \\
	                user\_article  &  \yes &  \yes &  44.8 & 10.0 \\ \\

	    create\_default\_ex...     &    &	    &  76.0 & 22.4 \\ 
	    create\_default\_ex...     &  \yes &    & 313.6 & 43.2 \\
	    create\_default\_ex...     &  \yes &  \yes & 385.3 & 16.9 \\

		\bottomrule

	\end{tabular}
	\caption{Descriptive statistics of the response times reported by running the benchmark}
	\label{tab:benchmark}

\end{table}

### 2018-04-09 overhead.tex
added benchmark measurements
nicer benchmarking image
a nice table with the benchmark numbers is in

@@
 
+\newpage
 \section{Overhead of the \tool}
@@
 
-To know the performance overhead of the \tool, we have implemented an automated benchmark system.
+To measure the performance overhead of the \tool, we have implemented an automated benchmarking system.
 It is open source and available online and can be tested by the reader.
-The benchmark does the following steps:
+The benchmark downloads the latest version of the \zee API and installs it in a Docker container.
+Then it calls several selected endpoints for 500 times each, tracking the response times.
+The endpoints are called in three different configurations:
 
-\begin{itemize}
-	\item downloads the latest version of the Zeeguu API and installs it in a Docker container
-	\item calls multiple endpoints for 500 times tracking the response times. the endpoints are called in three configurations
-	\begin{itemize}
-		\item with the dashboard disabled
-		\item with the dashboard enabled but with the outlier detection deactivated
-		\item with the dashboard enabled and the outlier tracking enabled always
-	\end{itemize}
-\end{itemize}
+	\begin{enumerate}
+		\item With the dashboard disabled
+		\item With the dashboard enabled but with the outlier detection deactivated
+		\item With the dashboard enabled and the outlier treshold set to zero. This helps us measure the time usually spent in an outlier.
+	\end{enumerate}
 
-\Fref{fig:bench} shows the results of calling every one of the three endpoints for 500 times.
+\Fref{fig:bench} shows using violin plots the distribution of times resulting from running the benchmark for three different endpoints on a quad-core machine, with Intel Core i5-4590 processor, @ 3.30GHz, 8G of RAM and 240GB SSD disk drive.
 
@@
 	\centering
-	\includegraphics[width=\linewidth]{benchmark}
+	\includegraphics[width=\linewidth]{benchmark2.pdf}
 	\caption{The distribution of response times when calling the three endpoints for 500 times in three conditions: no dashboard, dashboard but no outliers, dashboard and always outlier routine}
@@
 
-Based on running the benchmarking script we can report
+\input{overhead-table}
 

### 2018-04-09 overhead-table.tex
nicer and nicer ...

@@
 
-\newcommand{\yes}{YES}
+\newcommand{\yes}{\checkmark}
 
@@
 		\toprule
-		\bfseries Endpoint & \bfseries Dashboard & \bfseries Outliers & \bfseries Mean (ms) & \bfseries SD (ms) \\
+		\bfseries Endpoint & \bfseries Dash. & \bfseries All Out. & \bfseries Mean (ms) & \bfseries SD (ms) \\
 
@@
 
-	    create\_default\_ex...     &    &	    &  76.0 & 22.4 \\ 
+	    create\_default\_ex...     &    &	    &  276.0 & 22.4 \\ 
 	    create\_default\_ex...     &  \yes &    & 313.6 & 43.2 \\
@@
 	\end{tabular}
-	\caption{Descriptive statistics of the response times reported by running the benchmark}
+	\caption{Response times for three endpoints run 500 times in three different conditions each. Dash. = with dashboard. All Out. = every call being considered an outlier}
 	\label{tab:benchmark}

### 2018-04-09 overhead.tex
nicer and nicer ...

@@
 	\begin{enumerate}
-		\item With the dashboard disabled
+		\item With no dashboard installed
 		\item With the dashboard enabled but with the outlier detection deactivated
-		\item With the dashboard enabled and the outlier treshold set to zero. This helps us measure the time usually spent in an outlier.
+		\item With the dashboard enabled and the outlier treshold set to zero, thus effectively treating every request {\em like it were an outlier}\footnote{This forces all the requests to be treated as outliers, and thus provides insight into this situation, which otherwise would be hard to generate}.
 	\end{enumerate}
 
-\Fref{fig:bench} shows using violin plots the distribution of times resulting from running the benchmark for three different endpoints on a quad-core machine, with Intel Core i5-4590 processor, @ 3.30GHz, 8G of RAM and 240GB SSD disk drive.
+\Fref{fig:bench} and Table \ref{tab:benchmark} present with descriptive statistics and respectively violin plots the distribution of times resulting from running the benchmark for three different endpoints on a quad-core machine, with Intel Core i5-4590 processor, @ 3.30GHz, 8G of RAM and 240GB SSD disk drive.
 
@@
 	\includegraphics[width=\linewidth]{benchmark2.pdf}
-	\caption{The distribution of response times when calling the three endpoints for 500 times in three conditions: no dashboard, dashboard but no outliers, dashboard and always outlier routine}
+	\caption{The distribution of response times when calling the three endpoints for 500 times in three conditions: no dashboard, dashboard but no outliers, dashboard and every request treated as an outlier}
 	\label{fig:bench}

### 2018-04-09 overhead.tex
some more meaningful text about the overhead

@@
 
-\Fref{fig:bench} and Table \ref{tab:benchmark} present with descriptive statistics and respectively violin plots the distribution of times resulting from running the benchmark for three different endpoints on a quad-core machine, with Intel Core i5-4590 processor, @ 3.30GHz, 8G of RAM and 240GB SSD disk drive.
+\Fref{fig:bench} and Table \ref{tab:benchmark} present with violin plots and respectively descriptive statistics the distribution of times resulting from running the benchmark for three different endpoints on a quad-core machine, with Intel Core i5-4590 processor, @ 3.30GHz, 8G of RAM and 240GB SSD disk drive.
 
@@
 
-Observation -- not easy to benchmark this... it's flask on top of python communicating with
+	The data shows that the dashboard (without outliers) introduces for the faster endpoints an overhead of ~14ms and for the slower endpoint an overhead of ~40ms. 
+	In the case of an outlier, the overhead is doubled, so the design decision of only collecting the outlier information in the exceptional situations is correct.
 
-- this introspection, which looks up endpoints, happens only once at the startup of the service, so it is going to affect the API startup performance, but not the actual endpoint performance.
-In our case, the overhead here was quite small: TO MEASURE
+	Depending on the application this might be acceptable or this might be too much. In the case of the \zee case study, the maintainers find this acceptable.

### 2018-04-09 overhead.tex
first draft of the overhead is done.

@@
 
+	The three endpoints that are tested fall in two categories of complexity: 
+	\begin{itemize}
+		\item The two fast ones, are very simple. 
+		The first returns a list of language codes which are defined in the code, so it does not touch the DB. The second does a simple read from the database. 
+		\item The slower endpoint, does several complex writes to the database, so this is why it is much slower.
+However, in our case study, about half of the measured endpoints were at least as slow as this one, so its response time is representative.
+	\end{itemize}
+
 	The data shows that the dashboard (without outliers) introduces for the faster endpoints an overhead of ~14ms and for the slower endpoint an overhead of ~40ms. 
-	In the case of an outlier, the overhead is doubled, so the design decision of only collecting the outlier information in the exceptional situations is correct.
-
-	Depending on the application this might be acceptable or this might be too much. In the case of the \zee case study, the maintainers find this acceptable.
+	In the case of an outlier, the overhead is doubled, so the design decision of only collecting the extra information only for outliers is valid.
+	Depending on the application these numbers might be acceptable or this might be too much. In the case of the \zee case study, the maintainers find this acceptable.

### 2018-04-09 discussion.tex
...

@@
 
+- Performance.
+We have provided a mechanism for measuring performance, and showed a glimpse into the overhead imposed by the first version of the Dashboard.
+For our case study the overhead is acceptable, but for performance critical applications, it might be too large.
+One of the downsides of the current implementation is the fact that the dashboard runs in the same process as the main application and that Python only supports green threads.
+The performance could be greatly improved if instead the Dashboard would be run in a different process, and the information would be posted to it by using a messaging service.
+However, that would complicate the installation and configuration, and for this kind of architecture, there are also other alternatives.
+
+- We designed the dashboard to make it easy to adopt. However, we still have to run a study on the challenges and opportunities of the adoption of such a dashboard.
+
 -   Supporting situations where the API is deployed across multiple containers for example.

### 2018-04-09 discussion.tex
cleaning up the discussion section

@@
 
-\todo{use this subsection to discuss limitations (from the previous evaluation section) and add the need to be able to handle the horizontal scaling of the application; remove sectioning and summarize briefly limitations that have already been discussed in the VISSOFT paper/move material to other sections; consider renaming as limitations or something similar}
+  % \todo{use this subsection to discuss limitations (from the previous evaluation section) and add the need to be able to handle the horizontal scaling of the application; remove sectioning and summarize briefly limitations that have already been discussed in the VISSOFT paper/move material to other sections; consider renaming as limitations or something similar}
 
-- The main goal of the \tool design was to allow analytics to be collected and insight to be gained by making the smallest possible changes to a running API.
-%To allow the collection of evolutionary information
+  The main goal of the \tool design was to allow analytics to be collected and insight to be gained by making the smallest possible changes to a running API.
 
-- We can do this minimal configuration for any technology.
-We showed here how to do it for Python and Flask because they are some of the most popular web development technologies at the moment.
-however, nothing prevents generalization.
-sure, for different technologies, several components have to be adapted:
--- endpoint discovery -- this will be done differently for different technologies
--- deploying the dashboard as a new endpoint -- this will be also done differently for different technologies 
---   A further possible extension point here is supporting other version control systems (e.g. Mercurial). However, this is a straightforward extension.
-
-- could even start with this: NOTHING PREVENTS GENERALIZATION
-In this article we take as an example the Flask case study
-and of the \tool as a dashboard. 
-But there is nothing that prevents the generalization of 
-our approach to other languages and technologies.
-
-- Lanza has done nice stuff for source code evolution patterns. we need more work on automatic detection of service / endpoint evolution patterns. as exemplified by the utilization evolution.
-
-- A Special Type of Outlier: Exceptions. TODO: probably something for the journal extension ...
-  statistics about which endpoints fail most often might be useful too.
-  same information as the outlier maybe?
-
-- time as measured in days and commits are two sides of the same coin.
-
-- currently the system supports only one grouping. this need not be so, and there are many scenarios where one could envision multiple groupins in parallel.
-
-- threats to the validity of the constant-load performance testing. performance regressions? 
-- the endpoint where the unit-testing results can be uploaded is prone to being tampered with by Bob who would want to upload junk information to mess up with the dashboard.
-to solve this, a UUID can also be added that is generated by the server, and thus prevents anybody else but the rightful owner of the dashboard
-
-- user management. although not explained in the paper, the dashboard also comes with a user management system. and two classes of users. admins and guests.
-
-- Although predicting the performance of the application would require advanced statistical work, it can be succesfully used as an advanced warning system for the overall performance trend for given endpoints.
-
-- detecting minor versions which don't need to be tracked, since they didn't touch the performance of the system. E.g. a modification of the README, etc.
-
-- discuss later one situation in which opt-out might be desirable.
-
-- tracking other types of information: exceptions, etc. especially since logs usually get lost... nobody has time to look in them...
-
--  In this paper we present the integration with APIs written in Python and Flask hoping that this will not prevent the reader from seeing the more general idea; all the tools we show here for Flask can be applied to other API technologies (e.g.
+  In this paper we present the integration with APIs written in Python and Flask hoping that this will not prevent the reader from seeing the more general idea; all the tools we show here for Flask can be applied to other API technologies (e.g.
 Django) by simply providing a few back-end adapters in the right places.
 
-- We need more advanced outlier analysis tools... Instead of manually searching in the page to build stats about Google vs. Microsoft
+  We can use the same approach for other technologies.
+We showed here how to do it for Python and Flask because they are some of the most popular web development technologies at the moment.
+However, we hope that this will not prevent the reader from seeing the more general idea.
+All the interactive perspectives we show here for Flask can be applied to other service technologies (e.g.
+Django) by simply providing a few back-end adapters in the right places (e.g.
+endpoint discovery, the \code{/dashboard} endpoint deployment).
 
-- branches. the system does not care. it links to whatever commit is the current one. it could be that visualizations could be developed to comapre branches, but this optiona. future work.
+  However, the presented approach also comes with several limitations.
 
-- outlier detection and monitoring is important.
-manual exploration is critical.
-however, one needs a performant way of analyzing multiple logs.
-in the example we saw we had to use the search facility to investigate our hypothesis that Google Translate was slower than Microsoft Translate.
-Moreover, these trends might change too...
+  \subsection*{The Dashboard Overhead}
 
-- Performance.
     We have provided a mechanism for measuring performance, and showed a glimpse into the overhead imposed by the first version of the Dashboard.
 For our case study the overhead is acceptable, but for performance critical applications, it might be too large.
+
     One of the downsides of the current implementation is the fact that the dashboard runs in the same process as the main application and that Python only supports green threads.
@@
 
-- We designed the dashboard to make it easy to adopt. However, we still have to run a study on the challenges and opportunities of the adoption of such a dashboard.
+  \subsection*{Limitation of Visualizations }
 
--   Supporting situations where the API is deployed across multiple containers for example.
+    The visualizations for the user experiene perspectives as presented in Section \ref{sec:user} have been tested with several hundred users (of which about \activeUserCount were active during the course of the study), but the scalability of the visualizations must be further investigated for web services with tens of thousands of users.
 
--   DETECTING ENDPOINT RENAMES.  The history would normally be lost in the case of an API rename. (also known as the {\em provenance issue} in source code evolution analysis)
-    In theory we could infer that an API endpoint was renamed if it is not to be found anymore, and another one with very similar timing characteristics would be found.
-Such a detection can never be completely sure, but it would help with meeting the needs of the API maintainer...
+  \subsection*{Flexibility: Allowing Multiple Groupings}
+
+    To be agile, a tool has to be easily adaptable to different usage scenarios.
+
+    One of the current limitations is that one can only specify one type of grouping. 
+    However, there could be cases where multiple groupins would be desirable.
+
+    Grouping the information by client application that sends the request could be important as a complementary grouping even in the current \zee case study.
 
   \subsection*{Limitations of the Current Evolution Approach}
+
     This technique assumes that the web application code which is the target of the monitoring is deployed using \git in the following way:
@@
     A mechanism to control which versions are important for monitoring purposes is therefore required to be added.
+
+  \subsection*{Monitoring Failures}
+
+    A Special Type of Outlier: Exceptions. TODO: probably something for the journal extension ...statistics about which endpoints fail most often might be useful too.
+    same information as the outlier maybe?
+
+    - tracking other types of information: exceptions, etc. especially since logs usually get lost... nobody has time to look in them...
+
+  \subsection*{Multiple Containers}
+
+  -   Supporting situations where the API is deployed across multiple containers for example.
+
+  \subsection*{Performance Regressions}
+
+    - Although predicting the performance of the application would require advanced statistical work, it can be succesfully used as an advanced warning system for the overall performance trend for given endpoints.
+
+    - threats to the validity of the constant-load performance testing. performance regressions?
+
+    - time as measured in days and commits are two sides of the same coin.
+
+  \subsection*{Advanced Outlier Analysis}
+
+    - We need more advanced outlier analysis tools... Instead of manually searching in the page to build stats about Google vs. Microsoft
+
+    - outlier detection and monitoring is important.
+manual exploration is critical.
+however, one needs a performant way of analyzing multiple logs.
+in the example we saw we had to use the search facility to investigate our hypothesis that Google Translate was slower than Microsoft Translate.
+Moreover, these trends might change too...
+
+  \subsection*{Endpoint Provenance }
+
+    -   DETECTING ENDPOINT RENAMES.  The history would normally be lost in the case of an API rename. (also known as the {\em provenance issue} in source code evolution analysis)
+    In theory we could infer that an API endpoint was renamed if it is not to be found anymore, and another one with very similar timing characteristics would be found.
+Such a detection can never be completely sure, but it would help with meeting the needs of the API maintainer...
+
+  \subsection*{Ease of Use}
+
+    - We designed the dashboard to make it easy to adopt. However, we still have to run a study on the challenges and opportunities of the adoption of such a dashboard.
+
+% - the endpoint where the unit-testing results can be uploaded is prone to being tampered with by Bob who would want to upload junk information to mess up with the dashboard.
+to solve this, a UUID can also be added that is generated by the server, and thus prevents anybody else but the rightful owner of the dashboard
+
+% - user management. although not explained in the paper, the dashboard also comes with a user management system. and two classes of users. admins and guests.
+
+% - detecting minor versions which don't need to be tracked, since they didn't touch the performance of the system. E.g. a modification of the README, etc.
+
+% - discuss later one situation in which opt-out might be desirable.
+
+% - branches. the system does not care. it links to whatever commit is the current one. it could be that visualizations could be developed to comapre branches, but this optiona. future work.
+
 %
-
-  \subsection*{Limitation of Visualizations }
-
-    The visualizations for the user experiene perspectives as presented in Section \ref{sec:user} have been tested with several hundred users (of which about \activeUserCount were active during the course of the study), but the scalability of the visualizations must be further investigated for web services with tens of thousands of users.
-
-  \subsection*{Allowing Multiple Groupings}
-
-  One of the current limitations is that one can only specify one type of grouping. 
-  However, there could be cases where multiple groupins would be desirable.
-
-  Grouping the information by client application that sends the request could be important as a complementary grouping even in the current \zee case study.

### 2018-04-09 discussion.tex
progressing...

@@
 
-  \subsection*{Limitation of Visualizations }
+  \subsection*{Limitations of the Measurements}
+    One limitation of the benchmarking is that it does not use a more realistic mix of workload, but rather, measures individual endpoints.
+It might be that multiple concurrent and diverse endpoints would have different performance impact.
+In fact, based on the data in the dashboard one could actually construct a statistically relevant workload mix.
 
-    The visualizations for the user experiene perspectives as presented in Section \ref{sec:user} have been tested with several hundred users (of which about \activeUserCount were active during the course of the study), but the scalability of the visualizations must be further investigated for web services with tens of thousands of users.
+  \subsection*{Lack of Proof of Adoptability}
+
+    We designed the dashboard to make it easy to adopt and we think this is to a certain degree apparent.
+However, we have only presented a single use case where the API developers have used the dashboard.
+It would be worthwile to organize a dedicated study to observe the challenges and opportunities of the adoption of such a dashboard.
+It will also be important to see whether the perspectives presented here are relevant for other developers, and also which other perspectives are missing.
+
+  % \subsection*{Limitation of Visualizations }
+
+  %   The visualizations for the user experiene perspectives as presented in Section \ref{sec:user} have been tested with several hundred users (of which about \activeUserCount were active during the course of the study), but the scalability of the visualizations must be further investigated for web services with tens of thousands of users.
 
@@
 
-  \subsection*{Limitations of the Current Evolution Approach}
+  \subsection*{Limitations of the Current Approach to Evolution Monitoring}
 
@@
 
-  \subsection*{Ease of Use}
-
-    - We designed the dashboard to make it easy to adopt. However, we still have to run a study on the challenges and opportunities of the adoption of such a dashboard.
-
 % - the endpoint where the unit-testing results can be uploaded is prone to being tampered with by Bob who would want to upload junk information to mess up with the dashboard.

### 2018-04-09 discussion.tex
...

@@
 
-  However, the presented approach also comes with several limitations.
+  However, the presented approach also comes with several limitations. Some of these limitations might also represent challenges for the future builders of similar tools.
 
@@
 
-  \subsection*{Lack of Proof of Adoptability}
+  \subsection*{Lack of Empirical Proof of Ease of Adoptability}
 
@@
 
-  \subsection*{Flexibility: Allowing Multiple Groupings}
+  \subsection*{Lack of Flexibility in Allowing Multiple Groupings}
 
     To be agile, a tool has to be easily adaptable to different usage scenarios.
-
-    One of the current limitations is that one can only specify one type of grouping. 
-    However, there could be cases where multiple groupins would be desirable.
-
-    Grouping the information by client application that sends the request could be important as a complementary grouping even in the current \zee case study.
+However, currently one of the current limitations of \tool is that that one can only specify one type of grouping.
+This is insufficient for the cases where multiple groupings are desirable.
+In the \zee case study this need emerged in the context in which it would be good to grouping the requests also by the client application that sends the request.
 

### 2018-04-09 overhead.tex
...
Overhead edits (TODO left in-text)

--- 2018-04-09 overhead.tex
@@
 It is open source and available online and can be tested by the reader.
-The benchmark downloads the latest version of the \zee API and installs it in a Docker container.
+\va{@Mircea: Don't forget to add link} The benchmark downloads the latest version of the \zee API and installs it in a Docker container.
 Then it calls several selected endpoints for 500 times each, tracking the response times.
@@
 	\begin{enumerate}
-		\item With no dashboard installed
-		\item With the dashboard enabled but with the outlier detection deactivated
-		\item With the dashboard enabled and the outlier treshold set to zero, thus effectively treating every request {\em like it were an outlier}\footnote{This forces all the requests to be treated as outliers, and thus provides insight into this situation, which otherwise would be hard to generate}.
+		\item With no dashboard installed.
+		\item With the dashboard enabled but with the outlier detection deactivated.
+		\item With the dashboard enabled and the outlier threshold set to zero, thus effectively treating every request {\em like it were an outlier}.\footnote{This forces all the requests to be treated as outliers, and thus provides insight into this situation, which otherwise would be hard to generate}.
 	\end{enumerate}
 
-\Fref{fig:bench} and Table \ref{tab:benchmark} present with descriptive statistics and respectively violin plots the distribution of times resulting from running the benchmark for three different endpoints on a quad-core machine, with Intel Core i5-4590 processor, @ 3.30GHz, 8G of RAM and 240GB SSD disk drive.
+The last configuration is meant to evaluate the effect of the outlier detection as discussed in Section~\ref{sec:outliers} to the overall performance.
+%
+\Fref{fig:bench} and Table \ref{tab:benchmark} use violin plots and descriptive statistics, respectively, to present the distribution of response times resulting from running the benchmark for three different API endpoints on a quad-core machine, with Intel Core i5-4590 processor @3.30GHz, 8G of RAM and 240GB SSD disk drive.
 
@@
 
-Observation -- not easy to benchmark this... it's flask on top of python communicating with
+	The three endpoints that are tested fall in two categories of complexity: 
+	\begin{itemize}
+		\item The two fast ones are very simple. 
+		The first returns a list of language codes which are defined in the code, so it does not touch the \zee database. The second does a simple read from the database. 
+		\item The slower endpoint does several complex writes to the database, so this is why it is much slower.
+However, in our case study, about half of the measured endpoints were at least as slow as this one, so its response time is representative.
+	\end{itemize}
 
-- this introspection, which looks up endpoints, happens only once at the startup of the service, so it is going to affect the API startup performance, but not the actual endpoint performance.
-In our case, the overhead here was quite small: TO MEASURE
+	The data shows that the dashboard (without outliers) introduces for the faster endpoints an overhead of~14ms and for the slower endpoint an overhead of~40ms, on average. 
+	In the case of an outlier, the overhead is doubled, so the design decision of only collecting the extra information only for outliers seems to be validated. 
+	Depending on the monitored application these numbers might be acceptable or not.
+In the case of the \zee case study, the maintainers find this overhead acceptable since they do not affect the overall user experience in any significant manner.

### 2018-04-09 discussion.tex
...

@@
 
-  \subsection*{Monitoring Failures}
+  \subsection*{Endpoint Provenance }
 
-    A Special Type of Outlier: Exceptions. TODO: probably something for the journal extension ...statistics about which endpoints fail most often might be useful too.
-    same information as the outlier maybe?
+    The \tool can not detect endpoint renames, and thus the history is normally lost in the case of an API rename.
+The problem can be approached in a similar way to the {\em provenance problem} in source code evolution analysis)\cite{Davi11a} by infering that a newly introduced endpoint with similar patterns of request traffic might be the same as one that is not used anymore.
+Since this would still be a statistic approach, the integration in the tool would have to ask the user for a confirmation.
+    However, this would work better in a system where endpoint tracking is activated by default for all the endpoints.
+Otherwise, it would require the user to remember to always enable the tracking of the new endpoint.
 
-    - tracking other types of information: exceptions, etc. especially since logs usually get lost... nobody has time to look in them...
+  % \subsection*{Monitoring Failures}
+
+  %   A Special Type of Outlier: Exceptions. TODO: probably something for the journal extension ...statistics about which endpoints fail most often might be useful too.
+  %   same information as the outlier maybe?
+
+  %   - tracking other types of information: exceptions, etc. especially since logs usually get lost... nobody has time to look in them...
 
@@
 
-  \subsection*{Endpoint Provenance }
-
-    -   DETECTING ENDPOINT RENAMES.  The history would normally be lost in the case of an API rename. (also known as the {\em provenance issue} in source code evolution analysis)
-    In theory we could infer that an API endpoint was renamed if it is not to be found anymore, and another one with very similar timing characteristics would be found.
-Such a detection can never be completely sure, but it would help with meeting the needs of the API maintainer...
-
 % - the endpoint where the unit-testing results can be uploaded is prone to being tampered with by Bob who would want to upload junk information to mess up with the dashboard.

### 2018-04-09 discussion.tex
...

@@
   -   Supporting situations where the API is deployed across multiple containers for example.
+  \ml{@V: Do you want to add this? Or can we skip it?}
 
@@
 
-  \subsection*{Advanced Outlier Analysis}
+    \ml{@V: do you want to give a first stab to this?}
 
-    - We need more advanced outlier analysis tools... Instead of manually searching in the page to build stats about Google vs. Microsoft
+  % \subsection*{Advanced Outlier Analysis}
 
-    - outlier detection and monitoring is important.
+  %   - We need more advanced outlier analysis tools... Instead of manually searching in the page to build stats about Google vs. Microsoft
+
+  %   - outlier detection and monitoring is important.
 manual exploration is critical.

### 2018-04-09 architecture.tex
anonymization

@@
 
-  As discussed in the introductory section, in previous work~\cite{vogel2017low} we introduced \tool, a drop-in Python library that allows developers to monitor their Flask-based Python web applications with minimal effort.
+  As discussed in the introductory section, in previous work \withheld we introduced \tool, a drop-in Python library that allows developers to monitor their Flask-based Python web applications with minimal effort.
 %
-  The \tool is implemented for Python 3.6 and is available on the Python Package Index repository\footnote{\url{https://pypi.python.org/pypi/flask-monitoring-dashboard/1.8}} from where it can be installed by running \install from the command line.
+  The \tool is implemented for Python 3.6 and is available on the Python Package Index repository
+  % \footnote{\url{https://pypi.python.org/pypi/flask-monitoring-dashboard/1.8}} 
+  \footnote{\url{https://pypi.python.org/pypi/[\witheld]}} 
+  from where it can be installed by running \install from the command line. 
 %  

### 2018-04-09 introductory.tex
anonymization

@@
 
-  \zee\footnote{\url{https://github.com/zeeguu-ecosystem/}} is a platform and an ecosystem of applications for accelerating vocabulary acquisition in a foreign language \cite{Lungu16}. 
+  \zee
+  % \footnote{\url{https://github.com/zeeguu-ecosystem/}} 
+  is a platform and an ecosystem of applications for accelerating vocabulary acquisition in a foreign language \cite{lungu18-chi}. 
 %

### 2018-04-09 loc1.tex
nicer benchmarking image
anonymization

@@
 
-  \newpage
+  % \newpage
   \section{Service Utilization and Performance}
@@
 
-  A dashboard can be attached to an already existing Flask application with a single line of code~\cite{vogel2017low} as the following snippet shows\footnote{We do not count the import statement that enables that line.}:
+  A dashboard can be attached to an already existing Flask application with a single line of code \citationwithheld as the following snippet shows\footnote{We do not count the import statement that enables that line.}:
 
@@
 The information is also saved in a database such that it is persistent across application restarts.
-
-  \va{\Fref{fig:sep} needs to be updated --- it still has 2017 dates}
 

### 2018-04-09 discussion.tex
xxx
Discussion section done

@@
 
-  In this paper we present the integration with APIs written in Python and Flask hoping that this will not prevent the reader from seeing the more general idea; all the tools we show here for Flask can be applied to other API technologies (e.g.
+  %In this paper we present the integration with APIs written in Python and Flask hoping that this will not prevent the reader from seeing the more general idea; all the tools we show here for Flask can be applied to other API technologies (e.g.
 Django) by simply providing a few back-end adapters in the right places.
 
-  We can use the same approach for other technologies.
-We showed here how to do it for Python and Flask because they are some of the most popular web development technologies at the moment.
-However, we hope that this will not prevent the reader from seeing the more general idea.
-All the interactive perspectives we show here for Flask can be applied to other service technologies (e.g.
+  The presented approach was centered around an API written in Python and Flask, but, in principle, 
+  we can use the same approach for other technologies. We showed here how to do it for Python and Flask because they are some of the most popular web development technologies at the moment. However,
+  % we hope that this will not prevent the reader from seeing the more general idea. All 
+  all the interactive perspectives we developed can be equally applied to other service technologies (e.g.
 Django) by simply providing a few back-end adapters in the right places (e.g.
-endpoint discovery, the \code{/dashboard} endpoint deployment).
+endpoint discovery, the \code{/dashboard} endpoint deployment, etc.)
+  As such, our approach is more generic than the technologies that it relies upon for its implementation.
 
-  However, the presented approach also comes with several limitations. Some of these limitations might also represent challenges for the future builders of similar tools.
+  At the same time, the presented approach also comes with several limitations. Some of these limitations might also represent challenges for the future builders of similar tools.
 
@@
     One of the downsides of the current implementation is the fact that the dashboard runs in the same process as the main application and that Python only supports green threads.
-The performance could be greatly improved if instead the Dashboard would be run in a different process, and the information would be posted to it by using a messaging service.
-However, that would complicate the installation and configuration, and for this kind of architecture, there are also other alternatives.
+The performance could be greatly improved if instead the dashboard would run in a different process, %and the information would be posted to it by using a messaging service.
+    and a message queue was used to allow the decoupling of the monitoring itself from the persistence of measurements to the dashboard DB. 
+    However, that would complicate the installation and configuration of the \tool.
+    %, and for this kind of architecture, there are also other alternatives.
 
   \subsection*{Limitations of the Measurements}
-    One limitation of the benchmarking is that it does not use a more realistic mix of workload, but rather, measures individual endpoints.
-It might be that multiple concurrent and diverse endpoints would have different performance impact.
+    One limitation of the benchmarking we used in Section~\ref{sec:overhead} is that it does not use a more realistic mix of workload, but rather, measures individual endpoints.
+It might be that multiple concurrent and diverse endpoints would result into different performance impact.
 In fact, based on the data in the dashboard one could actually construct a statistically relevant workload mix.
+At this point, this constitutes future work.
 
@@
 However, we have only presented a single use case where the API developers have used the dashboard.
-It would be worthwile to organize a dedicated study to observe the challenges and opportunities of the adoption of such a dashboard.
+It would be worthwhile to organize a dedicated study to observe the challenges and opportunities of the adoption of such a dashboard across different applications.
 It will also be important to see whether the perspectives presented here are relevant for other developers, and also which other perspectives are missing.
@@
     To be agile, a tool has to be easily adaptable to different usage scenarios.
-However, currently one of the current limitations of \tool is that that one can only specify one type of grouping.
-This is insufficient for the cases where multiple groupings are desirable.
+However, currently one of the current limitations of \tool is that that one can specify only one type of grouping.
+This is insufficient for cases where multiple groupings are desirable.
 In the \zee case study this need emerged in the context in which it would be good to grouping the requests also by the client application that sends the request.
+This feature is currently under development.
 
-  \subsection*{Limitations of the Current Approach to Evolution Monitoring}
+  \subsection*{Limitations to Evolution Monitoring}
 
@@
     \begin{enumerate}
-      \item The deployment engineer pulls the latest version of the code from the integration server; this will result in a new commit being pointed at by the HEAD pointer. %than previously
+      \item The deployment engineer pulls the latest version of the code from the integration server; this will result in a new commit being pointed at by the \code{HEAD} pointer. %than previously
       \item The deployment engineer restarts the new version of the service.
-At this point, the \tool detects that a new HEAD is present in the local code base and consequently starts associating all the new data points with this new commit\footnote{The \tool detects the current version of the analyzed system the first time it is attached to the application object, and thus, assumes that the Flask application is restarted when a new version is deployed.
+At this point, the \tool detects that a new \code{HEAD} is present in the local code base and consequently starts associating all the new data points with this new commit\footnote{The \tool detects the current version of the analyzed system the first time it is attached to the application object, and thus, assumes that the Flask application is restarted when a new version is deployed.
 This is in tune with the current version of Flask, but if the web server will support dynamic updates in the future, this might have to be taken into account}.
@@
 
-      The advantage of this approach is the need for minimal configuration effort, as discussed in the presentation of the tool.
+      The advantage of this approach is the need for minimal configuration effort, as discussed in the presentation of the \tool.
 The disadvantage is that it will consider on equal ground the smallest of commits, even one that modifies a comment, and the shortest lived of commits, e.g.~a commit which was active only for a half an hour before a new version with a bug fix was deployed, with major and minor releases of the software.
 %as a distinct way of grouping the data points.
-    A mechanism to control which versions are important for monitoring purposes is therefore required to be added.
+    A mechanism to control which versions are important for monitoring purposes is therefore required to be added to the \tool.
 
@@
     The \tool can not detect endpoint renames, and thus the history is normally lost in the case of an API rename.
-The problem can be approached in a similar way to the {\em provenance problem} in source code evolution analysis)\cite{Davi11a} by infering that a newly introduced endpoint with similar patterns of request traffic might be the same as one that is not used anymore.
-Since this would still be a statistic approach, the integration in the tool would have to ask the user for a confirmation.
+The problem can be approached in a similar way to the {\em provenance problem} in source code evolution analysis)\cite{Davi11a} by inferring that a newly introduced endpoint with similar patterns of request traffic might be the same as one that is not used anymore.
+Since this would still be a statistical approach, the integration in the tool would have to ask the user for confirmation.
     However, this would work better in a system where endpoint tracking is activated by default for all the endpoints.
@@
 
-  \subsection*{Multiple Containers}
+  \subsection*{Performance Prediction}
 
-  -   Supporting situations where the API is deployed across multiple containers for example.
-  \ml{@V: Do you want to add this? Or can we skip it?}
+	In Section~\ref{sec:testing} we provided evidence towards our assumption that integration testing can be used for performance triangulation across service versions.
+The idea is to use a synthetic load based on unit testing to drive endpoints before they are deployed in production and observe whether performance improves or deteriorates on a version by version basis.
+The next logical step is to use this information for performance prediction purposes.
+A full blown performance prediction feature would require advanced statistical work, but it can be used as an advanced warning system for the overall performance trend for a given set of endpoints.
+As discussed in Section~\ref{sec:testing} however further evaluation of this feature is currently ongoing.
 
-  \subsection*{Performance Regressions}
+  \subsection*{Distributed Deployment}
 
-    - Although predicting the performance of the application would require advanced statistical work, it can be succesfully used as an advanced warning system for the overall performance trend for given endpoints.
+	Last but not least, the presented approach and case study assume that the monitored Flask application is deployed as a single node, i.e.
+without support for horizontal scaling by means of replication~\cite{vaquero2011dynamically}  --- or at least that the application owner is interested in monitoring each application deployment in isolation.
+This is a clear limitation of this work which we are already working towards addressing by allowing multiple application instances to be registered with the same dashboard, essentially providing a ``meta-dashboard'' approach through which application endpoint performance and utilization can be monitored and visualized in a unified way.
+Due to the addition of an orthogonal to the service evolution dimension that the \tool currently supports, this requires a significant effort of re-engineering on our part.
 
-    - threats to the validity of the constant-load performance testing. performance regressions?
+%	-   Supporting situations where the API is deployed across multiple containers for example.
+%	\ml{@V: Do you want to add this? Or can we skip it?}
 
-    - time as measured in days and commits are two sides of the same coin.
-
-    \ml{@V: do you want to give a first stab to this?}
+%    - threats to the validity of the constant-load performance testing. performance regressions? 
+%
+%
+%    - time as measured in days and commits are two sides of the same coin. 
+%
+%    \ml{@V: do you want to give a first stab to this?}
 

### 2018-04-09 introductory.tex

@@
   \item An advanced discussion on the integration with continuous integration environments in order to triangulate the live collected data with information from testing the endpoints with a constant load and introduce a mechanism for performance prediction is added (Section \ref{sec:testing})
-  \item A more extended discussion about lessons learned while developing the tool including measurements on the overhead incurred by the presented tool \ml{(... if we make it... )}
+  \item A more extended discussion about lessons learned while developing the tool including measurements on the overhead incurred by the presented tool (Sections \ref{sec:overhead} and \ref{sec:discussion})
 \end{itemize}
@@
       \includegraphics[width=0.7\linewidth]{zeeguu-architecture}
-      \caption{The architecture of the \zee ecosystem}
+      \caption{The architecture of the \zee ecosystem \cite{lungu18-chi}}
       \label{fig:zeeguuarch}
@@
 
-  We will use this \zee service API as a case study for this paper. 
-  The API has been deployed for several years without any monitoring
+  We will use this \zee service API as a case study for this paper. It is open source, and has been deployed for several years without any monitoring
   solution until June 2017, when the \tool has been deployed with it. 

### 2018-04-09 overhead.tex

@@
 To measure the performance overhead of the \tool, we have implemented an automated benchmarking system.
-It is open source and available online and can be tested by the reader.
-\va{@Mircea: Don't forget to add link} The benchmark downloads the latest version of the \zee API and installs it in a Docker container.
+It is open source and available online and can be tested by the reader\footnote{\em [The link will be added after deanonymization]}.
+The benchmark downloads the latest version of the \zee API and installs it in a Docker container.
 Then it calls several selected endpoints for 500 times each, tracking the response times.

### 2018-04-09 paper.tex
all iz well till section IV
nicer and nicer ...
Conclusions updated

@@
 
+\input{outliers}
+
 \input{grouping}
@@
 
-\todo{update as appropriate}
+In the previous sections we discussed our proposal for a monitoring dashboard for Flask-based web applications as the means for providing a low effort and flexible analytics solution to projects on a budget.
+The emphasis is in allowing application developers to gain insight into how the performance and utilization of their services evolves together with the application itself.
+For this purpose the \tool allows for integration with both \git and the Travis continuous integration platform for a seamless user experience.
+We also discussed more advanced features of the dashboard in the form of grouping results by user groups, and provided evidence towards the use of unit testing-sourced synthetic loads as a limited form of performance prediction.
+Finally we measured the overhead of the proposed solution within acceptable by the case study application owner limits, and discussed what we perceive as limitations of our proposal in Section~\ref{sec:discussion}.
 
-In this paper we have shown that it is possible to create a monitoring solution which provides basic insight into web service utilization and performance  with very little effort from the developer.
+The same section effectively acts also as a roadmap for further development of the \tool.
+Some of the tasks have already been identified as ongoing work (allowing different types of grouping, enabling performance prediction, developing a meta-dashboard) while others constitute short- to medium-term future work (identification of important versions to be monitored, support for endpoint provenance) with the ultimate goal of increasing the adoption of the \tool by as many projects as possible.
+This will allow us to evaluate our proposal in depth in the future.
+
+\va{@Mircea: see what you can add in the above, if any}
+
+%\todo{update as appropriate}
+%
+%In this paper we have shown that it is possible to create a monitoring solution which provides basic insight into web service utilization and performance  with very little effort from the developer.
 The user group that we are aiming for with this work is application developers using Flask and Python to build web applications with limited or no budget for implementing their own monitoring solutions.
@@
 We believe that the same architecture, and lessons can be applied to other frameworks and other languages.
-
-In the future, we plan to perform case studies with other sytstems, with the goal of discover other needs and to wean out the less useful visualizations in the \tool.
+%
+%In the future, we plan to perform case studies with other sytstems, with the goal of discover other needs and to wean out the less useful visualizations in the \tool.
 We plan to also extend the tool towards supporting multiple deployments of the same applications across multiple nodes (e.g.

### 2018-04-09 related-work.tex
cleaning up the discussion section
progressing slowly with the related work

@@
 
-Service monitoring data are increasingly being used and more across a variety of systems in which adaptability, flexibility, and environmental awareness are essential~\cite{pernici2016monitoring}.
+% testing... the performacen
+% there is plenty of work on testing for finding fauls
+% people: https://scholar.google.nl/citations?user=DWd972sAAAAJ&hl=en&oi=ao
+
+A rich body of work exists that studies the evolution and maintenance of APIs from the perspective of source code evolution \cite{dig2006apis, hora2015developers, hora2015apiwave}.
+However, since service monitoring data are increasingly being used across a variety of systems in which adaptability, flexibility, and environmental awareness are essential~\cite{pernici2016monitoring} in our work we focus on the analysis of the runtime aspects of services.
+There has been also a large body of work on using dynamic analysis in order to support the reverse engineering of software systems\cite{Corn09-dynamic}: in our work we focus more on understanding the usage of a system or service from the perspective of its clients.
+
 Monitoring can be performed on different types of system components and for several purposes.
@@
 
+Our 
+- Lanza has done nice stuff for source code evolution patterns. we need more work on automatic detection of service / endpoint evolution patterns. as exemplified by the utilization evolution.
+
 % \va{Mircea: Consider removing the rest for space...}

### 2018-04-09 discussion.tex
Conclusions updated
lay outing...

@@
 \section{Discussion}
+\label{sec:discussion}
 
@@
 
-    This technique assumes that the web application code which is the target of the monitoring is deployed using \git in the following way:
-
-    \begin{enumerate}
-      \item The deployment engineer pulls the latest version of the code from the integration server; this will result in a new commit being pointed at by the \code{HEAD} pointer. %than previously
-      \item The deployment engineer restarts the new version of the service.
-At this point, the \tool detects that a new \code{HEAD} is present in the local code base and consequently starts associating all the new data points with this new commit\footnote{The \tool detects the current version of the analyzed system the first time it is attached to the application object, and thus, assumes that the Flask application is restarted when a new version is deployed.
-This is in tune with the current version of Flask, but if the web server will support dynamic updates in the future, this might have to be taken into account}.
-    \end{enumerate}
-
-      The advantage of this approach is the need for minimal configuration effort, as discussed in the presentation of the \tool.
+      The advantage of the presented approach to evolution monitoring based on observing the \code{.git} folder is the need for minimal configuration effort, as discussed in the presentation of the \tool.
 The disadvantage is that it will consider on equal ground the smallest of commits, even one that modifies a comment, and the shortest lived of commits, e.g.~a commit which was active only for a half an hour before a new version with a bug fix was deployed, with major and minor releases of the software.

### 2018-04-09 evolution.tex
...
anonymization
lay outing...

@@
 \begin{lstlisting}[style=custompython]
-
 # LOC #3: provide the dashboard with 
@@
 
+    This technique assumes that the web application code which is the target of the monitoring is deployed using \git by pulling the latest version of the code from the integration server; this will result in a new commit being pointed at by the \code{HEAD} pointer.
+Thus, when the deployment engineer restarts the new version of the service, the \tool detects that a new \code{HEAD} is present in the local code base and consequently starts associating all the new data points with this new commit.
+The \tool detects the current version of the analyzed system the first time it is attached to the application object, and thus, assumes that the Flask application is restarted when a new version is deployed.
+This would not work in the presence of dynamic updates.
+
   \subsection*{Evolving Utilization}
+
+    \begin{figure}[h!]
+      \centering
+      \includegraphics[width=\linewidth]{utilization-evolution}
+      \caption{Evolving Endpoint Utilization Across System Versions}
+      \label{fig:mv-util}
+    \end{figure}
 
@@
 Otherwise a version that is deployed for many weeks would make all those deployed for a few days invisible}.
-
-    \begin{figure}
-      \centering
-      \includegraphics[width=0.9\linewidth]{utilization-evolution}
-      \caption{The Evolution of All the API Endpoints Utilization Across System Versions}
-      \label{fig:mv-util}
-    \end{figure}
 
@@
       \centering
-      \includegraphics[width=0.9\linewidth]{translation_endpoint_evolution_}
-      \caption{The Performance Evolution of the \epTranslations endpoint}
+      \includegraphics[width=0.97\linewidth]{translation_endpoint_evolution_}
+      \caption{Endpoint performance across three versions}
       \label{fig:tee}
@@
   \centering
-  \includegraphics[width=0.77\linewidth]{time_per_user_per_version}
+  \includegraphics[width=\linewidth]{time_per_user_per_version}
   \caption{This perspective shows that the evolution of response times for individual users (horizontal lines) across versions (the x-axis) for a given endpoint}

### 2018-04-09 grouping.tex
lay outing...

@@
 \begin{lstlisting}[style=custompython]  
-# LOC #2: configure the dashboard
-# to group requests by the user id
+# LOC #2: group requests by user id
 dashboard.config.group_by = 'User',
@@
   \centering
-  \includegraphics[width=.7\linewidth]{time_per_user}
-  \caption{The \epFeedItems shows a very high variability across users}
+  \includegraphics[width=.66\linewidth]{time_per_user}
+  \caption{And endpoint with a high variability across users}
   \label{fig:tpu}

### 2018-04-09 testing.tex
a nice table with the benchmark numbers is in
lay outing...

@@
         \centering
-        \includegraphics[width=0.65\columnwidth]{travis_builds-bw}
+        \includegraphics[width=0.85\columnwidth]{travis_builds-bw}
         \caption{Response times in 5 consequent Travis builds}
@@
   The concept of {\em preemptive monitoring} of the application performance by means of instrumenting integration unit testing as the synthetic load is similar to the idea of augmenting service monitoring with online testing~\cite{metzger2010proactive}, i.e.~testing service-based applications by using dedicated test input in parallel to its normal use and operation.
-The difference is that we take advantage of the capability of the CI framework to create an emulated ``live'' environment for integration testing purposes, and use unit testing as the dedicated test input in order to measure performance.
+The difference is that we take advantage of the capability of the CI framework to create an emulated ``live'' environment for integration testing purposes, and use unit testing as the dedicated test input.
+% in order to measure performance.
 
@@
 \Fref{fig:response_times_preemptive} shows the actual endpoint response times across five versions of the system (above) and the corresponding testing times for the synthetic load in the same five versions (below).
-The green line joining the medians hints to a correlation between the two sets of measurements.
 
@@
         \caption{The reported response time (in ms) per deployed version in the observation period: 
-        in the actual deployment (above) and using integration with Travis (below)}
+        in the actual deployment (above) and using integration with Travis (below). 
+        Green line hints to a correlation between the two sets of measurements. }
 
@@
   Computing Pearson correlation ($r(3)=.93, p=.02$) between the median values of the two datasets (cf.
-Table \ref{tab:correlations}) shows that there is indeed a correlation between the two.
+Table \ref{tab:correlations}) shows a correlation between the two.
 Further investigation on the characteristics of this observation is currently ongoing and it will be elaborated further in future work.

### 2018-04-09 discussion.tex
retitled

@@
 At this point, this constitutes future work.
-
-  \subsection*{Lack of Empirical Proof of Ease of Adoptability}
-
-    We designed the dashboard to make it easy to adopt and we think this is to a certain degree apparent.
-However, we have only presented a single use case where the API developers have used the dashboard.
-It would be worthwhile to organize a dedicated study to observe the challenges and opportunities of the adoption of such a dashboard across different applications.
-It will also be important to see whether the perspectives presented here are relevant for other developers, and also which other perspectives are missing.
 
@@
 
+  \subsection*{Need of Further Proof of Usefulness and Ease of Adoptability}
+
+    We designed the dashboard to make it easy to adopt and we think this is to a certain degree apparent.
+However, we have only presented a single use case where the API developers have used the dashboard.
+It has been successfull, since based on the dashboard, the developers have become aware of the limitations of their service APIs and have started improving it.
+However, we belive that it will be worthwhile to organize a dedicated study to observe the challenges and opportunities of the adoption of such a dashboard across different applications.
+It will also be important to see whether the perspectives presented here are relevant for other developers, and also which other perspectives are missing.
+
 %	-   Supporting situations where the API is deployed across multiple containers for example.

### 2018-04-09 paper.tex
lay outing...
retitled

@@
 
-%
-\title{Agile Monitoring of Evolving Web Services: For the Little Man}
+% An Agile Analytics Platform for Visualizing Evolving Flask-Based Python Web Services
+% A Dashboard for the Agile Monitoring of Evolving Web Services: For the Little Man
+% A Dashboard for Monitoring Flask \& Python-Based Web Service Evolution
+% Monitoring the Evolution of Flask \& Python-Based Web Services
+% A Dashboard for Monitoring the Evolution of Flask \& Python-Based Web Services
+\title{Monitoring the Evolution of Flask \& Python-Based Web Services}
 
@@
 The emphasis is in allowing application developers to gain insight into how the performance and utilization of their services evolves together with the application itself.
-For this purpose the \tool allows for integration with both \git and the Travis continuous integration platform for a seamless user experience.
-We also discussed more advanced features of the dashboard in the form of grouping results by user groups, and provided evidence towards the use of unit testing-sourced synthetic loads as a limited form of performance prediction.
-Finally we measured the overhead of the proposed solution within acceptable by the case study application owner limits, and discussed what we perceive as limitations of our proposal in Section~\ref{sec:discussion}.
-
-The same section effectively acts also as a roadmap for further development of the \tool.
-Some of the tasks have already been identified as ongoing work (allowing different types of grouping, enabling performance prediction, developing a meta-dashboard) while others constitute short- to medium-term future work (identification of important versions to be monitored, support for endpoint provenance) with the ultimate goal of increasing the adoption of the \tool by as many projects as possible.
+For this purpose the \tool allows for integration with both \git and the Travis CI platform.
+We also discussed more advanced features of the dashboard in the form of grouping results by user groups, and provided evidence towards the use of unit testing-sourced synthetic loads as a potential limited form of performance prediction.
+Finally we measured the overhead of the proposed solution within acceptable by the case study application owner limits.
+Section~\ref{sec:discussion} which discusses what we perceive as limitations of our proposal and also potential challenges for other similar tools, effectively acts also as a roadmap for further development, with the ultimate goal of increasing the adoption of the \tool by as many projects as possible.
 This will allow us to evaluate our proposal in depth in the future.
 
-\va{@Mircea: see what you can add in the above, if any}
+% Some of the tasks have already been identified as ongoing work (allowing different types of grouping, enabling performance prediction, developing a meta-dashboard) while others constitute short- to medium-term future work (identification of important versions to be monitored, support for endpoint provenance)
 

### 2018-04-09 related-work.tex
retitled

--- 2018-04-08 related-work.tex
@@
 
-Service monitoring data are increasingly being used and more across a variety of systems in which adaptability, flexibility, and environmental awareness are essential~\cite{pernici2016monitoring}.
-Monitoring can be performed on different types of system components and for several purposes.
-Service-oriented literature on the subject discussed the multitude of aspects related to monitoring for services; the interested reader is referred to ~\cite{ghezzi2007run},~\cite{metzger2010analytical}, or more recently~\cite{pernici2016monitoring} for an extensive presentation of the subject.
+% testing... the performacen
+% there is plenty of work on testing for finding fauls
+% people: https://scholar.google.nl/citations?user=DWd972sAAAAJ&hl=en&oi=ao
+
+A rich body of work exists that studies the evolution and maintenance of APIs from the perspective of source code evolution \cite{dig2006apis, hora2015developers, hora2015apiwave}.
+However, since services are increasingly being used across a variety of systems in which adaptability, flexibility, and environmental awareness are essential~\cite{pernici2016monitoring} in our work we focus on the analysis of the runtime aspects of service APIs.
+Our work is thus, more related to the work on dynamic analysis of software systems; however, most of this work has been done in order to support the reverse engineering of software systems\cite{Corn09-dynamic}.
 
@@
 While we don't implement the more advanced features of related monitoring solutions like QoS policies driving the monitoring, it presents nevertheless an easy to use approach towards improving the performance of web applications.
-The emphasis in this work is given in identifying the appropriate way to visualize the monitoring data.
 
@@
 
+Kieker -- has been presented by Rohr et al \cite{Rohr08} as a system targeted at monitoring the performance of Java services.
+The system provides a better performance in terms of overhead, but does not support integration with git and multi-version analysis as the system presented here.
+
 In their work Baresi and Guinea have proposed the Multi-layer Collection and Constraint Language (mlCCL) which allows them to define how to collect, aggregate, and analyze runtime data in a multi-layered system.
 They also present ECoWare, a framework for event correlation and aggregation that supports mlCCL, and provides a dashboard for on-line and off-line drill-down analyses of collected data \cite{Bare13-monitoring}.
+
+Monitoring can be performed on different types of system components and for several purposes beyond the ones presented here.
+Service-oriented literature on the subject discussed the multitude of aspects related to monitoring for services; the interested reader is referred to ~\cite{ghezzi2007run},~\cite{metzger2010analytical}, or more recently~\cite{pernici2016monitoring} for an extensive presentation of the subject.
+
+% Our 
+% - Lanza has done nice stuff for source code evolution patterns. we need more work on automatic detection of service / endpoint evolution patterns. as exemplified by the utilization evolution.
 

### 2018-04-09 testing.tex
retitled

@@
   \begin{lstlisting}[style=custompython]
-# LOC #4: to be added to the .travis.yml file
+
+# LOC #4: is added to the .travis.yml file
 python -m flask_monitoringdashboard.collect
@@
         \centering
-        \includegraphics[width=0.85\columnwidth]{travis_builds-bw}
+        \includegraphics[width=\columnwidth]{travis_builds-bw}
         \caption{Response times in 5 consequent Travis builds}

### 2018-04-09 evolution.tex
retitled
one more tiny detail

@@
       \centering
-      \includegraphics[width=\linewidth]{utilization-evolution}
+      \includegraphics[width=\linewidth]{utilization-evolution.pdf}
       \caption{Evolving Endpoint Utilization Across System Versions}
@@
 
-  With this view, several patterns could be visible on such a graph:
+  Several patterns could be visible on such a graph:
   \begin{itemize}
 
-    \item Endpoints being deprecated. In the case study, \epFeedItems stops being used somewhere after the half of the presented timeline.
+    \item Endpoints being deprecated. In the case study, \epFeedItemsColor stops being used somewhere after the half of the presented timeline.
 
-    \item Endpoints being introduced. In the case study, \epTopArticles appears soon after the half o the presented timeline.
+    \item Endpoints being introduced. In the case study, \epTopArticlesColor appears soon after the half o the presented timeline.
 
@@
   \centering
-  \includegraphics[width=\linewidth]{time_per_user_per_version}
+  \includegraphics[width=0.9\linewidth]{time_per_user_per_version}
   \caption{This perspective shows that the evolution of response times for individual users (horizontal lines) across versions (the x-axis) for a given endpoint}

### 2018-04-09 paper.tex
one more tiny detail
submitted version of the paper

@@
 % A Dashboard for Monitoring the Evolution of Flask \& Python-Based Web Services
-\title{Monitoring the Evolution of Flask \& Python-Based Web Services}
+% Continuously Monitoring the Evolution of Flask \& Python-Based Web Services
+% A Dashboard for Continuously Monitoring the Evolution of Flask \& Python-Based Web Services
+% something about visualizing
+% Continuously Monitoring Through Visualization the Evolution of Flask \& Python-Based Web Services
+% Visual Perspectives for Continuously Monitoring the Evolution of Flask \& Python-Based Web Services
+\title{Visualization-Focused Continuous Monitoring of Evolving Flask \& Python-Based Web Services}
 
@@
 
-In the previous sections we discussed our proposal for a monitoring dashboard for Flask-based web applications as the means for providing a low effort and flexible analytics solution to projects on a budget.
+In this paper we discussed our proposal for a monitoring dashboard for Flask-based web applications as the means for providing a low effort analytics solution.
 The emphasis is in allowing application developers to gain insight into how the performance and utilization of their services evolves together with the application itself.
-For this purpose the \tool allows for integration with both \git and the Travis CI platform.
-We also discussed more advanced features of the dashboard in the form of grouping results by user groups, and provided evidence towards the use of unit testing-sourced synthetic loads as a potential limited form of performance prediction.
+\tool allows for integration with both \git and the Travis CI platform.
+We also discussed more advanced features of the dashboard and provided evidence towards the use of unit testing-sourced synthetic loads as a potential limited form of performance prediction.
 Finally we measured the overhead of the proposed solution within acceptable by the case study application owner limits.
@@
 
+\newpage
 \bibliographystyle{abbrv}

### 2018-04-09 related-work.tex
submitted version of the paper

@@
 However, since services are increasingly being used across a variety of systems in which adaptability, flexibility, and environmental awareness are essential~\cite{pernici2016monitoring} in our work we focus on the analysis of the runtime aspects of service APIs.
-Our work is thus, more related to the work on dynamic analysis of software systems; however, most of this work has been done in order to support the reverse engineering of software systems\cite{Corn09-dynamic}.
+Our work is thus, more related to the work on dynamic analysis of software systems; however, most of this work has been done in order to support the reverse engineering and enhance the understanding of software systems\cite{Corn09-dynamic}.
 
 In this context, our work falls within the server-side run-time monitoring of services ~\cite{ghezzi2007run}. 
-While we don't implement the more advanced features of related monitoring solutions like QoS policies driving the monitoring, it presents nevertheless an easy to use approach towards improving the performance of web applications.
-
+% While we don't implement the more advanced features of related monitoring solutions like QoS policies driving the monitoring, it presents nevertheless an easy to use approach towards improving the performance of web applications.
 There is a long tradition of using visualization for gaining insight into software performance.
 Tools like Jinsight \cite{Pauw02a} and Web Services Navigator \cite{Pauw05} pioneered such an approach for Java and for Web Services that communicate with SOAP messages.
-Both have an ``omniscient'' view of the services / objects and their interactions.
+More recent tools present live visualizations for monitoring multiple interacting software systems utilizing the city metaphor for each software system \cite{fittkau2013live}.
+All these approaches have an ``omniscient'' view of the services / objects and their interactions.
 As opposed to them, in our work we present an analytics platform which focuses on monitoring a single Python web service from its own point of view.
@@
 Kieker -- has been presented by Rohr et al \cite{Rohr08} as a system targeted at monitoring the performance of Java services.
-The system provides a better performance in terms of overhead, but does not support integration with git and multi-version analysis as the system presented here.
+The system provides better performance in terms of overhead with respect to ours, but does not support integration with git and multi-version analysis as the system presented here.
+
+Similar to our discussion on pre-emptive monitoring, Cito et al.
+\cite{cito2015runtime} argue for bringing performance measurements back in the IDE to provide the developers with such information during the development process.
 
